---
title: "Problem Set 1: Frequentist hypothesis testing"
subtitle: "Paired, exact, and contingency-table tests"
format:
  html:
    toc: true
execute:
  enabled: false
---

## Null-hypothesis significance testing

A [null-hypothesis significance test](../../statistical-inference/null-hypotheses-and-test-statistics.qmd) has four parts. First, state a null hypothesis about a population parameter or a relation between variables. Second, define a statistic that measures departure from that hypothesis. Third, derive or select the statistic's sampling distribution under the null hypothesis. Fourth, calculate the probability, under that distribution, of a result at least as discrepant as the observed result. This probability is the p value. It is not the probability that the null hypothesis is true.

This assignment applies that sequence to five tests: the paired t test, the Wilcoxon signed-rank test, the exact sign test, Fisher's exact test, and the chi-squared test of independence. For each test, you will derive the statistic or its null probability and then implement the calculation in R. The acoustic data become relevant in Exercises 1–4. The morphological data become relevant in Exercises 5–6.

All six exercises are required. The points sum to 100.

::: {.callout-note title="Assignment files"}
[Download the complete Problem Set 1 package](../../downloads/ps1-assignment.zip){download="ps1-assignment.zip"}. The package contains the assignment source, the README, the preparation scripts, and the redistributable teaching tables.

The files are also available separately:

- [README and setup instructions](README.md)
- [`prepare_data.R`](scripts/prepare_data.R){download="prepare_data.R"}
- [`fetch_unimorph_data.py`](scripts/fetch_unimorph_data.py){download="fetch_unimorph_data.py"}
- [`prepare_unimorph_tables.R`](scripts/prepare_unimorph_tables.R){download="prepare_unimorph_tables.R"}
:::

Use `<-` for assignment, use `snake_case` object names, and record a seed before every random calculation. Every numerical answer must be produced by visible code.

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
```

## Exercise 1: Vowel formants and paired differences (10 points)

We'll use hypothesis tests to analyze **vowel formant frequencies**. Formants are acoustic correlates of vowel quality in speech production. The first formant (F1) correlates with vowel height: lower F1 tends to correspond to a higher vowel. The second formant (F2) correlates with vowel backness: higher F2 tends to correspond to a fronter vowel.

The data for Exercises 1–4 come from Hillenbrand, Getty, Clark, and Wheeler's study of American English vowels ([1995, JASA](https://doi.org/10.1121/1.411872)). The preparation script extracts the `phonTools::h95` teaching table. Each of the 45 adult men produced the same 12 vowels in an /hVd/ context. In particular, each man produced /i/ as in *heed* and /ɪ/ as in *hid*. Because the two measurements come from the same speaker, they form a [paired observation](../../statistical-inference/paired-observations.qmd).

Generate `data/hillenbrand_vowels.csv` by following the [README](README.md). The columns used here are `speaker_id`, `speaker_type`, `vowel`, `f1_hz`, and `f2_hz`.

### 1.1 Prepare the paired table

Load the table. Retain rows for adult men and vowels `i` and `I`. Verify that each retained speaker has one row for each vowel. Join the two vowel tables by `speaker_id` and `speaker_type`. Create one row per speaker with `f1_i_hz`, `f1_I_hz`, and

$$
d_s=\operatorname{F1}_{s,/i/}-\operatorname{F1}_{s,/ɪ/}.
$$

Here $d_s$ is the observed difference for speaker $s$.

```{r}
# Load Hillenbrand formant data.

# Examine the column names, classes, and vowel labels.

# Filter to adult men and the vowels i and I.

# Verify one row per speaker and vowel.

# Join the two vowel tables by speaker_id and speaker_type.

# Calculate difference_hz as F1 for i minus F1 for I.
```

### 1.2 Visualize vowel space

The **F1–F2 vowel space** provides an acoustic representation of vowel quality. Plot F2 on the horizontal axis and F1 on the vertical axis, with both axes reversed to mirror the conventional articulatory vowel space. Use the complete adult-male subset, not only /i/ and /ɪ/.

```{r}
# Classic F1-F2 vowel plot with inverted axes.
# 1. Filter to adult men.
# 2. Map f2_hz to x and f1_hz to y.
# 3. Color points by vowel.
# 4. Add one ellipse per vowel with stat_ellipse(level = .68).
# 5. Reverse both axes and label both measurements in Hz.
```

### 1.3 Summarize the differences

Calculate the number of speakers, the mean, the standard deviation, the median, and the number of negative, zero, and positive differences. Plot the differences with zero marked.

```{r}
# Descriptive statistics.
# 1. Count the speaker pairs.
# 2. Calculate the mean, standard deviation, and median of difference_hz.
# 3. Count negative, zero, and positive differences.

# Plot the paired differences.
# 1. Put one point at each observed difference.
# 2. Mark zero with a reference line.
# 3. Label the x-axis with the subtraction order and units.
```

Report the requested values in the subtraction order /i/ minus /ɪ/.

::: {.callout-note title="Your result"}
Replace this sentence with the sample size and requested summaries.
:::

## Exercise 2: Paired t test (20 points)

Let $D_s:\Omega\to\mathbb R$ be the random /i/ minus /ɪ/ difference for speaker $s$, and let $\mu_D=\mathbb E[D_s]$. Under the [paired normal model](../../statistical-inference/paired-mean-inference.qmd),

$$
D_s\mid\mu_D,\sigma_D\sim\mathcal N(\mu_D,\sigma_D^2).
$$

Test $H_0:\mu_D=0$ against $H_1:\mu_D\ne0$.

::: {.callout-note title="Required background"}
Review [the Student t distribution](../../random-variables-and-distributions/student-t-distribution.qmd), [null hypotheses and test statistics](../../statistical-inference/null-hypotheses-and-test-statistics.qmd), [paired observations](../../statistical-inference/paired-observations.qmd), and [inference for a paired mean difference](../../statistical-inference/paired-mean-inference.qmd) before beginning the derivation below. The two vowel measurements are paired because each speaker produced both vowels. Thus the paired t test applies the one-sample t procedure to the within-speaker differences.
:::

### 2.1 Derive the estimator and its standard error

Let $S$ be the number of speakers and $\overline D=S^{-1}\sum_{s=1}^{S}D_s$. Use [linearity of expectation](../../random-variables-and-distributions/linearity-of-expectation.qmd) and independence to derive $\mathbb E[\overline D]$ and $\operatorname{Var}(\overline D)$. Then replace $\sigma_D$ with the sample standard deviation $S_D$ to obtain the estimated standard error.

**Target:** derive $\mathbb E[\overline D]=\mu_D$, $\operatorname{Var}(\overline D)=\sigma_D^2/S$, and $\widehat{\operatorname{SE}}(\overline D)=S_D/\sqrt S$.

**Your LaTeX**

$$
\begin{aligned}
&\text{Enter the three derivations here.}
\end{aligned}
$$

### 2.2 Derive the test statistic

Standardize the difference between $\overline D$ and the null value. Because $\sigma_D$ is estimated by $S_D$, the null reference distribution is a [Student t distribution](../../random-variables-and-distributions/student-t-distribution.qmd) with $S-1$ degrees of freedom.

**Target:** derive the random statistic $\mathcal T$ and its observed value $t_{\mathrm{obs}}$.

**Your LaTeX**

$$
\begin{aligned}
\mathcal T&=\text{enter the random statistic},\\
t_{\mathrm{obs}}&=\text{enter the statistic as a function of }\overline d,s_d,\text{ and }S.
\end{aligned}
$$

### 2.3 Derive the two-sided p value

Define the event containing null outcomes at least as far from zero as the observed statistic. Use full set-builder notation. Then use the cumulative distribution function $F_{t_{S-1}}$ to write its probability.

**Target:** define $B_{\mathcal T}$ and derive $\mathbb P_0(B_{\mathcal T})=2[1-F_{t_{S-1}}(|t_{\mathrm{obs}}|)]$.

**Your LaTeX**

$$
\begin{aligned}
B_{\mathcal T}&=\{\omega\in\Omega\mid\text{enter the two-sided condition}\},\\
\mathbb P_0(B_{\mathcal T})&=\text{enter the tail probability}.
\end{aligned}
$$

### 2.4 Derive the confidence interval

A 95% [confidence interval](../../statistical-inference/confidence-intervals.qmd) is obtained by inverting the central 95% t reference region.

**Target:** write the lower and upper endpoints as functions of $\overline d$, $s_d$, $S$, and $t_{S-1,.975}$.

**Your LaTeX**

$$
\begin{aligned}
L&=\text{enter the lower endpoint},\\
U&=\text{enter the upper endpoint}.
\end{aligned}
$$

### 2.5 Implement the test

Implement the paired t-test calculation from scratch to verify R's `t.test()` output. Without calling `t.test()`, calculate $\overline d$, $s_d$, the estimated standard error, $t_{\mathrm{obs}}$, the two-sided p value, and the 95% confidence interval.

```{r}
# YOUR CODE HERE
# Implement the paired t-test.
#
# Steps:
# 1. Extract the speaker-level differences.
# 2. Compute the mean, standard deviation, and number of differences.
# 3. Compute the estimated standard error.
# 4. Compute the t statistic.
# 5. Compute the degrees of freedom.
# 6. Compute the two-sided p value using pt().
# 7. Compute the confidence-interval endpoints using qt().
```

Verify the result with `t.test(..., paired = TRUE)`.

```{r}
# Compare the manual estimate, statistic, degrees of freedom, p value,
# and confidence interval with t.test(..., paired = TRUE).
```

Report $\overline d$, the interval, $t_{\mathrm{obs}}$, the degrees of freedom, and the p value.

::: {.callout-note title="Your result"}
Replace this sentence with the requested test result.
:::

### 2.6 Choose your own vowel pair

Select another pair of vowels and test whether their mean F1 or F2 values differ. Use the adult-male measurements and preserve the paired speaker design. State the two vowels, the acoustic measurement, the subtraction order, and the hypotheses. Construct the speaker-level differences, calculate the paired t test from the differences, and verify it with `t.test(..., paired = TRUE)`.

```{r}
# YOUR CODE HERE
# 1. Select two vowels and either f1_hz or f2_hz.
# 2. Join their measurements by speaker_id and speaker_type.
# 3. Calculate one difference per speaker.
# 4. Plot the differences with zero marked.
# 5. Calculate and verify the paired t test.
```

Report the estimate, 95% confidence interval, t statistic, degrees of freedom, and p value. Relate the sign of the estimate to the stated subtraction order and the selected formant.

::: {.callout-note title="Your result"}
Replace this sentence with the requested vowel comparison.
:::

## Exercise 3: Wilcoxon signed-rank test (10 points)

The paired t statistic uses the numerical magnitudes of the differences. The Wilcoxon signed-rank statistic replaces each nonzero magnitude $|d_s|$ with its rank. It thus uses the order of the magnitudes rather than their spacing.

For $S$ nonzero differences, rank $|d_1|,\ldots,|d_S|$ from 1 to $S$, assigning the mean rank to tied magnitudes. Let $r_s$ be the rank assigned to $|d_s|$. Define

$$
w_+=\sum_{s=1}^{S}r_s\mathbb I(d_s>0),
$$

where $\mathbb I(d_s>0)$ equals one when $d_s>0$ and zero otherwise. Under the signed-rank null model, the differences are independent and symmetric about zero. Conditional on the observed absolute ranks, positive and negative signs are equally probable.

### 3.1 Construct the statistic

Use `rank(abs(difference_hz), ties.method = "average")` to calculate $r_s$. Calculate $w_+$ from the ranks and signs. Display `speaker_id`, `difference_hz`, the absolute difference, its rank, and its sign indicator.

```{r}
# YOUR CODE HERE
# 1. Remove zero differences, if any.
# 2. Calculate abs(difference_hz).
# 3. Rank the absolute differences with average ranks for ties.
# 4. Create the indicator for a positive difference.
# 5. Multiply each rank by its indicator and sum to obtain w_plus.
# 6. Display the speaker, difference, magnitude, rank, and indicator.
```

### 3.2 Derive the null mean

Let $B_s:\Omega\to\{0,1\}$ be the random positive-sign indicator under the null model. Then $\mathbb E[B_s]=.5$ and $\mathcal W_+=\sum_{s=1}^{S}r_sB_s$. Use linearity of expectation and $\sum_{s=1}^{S}r_s=S(S+1)/2$ to derive the null mean.

**Target:** derive $\mathbb E_0[\mathcal W_+]=S(S+1)/4$.

**Your LaTeX**

$$
\begin{aligned}
\mathbb E_0[\mathcal W_+]&=\text{enter the derivation here}.
\end{aligned}
$$

### 3.3 Implement the test

Run `wilcox.test()` on the paired F1 measurements with `paired = TRUE`, `alternative = "two.sided"`, `exact = FALSE`, and `correct = FALSE`. The observed integer-valued F1 measurements produce tied absolute differences, so this call uses the large-sample signed-rank reference approximation rather than exact enumeration.

```{r}
# Run the paired Wilcoxon signed-rank test.
# Compare the returned V statistic with the manual value of w_plus.
```

Report $w_+$, the p value, and the null hypothesis tested by the signed-rank procedure.

::: {.callout-note title="Your result"}
Replace this sentence with the requested signed-rank result.
:::

## Exercise 4: Exact sign test (10 points)

The [Bernoulli distribution](../../random-variables-and-distributions/bernoulli-distribution.qmd) represents whether each difference has the predicted sign. Define

$$
Z_s(\omega)=
\begin{cases}
1 & \text{if }D_s(\omega)<0,\\
0 & \text{if }D_s(\omega)\ge0.
\end{cases}
$$

Let $\pi_D=\mathbb P(Z_s=1)$. Test $H_0:\pi_D=.5$ against $H_1:\pi_D>.5$. There are no zero differences in these data.

### 4.1 Derive the null distribution and p value

Let $\mathcal K=\sum_{s=1}^{S}Z_s$ and $k_{\mathrm{obs}}=\sum_{s=1}^{S}z_s$. Use the [binomial distribution](../../random-variables-and-distributions/binomial-distribution.qmd) to derive the null distribution of $\mathcal K$. Then define the upper-tail event and expand its probability as a sum.

**Target:** derive $\mathcal K\mid H_0\sim\operatorname{Binomial}(S,.5)$ and $\mathbb P_0(\{\omega\in\Omega\mid\mathcal K(\omega)\ge k_{\mathrm{obs}}\})$.

**Your LaTeX**

$$
\begin{aligned}
\mathcal K\mid H_0&\sim\text{enter the null distribution},\\
B_{\mathcal K}&=\{\omega\in\Omega\mid\text{enter the upper-tail condition}\},\\
\mathbb P_0(B_{\mathcal K})&=\text{enter the binomial sum}.
\end{aligned}
$$

### 4.2 Implement the test

Calculate $k_{\mathrm{obs}}$, $k_{\mathrm{obs}}/S$, and the upper-tail probability with `pbinom(..., lower.tail = FALSE)`. Verify the calculation with `binom.test(..., alternative = "greater")` and report the one-sided 95% [exact confidence bound](../../statistical-inference/exact-binomial-intervals.qmd).

```{r}
# YOUR CODE HERE
# 1. Convert each difference to the indicator I(difference_hz < 0).
# 2. Sum the indicators to obtain k_observed.
# 3. Divide by the number of speakers to obtain the observed proportion.
# 4. Compute the upper-tail probability using pbinom().
# 5. Verify the calculation with binom.test().
# 6. Extract the one-sided 95% confidence bound.
```

Report the sign count, the observed proportion, the confidence bound, and the p value.

::: {.callout-note title="Your result"}
Replace this sentence with the requested sign-test result.
:::

## Exercise 5: Fisher's exact test (25 points)

We'll use Fisher's exact test to analyze **German genitive allomorphy**, a case of morphological variation in which the genitive singular ending on masculine and neuter nouns alternates between *-s* and *-es*. For instance, the genitive of *Tag* “day” may be realized as *Tages* or *Tags*. The traditional pedagogical rule states that stems ending in sibilants require *-es*. Syllable count has also been proposed to matter, with monosyllabic nouns tending to favor *-es* and polysyllabic nouns tending to favor *-s*.

The data come from the German [UniMorph](https://unimorph.github.io/) paradigm inventory described by [McCarthy et al. (2020)](https://aclanthology.org/2020.lrec-1.483/). The preparation scripts reduce the genitive records to one row per lemma. They record syllable class, whether a lemma ends in one of the strings `s`, `z`, `ß`, `sch`, or `x`, and whether a genitive form ending in *-es* is attested. These are variables produced by the scripts. UniMorph does not supply them as linguistic analyses.

The raw paradigm may contain both *-s* and *-es* forms for one lemma. Thus `es_attested` records attestation in the extracted paradigm rather than an exclusive suffix choice.

The teaching table contains a reproducible simple random sample of 120 lexemes. Its relevant columns are `lemma`, `syllable_class`, `ends_in_sibilant`, and `es_attested`.

::: {.callout-note title="Required background"}
Review [the hypergeometric distribution](../../random-variables-and-distributions/hypergeometric-distribution.qmd) and [Fisher's exact test](../../statistical-inference/fishers-exact-test.qmd) before beginning the manual calculation below. These notes develop the 2×2 contingency table, fixed margins, and the probability-ordering definition of a two-sided p value.
:::

### 5.1 Load and inspect the lexeme table

Load `data/unimorph_genitive_lexemes.csv`. Verify that `lemma` is unique. Examine the levels and marginal distributions of `syllable_class`, `ends_in_sibilant`, and `es_attested`.

```{r}
# Load genitive allomorphy data.

# Examine the data.

# Verify one row per lemma.

# Display the syllable-class, sibilant, and -es-attestation distributions.
```

### 5.2 Test syllable count and genitive *-es* attestation

In German genitive allomorphy, is there an association between syllable class and the attestation of an *-es* form? Construct the two-by-two table for `syllable_class` by `es_attested`, with `monosyllabic` before `polysyllabic` and `TRUE` before `FALSE`. Plot the table with a mosaic plot or grouped bar chart. Run `fisher.test()` and calculate the sample odds ratio. With this order, the odds ratio places the odds of *-es* attestation among monosyllabic lexemes in the numerator and the corresponding odds among polysyllabic lexemes in the denominator.

```{r}
# YOUR CODE HERE
# a) Create a contingency table for syllable_class by es_attested.

# b) Visualize the table using a mosaic plot or grouped bar chart.

# c) Conduct Fisher's exact test.

# d) Compute the sample odds ratio in the stated orientation.
```

Report the observed table, sample odds ratio, conditional estimate and interval returned by `fisher.test()`, and p value.

::: {.callout-note title="Your result"}
Replace this sentence with the requested syllable-class result.
:::

### 5.3 Construct the sibilant table

Is there an association between the string-based sibilant classification and the attestation of an *-es* form? Construct the table for `ends_in_sibilant` by `es_attested`, with the character value `"True"` before `"False"` in the rows and the logical value `TRUE` before `FALSE` in the columns. With this order, the sample odds ratio places the odds of *-es* attestation among sibilant-final lexemes in the numerator and the corresponding odds among other lexemes in the denominator.

```{r}
# YOUR CODE HERE
# 1. Set explicit "True" and "False" row levels.
# 2. Set explicit TRUE and FALSE column levels.
# 3. Construct the two-by-two contingency table.
# 4. Display all four cells and both sets of margins.
```

Report the four cells and both sets of margins.

::: {.callout-note title="Your result"}
Replace this sentence with the observed sibilant table and its margins.
:::

### 5.4 Derive the conditional null probability

[Fisher's exact test](../../statistical-inference/fishers-exact-test.qmd) tests independence in a two-by-two table while holding both observed margins fixed. Write the observed cells in row-major order as $a,b,c,d$. Let $A$ be the random upper-left cell count among all tables with those margins. Under the null hypothesis of independence, $A$ has a [hypergeometric distribution](../../random-variables-and-distributions/hypergeometric-distribution.qmd).

Derive the probability of the observed table by counting the ways to select its first-column total from the two rows and dividing by the number of selections from all observations.

**Target:** derive $\mathbb P_0(A=a)$ as a ratio of binomial coefficients containing $a,b,c,d$.

**Your LaTeX**

$$
\begin{aligned}
\mathbb P_0(A=a)&=\text{enter the ratio of binomial coefficients}.
\end{aligned}
$$

### 5.5 Implement the sibilant test

Calculate the sample odds ratio $ad/(bc)$ and the observed-table probability with `dhyper()`. Run `fisher.test()` to obtain the two-sided p value and conditional confidence interval. Fisher's two-sided p value sums the probabilities of all margin-compatible tables whose null probabilities are no greater than the observed table's probability.

```{r}
# YOUR CODE HERE
# 1. Extract a, b, c, and d in the declared row and column order.
# 2. Calculate the sample odds ratio ad / bc.
# 3. Calculate the observed-table probability using dhyper().
# 4. Conduct Fisher's exact test.
# 5. Extract the conditional estimate, confidence interval, and p value.
# 6. Compare the sample odds ratio with the syllable-class odds ratio.
```

Report the odds-ratio orientation, the sample odds ratio, the conditional estimate returned by `fisher.test()`, its confidence interval, and the p value. Put the sibilant and syllable-class sample odds ratios in the same table.

::: {.callout-note title="Your result"}
Replace this sentence with the requested sibilant result and odds-ratio comparison.
:::

### 5.6 Estimate sampling distributions across lexicon sizes

A [sampling distribution](../../statistical-inference/properties-of-estimators.qmd) is the distribution of a statistic across repeated samples from a fixed population. For this exercise, treat `data/unimorph_genitive_lexeme_population.csv` as the finite population.

For each $n\in\{12,20,40,80,120\}$, draw 500 simple random samples without replacement. Use one recorded seed. For every sample, record the odds ratio, Fisher p value, confidence-interval endpoints, whether any cell is zero, and whether either margin is zero. Use explicit factor levels when constructing every table.

```{r}
# YOUR CODE HERE
# 1. Write a function that accepts a sample size and replicate identifier.
# 2. Draw lexemes without replacement.
# 3. Construct the sibilant by -es-attestation table with explicit levels.
# 4. Record zero margins and zero cells.
# 5. Record the odds ratio, Fisher p value, and interval endpoints.
# 6. Return one tidy row.
# 7. Run 500 replicates at every requested sample size.
```

For each $n$, calculate the proportion of samples with a zero cell, the proportion with a zero margin, the proportion with a nonfinite odds ratio, the median finite log odds ratio, and the median finite confidence-interval width on the log scale. Plot the sampling distributions of the finite log odds ratios by $n$.

```{r}
# 1. Group the replicate results by sample size.
# 2. Calculate every requested summary.
# 3. Plot the finite log odds ratios by sample size.
```

Report the summary table.

::: {.callout-note title="Your result"}
Replace this sentence with the sample-size summary table.
:::

## Exercise 6: Chi-squared test of independence (25 points)

We'll use the chi-squared test to analyze **German plural allomorphy**. German plural marking combines several suffix classes with possible stem changes. The classes represented in this teaching table include *-e* as in *Tag* → *Tage*, *-en* as in *Frau* → *Frauen*, *-er* as in *Kind* → *Kinder*, *-s* as in *Auto* → *Autos*, and forms for which the preparation rule finds none of these overt suffixes.

The second UniMorph table has one row per noun lemma. The preparation rule retains one plural form and classifies its ending as `e`, `en`, `er`, `s`, or `ø`. Here `ø` means that the rule found none of the listed overt suffixes. It is a label produced by the script, not a claim that the plural has no exponence or that it contains umlaut.

The linguistic question is whether grammatical gender and the derived plural-suffix class are independent in the lexeme inventory represented by the table. This creates a contingency table larger than 2×2.

::: {.callout-note title="Required background"}
Review [the chi-squared test of independence](../../statistical-inference/chi-squared-test.qmd) and [the chi-squared distribution](../../random-variables-and-distributions/chi-squared-distribution.qmd) before beginning the derivations below. These notes develop the independence factorization, expected counts, Pearson's statistic, degrees of freedom, and the large-sample reference distribution.
:::

Load `data/unimorph_plural_lexemes.csv`. The relevant columns are `lemma`, `gender`, and `plural_suffix`. Construct the observed three-by-five table with gender ordered `FEM`, `MASC`, `NEUT` and suffix ordered `e`, `en`, `er`, `s`, `ø`.

```{r}
# Load plural suffix data.

# Examine the data and verify one row per lemma.

# Set the required gender and suffix factor levels.

# Create the gender by plural-suffix contingency table.
```

### 6.1 Derive the expected counts

Under independence, the expected count in cell $(i,j)$ is determined by its row total, column total, and the grand total. Follow the derivation in the [chi-squared-test notes](../../statistical-inference/chi-squared-test.qmd).

**Target:** derive $E_{ij}=O_{i+}O_{+j}/O_{++}$ and substitute the observed margins for the feminine-by-*en* cell.

**Your LaTeX**

$$
\begin{aligned}
E_{ij}&=\text{enter the general derivation},\\
E_{\mathrm{FEM},\mathrm{en}}&=\text{enter the numerical substitution and result}.
\end{aligned}
$$

### 6.2 Derive the statistic and reference distribution

Pearson's statistic sums the standardized squared difference between every observed and expected count. If the table has $R$ rows and $C$ columns, estimating the margins leaves $(R-1)(C-1)$ degrees of freedom.

**Target:** write $\mathcal X^2$, derive $\nu=(R-1)(C-1)$, define the upper-tail event in full set-builder notation, and write its probability under $\chi^2_\nu$.

**Your LaTeX**

$$
\begin{aligned}
\mathcal X^2&=\text{enter Pearson's statistic},\\
\nu&=\text{enter the degrees of freedom},\\
B_{\mathcal X^2}&=\{\omega\in\Omega\mid\text{enter the upper-tail condition}\},\\
\mathbb P_0(B_{\mathcal X^2})&=\text{enter the chi-squared tail probability}.
\end{aligned}
$$

### 6.3 Implement the test

Compute the chi-squared statistic manually to verify R's output. Without calling `chisq.test()`, calculate the complete expected table, every cell contribution, $x^2_{\mathrm{obs}}$, the degrees of freedom, and the p value. Verify that every expected count is at least five.

```{r}
# YOUR CODE HERE
# Steps:
# 1. Create the contingency table.
# 2. Compute expected counts for each cell.
# 3. Compute (O - E)^2 / E for each cell.
# 4. Sum the contributions to get the chi-squared statistic.
# 5. Compute the degrees of freedom.
# 6. Compute the p value using pchisq().
```

Verify the statistic, degrees of freedom, p value, and expected table with `chisq.test(..., correct = FALSE)`. Calculate the Pearson residuals $(O_{ij}-E_{ij})/\sqrt{E_{ij}}$ and identify the two cells with the largest absolute residuals.

```{r}
# Compare the manual result with chisq.test() output.

# Check the expected-count condition.

# Calculate and sort the Pearson residuals by absolute magnitude.
```

Report $x^2_{\mathrm{obs}}$, the degrees of freedom, the p value, the minimum expected count, and the two largest absolute residuals with their signs.

::: {.callout-note title="Your result"}
Replace this sentence with the requested chi-squared-test result.
:::

## Submission checklist

- The notebook runs from a fresh R session after the data-preparation steps.
- Every derivation is entered in its LaTeX block.
- Every computation is entered in the R block immediately following its instruction.
- Every requested result is entered in its result block.
- The paired table is joined by identifiers rather than aligned by row position.
- Every hypothesis names its parameter or independence relation.
- Every p value is defined relative to its null reference distribution.
- Every random calculation has a recorded seed.
