---
title: "Problem Set 1: Frequentist hypothesis testing"
subtitle: "Paired, exact, and contingency-table tests"
format:
  html:
    toc: true
execute:
  enabled: false
---

## Null-hypothesis significance testing

A [null-hypothesis significance test](../../statistical-inference/null-hypotheses-and-test-statistics.qmd) has four parts. First, state a null hypothesis about a population parameter or a relation between variables. Second, define a statistic that measures departure from that hypothesis. Third, derive or select the statistic's sampling distribution under the null hypothesis. Fourth, calculate the probability, under that distribution, of a result at least as discrepant as the observed result. This probability is the p value. It is not the probability that the null hypothesis is true.

This assignment applies that sequence to five tests: the paired t test, the Wilcoxon signed-rank test, the exact sign test, Fisher's exact test, and the chi-squared test of independence. For each test, you will derive the statistic or its null probability and then implement the calculation in R. The acoustic data become relevant in Exercises 1–4. The morphological data become relevant in Exercises 5–6.

All six exercises are required. The points sum to 100.

::: {.callout-note title="Assignment files"}
[Download the complete Problem Set 1 package](../../downloads/ps1-assignment.zip){download="ps1-assignment.zip"}. The package contains the assignment source, the README, the preparation scripts, and the redistributable teaching tables.

The files are also available separately:

- [README and setup instructions](README.md)
- [`prepare_data.R`](scripts/prepare_data.R){download="prepare_data.R"}
- [`fetch_unimorph_data.py`](scripts/fetch_unimorph_data.py){download="fetch_unimorph_data.py"}
- [`prepare_unimorph_tables.R`](scripts/prepare_unimorph_tables.R){download="prepare_unimorph_tables.R"}
:::

Use `<-` for assignment, use `snake_case` object names, and record a seed before every random calculation. Every numerical answer must be produced by visible code.

## Exercise 1: Construct paired differences (10 points)

The data for Exercises 1–4 come from Hillenbrand, Getty, Clark, and Wheeler's study of American English vowels ([1995, JASA](https://doi.org/10.1121/1.411872)). The preparation script extracts the `phonTools::h95` teaching table. Each of the 45 adult men produced /i/ as in *heed* and /ɪ/ as in *hid*. Because the two measurements come from the same speaker, they form a [paired observation](../../statistical-inference/paired-observations.qmd).

Generate `data/hillenbrand_vowels.csv` by following the [README](README.md). The columns used here are `speaker_id`, `speaker_type`, `vowel`, and `f1_hz`, where F1 is the first formant in hertz.

### 1.1 Prepare the paired table

Load the table. Retain rows for adult men and vowels `i` and `I`. Verify that each retained speaker has one row for each vowel. Join the two vowel tables by `speaker_id` and `speaker_type`. Create one row per speaker with `f1_i_hz`, `f1_I_hz`, and

$$
d_s=\operatorname{F1}_{s,/i/}-\operatorname{F1}_{s,/ɪ/}.
$$

Here $d_s$ is the observed difference for speaker $s$.

```{r}
# Load, filter, verify, join, and calculate difference_hz.
```

### 1.2 Summarize the differences

Calculate the number of speakers, the mean, the standard deviation, the median, and the number of negative, zero, and positive differences. Plot the differences with zero marked.

```{r}
# Calculate the requested summaries and plot the paired differences.
```

Report the requested values in the subtraction order /i/ minus /ɪ/.

::: {.callout-note title="Your result"}
Replace this sentence with the sample size and requested summaries.
:::

## Exercise 2: Paired t test (20 points)

Let $D_s:\Omega\to\mathbb R$ be the random /i/ minus /ɪ/ difference for speaker $s$, and let $\mu_D=\mathbb E[D_s]$. Under the [paired normal model](../../statistical-inference/paired-mean-inference.qmd),

$$
D_s\mid\mu_D,\sigma_D\sim\mathcal N(\mu_D,\sigma_D^2).
$$

Test $H_0:\mu_D=0$ against $H_1:\mu_D\ne0$.

### 2.1 Derive the estimator and its standard error

Let $S$ be the number of speakers and $\overline D=S^{-1}\sum_{s=1}^{S}D_s$. Use [linearity of expectation](../../random-variables-and-distributions/linearity-of-expectation.qmd) and independence to derive $\mathbb E[\overline D]$ and $\operatorname{Var}(\overline D)$. Then replace $\sigma_D$ with the sample standard deviation $S_D$ to obtain the estimated standard error.

**Target:** derive $\mathbb E[\overline D]=\mu_D$, $\operatorname{Var}(\overline D)=\sigma_D^2/S$, and $\widehat{\operatorname{SE}}(\overline D)=S_D/\sqrt S$.

**Your LaTeX**

$$
\begin{aligned}
&\text{Enter the three derivations here.}
\end{aligned}
$$

### 2.2 Derive the test statistic

Standardize the difference between $\overline D$ and the null value. Because $\sigma_D$ is estimated by $S_D$, the null reference distribution is a [Student t distribution](../../random-variables-and-distributions/student-t-distribution.qmd) with $S-1$ degrees of freedom.

**Target:** derive the random statistic $\mathcal T$ and its observed value $t_{\mathrm{obs}}$.

**Your LaTeX**

$$
\begin{aligned}
\mathcal T&=\text{enter the random statistic},\\
t_{\mathrm{obs}}&=\text{enter the statistic as a function of }\overline d,s_d,\text{ and }S.
\end{aligned}
$$

### 2.3 Derive the two-sided p value

Define the event containing null outcomes at least as far from zero as the observed statistic. Use full set-builder notation. Then use the cumulative distribution function $F_{t_{S-1}}$ to write its probability.

**Target:** define $B_{\mathcal T}$ and derive $\mathbb P_0(B_{\mathcal T})=2[1-F_{t_{S-1}}(|t_{\mathrm{obs}}|)]$.

**Your LaTeX**

$$
\begin{aligned}
B_{\mathcal T}&=\{\omega\in\Omega\mid\text{enter the two-sided condition}\},\\
\mathbb P_0(B_{\mathcal T})&=\text{enter the tail probability}.
\end{aligned}
$$

### 2.4 Derive the confidence interval

A 95% [confidence interval](../../statistical-inference/confidence-intervals.qmd) is obtained by inverting the central 95% t reference region.

**Target:** write the lower and upper endpoints as functions of $\overline d$, $s_d$, $S$, and $t_{S-1,.975}$.

**Your LaTeX**

$$
\begin{aligned}
L&=\text{enter the lower endpoint},\\
U&=\text{enter the upper endpoint}.
\end{aligned}
$$

### 2.5 Implement the test

Without calling `t.test()`, calculate $\overline d$, $s_d$, the estimated standard error, $t_{\mathrm{obs}}$, the two-sided p value, and the 95% confidence interval.

```{r}
# Implement the paired t calculation from the formulas above.
```

Verify the result with `t.test(..., paired = TRUE)`.

```{r}
# Verify the manual calculation.
```

Report $\overline d$, the interval, $t_{\mathrm{obs}}$, the degrees of freedom, and the p value.

::: {.callout-note title="Your result"}
Replace this sentence with the requested test result.
:::

## Exercise 3: Wilcoxon signed-rank test (10 points)

The paired t statistic uses the numerical magnitudes of the differences. The Wilcoxon signed-rank statistic replaces each nonzero magnitude $|d_s|$ with its rank. It thus uses the order of the magnitudes rather than their spacing.

For $S$ nonzero differences, rank $|d_1|,\ldots,|d_S|$ from 1 to $S$, assigning the mean rank to tied magnitudes. Let $r_s$ be the rank assigned to $|d_s|$. Define

$$
w_+=\sum_{s=1}^{S}r_s\mathbb I(d_s>0),
$$

where $\mathbb I(d_s>0)$ equals one when $d_s>0$ and zero otherwise. Under the signed-rank null model, the differences are independent and symmetric about zero. Conditional on the observed absolute ranks, positive and negative signs are equally probable.

### 3.1 Construct the statistic

Use `rank(abs(difference_hz), ties.method = "average")` to calculate $r_s$. Calculate $w_+$ from the ranks and signs. Display `speaker_id`, `difference_hz`, the absolute difference, its rank, and its sign indicator.

```{r}
# Construct the signed ranks and calculate w_plus.
```

### 3.2 Derive the null mean

Let $B_s:\Omega\to\{0,1\}$ be the random positive-sign indicator under the null model. Then $\mathbb E[B_s]=.5$ and $\mathcal W_+=\sum_{s=1}^{S}r_sB_s$. Use linearity of expectation and $\sum_{s=1}^{S}r_s=S(S+1)/2$ to derive the null mean.

**Target:** derive $\mathbb E_0[\mathcal W_+]=S(S+1)/4$.

**Your LaTeX**

$$
\begin{aligned}
\mathbb E_0[\mathcal W_+]&=\text{enter the derivation here}.
\end{aligned}
$$

### 3.3 Implement the test

Run `wilcox.test()` on the paired F1 measurements with `paired = TRUE`, `alternative = "two.sided"`, `exact = FALSE`, and `correct = FALSE`. The observed integer-valued F1 measurements produce tied absolute differences, so this call uses the large-sample signed-rank reference approximation rather than exact enumeration.

```{r}
# Run the paired Wilcoxon signed-rank test.
```

Report $w_+$, the p value, and the null hypothesis tested by the signed-rank procedure.

::: {.callout-note title="Your result"}
Replace this sentence with the requested signed-rank result.
:::

## Exercise 4: Exact sign test (10 points)

The [Bernoulli distribution](../../random-variables-and-distributions/bernoulli-distribution.qmd) represents whether each difference has the predicted sign. Define

$$
Z_s(\omega)=
\begin{cases}
1 & \text{if }D_s(\omega)<0,\\
0 & \text{if }D_s(\omega)\ge0.
\end{cases}
$$

Let $\pi_D=\mathbb P(Z_s=1)$. Test $H_0:\pi_D=.5$ against $H_1:\pi_D>.5$. There are no zero differences in these data.

### 4.1 Derive the null distribution and p value

Let $\mathcal K=\sum_{s=1}^{S}Z_s$ and $k_{\mathrm{obs}}=\sum_{s=1}^{S}z_s$. Use the [binomial distribution](../../random-variables-and-distributions/binomial-distribution.qmd) to derive the null distribution of $\mathcal K$. Then define the upper-tail event and expand its probability as a sum.

**Target:** derive $\mathcal K\mid H_0\sim\operatorname{Binomial}(S,.5)$ and $\mathbb P_0(\{\omega\in\Omega\mid\mathcal K(\omega)\ge k_{\mathrm{obs}}\})$.

**Your LaTeX**

$$
\begin{aligned}
\mathcal K\mid H_0&\sim\text{enter the null distribution},\\
B_{\mathcal K}&=\{\omega\in\Omega\mid\text{enter the upper-tail condition}\},\\
\mathbb P_0(B_{\mathcal K})&=\text{enter the binomial sum}.
\end{aligned}
$$

### 4.2 Implement the test

Calculate $k_{\mathrm{obs}}$, $k_{\mathrm{obs}}/S$, and the upper-tail probability with `pbinom(..., lower.tail = FALSE)`. Verify the calculation with `binom.test(..., alternative = "greater")` and report the one-sided 95% [exact confidence bound](../../statistical-inference/exact-binomial-intervals.qmd).

```{r}
# Implement and verify the exact sign test.
```

Report the sign count, the observed proportion, the confidence bound, and the p value.

::: {.callout-note title="Your result"}
Replace this sentence with the requested sign-test result.
:::

## Exercise 5: Fisher's exact test (25 points)

The data for this exercise come from the German [UniMorph](https://unimorph.github.io/) paradigm inventory described by [McCarthy et al. (2020)](https://aclanthology.org/2020.lrec-1.483/). The preparation scripts reduce the genitive records to one row per lemma. They record whether a lexeme ends in a string classified as a sibilant and whether a genitive form ending in *-es* is attested. These are variables produced by the scripts, not annotations supplied by UniMorph.

The teaching table contains a reproducible simple random sample of 120 lexemes. Its relevant columns are `lemma`, `ends_in_sibilant`, and `es_attested`.

### 5.1 Construct the two-by-two table

Load `data/unimorph_genitive_lexemes.csv`. Verify that `lemma` is unique. Construct the table for `ends_in_sibilant` by `es_attested`, with `FALSE` before `TRUE` on both dimensions.

```{r}
# Load the lexemes, verify uniqueness, and construct the table.
```

Report the four cells and both sets of margins.

::: {.callout-note title="Your result"}
Replace this sentence with the observed table and its margins.
:::

### 5.2 Derive the conditional null probability

[Fisher's exact test](../../statistical-inference/fishers-exact-test.qmd) tests independence in a two-by-two table while holding both observed margins fixed. Write the observed cells in row-major order as $a,b,c,d$. Let $A$ be the random upper-left cell count among all tables with those margins. Under the null hypothesis of independence, $A$ has a [hypergeometric distribution](../../random-variables-and-distributions/hypergeometric-distribution.qmd).

Derive the probability of the observed table by counting the ways to select its first-column total from the two rows and dividing by the number of selections from all observations.

**Target:** derive $\mathbb P_0(A=a)$ as a ratio of binomial coefficients containing $a,b,c,d$.

**Your LaTeX**

$$
\begin{aligned}
\mathbb P_0(A=a)&=\text{enter the ratio of binomial coefficients}.
\end{aligned}
$$

### 5.3 Implement the test

Calculate the sample odds ratio $ad/(bc)$ and the observed-table probability with `dhyper()`. Run `fisher.test()` to obtain the two-sided p value and conditional confidence interval. Fisher's two-sided p value sums the probabilities of all margin-compatible tables whose null probabilities are no greater than the observed table's probability.

```{r}
# Calculate the odds ratio, the observed-table probability, and Fisher's test.
```

Report the odds-ratio orientation, the sample odds ratio, the conditional estimate returned by `fisher.test()`, its confidence interval, and the p value.

::: {.callout-note title="Your result"}
Replace this sentence with the requested Fisher-test result.
:::

### 5.4 Estimate sampling distributions across lexicon sizes

A [sampling distribution](../../statistical-inference/properties-of-estimators.qmd) is the distribution of a statistic across repeated samples from a fixed population. For this exercise, treat `data/unimorph_genitive_lexeme_population.csv` as the finite population.

For each $n\in\{12,20,40,80,120\}$, draw 500 simple random samples without replacement. Use one recorded seed. For every sample, record the odds ratio, Fisher p value, confidence-interval endpoints, whether any cell is zero, and whether either margin is zero. Use explicit factor levels when constructing every table.

```{r}
# Write one function that returns one tidy row, then run the complete sweep.
```

For each $n$, calculate the proportion of samples with a zero cell, the proportion with a zero margin, the proportion with a nonfinite odds ratio, the median finite log odds ratio, and the median finite confidence-interval width on the log scale. Plot the sampling distributions of the finite log odds ratios by $n$.

```{r}
# Summarize and plot the empirical sampling distributions.
```

Report the summary table.

::: {.callout-note title="Your result"}
Replace this sentence with the sample-size summary table.
:::

## Exercise 6: Chi-squared test of independence (25 points)

The second UniMorph table has one row per noun lemma. The preparation rule retains one plural form and classifies its ending as `e`, `en`, `er`, `s`, or `ø`. Here `ø` means that the rule found none of the listed overt suffixes. It is a label produced by the script, not a complete analysis of German plural exponence.

Load `data/unimorph_plural_lexemes.csv`. The relevant columns are `lemma`, `gender`, and `plural_suffix`. Construct the observed three-by-five table with gender ordered `FEM`, `MASC`, `NEUT` and suffix ordered `e`, `en`, `er`, `s`, `ø`.

```{r}
# Load the lexemes, verify lemma uniqueness, and construct the observed table.
```

### 6.1 Derive the expected counts

Under independence, the expected count in cell $(i,j)$ is determined by its row total, column total, and the grand total. Follow the derivation in the [chi-squared-test notes](../../statistical-inference/chi-squared-test.qmd).

**Target:** derive $E_{ij}=O_{i+}O_{+j}/O_{++}$ and substitute the observed margins for the feminine-by-*en* cell.

**Your LaTeX**

$$
\begin{aligned}
E_{ij}&=\text{enter the general derivation},\\
E_{\mathrm{FEM},\mathrm{en}}&=\text{enter the numerical substitution and result}.
\end{aligned}
$$

### 6.2 Derive the statistic and reference distribution

Pearson's statistic sums the standardized squared difference between every observed and expected count. If the table has $R$ rows and $C$ columns, estimating the margins leaves $(R-1)(C-1)$ degrees of freedom.

**Target:** write $\mathcal X^2$, derive $\nu=(R-1)(C-1)$, define the upper-tail event in full set-builder notation, and write its probability under $\chi^2_\nu$.

**Your LaTeX**

$$
\begin{aligned}
\mathcal X^2&=\text{enter Pearson's statistic},\\
\nu&=\text{enter the degrees of freedom},\\
B_{\mathcal X^2}&=\{\omega\in\Omega\mid\text{enter the upper-tail condition}\},\\
\mathbb P_0(B_{\mathcal X^2})&=\text{enter the chi-squared tail probability}.
\end{aligned}
$$

### 6.3 Implement the test

Without calling `chisq.test()`, calculate the complete expected table, every cell contribution, $x^2_{\mathrm{obs}}$, the degrees of freedom, and the p value. Verify that every expected count is at least five.

```{r}
# Implement the chi-squared test from the formulas above.
```

Verify the statistic, degrees of freedom, p value, and expected table with `chisq.test(..., correct = FALSE)`. Calculate the Pearson residuals $(O_{ij}-E_{ij})/\sqrt{E_{ij}}$ and identify the two cells with the largest absolute residuals.

```{r}
# Verify the test and calculate the Pearson residuals.
```

Report $x^2_{\mathrm{obs}}$, the degrees of freedom, the p value, the minimum expected count, and the two largest absolute residuals with their signs.

::: {.callout-note title="Your result"}
Replace this sentence with the requested chi-squared-test result.
:::

## Submission checklist

- The notebook runs from a fresh R session after the data-preparation steps.
- Every derivation is entered in its LaTeX block.
- Every computation is entered in the R block immediately following its instruction.
- Every requested result is entered in its result block.
- The paired table is joined by identifiers rather than aligned by row position.
- Every hypothesis names its parameter or independence relation.
- Every p value is defined relative to its null reference distribution.
- Every random calculation has a recorded seed.
