---
title: "Problem Set 1: Paired acoustic measurements"
subtitle: "Statistical Methods in Linguistics"
format:
  html:
    toc: true
execute:
  enabled: false
---

## The question

The first formant (F1) is an acoustic correlate of vowel height: higher vowels tend to have lower F1. We will ask whether F1 differs between /i/ as in *heed* and /ɪ/ as in *hid* in the Hillenbrand vowel sample.

Every recorded speaker produced both vowels. Thus the theoretically relevant predictor is vowel category, while speaker identifies which two measurements form a pair. The focal method is [inference for a paired mean difference](../../statistical-inference/paired-mean-inference.qmd): construct one within-speaker difference and estimate its population mean.

By the end of the assignment, you should be able to (i) identify a contrast-bearing unit from a repeated design, (ii) compute and interpret a paired t interval, and (iii) explain why an analysis that ignores the pairing answers the question less efficiently.

::: {.callout-note title="Reading"}
Read Hillenbrand, Getty, Clark, and Wheeler (1995), [“Acoustic characteristics of American English vowels”](https://doi.org/10.1121/1.411872). Focus on the speaker sample, recording procedure, and formant measurements. The reading is included because the paired analysis follows from the recording design: each speaker contributes both vowel measurements, and the within-speaker contrast is the unit that bears on vowel category.
:::

Before beginning, review [turning linguistic records into data](../../foundations/linguistic-data.qmd), [paired observations](../../statistical-inference/paired-observations.qmd), and [inference for a paired mean difference](../../statistical-inference/paired-mean-inference.qmd). These pages establish the unit of analysis, the pairing relation, and the confidence interval used below.

## Data and source

The data derive from Hillenbrand, Getty, Clark, and Wheeler's study of American English vowels ([1995, JASA](https://doi.org/10.1121/1.411872)) and are loaded from the [`phonTools::h95` teaching table](https://search.r-project.org/CRAN/refmans/phonTools/html/h95.html). The preparation script retains all 12 vowels for 139 speakers. You will select the two vowels required by the question.

Generate `data/hillenbrand_vowels.csv` by following the instructions in `README.md`. Then load it.

```{r}
vowels <- read.csv("data/hillenbrand_vowels.csv")
head(vowels)
```

The relevant columns are:

- `speaker_id`: a unique speaker identifier;
- `speaker_type`: boy, girl, man, or woman;
- `vowel`: the X-SAMPA vowel category;
- `f1_hz`: steady-state F1 in Hz; and
- `duration_ms`, `f0_hz`, `f2_hz`, and `f3_hz`: measurements not needed for the focal analysis.

## Task 1: establish the design (20 points)

1. State the measurement unit, the pair identifier, and the unit of analysis for the contrast between /i/ and /ɪ/.
2. Verify that each speaker has exactly one row for `i` and one row for `I`. Report the number of complete pairs.
3. Explain why 278 vowel tokens do not provide 278 independent observations of the vowel contrast.

```{r}
# Inspect the speaker-by-vowel counts and verify the paired subset.
```

## Task 2: construct the within-speaker response (20 points)

Create a table with one row per speaker and the columns `f1_i_hz`, `f1_I_hz`, and

$$
d_s=\operatorname{F1}_{s,/i/}-\operatorname{F1}_{s,/ɪ/}.
$$

Join the two vowel tables by `speaker_id` and `speaker_type`; do not rely on their current row order. Plot the differences. Your plot should make the zero point and the direction of the contrast visible.

```{r}
# Construct the paired table by joining on identifiers.
```

```{r}
# Plot the within-speaker differences.
```

Describe the distribution in two or three sentences. Address its center, variation, and any observations that might strongly affect the mean.

## Task 3: inference for the paired mean difference (30 points) {#task-3-paired-mean-inference-30-points}

Compute the following quantities from the difference vector without calling `t.test()`:

1. the mean difference $\overline{d}$;
2. the standard deviation $s_d$;
3. the standard error $s_d/\sqrt{S}$;
4. the 95% t confidence interval with $S-1$ degrees of freedom; and
5. the t statistic and two-sided p value for $H_0:\mu_d=0$.

```{r}
# Manual paired-mean calculation.
```

Verify the calculation with `t.test(..., paired = TRUE)`. Show that its estimate, interval, statistic, and p value agree with your manual result to numerical precision.

```{r}
# Software verification.
```

Interpret the estimate and interval in Hz. Your interpretation must preserve the subtraction order; do not report only that the vowels are “significantly different.”

## Task 4: diagnose the independence error (20 points)

Run the following deliberately inadequate analysis, which treats the two vowel samples as independent.

```{r}
unpaired_fit <- t.test(
  pairs$f1_i_hz,
  pairs$f1_I_hz,
  paired = FALSE
)
unpaired_fit
```

1. Compare its point estimate, standard error, and 95% interval with the paired result.
2. Compute the within-speaker correlation between the two F1 measurements.
3. Use

$$
s_d^2=s_i^2+s_I^2-2s_{iI}
$$

to explain why pairing changes the standard error in this dataset.
4. Explain why the paired analysis is required by the design even if the correlation had been near zero.

## Task 5: scope the claim (10 points)

Write a four-to-six sentence results paragraph. It must state the unit, estimate, interval, and direction of the contrast. End by stating a defensible generalization target without describing the study speakers as a probability sample of all American English speakers.

## LING 414 extension: resample the correct unit

Construct a 95% percentile [bootstrap interval](../../statistical-inference/bootstrap-confidence-intervals.qmd) for the mean difference. Resample whole speakers by sampling rows of the paired table, not individual vowel tokens. Use at least 5,000 bootstrap replicates and report the random seed.

Compare the bootstrap and t intervals. If they are similar, say what feature of the observed difference distribution makes that agreement plausible. If they differ, identify which feature may account for the difference.

## Submission checklist

- The notebook runs from a fresh R session after the data-preparation step.
- Every numerical answer is produced by visible code.
- The paired table is joined by identifiers rather than aligned by row position.
- Figures label the response and its units.
- Interpretations name the subtraction order and target population.
