---
title: "Rank one structure"
---

Suppose three speakers are measured on three speech-rate variables. The speakers differ in their overall rates, but every variable orders them in exactly the same proportional way. The nine cells may then contain only one shared pattern repeated at different strengths.

A nonzero matrix has [**rank one**](https://online.stat.psu.edu/stat508/Lesson07.html) when it can be written as the outer product of one vector with an entry for each row and another vector with an entry for each column. This representation separates variation across objects from variation across variables without assigning either vector a new substantive meaning.

## Constructing one outer product

Let

$$
\mathbf{a}
=
\begin{bmatrix}
1\\2\\3
\end{bmatrix}
$$

record the relative rate of three speakers, and let

$$
\mathbf{b}^{\mathsf T}
=
\begin{bmatrix}
1&2&3
\end{bmatrix}
$$

record the relative contribution of three rate measurements. Their outer product is

$$
\mathbf{a}\mathbf{b}^{\mathsf T}
=
\begin{bmatrix}
a_1b_1&a_1b_2&a_1b_3\\
a_2b_1&a_2b_2&a_2b_3\\
a_3b_1&a_3b_2&a_3b_3
\end{bmatrix}.
$$

Substituting the vector entries gives

$$
\mathbf{X}
=
\begin{bmatrix}
1(1)&1(2)&1(3)\\
2(1)&2(2)&2(3)\\
3(1)&3(2)&3(3)
\end{bmatrix}
=
\begin{bmatrix}
1&2&3\\
2&4&6\\
3&6&9
\end{bmatrix}.
$$

Every cell obeys the same rule:

$$
x_{ij}=a_i b_j.
$$

The entry $x_{23}=6$ is not a separate parameter. It follows from $a_2=2$ and $b_3=3$.

## Reading the two vectors

The vector $\mathbf{a}$ has one entry per row. In the speech-rate interpretation, it says how strongly each speaker expresses the common pattern. The entry for speaker 3 is three times the entry for speaker 1.

The vector $\mathbf{b}$ has one entry per column. It says how each measurement participates in the pattern. The third measurement has three times the column weight of the first.

These interpretations are relative. We could multiply $\mathbf{a}$ by 2 and divide $\mathbf{b}$ by 2:

$$
(2\mathbf{a})
\left(
\frac{1}{2}\mathbf{b}
\right)^{\mathsf T}
=
\mathbf{a}\mathbf{b}^{\mathsf T}.
$$

Thus, the individual vector scales are not identified by the outer product. Their product is. Later decompositions use a normalization convention to separate direction from magnitude.

## Deriving proportional rows and columns

The $i$th row of a rank one matrix is

$$
\mathbf{x}_i^{\mathsf T}
=
a_i\mathbf{b}^{\mathsf T}.
$$

Every row is thus a multiple of the same row pattern $\mathbf{b}^{\mathsf T}$. In the worked matrix,

$$
\mathbf{x}_2^{\mathsf T}
=
2\mathbf{x}_1^{\mathsf T}
$$

and

$$
\mathbf{x}_3^{\mathsf T}
=
3\mathbf{x}_1^{\mathsf T}.
$$

The same reasoning applies to columns:

$$
\mathbf{x}_{\cdot j}
=
b_j\mathbf{a}.
$$

Every column is a multiple of $\mathbf{a}$. This double proportionality is what lets one outer product generate the complete matrix.

If two nonzero rows are not proportional, one outer product cannot represent both. The matrix then has rank greater than one.

## Counting what has been compressed

The $3\times 3$ matrix contains nine cells. Its rank one representation uses three entries in $\mathbf{a}$ and three entries in $\mathbf{b}$, subject to the scaling ambiguity described above.

The important reduction is structural rather than merely numerical. Instead of treating every cell as unrelated, the representation states that all cells arise from row strength times column participation.

For an $n\times p$ matrix, this rule uses an $n$-entry row vector and a $p$-entry column vector. When $n$ and $p$ are large, one shared pattern may summarize a substantial collection of cells.

## Computing the outer product in R

The function `outer()` implements the cell rule $x_{ij}=a_i b_j$.

```{r}
#| eval: false

a <- c(1, 2, 3)
b <- c(1, 2, 3)

X_rank_one <- outer(a, b)
X_rank_one
```

Matrix multiplication gives the same result when the vector shapes are explicit.

```{r}
#| eval: false

a_column <- matrix(a, ncol = 1)
b_row <- matrix(b, nrow = 1)

a_column %*% b_row
```

We can check the numerical rank with a QR decomposition.

```{r}
#| eval: false

qr(X_rank_one)$rank
```

The result is one. The computation verifies the algebraic property of this matrix. It does not supply a linguistic interpretation of the vectors.

## Distinguishing exact structure from approximation

Real linguistic measurements will almost never satisfy exact proportionality. Suppose the bottom-right entry is measured as 9.1 rather than 9. The displayed rows are no longer exact multiples, and the numerical matrix may have rank greater than one.

This does not make rank one structure useless. It changes the question from exact representation to approximation: how much of the matrix can one outer-product pattern reconstruct, and where are the remaining errors? The [next page](singular-value-decomposition.qmd) supplies an ordered way to answer that question.

We should not use that later machinery prematurely. At this point, rank one means exactly that one outer product generates the nonzero matrix.

## Rank is not a psychological construct

A potential interpretation is that one rank one pattern proves the existence of one mental variable. This interpretation does not follow from the algebra.

Matrix rank is a property of the observed numerical representation. A shared recording artifact could create proportional columns. A theoretical construct could produce a matrix with higher observed rank because the measurements contain additional variation. And changing the chosen columns could change the rank without changing the speakers.

A low-rank pattern may motivate a substantive hypothesis, but the factorization alone does not identify a linguistic construct or the process that generated the pattern.

Numerical rank also depends on a tolerance. Tiny rounding errors can turn an intended zero into a very small nonzero value. Software thus decides whether small directions count as effectively zero. Exact algebra and numerical computation should not be conflated.

::: {.callout-note title="Problem Set 5 connection"}
[Problem Set 5, Task 6](../problem-sets/ps5/ps5.qmd#task-6-select-the-rank-on-validation-cells-15-points) builds a reconstruction by adding rank one patterns to a segment prevalence baseline. Every added term must be interpreted as the combination of one inventory score vector and one segment loading vector.
:::

## Checking the common pattern

1. In the worked matrix, which entries of $\mathbf{a}$ and $\mathbf{b}$ generate $x_{32}=6$?
2. Why is every row of $\mathbf{a}\mathbf{b}^{\mathsf T}$ proportional to $\mathbf{b}^{\mathsf T}$?
3. What happens to the outer product if $\mathbf{a}$ is doubled and $\mathbf{b}$ is halved?
4. Why would changing one cell from 9 to 9.1 usually remove exact rank one structure?
5. What additional evidence would be needed to interpret the common pattern as a psychological construct?

A rank one matrix is generated by one vector for its rows and one vector for its columns. The outer product explains every cell through their multiplication, but it does not identify the linguistic construct that produced the pattern. The [next page](singular-value-decomposition.qmd) represents any real matrix as an ordered sum of such rank one patterns.
