---
title: "Expectation-maximization"
---

If the component label of every vowel token's acoustic vector were observed, we could estimate each component from its assigned tokens. If the component parameters were known, we could calculate the posterior membership of every token. In a mixture model, neither side is known at the start. We need an algorithm that lets each provisional answer update the other.

The [**expectation-maximization (EM) algorithm**](https://scikit-learn.org/stable/modules/mixture.html) fits a latent-variable model by alternating between two calculations. The expectation step forms the conditional expectation of the complete-data log likelihood under the current parameters. For a Gaussian mixture, this calculation requires the responsibilities from the preceding page. The maximization step then chooses parameters that maximize that expected complete-data log likelihood.

## Beginning with provisional parameters

Let the current mixture parameters be

$$
\boldsymbol{\theta}^{(t)}
=
\left\{
\pi_k^{(t)},
\boldsymbol{\mu}_k^{(t)},
\boldsymbol{\Sigma}_k^{(t)}
\right\}_{k=1}^{K},
$$

where $t$ indexes the current iteration. These values may come from random starting assignments or another declared initialization rule.

The superscript does not index data. It records the repeated algorithmic update.

## Expanding the complete-data objective

Why is the first update called an expectation step? Suppose for a moment that every component label $Z_i$ were observed. Define the indicator

$$
I_{ik}
=
\mathbf{1}[Z_i=k].
$$

For each token, exactly one of the $K$ indicators would equal one. The complete-data log likelihood could then be written as

$$
\ell_{\mathrm{complete}}(\boldsymbol{\theta})
=
\sum_{i=1}^{N}\sum_{k=1}^{K}
I_{ik}
\left[
\log\pi_k
+
\log\mathcal{N}
(\mathbf{x}_i;\boldsymbol{\mu}_k,\boldsymbol{\Sigma}_k)
\right].
$$

This expression is compressed bookkeeping. The indicator retains the weight-and-density term for the token's component and multiplies every other component term by zero. But the $I_{ik}$ values are not observed, so we cannot maximize this objective directly.

## The expectation step uses responsibilities

Given $\boldsymbol{\theta}^{(t)}$, compute

$$
r_{ik}^{(t)}
=
P(Z_i=k\mid\mathbf{x}_i,\boldsymbol{\theta}^{(t)}).
$$

Using the formula from the previous page,

$$
r_{ik}^{(t)}
=
\frac{
\pi_k^{(t)}
\mathcal{N}
(\mathbf{x}_i;
\boldsymbol{\mu}_k^{(t)},
\boldsymbol{\Sigma}_k^{(t)})
}{
\sum_{j=1}^{K}
\pi_j^{(t)}
\mathcal{N}
(\mathbf{x}_i;
\boldsymbol{\mu}_j^{(t)},
\boldsymbol{\Sigma}_j^{(t)})
}.
$$

Every row of responsibilities sums to one. A token with $(0.8,0.2)$ contributes mostly to component 1 in the next update but still contributes to component 2.

These responsibilities provide the conditional probabilities needed to average over the unobserved component labels. This average is the expectation in the name of the algorithm. It replaces each unknown hard label with a probability vector under the current fit.

More specifically,

$$
E[I_{ik}\mid\mathbf{x}_i,\boldsymbol{\theta}^{(t)}]
=
r_{ik}^{(t)}.
$$

Replacing each unknown indicator by this conditional expectation gives the EM auxiliary objective

$$
Q(\boldsymbol{\theta}\mid\boldsymbol{\theta}^{(t)})
=
\sum_{i=1}^{N}\sum_{k=1}^{K}
r_{ik}^{(t)}
\left[
\log\pi_k
+
\log\mathcal{N}
(\mathbf{x}_i;\boldsymbol{\mu}_k,\boldsymbol{\Sigma}_k)
\right].
$$

The maximization step holds the responsibilities fixed and chooses new parameters to increase this expected complete-data log likelihood. This is the bridge between the membership probabilities and the weighted updates below.

## The maximization step updates the weights

The effective number of observations assigned to component $k$ is

$$
N_k^{(t)}
=
\sum_{i=1}^{N}
r_{ik}^{(t)}.
$$

The updated mixing weight is

$$
\pi_k^{(t+1)}
=
\frac{N_k^{(t)}}{N}.
$$

The weights still sum to one because

$$
\sum_{k=1}^{K}
N_k^{(t)}
=
\sum_{i=1}^{N}
\sum_{k=1}^{K}
r_{ik}^{(t)}
=
N.
$$

Thus, dividing each effective count by $N$ gives a valid set of mixture proportions.

## Updating the component means

The updated mean is the responsibility-weighted average

$$
\boldsymbol{\mu}_k^{(t+1)}
=
\frac{1}{N_k^{(t)}}
\sum_{i=1}^{N}
r_{ik}^{(t)}\mathbf{x}_i.
$$

Tokens with large responsibility for component $k$ have greater influence on its center. Tokens with responsibility near zero contribute little.

Suppose one one-dimensional component has responsibilities 0.9, 0.6, and 0.1 for observations 2, 4, and 10. Its effective count is

$$
N_k
=
0.9+0.6+0.1
=
1.6.
$$

The weighted sum is

$$
0.9(2)+0.6(4)+0.1(10)
=
1.8+2.4+1.0
=
5.2.
$$

The updated mean is

$$
\mu_k^{\mathrm{new}}
=
\frac{5.2}{1.6}
=
3.25.
$$

The observation at 10 is far from the new mean, but it receives little weight because its current responsibility is only 0.1.

## Updating the covariance

For an unrestricted Gaussian component, the covariance update is

$$
\boldsymbol{\Sigma}_k^{(t+1)}
=
\frac{1}{N_k^{(t)}}
\sum_{i=1}^{N}
r_{ik}^{(t)}
\left(
\mathbf{x}_i-
\boldsymbol{\mu}_k^{(t+1)}
\right)
\left(
\mathbf{x}_i-
\boldsymbol{\mu}_k^{(t+1)}
\right)^{\mathsf T}.
$$

This is a responsibility-weighted average of outer products around the updated mean. Covariance restrictions change this update. A shared covariance pools information across components, while a spherical covariance retains only a restricted scale.

The chosen update must match the Gaussian component geometry declared in the model.

## Reproducing the mean update in R

```{r}
#| eval: false

x <- c(2, 4, 10)
responsibility <- c(0.9, 0.6, 0.1)

effective_count <- sum(responsibility)

updated_mean <-
  sum(responsibility * x) /
  effective_count

effective_count
updated_mean
```

A one-dimensional covariance update follows the same weighting rule.

```{r}
#| eval: false

updated_variance <-
  sum(
    responsibility *
    (x - updated_mean)^2
  ) /
  effective_count

updated_variance
```

These lines implement one maximization step after the responsibilities have been supplied. A complete EM implementation alternates this update with a new expectation step.

## Alternating until the fit stabilizes

After the maximization step, recompute responsibilities using the updated parameters. Continue until the observed-data log likelihood or parameter changes fall below a declared tolerance.

The observed-data log likelihood is

$$
\ell(\boldsymbol{\theta})
=
\sum_{i=1}^{N}
\log
\left[
\sum_{k=1}^{K}
\pi_k
\mathcal{N}
(\mathbf{x}_i;
\boldsymbol{\mu}_k,
\boldsymbol{\Sigma}_k)
\right].
$$

Exact EM updates do not decrease this likelihood. The expectation step constructs the expected complete-data log likelihood under the current parameters, and an exact maximization step selects parameters that maximize that objective. This is a likelihood-optimization result of the sort introduced in [maximum likelihood](../statistical-inference/maximum-likelihood.qmd), now applied when component labels are unobserved.

This monotonicity is an algorithmic guarantee. It is not a guarantee that the largest possible likelihood has been found.

## Several starts are necessary

Mixture likelihoods can have several local maxima. Different provisional parameters may lead EM to different component locations or covariance structures.

The expectation-maximization algorithm can converge to different local optima from different initializations. Fit the same declared model from several starts, retain the valid solution with the largest attained likelihood, and record whether substantively different solutions recur.

A fixed random seed makes the collection of starts reproducible. It does not make one start sufficient.

## Monitoring for collapsing components

EM may move one component toward a small number of observations while shrinking its covariance. The likelihood can become extremely large as the covariance approaches singularity.

This is not evidence for a sparsely represented vowel category. It is the likelihood singularity from the previous page appearing during fitting. The implementation should enforce appropriate covariance restrictions or regularization and flag degenerate solutions.

Retaining the fit with the largest likelihood could select the collapsed solution. This is why validity checks precede likelihood comparison. An invalid degenerate solution is not a candidate interpretation.

## EM estimates a local representation

At convergence, the responsibilities and parameters agree under the alternating updates. The result is a local maximum or stationary solution for the declared mixture.

It does not establish the correct component count, covariance family, or linguistic category interpretation. Those are comparison and interpretation questions, not consequences of convergence. In particular, an algorithmically stable acoustic component is not thereby a phonological category.

## Checking the alternating algorithm

1. Which quantities are calculated in the expectation step?
2. Why does an observation with responsibility 0.1 still affect a component mean?
3. How is the updated mean 3.25 obtained in the worked calculation?
4. Why do updated mixing weights sum to one?
5. What does monotonic observed-data likelihood guarantee, and what does it not guarantee?
6. Why should the same mixture be fit from several initializations?
7. Why must a covariance-collapse fit be rejected before comparing attained likelihoods?

EM replaces unobserved component indicators with their conditional expectations, then alternates posterior membership calculation with responsibility-weighted parameter updates. It improves the observed-data likelihood from a provisional starting point but may reach a local or degenerate solution. The [next page](choosing-a-mixture.qmd) compares component counts and covariance assumptions using BIC's penalty for fitted flexibility.
