---
title: "Posterior membership"
---

A Gaussian mixture assigns a density to every vowel token. We often also want to know how the weighted component contributions divide for one particular token. A token near overlapping vowel regions may be compatible with several components even when one contributes most.

The [**posterior probability of component membership**](https://scikit-learn.org/stable/modules/mixture.html) gives the probability of each component label conditional on the observed measurements and fitted mixture parameters. It applies the [Bayes' rule introduced earlier](../foundations/bayes-rule.qmd) to the weighted densities from the previous page, producing probabilities that sum to one. In mixture-model software, these conditional probabilities are often called responsibilities.

We first derive the formula from Bayes' rule. We then finish the two-component density calculation by hand and in R, distinguish soft from hard assignment, and explain why membership probability is conditional on the fitted model rather than a probability that a phonological analysis is true.

## Introduce the unobserved label

Let

$$
Z_i\in\{1,\ldots,K\}
$$

be the unobserved component label for observation $i$. Before observing $\mathbf{x}_i$, the mixture gives

$$
P(Z_i=k)=\pi_k.
$$

Conditional on component $k$, the observation density is

$$
p(\mathbf{x}_i\mid Z_i=k)
=
\mathcal{N}
(\mathbf{x}_i;
\boldsymbol{\mu}_k,
\boldsymbol{\Sigma}_k).
$$

Bayes' rule gives

$$
P(Z_i=k\mid\mathbf{x}_i)
=
\frac{
P(Z_i=k)
p(\mathbf{x}_i\mid Z_i=k)
}{
p(\mathbf{x}_i)
}.
$$

Substituting the mixture terms yields

$$
P(Z_i=k\mid\mathbf{x}_i)
=
\frac{
\pi_k
\mathcal{N}
(\mathbf{x}_i;
\boldsymbol{\mu}_k,
\boldsymbol{\Sigma}_k)
}{
\sum_{j=1}^{K}
\pi_j
\mathcal{N}
(\mathbf{x}_i;
\boldsymbol{\mu}_j,
\boldsymbol{\Sigma}_j)
}.
$$

We write this probability as

$$
r_{ik}
=
P(Z_i=k\mid\mathbf{x}_i).
$$

The quantity $r_{ik}$ is the responsibility of component $k$ for observation $i$.

## Finish the previous calculation

The two weighted component contributions were

$$
0.048
\qquad\text{and}\qquad
0.008.
$$

Their sum, the mixture density, was 0.056. The first posterior membership is

$$
\begin{aligned}
r_{i1}
&=
\frac{0.048}{0.056}\\
&\approx0.857.
\end{aligned}
$$

The second is

$$
\begin{aligned}
r_{i2}
&=
\frac{0.008}{0.056}\\
&\approx0.143.
\end{aligned}
$$

They sum to one:

$$
0.857+0.143=1.
$$

Component 1 accounts for most of the fitted density at this observation, but component 2 remains possible under the model.

## Why the probabilities sum to one

The denominator is the sum of all component contributions. Thus,

$$
\sum_{k=1}^{K}r_{ik}
=
\frac{
\sum_{k=1}^{K}
\pi_k
\mathcal{N}
(\mathbf{x}_i;
\boldsymbol{\mu}_k,
\boldsymbol{\Sigma}_k)
}{
\sum_{j=1}^{K}
\pi_j
\mathcal{N}
(\mathbf{x}_i;
\boldsymbol{\mu}_j,
\boldsymbol{\Sigma}_j)
}
=
1.
$$

Membership is thus a normalization of the weighted densities. It does not require a separate distance rule.

## Calculate membership in R

```{r}
#| eval: false

weighted_contribution <- c(0.048, 0.008)

responsibility <-
  weighted_contribution /
  sum(weighted_contribution)

responsibility
sum(responsibility)
```

For many observations, store weighted contributions in a matrix with one row per token and one column per component. Divide each row by its sum.

```{r}
#| eval: false

responsibility_matrix <-
  weighted_density_matrix /
  rowSums(weighted_density_matrix)

head(responsibility_matrix)
```

The row normalization implements the denominator in Bayes' rule.

## Hard assignment discards uncertainty

A hard assignment chooses

$$
\widehat{z}_i
=
\underset{k}{\operatorname{argmax}}
r_{ik}.
$$

The worked token is assigned to component 1. But the hard label hides the difference between memberships $(0.51,0.49)$ and $(0.99,0.01)$. Both yield label 1 even though the first observation lies near a modeled boundary and the second does not.

For Peterson and Barney vowel tokens, retaining the complete membership vector can reveal which intended vowels occupy overlapping formant regions and which individual tokens the fitted mixture represents uncertainly.

Converting posterior membership probabilities to one hard label removes information about component overlap.

## Keep assignment and uncertainty separate

A useful descriptive table contains both the most probable component and its probability.

| token | assigned component | largest membership |
|:--|--:|--:|
| A | 1 | 0.98 |
| B | 1 | 0.54 |
| C | 2 | 0.91 |

Token B deserves inspection because its assignment is weak under the fitted model. A plot colored only by hard component would conceal this difference.

The complete probability vector is preferable when more than two components overlap. The largest probability alone does not show whether the remaining mass belongs to one close competitor or several.

## Component numbers are arbitrary

If every parameter associated with component 1 is exchanged with the corresponding parameter for component 2, the mixture density is unchanged. Membership columns are exchanged as well.

This permutation nonidentifiability is known as [**label switching**](https://mc-stan.org/docs/stan-users-guide/finite-mixtures.html). The number attached to a component has no stable linguistic meaning. Across separate fits, component 1 may correspond to a differently numbered component with similar parameters.

Interpret a component through its mean, covariance, membership pattern, and relation to external variables. Do not interpret the integer label itself.

## Membership is conditional on the model

A posterior membership of 0.857 is not the probability that a token truly belongs to a phoneme. It is the probability of the fitted component label conditional on the measurements, component count, Gaussian form, covariance restrictions, transformations, and estimated parameters.

If the model omits a relevant formant or uses an unsuitable covariance structure, the membership may be sharply estimated under a poor representation. High numerical certainty does not validate the component interpretation.

These memberships are not arbitrary. They are exact consequences of the declared mixture. Their interpretation should retain those conditions and be checked against external phonetic information.

## Check the normalized contributions

1. Which quantity appears in the denominator of every membership probability?
2. Why do the memberships for one token sum to one?
3. How are 0.857 and 0.143 obtained from the previous weighted densities?
4. What information is lost when $(0.51,0.49)$ becomes hard label 1?
5. Why can component 1 in one fit correspond to component 3 in another fit?
6. Why is a membership of 0.99 not a 99 percent probability that a phoneme analysis is true?

## Before fitting the parameters

Posterior membership normalizes weighted component densities through Bayes' rule. It retains overlap that a hard assignment discards, while remaining conditional on the declared mixture model. The [next page](expectation-maximization.qmd) uses these responsibilities to alternate between membership calculation and parameter updates.
