---
title: "Within-cluster loss"
---

K-means requires the number of centroids $K$ before fitting. A small $K$ may force distinct speaker profiles into one group. A large $K$ may assign separate centroids to irregularities that do not recur. Recall from [training and test data](../model-criticism-and-prediction/training-and-test-data.qmd) that training loss rewards flexibility on the observations used for fitting. It thus cannot identify a true group count.

Plotting the fitted within-cluster loss against $K$ produces the standard [**elbow plot**](https://online.stat.psu.edu/stat505/Lesson14). The [**elbow method**](https://online.stat.psu.edu/stat505/Lesson14) looks for a point after which another centroid produces a much smaller reduction than earlier centroids did.

## Defining the loss at each group count

For a fitted K-means partition with $K$ centroids, write

$$
W_K
=
\sum_{i=1}^{N}
\left\lVert
\mathbf{x}_i-\boldsymbol{\mu}_{z_i,K}
\right\rVert^2.
$$

The subscript $K$ records that both the assignments and centroids were fit under that candidate count. Lower $W_K$ means that the rows lie closer to their assigned centroids on the training representation.

To measure what the next centroid adds, define the reduction

$$
\Delta_K
=
W_{K-1}-W_K
$$

for $K\geq2$. A larger $\Delta_K$ means that moving from $K-1$ to $K$ centroids substantially improves fitted compactness.

## Why the optimum curve cannot rise

Suppose we have a fitted solution with $K$ centroids. A model with $K+1$ centroids can reproduce that solution by placing the additional centroid on an existing centroid and leaving the assignments unchanged.

Since the $K+1$ model can do at least this well and may find a better partition,

$$
W_{K+1}\leq W_K.
$$

Thus, training loss cannot tell us to stop by reaching a minimum and then rising. Its minimum occurs at the largest permitted $K$, including the degenerate case of one centroid per distinct row.

The globally minimized within-cluster loss decreases monotonically as $K$ increases because greater flexibility is evaluated on the same observations used to fit it. A reported K-means curve can still rise when the fitting algorithm reaches a poorer [local minimum](k-means.qmd#initial-values-can-change-the-solution) for one value of $K$. Multiple starts reduce that computational problem but do not change the monotonicity of the global optimum.

## Calculating an elbow curve

Suppose repeated K-means fits give

| $K$ | $W_K$ |
|---:|---:|
| 1 | 500 |
| 2 | 280 |
| 3 | 190 |
| 4 | 165 |
| 5 | 148 |

The first reduction is

$$
\Delta_2
=
500-280
=
220.
$$

The remaining reductions are

$$
\Delta_3=280-190=90,
$$

$$
\Delta_4=190-165=25,
$$

and

$$
\Delta_5=165-148=17.
$$

The move from two to three centroids still removes 90 squared units. Later moves remove only 25 and 17. A reader might place the elbow near $K=3$ because the rate of improvement becomes much smaller after that point.

This interpretation is a judgment about the shape of the curve. The number 3 is not produced by a universal cutoff.

## Fitting and plotting the curve in R

Fit every candidate with several initializations so that a poor start is less likely to distort one point on the curve.

```{r}
#| eval: false

candidate_k <- 1:10

set.seed(214)

within_loss <- vapply(
  candidate_k,
  function(k) {
    kmeans(
      X_scaled,
      centers = k,
      nstart = 50
    )$tot.withinss
  },
  numeric(1)
)
```

Plot the loss rather than printing only the selected value.

```{r}
#| eval: false

plot(
  candidate_k,
  within_loss,
  type = "b",
  xlab = "Number of centroids",
  ylab = "Within-cluster sum of squares"
)
```

The reductions can be displayed directly.

```{r}
#| eval: false

data.frame(
  added_centroid = candidate_k[-1],
  loss_reduction = -diff(within_loss)
)
```

The code uses the same representation, scaling, starting policy, and fitting settings at every $K$. Changing these choices midway would make the points belong to different procedures.

## The elbow may be ambiguous

Some curves decline smoothly without a visible bend. Two readers may also locate different elbows on the same curve. This is not a computational error. The elbow rule does not define how large the reduction ratio must be.

An elbow is often ambiguous. The plot can narrow attention to a range of plausible counts without selecting one mechanically.

A potential response is to search for a more elaborate curve transformation until a bend appears. That changes the rule after seeing the result. A better report shows the original loss curve, states the candidate range, and treats the elbow as one diagnostic.

## Loss values depend on the representation

Within-cluster loss depends on the feature units, number of columns, and number of rows. A loss of 190 has no standalone linguistic scale.

If formants are measured in hertz rather than kilohertz, raw squared distances change. If a feature is duplicated, it receives additional influence. If the sample contains more tokens, the sum has more terms.

Thus, loss curves from differently constructed representations should not be compared as though their vertical axes measured one common quantity. Refit and reinterpret the complete curve under each declared representation.

## Training compactness does not imply performance on held-out rows

The elbow describes the rows used to estimate the centroids. It does not show that the partition is stable under another sample or that the centroids represent new speakers.

Increasing $K$ can create a small cluster around one unusual training row. This reduces $W_K$ but may not improve representation of any new row. Held-out validation later in the module will evaluate performance on rows excluded from fitting.

The elbow also says nothing about whether rows are closer to their assigned cluster than to an alternative. The [next page](silhouette-scores.qmd) adds that row-level comparison.

## Checking the loss curve

1. Why can $W_{K+1}$ not exceed $W_K$ for an optimally fit solution?
2. Which reduction in the worked table is largest?
3. Why might a reader place the elbow near $K=3$?
4. What makes the elbow rule judgment-dependent?
5. Why should losses from raw and standardized representations not be placed on one common scale?
6. What property of new rows is not measured by the fitted loss curve?

The elbow plot shows how training compactness changes as centroids are added. Its monotonicity makes a visual elbow possible but prevents training loss from identifying a final group count by itself. The [next page](silhouette-scores.qmd) evaluates each row relative to both its assigned cluster and its nearest alternative cluster.
