---
title: "Power by simulation"
---

Suppose we plan the within-participant acceptability experiment from the earlier pages. Each participant judges counterbalanced active and passive descriptions in matched contexts, and the primary analysis compares the conditions while allowing the contrast to vary across participants and sentence items. How many participants and items are needed?

There is no answer until we specify the estimand, a range of condition differences that matter, variation across participants and items, the assignment and counterbalancing scheme, and the decision the analysis must make. A simulation-based design analysis repeatedly generates data under those assumptions, fits the planned model, and applies the planned decision rule. It reuses the distinction between a [null hypothesis and test statistic](../statistical-inference/null-hypotheses-and-test-statistics.qmd), but now evaluates the decision rule under a stated nonzero effect and complete design.

## Power is conditional on a data-generating process

Let $G$ denote an assumed data-generating process, and let $A(Y)$ denote the decision returned by the planned analysis for simulated data $Y$. If $A(Y)=1$ means that the decision rule is met, power is

$$
\operatorname{Power}(G)=P_G\{A(Y)=1\}.
$$

The subscript $G$ is essential. Power is not a fixed property of a sample size. It depends on the assumed effect, variation, design, exclusion process, analysis, and decision rule.

We begin with a deliberately simplified case. Suppose the analysis will use one active-minus-passive mean difference $D_i$ for each participant, treating the particular item set as fixed. A simple generating model is

$$
D_i\sim\operatorname{Normal}(\delta,\sigma_D^2),
$$

where $\delta$ is the mean condition difference and $\sigma_D$ is the standard deviation of participant differences. Both quantities must be supplied before simulation begins. This model is useful for seeing the simulation logic, but it cannot answer a design question about generalizing to new sentence items because item sampling does not appear in $G$.

## Working through the simulation estimate by hand

Imagine simulating 20 complete studies. The planned decision rule is met in 16 of them. The estimated power is

$$
\widehat{\operatorname{Power}}=\frac{16}{20}=.80.
$$

This estimate has Monte Carlo error because it is based on a finite number of simulated studies. As in the earlier discussion of [Monte Carlo integration](../statistical-inference/monte-carlo-integration.qmd), the simulation average has its own numerical uncertainty. Its approximate Monte Carlo standard error is

$$
\operatorname{MCSE}
=\sqrt{\frac{\widehat{\operatorname{Power}}(1-\widehat{\operatorname{Power}})}{S}},
$$

where $S$ is the number of simulations. With $S=20$ and estimated power $.80$,

$$
\operatorname{MCSE}
=\sqrt{\frac{.8(.2)}{20}}
\approx .089.
$$

An estimate of $.80$ with Monte Carlo standard error $.089$ is too imprecise for a fine distinction between candidate designs. If the same estimate came from 1,000 simulations, the standard error would be about $.013$.

## Simulating the exact planned comparison

The following function simulates participant differences, runs the planned one-sample test of those differences, and records whether the two-sided decision rule is met.

```{r}
simulate_study <- function(n_participants, effect, sd_difference) {
  difference <- rnorm(
    n = n_participants,
    mean = effect,
    sd = sd_difference
  )

  test <- t.test(difference, mu = 0)

  data.frame(
    estimate = mean(difference),
    p_value = test$p.value,
    decision = test$p.value < .05
  )
}
```

Each argument corresponds to part of the formal statement. `n_participants` determines how many $D_i$ values are drawn. `effect` supplies $\delta$, and `sd_difference` supplies $\sigma_D$. The analysis then operates on the same participant summaries that this simplified plan would produce. It does not yet reproduce trial-level ordinal ratings or crossed participant and item variation.

We estimate power by repeating the entire function.

```{r}
set.seed(903)

simulations <- replicate(
  n = 1000,
  expr = simulate_study(
    n_participants = 40,
    effect = 0.35,
    sd_difference = 0.80
  )$decision
)

estimated_power <- mean(simulations)
mcse <- sqrt(estimated_power * (1 - estimated_power) / length(simulations))

c(power = estimated_power, mcse = mcse)
```

The printed probability is conditional on a mean rating difference of .35 scale points and a participant-difference standard deviation of .80. The calculation treats the participant summaries as approximately continuous. It is not a claim that those assumptions are true, and $.35$ is not meaningful merely because it is nonzero. The range should come from related evidence, a blinded pilot, or a smallest difference that would change the linguistic argument.

## Power analysis is a curve, not one number

A single assumed effect gives a single conditional answer. A more useful design analysis varies assumptions that remain uncertain. For instance, we can repeat the simulation across several participant counts and effect sizes while holding the decision rule fixed.

```{r}
set.seed(904)

designs <- expand.grid(
  n_participants = c(24, 32, 40, 48),
  effect = c(0.20, 0.35, 0.50)
)

designs$power <- mapply(
  FUN = function(n, delta) {
    mean(replicate(
      1000,
      simulate_study(n, delta, sd_difference = 0.80)$decision
    ))
  },
  n = designs$n_participants,
  delta = designs$effect
)

designs
```

This table describes a [**power curve**](https://appliedstatisticsforlinguists.org/bwinter_stats_proofs.pdf#page=187) across assumptions. A design that performs adequately only at the largest effect is more fragile than one that performs similarly across a plausible range. The decision should also consider whether the smaller effects would change a claim about the voice alternation, not merely whether they would be difficult to detect.

## Power follows the complete procedure

The simulated design and planned study must match. If the study will analyze participant differences, a simulation of unrelated trial values answers a different question. If the study will exclude participants below a prespecified accuracy threshold, the simulation should apply that rule and record how often it removes participants. If condition order can affect the response, plausible order effects should be included when they bear on the target.

The estimated power will not describe the planned study when the simulated analysis is simpler than the planned analysis or when the generated data omit a design feature that determines the comparison. This matters sharply in psycholinguistic experiments: [Westfall, Kenny, and Judd](https://pubmed.ncbi.nlm.nih.gov/25111580/) show that adding participants cannot compensate indefinitely for too few sampled stimuli when participants respond to samples of items. The [mixed-effects models introduced earlier](../regression-models/random-effects.qmd) allow richer data generators with participant and item intercepts and condition slopes. The principle does not change: simulate the list design, generate responses at the recorded unit, apply exclusions, fit the planned analysis, and apply the planned decision.

Power also does not guarantee a successful individual study. A study with power $.80$ still fails its decision rule with probability $.20$ under the assumed process. And power says nothing about bias when the data-generating assumptions are wrong or about whether an acceptability contrast has a unique grammatical explanation. Simulation exposes assumptions; it does not certify them.

::: {.callout-note title="Audit the design simulation"}
1. Which quantities define $G$ in the participant-difference simulation, and which item-level quantities are absent?
2. What uncertainty is measured by the Monte Carlo standard error?
3. Why does increasing the number of simulations not add participants to the planned study?
4. What target and dependence structure change if the simulation draws independent trials but the analysis compares participant averages?
5. Why should power be reported across plausible effect sizes?
:::

Simulation-based design analysis evaluates a complete planned procedure under a stated data-generating process. Estimated power is the proportion of simulated studies meeting the decision rule, and its Monte Carlo error depends on the number of simulations. Participant count, item count, and simulation count do three different jobs. The design pages have now fixed the comparison and its repeated units, which tells us how to apply the earlier models for [grouped linguistic data](../regression-models/structured-data.qmd) without confusing additional rows with additional assignments.
