---
title: "Random assignment"
---

Suppose listeners receive either multi-talker exposure to Chinese-accented English or a control exposure and then identify keywords produced by a previously unheard Chinese-accented talker. If listeners choose their exposure, those who choose the accented-speech materials may already have more experience with that accent. A later recognition difference could then reflect exposure, prior experience, or both.

The [**assignment mechanism**](https://openstax.org/books/introductory-statistics-2e/pages/1-4-experimental-design-and-ethics) is the probability rule that selects a condition for each [assignment unit](units-of-assignment.qmd). Under random assignment, the randomizer, rather than listener choice or experimenter judgment, determines who receives which condition. Basically, this breaks the systematic link between condition and what the enrolled listeners would have done under either condition. More specifically, it makes assignment independent of those potential responses under the stated randomization procedure. This is the **assignment-identification link** (AIL).

## Stating the mechanism before drawing an assignment

Let $Z_i$ denote the condition assigned to listener $i$, with $Z_i=1$ for multi-talker exposure and $Z_i=0$ for control exposure. Under independent assignment with equal probabilities,

$$
P(Z_i=1)=P(Z_i=0)=.5.
$$

This statement describes the procedure before assignments are drawn. It does not say that every realized sample will contain the same number of participants in each condition. With eight participants, an independent mechanism can produce a split of four and four, five and three, or even eight and zero. These outcomes have different probabilities, but they are all possible under the stated rule.

The distinction between the procedure and its realized result matters. Under repeated use of the assignment mechanism, expected group means for a preassignment characteristic are equal. Random assignment does not guarantee that age, experience with Chinese-accented English, hearing threshold, or another characteristic will have identical distributions in one finite sample.

## Working through one realized assignment

Suppose eight participants are assigned by eight independent fair draws, producing

| listener | assigned exposure |
|:--|:--|
| P1 | control |
| P2 | multi-talker |
| P3 | multi-talker |
| P4 | control |
| P5 | control |
| P6 | control |
| P7 | multi-talker |
| P8 | control |

The realized split is three multi-talker and five control listeners. That imbalance does not show that the mechanism failed. We would instead ask whether each listener had probability $.5$ of receiving either condition and whether the recorded assignments match the generated assignment table.

A reproducible implementation stores the [random seed](../random-variables-and-distributions/random-seeds.qmd) and the table it produces.

```{r}
set.seed(418)

assignment <- data.frame(
  listener = paste0("P", 1:8),
  exposure = sample(
    c("multi_talker", "control"),
    size = 8,
    replace = TRUE
  )
)

assignment
table(assignment$exposure)
```

The vector passed to `sample()` gives the two possible conditions. `replace = TRUE` makes the draws independent under this implementation, so the first listener's assignment does not change the second listener's condition probabilities. Saving only the seed is not enough when software versions or upstream code may change. Save the realized table as part of the study record.

## Restricted random assignment

Some studies require exact balance. We may want four listeners in each exposure condition or equal numbers in each stimulus list. The assignment can remain random if the procedure samples from the assignments satisfying that restriction.

```{r}
set.seed(418)

balanced_assignment <- data.frame(
  listener = paste0("P", 1:8),
  exposure = sample(rep(c("multi_talker", "control"), each = 4))
)

balanced_assignment
table(balanced_assignment$exposure)
```

Here, the multiset of four multi-talker and four control labels is shuffled across listeners. Each listener has the same chance of receiving either condition, while the total condition counts are fixed. This is a restricted assignment mechanism, not an observational matching procedure performed after outcomes are known.

## Random assignment is not random sampling

Random assignment and random sampling answer different questions. As discussed in [populations and samples](../statistical-inference/populations-and-samples.qmd), sampling concerns how units enter the study and thus which population they may represent. Assignment concerns how conditions are allocated among the recruited units and thus which condition comparisons receive a causal interpretation.

A convenience sample of university students can be randomly assigned to exposure conditions. The design may support a causal comparison for the recruited listeners while leaving generalization to other ages, language histories, accents, talkers, or sentence materials uncertain. Conversely, a broad probability sample without random condition assignment may describe a population well but still leave a condition difference confounded. Randomization addresses assignment; it does not create representative samples of listeners or linguistic materials.

## Concealing future assignments

Alternating multi-talker, control, multi-talker, control is balanced, but it is predictable. If an experimenter knows the next condition while scheduling participants, time of day, recruitment source, or a judgment about eligibility may become associated with condition. The sequence then fails to protect the comparison from allocation decisions.

Predictable assignment can allow recruiters to change who enters a condition. Allocation concealment and a complete assignment record help prevent that problem. A second failure occurs when listeners are moved between conditions after assignment without recording the move. The primary **intention-to-treat contrast** compares listeners according to their assigned condition, whether or not exposure proceeded exactly as planned; departures and any per-protocol contrast require separate reporting.

::: {.callout-note title="Audit the assignment mechanism"}
1. Does an exact four and four condition split imply that assignment was random? What additional information is required?
2. Which claim is supported by random assignment but not by random sampling?
3. Why can a realized imbalance occur under a valid independent mechanism?
4. What should be saved in addition to the randomization seed?
:::

A random assignment mechanism is a probability rule for allocating conditions to assignment units. It supports the AIL under the procedure, while realized groups may still differ by chance. With the unit and mechanism fixed, we can now ask whether each participant contributes to [more than one condition](within-participant-designs.qmd) or to only one.
