---
title: "Within-participant designs"
---

Suppose an acceptability experiment asks whether active and passive descriptions differ in acceptability in a particular discourse context. The linguistic question concerns the English voice alternation and its fit to that context. The measured response, though, is a rating: it is not a direct observation of whether either sentence is generated by a participant's grammar. [Acceptability and grammaticality must be kept distinct](https://escholarship.org/uc/item/88x529wr) because ratings may also reflect plausibility, discourse fit, task interpretation, and scale use.

Participants also differ in how they use a seven-point scale. One participant may avoid the endpoints, while another may use them often. Comparing different people across conditions would allow those stable response habits to obscure the construction contrast.

In a [**within-participant design**](https://opentextbc.ca/researchmethods/chapter/experimental-design/), each participant contributes observations to more than one condition. The condition comparison can then be made inside each participant before it is averaged across participants. This is the same pairing logic introduced for [paired observations](../statistical-inference/paired-observations.qmd), now treated as a design choice.

## Each participant supplies a contrast

Let $Y_{i1}$ be participant $i$'s mean rating across active items and $Y_{i0}$ the same participant's mean across passive items. These should be different lexical items whose versions have been [counterbalanced](counterbalancing.qmd); the participant need not rate both versions of one sentence. The participant's condition difference is

$$
D_i=Y_{i1}-Y_{i0}.
$$

For $n$ participants, the estimated mean condition difference is

$$
\bar D=\frac{1}{n}\sum_{i=1}^{n}D_i.
$$

Stable participant characteristics appear in both $Y_{i1}$ and $Y_{i0}$. If participant $i$ tends to add a response shift $a_i$ in both conditions, then

$$
D_i=(\mu_1+a_i)-(\mu_0+a_i)=\mu_1-\mu_0.
$$

The shared additive shift cancels. Notice the scope of that statement. It does not remove condition-specific response habits, and it does not remove item variation. It explains why a within-participant design *may* estimate the participant-level contrast more precisely than a design that compares different groups with the same number of participants.

## Working through four participants

To keep the arithmetic visible, consider each participant's mean active rating and mean passive rating after averaging over their counterbalanced items.

| participant | active | passive | active minus passive |
|:--|--:|--:|--:|
| P1 | 6 | 5 | 1 |
| P2 | 7 | 5 | 2 |
| P3 | 5 | 4 | 1 |
| P4 | 6 | 6 | 0 |

The four participant differences are $1$, $2$, $1$, and $0$, so

$$
\bar D=\frac{1+2+1+0}{4}=1.
$$

The estimate says that these participants gave active items ratings one scale point higher on average than passive items in the tested contexts. It does not say that every participant showed a one-point difference: P2 showed a two-point difference, while P4 showed none. And it does not establish that passives are ungrammatical. The response is acceptability under the task and materials.

The calculation in R follows the definition directly.

```{r}
ratings <- data.frame(
  participant = paste0("P", 1:4),
  active = c(6, 7, 5, 6),
  passive = c(5, 5, 4, 6)
)

ratings$difference <- ratings$active - ratings$passive
ratings
mean(ratings$difference)
```

If one aggregate response per participant and condition is the intended analysis, uncertainty can be computed from these participant differences using the earlier method for [a paired mean difference](../statistical-inference/paired-mean-inference.qmd).

```{r}
t.test(ratings$active, ratings$passive, paired = TRUE)
```

The argument `paired = TRUE` is not a software convenience. It tells R that the two values in each row belong to the same participant and that uncertainty should be calculated from the differences.

## Repetition creates dependence

The fact that the comparison is made within participants does not make the original responses independent. Responses from one participant share response habits, experience, fatigue, and other characteristics. If sentence items are also repeated across participants, responses to one item share lexical content, plausibility, and structural properties.

For one response per participant and condition, the difference score carries the participant pairing into the analysis. With many trials per participant, one may first average within each participant and condition, then compare the participant averages. This simpler analysis changes what variation is retained. The earlier [mixed effects model sequence](../regression-models/structured-data.qmd) introduced methods that preserve trial level responses while representing repeated participants and items.

## Representing carryover effects

A condition encountered early may change a later response. This change is a [**carryover effect**](https://opentextbc.ca/researchmethods/chapter/experimental-design/). Participants may learn the task, become fatigued, or infer the experimental contrast.

Suppose every participant sees active items early and passive items late. A lower mean for passive items could reflect voice, later position, or both. Randomizing trial order distributes position across conditions, while order counterbalancing can ensure that each condition appears equally often in relevant positions. Neither step makes carryover impossible. The study should record trial order and state whether adaptation, fatigue, or discovery of the manipulation is plausible.

Possible carryover does not automatically rule out a within-participant design. Short, reversible manipulations often permit repeated conditions, and the increase in precision may be substantial. But training, adaptation, or explicit instruction may change the participant in a way that cannot be reset. Those cases motivate a design in which each participant receives only one condition.

::: {.callout-note title="Audit the within-participant contrast"}
1. Which quantity is averaged across participants, the two condition means separately or each participant's difference?
2. What participant characteristic cancels if it adds the same shift in both conditions?
3. Why does pairing fail if the active and passive vectors are sorted independently before subtraction?
4. What design record is needed to diagnose a carryover effect?
:::

A within-participant design turns each participant into a condition comparison across counterbalanced items. Stable additive participant shifts may cancel, but condition effects can still differ across participants and items, and earlier trials can alter later responses. When the manipulation cannot be repeated without changing its meaning, the appropriate alternative is a [between-participant design](between-participant-designs.qmd).
