---
title: "Within-participant designs"
---

Suppose an acceptability experiment asks whether a passive sentence receives a lower rating than its active counterpart. Participants differ greatly in how they use a seven point scale. One participant may avoid the endpoints, while another may use them often. Comparing different people across conditions would allow those stable response habits to obscure the construction contrast.

In a [**within-participant design**](https://opentextbc.ca/researchmethods/chapter/experimental-design/), each participant contributes observations to more than one condition. The condition comparison can then be made inside each participant before it is averaged across participants. This is the same pairing logic introduced for [paired observations](../statistical-inference/paired-observations.qmd), now treated as a design choice.

## Each participant supplies a contrast

Let $Y_{i1}$ be participant $i$'s response in condition 1 and $Y_{i0}$ the response in condition 0. The participant's condition difference is

$$
D_i=Y_{i1}-Y_{i0}.
$$

For $n$ participants, the estimated mean condition difference is

$$
\bar D=\frac{1}{n}\sum_{i=1}^{n}D_i.
$$

Stable participant characteristics appear in both $Y_{i1}$ and $Y_{i0}$. If participant $i$ tends to add a response shift $a_i$ in both conditions, then

$$
D_i=(\mu_1+a_i)-(\mu_0+a_i)=\mu_1-\mu_0.
$$

The shared shift cancels. This cancellation explains why a within-participant design can estimate a condition contrast more precisely than a design that compares different groups with the same number of participants.

## Work through four participants

Consider one active and one passive rating from each of four participants.

| participant | active | passive | active minus passive |
|:--|--:|--:|--:|
| P1 | 6 | 5 | 1 |
| P2 | 7 | 5 | 2 |
| P3 | 5 | 4 | 1 |
| P4 | 6 | 6 | 0 |

The four participant differences are $1$, $2$, $1$, and $0$, so

$$
\bar D=\frac{1+2+1+0}{4}=1.
$$

The estimate says that these participants rated the active version one scale point higher on average than the passive version. It does not say that every participant showed a one point difference. P2 showed a two point difference, while P4 showed none.

The calculation in R follows the definition directly.

```{r}
ratings <- data.frame(
  participant = paste0("P", 1:4),
  active = c(6, 7, 5, 6),
  passive = c(5, 5, 4, 6)
)

ratings$difference <- ratings$active - ratings$passive
ratings
mean(ratings$difference)
```

If one aggregate response per participant and condition is the intended analysis, uncertainty can be computed from these participant differences using the earlier method for [a paired mean difference](../statistical-inference/paired-mean-inference.qmd).

```{r}
t.test(ratings$active, ratings$passive, paired = TRUE)
```

The argument `paired = TRUE` is not a software convenience. It tells R that the two values in each row belong to the same participant and that uncertainty should be calculated from the differences.

## Repetition creates dependence

The fact that the comparison is made within participants does not make the original responses independent. Responses from one participant share response habits, experience, fatigue, and other characteristics. If sentence items are also repeated across participants, responses to one item share lexical and structural properties.

For one response per participant and condition, the difference score carries the participant pairing into the analysis. With many trials per participant, one may first average within each participant and condition, then compare the participant averages. This simpler analysis changes what variation is retained. The earlier [mixed effects model sequence](../regression-models/structured-data.qmd) introduced methods that preserve trial level responses while representing repeated participants and items.

## Represent carryover effects

A condition encountered early may change a later response. This change is a [**carryover effect**](https://opentextbc.ca/researchmethods/chapter/experimental-design/). Participants may learn the task, become fatigued, or infer the experimental contrast.

Suppose every participant sees the active sentence before the passive sentence. A lower passive rating could reflect construction, later position, or both. Randomizing order distributes position across conditions, while counterbalancing can ensure that each condition appears equally often in relevant positions. Neither step makes carryover impossible. The study should record trial order and state whether a lasting effect is plausible.

Possible carryover does not automatically rule out a within-participant design. Short, reversible manipulations often permit repeated conditions, and the increase in precision may be substantial. But training, adaptation, or explicit instruction may change the participant in a way that cannot be reset. Those cases motivate a design in which each participant receives only one condition.

::: {.callout-note title="Audit the within-participant contrast"}
1. Which quantity is averaged across participants, the two condition means separately or each participant's difference?
2. What participant characteristic cancels if it adds the same shift in both conditions?
3. Why does pairing fail if the active and passive vectors are sorted independently before subtraction?
4. What design record is needed to diagnose a carryover effect?
:::

## Summing up

A within-participant design turns each participant into a condition comparison. Stable participant shifts may cancel, but condition effects can still differ across participants and earlier conditions can alter later responses. When the manipulation cannot be repeated without changing its meaning, the appropriate alternative is a [between-participant design](between-participant-designs.qmd).
