---
title: "From outcomes to distributions"
subtitle: "Statistical Methods in Linguistics"
author: "Aaron Steven White"
institute: "University of Rochester"
date: "2026-09-14"
date-format: "[September 14, 16, 21 and 23,] YYYY"
execute:
  enabled: true
  echo: false
  warning: false
  message: false
title-slide-attributes:
  data-background-color: "#113A6A"
format:
  revealjs:
    css: ../shared/theme.css
    width: 1280
    height: 720
    margin: 0
    self-contained: true
    embed-resources: true
    transition: none
    slide-number: false
    controls: true
    progress: true
    center: false
    incremental: false
    html-math-method: katex
---

```{r}
#| include: false
course_blue <- "#2166AC"
course_red <- "#B2182B"
course_gray <- "#4D4D4D"

plot_discrete_pmf <- function(x, mass, main, xlab = "Value", axis_labels = x) {
  plot(
    x, mass,
    type = "h", lwd = 5, lend = 1,
    col = course_red,
    xlab = xlab, ylab = "Probability",
    main = main, ylim = c(0, 1.08 * max(mass)),
    xaxt = "n", bty = "l"
  )
  axis(1, at = x, labels = axis_labels)
  points(x, mass, pch = 19, cex = 1.3, col = course_red)
}
```

## From outcomes to measured values {.statement-slide}

A probability measure assigns probability to events. A statistical analysis usually concerns one or more values assigned to each outcome.

## This Module {.compact}

1. **September 14:** Random variables, PMFs, densities, and CDFs
2. **September 16:** Expectations, variance, and moments
3. **September 21:** Standard distribution families
4. **September 23:** Joint, marginal, and conditional distributions

## Random variables and distributions {.section-slide}

<span class="section-subtitle">September 14</span>

## Several variables may be evaluated at one outcome

| outcome | word | speaker | duration |
|:--|:--|:--|--:|
| $\omega_1$ | *heed* | s01 | 112 ms |
| $\omega_2$ | *hid* | s01 | 83 ms |
| $\omega_3$ | *heed* | s02 | 126 ms |

A duration variable assigns one value to each complete token outcome.

## Defining a random variable

Given a probability space $(\Omega,\mathcal F,\mathbb{P})$ and a measurable space $(\mathcal X,\mathcal G)$, a **random variable** is a measurable function

$$
X:\Omega\rightarrow\mathcal X.
$$

The domain contains outcomes. The codomain contains possible values.

::: {.notes}
[Sources]

Siegrist, Probability, Mathematical Statistics, and Stochastic Processes, Probability Spaces Revisited: https://stats.libretexts.org/Bookshelves/Probability_Theory/Probability_Mathematical_Statistics_and_Stochastic_Processes_%28Siegrist%29/02%3A_Probability_Spaces/2.09%3A_Probability_Spaces_Revisited
:::

## Defining the duration variable

For each token $\omega\in\Omega$, define

$$
D(\omega)\equiv\text{duration of token }\omega\text{ in milliseconds}.
$$

Thus $D(\omega_1)=112$, $D(\omega_2)=83$, and $D(\omega_3)=126$.

## Value statements denote events

For a measurable set $B\subseteq\mathcal X$, define the shorthand

$$
\{X\in B\}
\equiv X^{-1}(B)
=\{\omega\in\Omega:X(\omega)\in B\}.
$$

This set contains outcomes, not bare values.

## A threshold event

For the three-outcome illustration, let $\Omega\equiv\{\omega_1,\omega_2,\omega_3\}$. Then

$$
\{D>100\}
\equiv\{\omega\in\Omega:D(\omega)>100\}
=\{\omega_1,\omega_3\}.
$$

The notation $D>100$ is shorthand for an event in $\mathcal F$.

## Measurability makes value probabilities well defined

The function $X$ is measurable when

$$
B\in\mathcal G
\quad\Longrightarrow\quad
X^{-1}(B)\in\mathcal F,
$$

where $\mathcal G$ is the sigma-algebra on $\mathcal X$.

## The pronoun enumeration from the original notes

Let the case-coded pronoun outcomes be ordered as

$$
\begin{aligned}
(&\textit{us},\textit{they},\textit{them},\textit{you}_{[-\mathrm{acc}]},
\textit{he},\textit{I},\textit{it}_{[-\mathrm{acc}]},\\
&\textit{me},\textit{him},\textit{she},\textit{we},
\textit{it}_{[+\mathrm{acc}]},\textit{you}_{[+\mathrm{acc}]},\textit{her}).
\end{aligned}
$$

Define $V(\omega)$ as the position of $\omega$ in this ordered list. This definition fixes the correspondence between outcomes, integer codes, and probabilities.

## Preimages are events

Under the enumeration $V$,

$$
V^{-1}(\{1,2,3\})
=\{\textit{us},\textit{they},\textit{them}\}.
$$

More generally,

$$
V^{-1}(B)
\equiv\{\omega\in\Omega:V(\omega)\in B\}.
$$

The measurability condition requires this preimage to belong to $\mathcal F$ for every $B\in\mathcal G$.

## Distribution of a random variable

The **distribution** of $X$ assigns each $B\in\mathcal G$ the probability

$$
\mathbb{P}(X\in B)
\equiv
\mathbb{P}\!\left(X^{-1}(B)\right).
$$

As a function of $B$, this assignment is a probability measure on $(\mathcal X,\mathcal G)$. We retain $\mathbb{P}$ for the probability measure on $(\Omega,\mathcal F)$ and introduce no second measure symbol.

::: {.notes}
[Sources]

Durrett, *Probability: Theory and Examples*, Sections 1.2--1.3: https://services.math.duke.edu/~rtd/PTE/PTE5_011119.pdf

ETH Zurich, *Probability Theory*, Definition 2.4 and Remark 2.9: https://metaphor.ethz.ch/x/2025/hs/401-3601-00L/main.pdf
:::

## Discrete random variables

A random variable $X$ is **discrete** when some finite or countably infinite set $\mathcal S$ satisfies $\mathbb{P}(X\in\mathcal S)=1$.

Grammatical person and utterance length are discrete. The first has finite support; the second may have countably infinite support.

::: {.notes}
[Sources]

Nicenboim, Schad, and Vasishth, Introduction to Bayesian Data Analysis for Cognitive Science, Chapter 2: https://bruno.nicenboim.me/bayescogsci/ch-intro.html
:::

## Defining a probability mass function

For a discrete random variable $X$, its **probability mass function** (PMF) is defined by

$$
p_X:\mathcal X\to[0,1],
\qquad
p_X(x)
\equiv\mathbb{P}(X=x).
$$

The subscript names the random variable; the argument is a possible value.

Its support is

$$
\operatorname{supp}(X)
\equiv
\{x\in\mathcal X:p_X(x)>0\}.
$$

## Requirements for a PMF

Its PMF satisfies

$$
p_X(x)\geq0
\qquad\text{and}\qquad
\sum_{x\in\operatorname{supp}(X)}p_X(x)=1.
$$

For any $A\subseteq\mathcal X$,

$$
\mathbb{P}(X\in A)
=\sum_{x\in A\cap\operatorname{supp}(X)}p_X(x).
$$

## Adding probability mass

| syllables $s$ | 1 | 2 | 3 | 4 |
|:--|--:|--:|--:|--:|
| $p_S(s)$ | .42 | .33 | .17 | .08 |

Because $\{S=3\}$ and $\{S=4\}$ are disjoint,

$$
\mathbb{P}(S\geq3)=.17+.08=.25.
$$

## Absolutely continuous random variables

A real-valued random variable $X$ is **absolutely continuous** when there is a function $f_X:\mathbb R\to[0,\infty)$ such that, for every Borel set $A\subseteq\mathbb R$,

$$
\mathbb{P}(X\in A)=\int_A f_X(x)\,\mathrm{d}x.
$$

The function $f_X$ is a probability density function.

::: {.notes}
[Sources]

Pfeiffer, Applied Probability, Distribution and Density Functions: https://stats.libretexts.org/Bookshelves/Probability_Theory/Applied_Probability_%28Pfeiffer%29/07%3A_Distribution_and_Density_Functions/7.01%3A_Distribution_and_Density_Functions
:::

## Requirements for a density

The density of $X$ satisfies

$$
f_X(x)\geq0
\qquad\text{and}\qquad
\int_{-\infty}^{\infty}f_X(x)\,\mathrm{d}x=1.
$$

Conversely, any measurable nonnegative function with integral one defines an absolutely continuous probability distribution by integration.

## A uniform duration model

Suppose

$$
f_D(d)\equiv
\begin{cases}
1/100,&50\leq d\leq150,\\
0,&\text{otherwise}.
\end{cases}
$$

Then

$$
\mathbb{P}(80\leq D\leq100)=\int_{80}^{100}\frac{1}{100}\,\mathrm{d}u=.20.
$$

## Points have no mass under a density

For an absolutely continuous random variable,

$$
\mathbb{P}(X=x)=\int_x^x f_X(t)\,\mathrm{d}t=0.
$$

Thus $f_X(x)>0$ does not imply $\mathbb{P}(X=x)>0$.

## Defining a cumulative distribution function

Every real-valued random variable has a **cumulative distribution function** (CDF), defined by

$$
F_X:\mathbb R\to[0,1],
\qquad
F_X(x)
\equiv\mathbb{P}(X\leq x).
$$

The same definition applies to discrete, continuous, and mixed distributions.

::: {.notes}
[Sources]

Nicenboim, Schad, and Vasishth, Introduction to Bayesian Data Analysis for Cognitive Science, Chapter 2: https://bruno.nicenboim.me/bayescogsci/ch-intro.html
:::

## Computing a CDF

For a discrete $X$,

$$
F_X(x)=\sum_{t\leq x}p_X(t).
$$

For an absolutely continuous $X$,

$$
F_X(x)=\int_{-\infty}^{x}f_X(t)\,\mathrm{d}t.
$$

These are consequences of the CDF definition.

## What the representations mean {.statement-slide}

A PMF assigns mass to values. A PDF assigns density whose integral is probability. A CDF gives the probability at or below a threshold.

## Expectations and variation {.section-slide}

<span class="section-subtitle">September 16</span>

## Expectations average with respect to probability

For a random variable $X$ satisfying $\mathbb E[|X|]<\infty$, define

$$
\mathbb E[X]
\equiv
\begin{cases}
\sum_x x\,p_X(x),&X\text{ discrete},\\[2mm]
\int_{-\infty}^{\infty}x f_X(x)\,\mathrm{d}x,&X\text{ absolutely continuous}.
\end{cases}
$$

::: {.notes}
[Sources]

Siegrist, Probability, Mathematical Statistics, and Stochastic Processes, Expected Value: https://stats.libretexts.org/Bookshelves/Probability_Theory/Probability_Mathematical_Statistics_and_Stochastic_Processes_%28Siegrist%29/04%3A_Expected_Value/4.01%3A_Definitions_and_Basic_Properties
:::

## A weighted average of word lengths

If $L$ takes values $1$, $2$, and $4$ with masses $.25$, $.50$, and $.25$, then

$$
\begin{aligned}
\mathbb E[L]
&=1(.25)+2(.50)+4(.25)\\
&=2.25.
\end{aligned}
$$

An expectation need not be a possible value.

## Expectations of functions

When $\mathbb E[|g(X)|]<\infty$, define

$$
\mathbb E[g(X)]
\equiv\sum_x g(x)p_X(x)
$$

in the discrete case, with the analogous integral in the continuous case.

In general, $\mathbb E[g(X)]\neq g(\mathbb E[X])$.

## Deriving linearity for a finite discrete model

For constants $a$ and $b$,

$$
\begin{aligned}
\mathbb E[aX+b]
&=\sum_x (ax+b)p_X(x)\\
&=a\sum_x xp_X(x)+b\sum_xp_X(x)\\
&=a\mathbb E[X]+b.
\end{aligned}
$$

## Linearity for sums

If $\mathbb E[|X_i|]<\infty$ for every $i$,

$$
\mathbb E\!\left[\sum_{i=1}^{n}a_iX_i\right]
=\sum_{i=1}^{n}a_i\mathbb E[X_i].
$$

This equality does not require independence. Dependence can change the distribution and variance of the sum.

## A distribution need not have a finite mean

Let

$$
p_L(l)\equiv\frac{6}{\pi^2l^2},
\qquad l=1,2,\ldots
$$

Then

$$
\sum_{l=1}^{\infty}p_L(l)
=\frac{6}{\pi^2}\sum_{l=1}^{\infty}\frac1{l^2}
=1,
$$

but

$$
\mathbb E[L]
=\frac{6}{\pi^2}\sum_{l=1}^{\infty}\frac1l
=\infty.
$$

## The original continuous counterexample: Cauchy

For $X\sim\operatorname{Cauchy}(x_0,\gamma)$ with $\gamma>0$,

$$
f_X(x)
\equiv
\frac{1}{\pi\gamma\left[1+\left(\frac{x-x_0}{\gamma}\right)^2\right]}.
$$

```{r}
#| fig-width: 8
#| fig-height: 4.2
#| fig-align: center
x_cauchy <- seq(-6, 6, length.out = 1000)
plot(
  x_cauchy, dcauchy(x_cauchy, 0, 1),
  type = "l", lwd = 3, col = course_blue,
  xlab = "x", ylab = "Density", main = "PDF of Cauchy(0, 1)", bty = "l"
)
```

The positive and negative parts of $\int_{-\infty}^{\infty}x f_X(x)\,\mathrm dx$ are both infinite. Thus $\mathbb E[X]$ is undefined; the symmetric principal value zero is not an expectation.

## Defining variance

For a random variable with finite second moment, define

$$
\operatorname{Var}(X)
\equiv\mathbb E\!\left[(X-\mathbb E[X])^2\right].
$$

Variance is an expected squared deviation from the mean.

## Deriving the computational identity

Writing $\mu\equiv\mathbb E[X]$,

$$
\begin{aligned}
\operatorname{Var}(X)
&=\mathbb E[X^2-2\mu X+\mu^2]\\
&=\mathbb E[X^2]-2\mu\mathbb E[X]+\mu^2\\
&=\mathbb E[X^2]-2\mu^2+\mu^2\\
&=\mathbb E[X^2]-\mu^2\\
&=\mathbb E[X^2]-\mathbb E[X]^2.
\end{aligned}
$$

## Standard deviation

The **standard deviation** is defined by

$$
\sigma_X
\equiv\sqrt{\operatorname{Var}(X)}.
$$

If $X$ is measured in milliseconds, $\operatorname{Var}(X)$ has units ms$^2$ and $\sigma_X$ has units ms.

## Defining central moments

When it exists, the $k$th **central moment** is defined by

$$
\mu_k
\equiv\mathbb E\!\left[(X-\mathbb E[X])^k\right].
$$

Thus $\mu_1=0$ and $\mu_2=\operatorname{Var}(X)$.

## What summaries retain {.statement-slide}

An expectation summarizes a specified function of a distribution. Its meaning depends on that function, the existence of the expectation, and the measurement units.

## Standard distribution families {.section-slide}

<span class="section-subtitle">September 21</span>

## Distribution notation

The statement

$$
X\sim\mathcal D(\theta)
$$

means that the distribution of $X$ belongs to family $\mathcal D$ with parameter $\theta$. The symbol $\sim$ does not mean numerical equality.

## The categorical distribution

For labels $c_1,\ldots,c_K$ and $\boldsymbol\pi\in[0,1]^K$ with $\sum_{k=1}^K\pi_k=1$,

$$
V\sim\operatorname{Categorical}(\boldsymbol\pi)
$$

means

$$
p_V(c_k)
\equiv\mathbb{P}(V=c_k)
=\pi_k.
$$

::: {.notes}
[Sources]

Stan Functions Reference, Categorical Distribution: https://mc-stan.org/docs/functions-reference/categorical-distribution.html
:::

## The pronoun categorical distribution

The original example uses the ordered pronoun labels from the enumeration $V$ and the probability vector

$$
\boldsymbol\pi
\equiv(.03,.09,.03,.12,.07,.28,.07,.05,.02,.02,.07,.08,.05,.02).
$$

The $k$th entry is the probability of the $k$th pronoun label. Reordering the labels without applying the same permutation to $\boldsymbol\pi$ changes the distribution.

## PMF for the pronoun categorical distribution

```{r}
#| fig-width: 10.8
#| fig-height: 4.5
#| fig-align: center
pronoun_labels <- c(
  "us", "they", "them", "you[-acc]", "he", "I", "it[-acc]",
  "me", "him", "she", "we", "it[+acc]", "you[+acc]", "her"
)
pronoun_mass <- c(.03, .09, .03, .12, .07, .28, .07,
                  .05, .02, .02, .07, .08, .05, .02)
plot(
  seq_along(pronoun_labels), pronoun_mass,
  type = "h", lwd = 5, lend = 1, col = course_red,
  xaxt = "n", xlab = "Pronoun", ylab = "Probability",
  main = "PMF for the categorical distribution on pronouns",
  ylim = c(0, .31), bty = "l"
)
axis(1, at = seq_along(pronoun_labels), labels = pronoun_labels,
     las = 2, cex.axis = .7)
points(seq_along(pronoun_labels), pronoun_mass,
       pch = 19, cex = 1.1, col = course_red)
```

## The Bernoulli distribution

Define the accusative-case indicator

$$
X(\omega)
\equiv
\begin{cases}
1,&\omega\in A,\\
0,&\omega\notin A.
\end{cases}
$$

The range is $\{0,1\}$ even though the sample space contains fourteen pronoun outcomes. For $0\leq\pi\leq1$,

$$
X\sim\operatorname{Bernoulli}(\pi)
$$

means

$$
p_X(x)
\equiv\mathbb{P}(X=x)
=\pi^x(1-\pi)^{1-x},
\qquad x\in\{0,1\}.
$$

## Bernoulli distribution on pronoun case

Using the original empirical value $\pi=.27$,

$$
p_X(1)=.27,
\qquad
p_X(0)=.73.
$$

The parameter is the probability of accusative case because the definition of $X$ assigns accusative outcomes the value $1$.

```{r}
#| fig-width: 7.5
#| fig-height: 4
#| fig-align: center
plot_discrete_pmf(
  0:1, c(.73, .27),
  "PMF for the Bernoulli distribution on pronoun case",
  "Case", c("[-acc]", "[+acc]")
)
```

## The binomial distribution

Let $Y_1,\ldots,Y_n$ be independent Bernoulli variables with common parameter $\pi$, and define $K\equiv\sum_{i=1}^nY_i$. Then

$$
K\sim\operatorname{Binomial}(n,\pi),
$$

with support $\{0,\ldots,n\}$.

## Deriving the binomial PMF

Each sequence with $k$ ones has probability $\pi^k(1-\pi)^{n-k}$, and there are ${n\choose k}$ such sequences. Thus

$$
p_K(k)
\equiv\mathbb{P}(K=k)
={n\choose k}\pi^k(1-\pi)^{n-k}.
$$

## Sampling without replacement

Suppose a finite archive contains $N$ clauses, $M$ of which are passive. If $n$ distinct clauses are sampled without replacement and $K$ counts passives, then

$$
p_K(k)
\equiv\mathbb{P}(K=k)
=\frac{{M\choose k}{N-M\choose n-k}}{{N\choose n}}.
$$

This is the hypergeometric PMF for

$$
\max(0,n-N+M)\leq k\leq\min(n,M).
$$

## A geometric series defines a PMF

The original notes begin with

$$
\sum_{k=0}^{\infty}\frac{1}{2^{k+1}}=1.
$$

Thus

$$
p_K(k)\equiv\frac{1}{2^{k+1}},
\qquad k=0,1,2,\ldots
$$

is a PMF with countably infinite support.

```{r}
#| fig-width: 7.5
#| fig-height: 4
#| fig-align: center
k_plot <- 0:9
plot_discrete_pmf(
  k_plot, 2^(-(k_plot + 1)),
  "PMF defined by a geometric series", "k"
)
```

## The geometric family

Let $K$ count failures before the first success in independent Bernoulli trials with common success probability $\pi$. Define

$$
K\sim\operatorname{Geometric}(\pi)
$$

by the PMF

$$
p_K(k)
\equiv\mathbb{P}(K=k)
=(1-\pi)^k\pi,
\qquad k=0,1,2,\ldots
$$

This is the convention used by `dgeom` in R. The total-trial count is $T\equiv K+1$.

## The original geometric parameter sequence: $\pi=.5$

```{r}
#| fig-width: 7.5
#| fig-height: 4.2
#| fig-align: center
plot_discrete_pmf(
  k_plot, dgeom(k_plot, prob = .5),
  "PMF of Geometric(.5)", "Failures before the first success"
)
```

## As $\pi$ decreases, the distribution spreads rightward

```{r}
#| fig-width: 7.5
#| fig-height: 4.2
#| fig-align: center
plot_discrete_pmf(
  k_plot, dgeom(k_plot, prob = .1),
  "PMF of Geometric(.1)", "Failures before the first success"
)
```

## As $\pi$ increases, mass concentrates at zero

```{r}
#| fig-width: 7.5
#| fig-height: 4.2
#| fig-align: center
plot_discrete_pmf(
  k_plot, dgeom(k_plot, prob = .9),
  "PMF of Geometric(.9)", "Failures before the first success"
)
```

## Every geometric PMF decreases from its first support point

For every $k\geq0$,

$$
\frac{p_K(k+1)}{p_K(k)}
=\frac{(1-\pi)^{k+1}\pi}{(1-\pi)^k\pi}
=1-\pi
<1.
$$

Thus $p_K(k+1)<p_K(k)$. If word length is $L\equiv K+1$, every geometric model has its mode at $L=1$.

## The CMUdict word-length data

The original worked example uses every noncomment pronunciation entry in CMU Pronouncing Dictionary 0.7b. For an entry $\omega$, define

$$
L(\omega)\equiv\text{number of phoneme symbols in }\omega.
$$

Alternate pronunciations occupy separate dictionary entries, as they did in the original computation.

```{r}
#| include: false
cmudict_candidates <- c(
  "data/cmudict-0.7b",
  "random-variables-and-distributions/data/cmudict-0.7b",
  "decks/random-variables-and-distributions/data/cmudict-0.7b"
)
cmudict_path <- cmudict_candidates[file.exists(cmudict_candidates)][1]
cmudict_lines <- readLines(cmudict_path, encoding = "latin1", warn = FALSE)
cmudict_entries <- cmudict_lines[
  nzchar(cmudict_lines) & !startsWith(cmudict_lines, ";;;")
]
cmudict_fields <- strsplit(trimws(cmudict_entries), "[[:space:]]+")
word_lengths <- lengths(cmudict_fields) - 1L
word_failures <- word_lengths - 1L

empirical_length_mass <- prop.table(table(word_lengths))
length_values <- as.integer(names(empirical_length_mass))
geom_pi_hat <- 1 / mean(word_lengths)

nb_mu_hat <- mean(word_failures)
nb_profile <- optimize(
  function(log_r) {
    r <- exp(log_r)
    -sum(dnbinom(word_failures, size = r, mu = nb_mu_hat, log = TRUE))
  },
  interval = c(log(.01), log(10000))
)
nb_r_hat <- exp(nb_profile$minimum)
nb_pi_hat <- nb_r_hat / (nb_r_hat + nb_mu_hat)
```

## Empirical distribution of word length in phonemes

```{r}
#| fig-width: 8.5
#| fig-height: 4.6
#| fig-align: center
barplot(
  empirical_length_mass,
  names.arg = length_values,
  col = course_blue, border = NA,
  xlab = "Word length in phonemes", ylab = "Relative frequency",
  main = "CMUdict 0.7b pronunciation entries",
  space = 0.15, las = 1
)
```

## The geometric family cannot reproduce the observed mode

The empirical PMF has its mode at six phonemes. The fitted geometric PMF must have its mode at one phoneme because it is strictly decreasing.

```{r}
#| fig-width: 8.5
#| fig-height: 4.6
#| fig-align: center
plot(
  length_values, as.numeric(empirical_length_mass),
  type = "h", lwd = 5, lend = 1, col = course_blue,
  xlab = "Word length in phonemes", ylab = "Probability / relative frequency",
  main = "Empirical and fitted geometric PMFs", bty = "l"
)
points(length_values, as.numeric(empirical_length_mass),
       pch = 19, col = course_blue)
lines(length_values, dgeom(length_values - 1, prob = geom_pi_hat),
      type = "b", pch = 17, col = course_red, lwd = 2)
legend("topright", c("Empirical", sprintf("Geometric(%.3f)", geom_pi_hat)),
       col = c(course_blue, course_red), lty = 1, pch = c(19, 17), bty = "n")
```

## The negative binomial generalizes the geometric

Let $K$ count failures before the $r$th success, where $r\in\{1,2,\ldots\}$. Define

$$
K\sim\operatorname{NegBin}(r,\pi)
$$

by

$$
p_K(k)
\equiv\mathbb{P}(K=k)
={k+r-1\choose k}(1-\pi)^k\pi^r,
\qquad k=0,1,2,\ldots
$$

When $r=1$, the binomial coefficient equals one and this PMF is exactly geometric.

## The original negative-binomial sequence: $r=1$, $\pi=.5$

```{r}
#| fig-width: 7.5
#| fig-height: 4.2
#| fig-align: center
k_nb <- 0:20
plot_discrete_pmf(
  k_nb, dnbinom(k_nb, size = 1, prob = .5),
  "PMF of NegBin(1, .5)", "Failures before the first success"
)
```

## Increasing $r$ permits an interior mode

With $\pi$ fixed, increasing $r$ increases

$$
\mathbb E[K]=r\frac{1-\pi}{\pi}.
$$

```{r}
#| fig-width: 7.5
#| fig-height: 4.2
#| fig-align: center
plot_discrete_pmf(
  k_nb, dnbinom(k_nb, size = 5, prob = .5),
  "PMF of NegBin(5, .5)", "Failures before the fifth success"
)
```

## With $r=5$, smaller $\pi$ moves mass farther right

```{r}
#| fig-width: 7.5
#| fig-height: 4.2
#| fig-align: center
k_nb_wide <- 0:60
plot_discrete_pmf(
  k_nb_wide, dnbinom(k_nb_wide, size = 5, prob = .1),
  "PMF of NegBin(5, .1)", "Failures before the fifth success"
)
```

## With $r=5$, larger $\pi$ concentrates mass near zero

```{r}
#| fig-width: 7.5
#| fig-height: 4.2
#| fig-align: center
plot_discrete_pmf(
  0:12, dnbinom(0:12, size = 5, prob = .9),
  "PMF of NegBin(5, .9)", "Failures before the fifth success"
)
```

## A larger $r$ can offset a large $\pi$

```{r}
#| fig-width: 7.5
#| fig-height: 4.2
#| fig-align: center
plot_discrete_pmf(
  0:18, dnbinom(0:18, size = 40, prob = .9),
  "PMF of NegBin(40, .9)", "Failures before the fortieth success"
)
```

## Fitting the word-length distribution

Set $K\equiv L-1$ so that the observed support begins at zero. Under the negative-binomial parameterization,

$$
\mathbb E[K]=r\frac{1-\pi}{\pi},
\qquad
\operatorname{Var}(K)=r\frac{1-\pi}{\pi^2}.
$$

The CMUdict data are underdispersed relative to this family: the empirical variance of $K$ is smaller than its empirical mean. The likelihood thus increases toward the Poisson boundary $r\to\infty$.

## Empirical, geometric, and negative-binomial-limit CDFs

```{r}
#| fig-width: 8.5
#| fig-height: 4.7
#| fig-align: center
cdf_x <- 1:20
plot(
  cdf_x, ecdf(word_lengths)(cdf_x),
  type = "s", lwd = 3, col = course_blue,
  xlab = "Word length in phonemes",
  ylab = "Cumulative probability / relative frequency",
  main = "CDF of word length in phonemes", ylim = c(0, 1), bty = "l"
)
lines(cdf_x, pgeom(cdf_x - 1, prob = geom_pi_hat),
      type = "s", lwd = 2, col = course_red)
lines(cdf_x, pnbinom(cdf_x - 1, size = nb_r_hat, prob = nb_pi_hat),
      type = "s", lwd = 2, lty = 2, col = course_gray)
legend(
  "bottomright",
  c("Empirical", "Fitted geometric", "Negative-binomial limit"),
  col = c(course_blue, course_red, course_gray),
  lty = c(1, 1, 2), lwd = c(3, 2, 2), bty = "n"
)
```

The negative-binomial generalization removes the geometric family's boundary-mode restriction. For these data, its maximum-likelihood fit approaches its Poisson limit rather than a finite $r$.

## Poisson counts require a declared exposure

If events occur at rate $\lambda>0$ per exposure unit, a Poisson model for exposure $t>0$ specifies

$$
Y_t\sim\operatorname{Poisson}(\lambda t),
$$

with

$$
p_{Y_t}(y)
\equiv\mathbb{P}(Y_t=y)
=e^{-\lambda t}\frac{(\lambda t)^y}{y!},
\qquad y=0,1,2,\ldots
$$

## A Poisson distribution is not a complete process claim

The count model implies

$$
\mathbb E[Y_t]=\operatorname{Var}(Y_t)=\lambda t.
$$

A homogeneous Poisson process additionally specifies $N(0)=0$ and stationary independent increments with rate $\lambda$.

## The continuous uniform distribution

For $a<b$,

$$
X\sim\operatorname{Uniform}(a,b)
$$

means

$$
f_X(x)\equiv
\begin{cases}
1/(b-a),&a\leq x\leq b,\\
0,&\text{otherwise}.
\end{cases}
$$

## Density and probability under $\operatorname{Uniform}(0,1)$

```{r}
#| fig-width: 8
#| fig-height: 4.4
#| fig-align: center
x_uniform <- seq(-1, 2, length.out = 800)
plot(
  x_uniform, dunif(x_uniform, 0, 1),
  type = "l", lwd = 3, col = course_blue,
  xlab = "x", ylab = "Density", main = "PDF of Uniform(0, 1)",
  ylim = c(0, 1.15), bty = "l"
)
```

For $X\sim\operatorname{Uniform}(0,1)$,

$$
\mathbb P(.25<X<.75)
=\int_{.25}^{.75}1\,\mathrm dx
=.50.
$$

## Probability is area under the density

```{r}
#| fig-width: 8
#| fig-height: 4.4
#| fig-align: center
plot(
  x_uniform, dunif(x_uniform, 0, 1),
  type = "l", lwd = 3, col = course_blue,
  xlab = "x", ylab = "Density",
  main = "Probability mass of (.25, .75) under Uniform(0, 1)",
  ylim = c(0, 1.15), bty = "l"
)
polygon(c(.25, .25, .75, .75), c(0, 1, 1, 0),
        col = adjustcolor(course_red, alpha.f = .45), border = NA)
lines(x_uniform, dunif(x_uniform, 0, 1), lwd = 3, col = course_blue)
```

## CDF of $\operatorname{Uniform}(0,1)$

```{r}
#| fig-width: 8
#| fig-height: 4.4
#| fig-align: center
plot(
  x_uniform, punif(x_uniform, 0, 1),
  type = "l", lwd = 3, col = course_blue,
  xlab = "x", ylab = expression(F[X](x)),
  main = "CDF of Uniform(0, 1)", ylim = c(0, 1), bty = "l"
)
```

## The beta distribution

For $\alpha,\beta>0$,

$$
X\sim\operatorname{Beta}(\alpha,\beta)
$$

has topological support $[0,1]$ and density

$$
f_X(x)\equiv
\begin{cases}
\dfrac{x^{\alpha-1}(1-x)^{\beta-1}}{B(\alpha,\beta)},&0<x<1,\\
0,&\text{otherwise}.
\end{cases}
$$

## $\operatorname{Beta}(1,1)$ is uniform on $(0,1)$

```{r}
#| fig-width: 8
#| fig-height: 4.4
#| fig-align: center
x_beta <- seq(.001, .999, length.out = 800)
plot(
  x_beta, dbeta(x_beta, 1, 1),
  type = "l", lwd = 3, col = course_blue,
  xlab = "x", ylab = "Density", main = "PDF of Beta(1, 1)",
  ylim = c(0, 1.15), bty = "l"
)
```

## Symmetric beta densities with $\alpha=\beta>1$

```{r}
#| fig-width: 8
#| fig-height: 4.4
#| fig-align: center
plot(x_beta, dbeta(x_beta, 3, 3), type = "l", lwd = 3,
     col = course_blue, xlab = "x", ylab = "Density",
     main = "Symmetric beta densities", bty = "l")
lines(x_beta, dbeta(x_beta, 5, 5), lwd = 3, col = course_red)
lines(x_beta, dbeta(x_beta, 10, 10), lwd = 3, col = course_gray)
legend("topleft", c("Beta(3, 3)", "Beta(5, 5)", "Beta(10, 10)"),
       col = c(course_blue, course_red, course_gray), lty = 1, lwd = 3, bty = "n")
```

## Asymmetric beta densities

```{r}
#| fig-width: 8
#| fig-height: 4.4
#| fig-align: center
plot(x_beta, dbeta(x_beta, 5, 3), type = "l", lwd = 3,
     col = course_blue, xlab = "x", ylab = "Density",
     main = "Asymmetric beta densities", bty = "l")
lines(x_beta, dbeta(x_beta, 3, 5), lwd = 3, col = course_red)
legend("top", c("Beta(5, 3)", "Beta(3, 5)"),
       col = c(course_blue, course_red), lty = 1, lwd = 3, bty = "n")
```

## Boundary modes in the beta family

```{r}
#| fig-width: 8
#| fig-height: 4.4
#| fig-align: center
plot(x_beta, dbeta(x_beta, 5, 1), type = "l", lwd = 3,
     col = course_blue, xlab = "x", ylab = "Density",
     main = "Beta densities with one boundary mode", bty = "l")
lines(x_beta, dbeta(x_beta, 1, 5), lwd = 3, col = course_red)
legend("top", c("Beta(5, 1)", "Beta(1, 5)"),
       col = c(course_blue, course_red), lty = 1, lwd = 3, bty = "n")
```

## Two boundary modes when $\alpha,\beta<1$

```{r}
#| fig-width: 8
#| fig-height: 4.4
#| fig-align: center
plot(x_beta, dbeta(x_beta, .5, .5), type = "l", lwd = 3,
     col = course_blue, xlab = "x", ylab = "Density",
     main = "Beta densities with two boundary modes", ylim = c(0, 8), bty = "l")
lines(x_beta, dbeta(x_beta, .6, .4), lwd = 3, col = course_red)
lines(x_beta, dbeta(x_beta, .4, .6), lwd = 3, col = course_gray)
legend("top", c("Beta(.5, .5)", "Beta(.6, .4)", "Beta(.4, .6)"),
       col = c(course_blue, course_red, course_gray), lty = 1, lwd = 3, bty = "n")
```

## Interpreting beta parameters

For $X\sim\operatorname{Beta}(\alpha,\beta)$,

$$
\mathbb E[X]=\frac{\alpha}{\alpha+\beta}.
$$

Holding this ratio fixed while increasing $\alpha+\beta$ concentrates the distribution around the same mean.

## The normal distribution

For $\sigma>0$,

$$
X\sim\mathcal N(\mu,\sigma^2)
$$

means

$$
f_X(x)\equiv\frac{1}{\sigma\sqrt{2\pi}}
\exp\!\left[-\frac{(x-\mu)^2}{2\sigma^2}\right].
$$

## PDF of the standard normal distribution

```{r}
#| fig-width: 8
#| fig-height: 4.4
#| fig-align: center
x_normal <- seq(-6, 6, length.out = 1000)
plot(
  x_normal, dnorm(x_normal, 0, 1),
  type = "l", lwd = 3, col = course_blue,
  xlab = "x", ylab = "Density", main = "PDF of N(0, 1)", bty = "l"
)
```

## The mean controls location; the standard deviation controls scale

```{r}
#| fig-width: 8
#| fig-height: 4.4
#| fig-align: center
plot(x_normal, dnorm(x_normal, 0, 1), type = "l", lwd = 3,
     col = course_blue, xlab = "x", ylab = "Density",
     main = "Normal location and scale", bty = "l")
lines(x_normal, dnorm(x_normal, 1, 1), lwd = 3, col = course_red)
lines(x_normal, dnorm(x_normal, 0, 2), lwd = 3, col = course_gray)
legend("topleft", c("N(0, 1)", "N(1, 1)", "N(0, 2^2)"),
       col = c(course_blue, course_red, course_gray), lty = 1, lwd = 3, bty = "n")
```

## Standardizing a normal variable

Define

$$
Z\equiv\frac{X-\mu}{\sigma}.
$$

If $X\sim\mathcal N(\mu,\sigma^2)$, then $Z\sim\mathcal N(0,1)$. Standardization changes units, not relative position within this family.

## Standard normal CDF

```{r}
#| fig-width: 8
#| fig-height: 4.4
#| fig-align: center
plot(
  x_normal, pnorm(x_normal, 0, 1),
  type = "l", lwd = 3, col = course_blue,
  xlab = "x", ylab = expression(Phi(x)),
  main = "CDF of N(0, 1)", ylim = c(0, 1), bty = "l"
)
```

## A beta CDF can have the same sigmoidal form

```{r}
#| fig-width: 8
#| fig-height: 4.4
#| fig-align: center
plot(
  x_beta, pbeta(x_beta, 5, 5),
  type = "l", lwd = 3, col = course_blue,
  xlab = "x", ylab = expression(F[X](x)),
  main = "CDF of Beta(5, 5)", ylim = c(0, 1), bty = "l"
)
```

## The chi-squared distribution

For $\nu\in\{1,2,\ldots\}$, let $Z_1,\ldots,Z_\nu$ be independent standard normal variables and define

$$
Q\equiv\sum_{i=1}^{\nu}Z_i^2.
$$

Then $Q\sim\chi^2_\nu$, with $\mathbb E[Q]=\nu$ and $\operatorname{Var}(Q)=2\nu$.

## Student's t distribution

Let $Z\sim\mathcal N(0,1)$ and $U\sim\chi^2_\nu$ be independent. Define

$$
T\equiv\frac{Z}{\sqrt{U/\nu}}.
$$

Then $T\sim t_\nu$. Because $U/\nu$ can be close to zero, extreme values of $T$ are more probable than extreme values under a standard normal distribution.

::: {.notes}
[Sources]

Penn State STAT 414, Student's t Distribution: https://online.stat.psu.edu/stat414/Lesson26
:::

## Choosing a family {.statement-slide}

A distribution family specifies a support and a mass or density function. A process interpretation requires additional assumptions about how observations are generated.

## Inverse transform sampling

For [inverse transform sampling](https://bookdown.org/richard_g_everitt/notes/chapsimulation.html), define the generalized inverse of a CDF $F$ by

$$
F^{-1}(u)
\equiv\inf\{x\in\mathbb R:F(x)\geq u\},
\qquad 0<u<1.
$$

If $U\sim\operatorname{Uniform}(0,1)$, define $X\equiv F^{-1}(U)$.

::: {.notes}
[Sources]

Everitt, Computer Intensive Statistics, Simulation-Based Inference: https://bookdown.org/richard_g_everitt/notes/chapsimulation.html
:::

## Verifying the target distribution

The generalized inverse satisfies $F^{-1}(U)\leq x$ exactly when $U\leq F(x)$. Thus

$$
\begin{aligned}
\mathbb{P}(X\leq x)
&=\mathbb{P}(F^{-1}(U)\leq x)\\
&=\mathbb{P}(U\leq F(x))\\
&=F(x).
\end{aligned}
$$

## Discrete inverse transform sampling

For ordered values $x_1,\ldots,x_K$, define

$$
c_k\equiv\sum_{j=1}^kp_X(x_j).
$$

Draw $U\sim\operatorname{Uniform}(0,1)$ and return the first $x_k$ for which $U\leq c_k$.

## Random seeds reproduce computations

A random seed fixes the initial state of a pseudorandom number generator. Reusing the seed reproduces the same sequence of draws.

It does not validate the distribution, independence assumptions, or sampling representation.

## Joint distributions {.section-slide}

<span class="section-subtitle">September 23</span>

## Several variables use one probability space

Let $X:\Omega\rightarrow\mathcal X$ and $Y:\Omega\rightarrow\mathcal Y$ be random variables on the same probability space.

The pair $(X(\omega),Y(\omega))$ records two values from the same outcome $\omega$.

## Defining a joint PMF

For discrete $X$ and $Y$, define

$$
p_{X,Y}(x,y)
\equiv\mathbb{P}(X=x,Y=y).
$$

The comma denotes the joint event $\{X=x\}\cap\{Y=y\}$.

## A joint linguistic distribution

Let $P$ code whether a dialogue turn contains a politeness marker and $R$ whether it ends in a final pitch rise.

| | $R=1$ | $R=0$ | total |
|:--|--:|--:|--:|
| $P=1$ | .30 | .20 | .50 |
| $P=0$ | .10 | .40 | .50 |
| total | .40 | .60 | 1 |

## Reading a joint cell

The upper-left cell states

$$
p_{P,R}(1,1)
=.30.
$$

It assigns probability to the paired values for one dialogue turn.

## Marginalization is a partition argument

For fixed $p$, the events $\{P=p,R=r\}$ partition $\{P=p\}$ as $r$ ranges over the support of $R$:

$$
\{P=p\}
=\biguplus_{r\in\operatorname{supp}(R)}
 \{P=p,R=r\}.
$$

Countable additivity gives

$$
\mathbb{P}(P=p)
=\sum_r\mathbb{P}(P=p,R=r).
$$

Applying the PMF definitions to both sides yields

$$
p_P(p)
=\sum_r p_{P,R}(p,r).
$$

## Computing a marginal PMF

For turns with a politeness marker,

$$
\begin{aligned}
p_P(1)
&=p_{P,R}(1,1)+p_{P,R}(1,0)\\
&=.30+.20\\
&=.50.
\end{aligned}
$$

We sum out pitch rise while holding politeness fixed.

## What marginalization removes

The marginal PMFs $p_P$ and $p_R$ retain the row and column totals. They do not retain how probability is paired within the four cells.

Different joint PMFs can have the same marginals.

## Marginal density

If $X$ and $Y$ have joint density $f_{X,Y}$, define the marginal density of $X$ by

$$
f_X(x)
\equiv\int_{-\infty}^{\infty}f_{X,Y}(x,y)\,\mathrm{d}y.
$$

The integral removes the $Y$ coordinate.

## Defining a conditional PMF

For any $y$ with $p_Y(y)>0$, define

$$
p_{X\mid Y}(x\mid y)
\equiv\mathbb{P}(X=x\mid Y=y)
=\frac{p_{X,Y}(x,y)}{p_Y(y)}.
$$

## Deriving the chain rule for random variables

Multiply the conditional-PMF definition by $p_Y(y)$:

$$
p_{X,Y}(x,y)
=p_{X\mid Y}(x\mid y)p_Y(y).
$$

This is the two-variable chain rule.

## Defining independence through conditional invariance

For every $y$ with $p_Y(y)>0$, independence requires

$$
p_{X\mid Y}(x\mid y)=p_X(x)
$$

for every $x$. Learning $Y=y$ does not change the distribution of $X$.

## Deriving the independence factorization

Substitute conditional invariance into the chain rule:

$$
\begin{aligned}
p_{X,Y}(x,y)
&=p_{X\mid Y}(x\mid y)p_Y(y)\\
&=p_X(x)p_Y(y).
\end{aligned}
$$

The product form is the equivalent unrestricted criterion for $X\perp\!\!\!\perp Y$.

## Defining conditional independence

For values with $p_{Y,Z}(y,z)>0$, $X$ and $Y$ are conditionally independent given $Z$ when

$$
p_{X\mid Y,Z}(x\mid y,z)=p_{X\mid Z}(x\mid z).
$$

We write $X\perp\!\!\!\perp Y\mid Z$.

## Conditional independence factorization

For every $z$ with $p_Z(z)>0$, the conditional chain rule gives the equivalent criterion

$$
X\perp\!\!\!\perp Y\mid Z
\quad\Longleftrightarrow\quad
p_{X,Y\mid Z}(x,y\mid z)
=p_{X\mid Z}(x\mid z)p_{Y\mid Z}(y\mid z).
$$

The conditioning variable remains in both factors.

## Defining conditional expectation

For discrete $X$ and any $y$ with $p_Y(y)>0$, define

$$
\mathbb E[X\mid Y=y]
\equiv\sum_x x\,p_{X\mid Y}(x\mid y).
$$

This is the mean of the conditional distribution at $Y=y$.

## Deriving the law of total expectation

For integrable discrete variables,

$$
\begin{aligned}
\mathbb E[X]
&=\sum_xx\sum_yp_{X,Y}(x,y)\\
&=\sum_{y:p_Y(y)>0}\sum_xx\,p_{X,Y}(x,y)\\
&=\sum_{y:p_Y(y)>0}
  \left[\sum_xx\,p_{X\mid Y}(x\mid y)\right]p_Y(y)\\
&=\sum_{y:p_Y(y)>0}\mathbb E[X\mid Y=y]p_Y(y).
\end{aligned}
$$

## Defining covariance

For variables with finite second moments, define

$$
\operatorname{Cov}(X,Y)
\equiv\mathbb E\!\left[(X-\mathbb E[X])(Y-\mathbb E[Y])\right].
$$

Its units are the product of the units of $X$ and $Y$.

## Deriving the covariance identity

Expanding the product and applying linearity gives

$$
\operatorname{Cov}(X,Y)
=\mathbb E[XY]-\mathbb E[X]\mathbb E[Y].
$$

This identity does not define covariance; it follows from the preceding definition.

## Correlation

When both standard deviations are positive, define

$$
\rho_{XY}
\equiv
\frac{\operatorname{Cov}(X,Y)}
{\sigma_X\sigma_Y}.
$$

Correlation is unitless and satisfies $-1\leq\rho_{XY}\leq1$.

## Correlation has a limited interpretation

Correlation summarizes linear association under the joint distribution. It does not determine the complete dependence structure and does not identify a causal direction.

## Module Summary {.compact}

1. A random variable maps outcomes to values and induces a distribution.
2. PMFs assign mass; PDFs integrate to probability; CDFs accumulate probability.
3. Expectations summarize specified functions when they exist.
4. Distribution families fix a support and probability form, sometimes under process assumptions.
5. Joint distributions determine marginals, conditionals, dependence, and paired summaries.
