# Continuous Random Variables {.center background-color="#2166ac"}

## Continuous Random Variables

A **continuous random variable** is a random variable whose range is uncountable.

::: {.fragment}
One example of a continuous random variable (mentioned earlier) is one where $\Omega$ is the set of all pairs of first and second formant values.
:::

::: {.fragment}
In this case, we'll assume that $\Omega$ is just all pairs of positive real numbers $\mathbb{R}_+^2$.
:::

---

## Formant Values Example

```{r}
#| echo: true
# Formant values as pairs of positive reals
library(tidyverse)

# Example formant measurements
formant_data <- tibble(
  speaker = paste0("S", 1:5),
  f1 = c(280, 320, 295, 305, 290),  # F1 values
  f2 = c(2250, 2180, 2300, 2150, 2200)  # F2 values
)

print(formant_data)
```

::: {.fragment}
Range: $\Omega = \mathbb{R}_+^2$ (all pairs of positive real numbers)
:::

---

## The Identity Function Example

If we assume that the random variable $F: \mathbb{R}_+^2 \rightarrow \mathbb{R}^2$ is the identity function $F(\mathbf{x}) = \mathbf{x}$, we get that $F$ is a continuous random variable, since $\mathbb{R}$ is uncountable and $F^{-1}(E) = E \in \mathcal{F}$.

```{r}
#| echo: true
# F is the identity function: F(x) = x
F <- function(formant_pair) {
  return(formant_pair)  # Identity: returns the input unchanged
}

# Example: F applied to a formant pair
example_pair <- c(f1 = 300, f2 = 2200)
F(example_pair)
```

::: {.fragment}
**Key insight**: $F$ maps formant measurements to themselves: it's the identity function.
:::

---

## Borel σ-Algebras

The event space for $\Omega = \mathbb{R}_+^2$ is analogous to the Borel $\sigma$-algebra for $\mathbb{R}$.

::: {.fragment}
**What you need to know about Borel σ-algebras**: 

- They contain every real interval
- $E \in \mathcal{G}$ will always be an interval (not just a single real number!)
- Basically, it contains all pairs of real intervals
:::

::: {.fragment}
The technical details aren't really going to be important for our purposes beyond knowing that $\mathbb{R}_+^2$ is going to act like $\mathbb{R}$ in the ways we care about.
:::

---

## Visualizing Continuous Data

```{r}
#| echo: false
#| fig-height: 5
library(tidyverse)
library(ggplot2)

# Simulate F1 values for /a/ vowel
set.seed(42)
f1_values <- rnorm(200, 700, 50)  # mean=700Hz, sd=50Hz for /a/
f1_data <- tibble(f1 = f1_values)

ggplot(f1_data, aes(x = f1)) +
  geom_histogram(aes(y = after_stat(density)), 
                 bins = 20, fill = "#2166ac", alpha = 0.7) +
  geom_density(color = "#b2182b", linewidth = 1.2) +
  labs(title = "F1 Distribution for /a/",
       x = "F1 (Hz)", y = "Density") +
  theme_minimal()
```

Continuous: Any value in the range is possible!

---

## 2D Example: F1-F2 Space

```{r}
#| echo: false
#| fig-height: 5
# Simulate vowel formants
set.seed(42)
vowel_data <- bind_rows(
  tibble(vowel = "i", 
         f1 = rnorm(50, 300, 30),
         f2 = rnorm(50, 2300, 150)),
  tibble(vowel = "a", 
         f1 = rnorm(50, 700, 50),
         f2 = rnorm(50, 1200, 100)),
  tibble(vowel = "u", 
         f1 = rnorm(50, 350, 30),
         f2 = rnorm(50, 900, 80))
)

ggplot(vowel_data, aes(x = f2, y = f1, color = vowel)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_x_reverse() + scale_y_reverse() +
  scale_color_manual(values = c("i" = "#e41a1c", 
                                "a" = "#377eb8", 
                                "u" = "#4daf4a")) +
  labs(title = "Vowel Space: F1-F2 Measurements",
       x = "F2 (Hz)", y = "F1 (Hz)") +
  theme_minimal()
```

$F: \mathbb{R}^2_+ \rightarrow \mathbb{R}^2$ (identity function on formant pairs)

---

## Why Can't We Use PMFs for Continuous RVs?

Once we move to working with random variables that have an uncountable number of values–as in the case of our formant value example above–we can no longer assign a non-zero probability to every value that variable takes.

::: {.fragment}
**The intuition**: There are just too many numbers (e.g. too many possible formant values); and if we assigned non-zero probability to more than countably many of them, we'd end up with a sum that doesn't satisfy the assumption of σ-additivity.
:::

::: {.fragment}
**Key fact**: Every possibility in a real-valued sample space has probability zero: $\forall x \in \Omega: \mathbb{P}(x) = 0$.
:::

::: {.fragment}
This means that a PMF isn't going to be useful here.
:::

---

## Preview: Probability Density Functions

What we work with instead is a **probability density function (PDF)** $f_X: \Omega \rightarrow \mathbb{R}_+$.

::: {.fragment}
**Important distinction**: Where the PMF tells you the probability of a particular possibility, the PDF does **not**.
:::

::: {.fragment}
The PDF can be used to compute a probability: specifically, the probability of $X$ taking on a value in some set:
$$\mathbb{P}(X \in A) = \int_A f_X(x) \, \mathrm{d}x$$
:::

::: {.fragment}
The next topic develops [probability density functions](probability-density-functions.qmd) in detail.
:::

---

## Summary: Discrete vs Continuous

| | **Discrete** | **Continuous** |
|---|---|---|
| **Range** | Countable | Uncountable |
| **Examples** | Pronoun count, word length | Formant values, VOT |
| **Individual P(X=x)** | Can be > 0 | Always = 0 |
| **Distribution** | PMF | PDF |
| **Probabilities** | Sums | Integrals |

::: {.fragment}
**Next time**: We'll see families of discrete distributions in detail!
:::
