# Model Types

## Three types of models

We distinguish three types of models based on how they handle group structure:

1. **Fixed effects models**: All predictors are fixed; groups ignored or included as fixed effects
2. **Random effects models**: Only random effects (no fixed predictors); ANOVA-style
3. **Mixed effects models**: Both fixed predictors AND random effects

---

## Fixed effects model

**Definition**: All variation is either explained by fixed predictors or treated as residual.

$$Y_{ij} = \beta_0 + \beta_1 x_{ij} + \epsilon_{ij}$$

**Assumptions**:

- Observations are independent (or independence after conditioning on $x$)
- All systematic variation captured by $\beta$ coefficients
- No group structure or grouping ignored

**Example**: Problem Set 3 models

- Aggregated MALD data: one observation per word (averaged over participants)
- Fixed effects for frequency
- Grouping structure (participants, items) removed by aggregation

---

## Random effects model

**Definition**: Group means vary, but no fixed predictors beyond the intercept.

$$Y_{ij} = \mu + u_i + \epsilon_{ij}$$

where:

- $\mu$: Overall population mean (grand mean)
- $u_i \sim N(0, \sigma_u^2)$: Random effect for group $i$
- $\epsilon_{ij} \sim N(0, \sigma^2)$: Residual error

**Example**: One-way random effects ANOVA

- Measuring RT for different participants, no predictors
- Question: How much do participants vary?
- Estimate $\sigma_u^2$ (between-participant variance)

---

## Mixed effects model

**Definition**: Combines fixed predictors (systematic component) and random effects (variance component).

$$Y_{ij} = \beta_0 + \beta_1 x_{ij} + u_i + \epsilon_{ij}$$

where:

- $\beta_0, \beta_1$: Fixed effects (population-level coefficients)
- $u_i \sim N(0, \sigma_u^2)$: Random intercepts
- $\epsilon_{ij} \sim N(0, \sigma^2)$: Residual error

**Example**: MALD lexical decision

- Fixed effects: frequency (predictor of interest)
- Random effects: participants, items (nuisance variation to control for)

---

## Visual comparison: Fixed vs. Mixed

```{r model-type-comparison}
#| echo: false
#| fig-width: 12
#| fig-height: 5

# Load MALD data
mald <- read_csv("data/mald_trials.csv", show_col_types = FALSE)

# Center predictor
mald <- mald %>%
    mutate(log_freq_c = log_freq_subtlex - mean(log_freq_subtlex))

# Subset for visualization
set.seed(42)
subjects_subset <- sample(unique(mald$subject), 6)
mald_subset <- mald %>% filter(subject %in% subjects_subset)

# Fixed effects model (ignores subjects)
p1 <- ggplot(mald_subset, aes(x = log_freq_c, y = log_rt)) +
    geom_point(alpha = 0.3, size = 1) +
    geom_smooth(method = "lm", se = TRUE, color = "red", linewidth = 1.5) +
    labs(x = "Log Frequency (centered)", y = "Log RT",
         title = "Fixed Effects Model",
         subtitle = "Single regression line (ignores participant structure)") +
    theme_minimal(base_size = 14)

# Mixed effects model (accounts for subjects)
p2 <- ggplot(mald_subset, aes(x = log_freq_c, y = log_rt, color = subject)) +
    geom_point(alpha = 0.5, size = 1) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
    geom_smooth(aes(group = 1), method = "lm", se = TRUE,
                color = "black", linewidth = 1.5) +
    scale_color_brewer(palette = "Set2") +
    labs(x = "Log Frequency (centered)", y = "Log RT",
         title = "Mixed Effects Model",
         subtitle = "Subject-specific intercepts (colored) + population average (black)") +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

Mixed model accounts for participant-level variation, improving both fit and inference.

---

## Model comparison: ANOVA and likelihood ratio tests

Once we fit multiple models, how do we compare them?

**ANOVA** = Analysis of Variance

- Statistical method for comparing **nested models**
- Nested: One model is a special case of another (e.g., M1 ⊂ M3)

**Likelihood ratio test**:

$$\text{LR} = -2(\log L_{\text{simpler}} - \log L_{\text{complex}})$$

- $\text{LR} \sim \chi^2$ with df = difference in number of parameters
- Null hypothesis: Simpler model is sufficient
- Reject if $p < \alpha$ → Complex model significantly better

**In R**:

```r
anova(model_simple, model_complex)
```

**Example**: Comparing M1 (subject REs) vs M3 (subject + item REs)

---

## When to use each model type

```{r decision-tree}
#| echo: false
#| fig-width: 10
#| fig-height: 7

library(ggplot2)
library(grid)

# Create corrected decision tree
tree_data <- tibble(
    x = c(5, 3, 7, 2, 4, 1.5, 2.5),
    y = c(7, 5, 5, 3, 3, 1, 1),
    label = c(
        "Grouping structure\nin your data?",
        "Want to generalize\nbeyond observed groups?",
        "FIXED EFFECTS\nMODEL",
        "Have predictors\nbeyond group means?",
        "FIXED EFFECTS\nMODEL\n(groups as fixed)",
        "MIXED EFFECTS\nMODEL",
        "RANDOM EFFECTS\nMODEL\n(ANOVA)"
    ),
    type = c("question", "question", "answer", "question", "answer", "answer", "answer")
)

tree_edges <- tibble(
    x = c(5, 5, 3, 3, 2, 2),
    y = c(7, 7, 5, 5, 3, 3),
    xend = c(3, 7, 2, 4, 1.5, 2.5),
    yend = c(5, 5, 3, 3, 1, 1),
    label_text = c("YES", "NO", "YES", "NO", "YES", "NO"),
    label_x = c(4, 6, 2.5, 3.5, 1.75, 2.25),
    label_y = c(6, 6, 4, 4, 2, 2)
)

ggplot() +
    geom_segment(data = tree_edges,
                 aes(x = x, y = y, xend = xend, yend = yend),
                 linewidth = 1, color = "gray30") +
    geom_label(data = tree_edges,
               aes(x = label_x, y = label_y, label = label_text),
               size = 4, fill = "white", label.size = 0) +
    geom_point(data = tree_data,
               aes(x = x, y = y, color = type, size = type)) +
    geom_label(data = tree_data,
               aes(x = x, y = y, label = label, fill = type),
               size = 4, fontface = "bold", label.padding = unit(0.5, "lines")) +
    scale_color_manual(values = c("answer" = "darkgreen", "question" = "steelblue")) +
    scale_fill_manual(values = c("answer" = "lightgreen", "question" = "lightblue")) +
    scale_size_manual(values = c("answer" = 8, "question" = 10)) +
    scale_x_continuous(limits = c(0, 8.5)) +
    scale_y_continuous(limits = c(0, 8)) +
    labs(title = "Decision Tree: Which Model Type?",
         subtitle = "Follow the path based on your data structure and research goals") +
    theme_void(base_size = 14) +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
          plot.subtitle = element_text(hjust = 0.5, size = 12))
```

---

## Fixed effects: When to use

Use **fixed effects** when:

**No group structure**:

- Observations are truly independent
- No repeated measures or clustering

**Groups are fixed**:

- Small number of groups (e.g., 2-3 experimental conditions)
- You care about those specific groups, not a larger population
- Example: Comparing British vs. American English (fixed levels)

**Example**: [Simple linear regression](simple-linear-regression.qmd)

- Self-paced reading with surprisal predicting RT
- Each word presented once to each participant
- Participants treated as a single pooled sample (or controlled via fixed effects)

---

## Random effects: When to use

Use **random effects** when:

**Generalizing beyond sample**:

- Groups are a random sample from a population
- Inferences should extend to new, unseen groups
- Example: 30 participants from population of speakers

**Many groups**:

- Large number of group levels (>5-10)
- Estimating fixed effects for each would be inefficient

**Variance is the quantity of interest**:

- Research question: "How much do groups vary?"
- Example: "How much do dialects vary in vowel duration?"

---

## Mixed effects: When to use

Use **mixed effects** when:

**Both fixed predictors AND grouping structure**:

- You have predictors of interest (fixed effects)
- AND observations are clustered within groups (random effects)

**Most common in linguistics**:

- **Repeated measures**: Same participants tested multiple times
- **Crossed effects**: Participants × Items design
- **Hierarchical structure**: Words nested within semantic classes

**Examples**:

- Lexical decision: Participants respond to multiple words
- Sentence processing: Multiple sentences per participant
- Corpus data: Words nested within documents, documents within genres

---

## Residual patterns: Fixed vs. Mixed

```{r residuals-comparison-types}
#| echo: false
#| fig-width: 12
#| fig-height: 5

# Load models
load("data/model_results/all_models.RData")

# Get residuals
mald$resid_m0 <- residuals(m0)
mald$resid_m3 <- residuals(m3)

# Randomly select 10 subjects for clarity
set.seed(42)
subjects_sample <- sample(unique(mald$subject), 10)

# M0 residuals by subject
p1 <- mald %>%
    filter(subject %in% subjects_sample) %>%
    ggplot(aes(x = subject, y = resid_m0)) +
    geom_boxplot(fill = "coral", alpha = 0.6) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    labs(x = "Subject", y = "Residuals",
         title = "Fixed Effects Model (M0)",
         subtitle = "Systematic patterns remain: Some subjects always positive, others negative") +
    theme_minimal(base_size = 14) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

# M3 residuals by subject
p2 <- mald %>%
    filter(subject %in% subjects_sample) %>%
    ggplot(aes(x = subject, y = resid_m3)) +
    geom_boxplot(fill = "steelblue", alpha = 0.6) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    labs(x = "Subject", y = "Residuals",
         title = "Mixed Effects Model (M3)",
         subtitle = "Random effects absorbed subject-level variation: Residuals centered at 0") +
    theme_minimal(base_size = 14) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

Mixed model removes systematic patterns, satisfying independence assumption.

---

## Model notation summary

**Fixed effects model** (all variation residual):

$$Y_i = \beta_0 + \beta_1 x_i + \epsilon_i$$

**Random effects model** (group means vary):

$$Y_{ij} = \mu + u_i + \epsilon_{ij}, \quad u_i \sim N(0, \sigma_u^2)$$

**Mixed effects model** (both fixed and random):

$$Y_{ij} = \beta_0 + \beta_1 x_{ij} + u_i + \epsilon_{ij}, \quad u_i \sim N(0, \sigma_u^2)$$

**Crossed random effects** (multiple grouping factors):

$$Y_{ijk} = \beta_0 + \beta_1 x_{ijk} + u_i + v_j + \epsilon_{ijk}$$

where $u_i$ and $v_j$ are independent random effects.

---

## Summary: Model types

| Model Type | Formula | Use Case |
|------------|---------|----------|
| **Fixed** | $Y_i = \beta_0 + \beta_1 x_i + \epsilon_i$ | No groups; groups as fixed levels |
| **Random** | $Y_{ij} = \mu + u_i + \epsilon_{ij}$ | Variance of interest; no predictors |
| **Mixed** | $Y_{ij} = \beta_0 + \beta_1 x_{ij} + u_i + \epsilon_{ij}$ | Fixed predictors + grouping structure |

**Key decision**: Do you want to generalize beyond observed groups? YES → Random effects

---
