# Frequentist Approach

## We already derived OLS = MLE

In Section 3, we showed:

- Under normality assumption, **OLS is the MLE**
- MLEs have good statistical properties (consistency, asymptotic normality)
- We know the formulas: $\hat\beta_1 = \text{Cov}(X,Y)/\text{Var}(X)$, $\hat\beta_0 = \bar{y} - \hat\beta_1\bar{x}$

Now we fit the model in R and interpret results.

---

## Fitting in R

```{r fit-lm}
#| echo: true
#| code-fold: false

# Fit linear model: log RT ~ surprisal
model <- lm(log_rt ~ surprisal, data = word_data)

# View summary
summary(model)
```

---

## Interpreting the output

```{r extract-coefs}
#| echo: true
#| code-fold: false

# Extract coefficients
coef(model)

# Extract confidence intervals
confint(model)
```

The slope coefficient tells us how log RT changes per bit increase in surprisal.

A positive coefficient confirms that higher surprisal predicts slower reading.

---

## Visualizing the fit

```{r plot-fit}
#| echo: false
#| fig-width: 8
#| fig-height: 5

ggplot(word_data, aes(x = surprisal, y = log_rt)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(x = "Surprisal (bits)",
       y = "Log Reaction Time (log seconds)",
       title = "Fitted Regression Line with 95% Confidence Band")
```

---

# Confidence Intervals

## Connecting to prior material

**Recall from [confidence intervals](../statistical-inference/confidence-intervals.qmd)**: intervals for the Bernoulli parameter $\pi$

- **Clopper-Pearson method**: Exact confidence intervals using binomial distribution
  - Complex formulas involving binomial CDFs
  - Required computing quantiles: $\inf\{\theta \mid P(X \geq k) > (1-\gamma)/2\}$

**Recall from Problem Set 1**: Bootstrap CIs for Poisson $\lambda$

- Resampled data 10,000 times with replacement
- Computed $\hat\lambda$ on each resample
- Took 2.5% and 97.5% quantiles of bootstrap distribution

---

## Why regression CIs are different

**Challenge**: In regression, we're estimating multiple parameters simultaneously: $\beta_0, \beta_1, \sigma$

- For Bernoulli: One parameter $\pi$, known sampling distribution
- For regression: Multiple parameters, more complex dependencies

**Good news**: Under the normality assumption, we know the exact sampling distribution of $\hat{\boldsymbol{\beta}}$

**Key result**: $$\frac{\hat\beta_j - \beta_j}{SE(\hat\beta_j)} \sim t_{N-p-1}$$

This gives us a simple closed-form formula for confidence intervals (no complex binomial quantiles, no bootstrap needed!)

---

## Confidence intervals for regression coefficients

A $(1-\alpha) \times 100\%$ confidence interval for $\beta_j$ is:

$$\hat\beta_j \pm t_{\alpha/2, N-p-1} \times SE(\hat\beta_j)$$

where $t_{\alpha/2, N-p-1}$ is the critical value from the $t$-distribution.

```{r confint-regression}
#| echo: true
#| code-fold: false

# 95% confidence intervals for all coefficients
confint(model, level = 0.95)
```

---

## Interpreting confidence intervals

For our model: `log_rt ~ surprisal`

Suppose the 95% CI for $\beta_1$ (surprisal) is $[0.012, 0.018]$.

**Interpretation**: We are 95% confident that the true effect of surprisal lies in this interval.

**In context**: Each additional bit of surprisal increases log RT by between 0.012 and 0.018 log seconds.

**Connection to hypothesis testing**: Since the interval doesn't contain 0, we can reject $H_0: \beta_1 = 0$ at $\alpha = 0.05$.

---

# Regression Diagnostics

## The importance of assumptions

OLS inference validity depends on our three assumptions:

1. **Independence**: Observations are independent
2. **Conditional normality**: $Y_i \mid X_i \sim \text{Normal}(\beta_0 + \beta_1 x_i, \sigma^2)$
3. **Homoscedasticity**: Constant variance $\sigma^2$

We check these using residuals: $e_i = y_i - \hat{y}_i$

---

## Residuals

The residual for observation $i$ is:

$$e_i = y_i - \hat{y}_i = y_i - (\hat\beta_0 + \hat\beta_1 x_i)$$

Residuals estimate the true errors $\epsilon_i$.

If assumptions hold, residuals should behave like random noise.

---

## Extracting residuals

```{r extract-residuals}
#| echo: true
#| code-fold: false

# Extract residuals and fitted values
residuals <- residuals(model)
fitted_values <- fitted(model)

# Summary statistics
summary(residuals)
```

---

# Normality Assumption

## Why normality matters

Normality of $\epsilon_i$ is required for:

1. Valid hypothesis tests ($t$-tests, $F$-tests)
2. Accurate confidence intervals
3. Valid prediction intervals

Departures invalidate statistical inference.

---

## Q-Q plot

```{r qq-plot}
#| echo: true
#| code-fold: false
#| fig-width: 7
#| fig-height: 5

qqnorm(residuals, main = "Normal Q-Q Plot")
qqline(residuals, col = "red", lwd = 2)
```

If residuals are normal, points should fall on the line. Systematic deviations indicate non-normality.

---

## Histogram of residuals

```{r histogram}
#| echo: true
#| code-fold: false
#| fig-width: 7
#| fig-height: 5

hist(residuals, breaks = 30, freq = FALSE,
     main = "Distribution of Residuals",
     xlab = "Residuals",
     col = "lightblue", border = "white")
curve(dnorm(x, mean = mean(residuals), sd = sd(residuals)),
      add = TRUE, col = "red", lwd = 2)
```

---

## Shapiro-Wilk test

**Test**: $H_0$: Residuals $e_1, \ldots, e_N \sim \text{Normal}(\mu, \sigma^2)$

**Test statistic**: $$W = \frac{(\sum_{i=1}^N a_i e_{(i)})^2}{\sum_{i=1}^N (e_i - \bar{e})^2}$$

where:
- $e_{(i)}$ are ordered residuals
- $a_i$ are constants derived from expected order statistics of standard normal distribution

**What W measures**: Correlation between ordered residuals and expected normal quantiles

- $W \in [0, 1]$; values close to 1 indicate normality
- Small $W$ = evidence against normality

**Decision**: Reject $H_0$ if $p < \alpha$

**Limitation**: Sample size $N \leq 5000$

---

```{r shapiro-test}
#| echo: true
#| code-fold: false

# Shapiro-Wilk test has max sample size of 5000
if (length(residuals) <= 5000) {
  shapiro.test(residuals)
} else {
  # Use random subsample for large datasets
  set.seed(42)
  subsample <- sample(residuals, 5000)
  cat("Note: Using random subsample of 5000 observations (N =", length(residuals), ")\n")
  shapiro.test(subsample)
}
```

---

## Kolmogorov-Smirnov test

**Test**: $H_0$: Residuals follow $\text{Normal}(\mu, \sigma^2)$

**Test statistic**: $$D = \sup_x |F_N(x) - \Phi(x)|$$

where:
- $F_N(x)$ = empirical CDF (proportion of residuals $\leq x$)
- $\Phi(x)$ = standard normal CDF

**What D measures**: Maximum vertical distance between empirical and theoretical CDFs

- **Intuition**: How far is the "worst" point from the normal distribution?
- Small $D$ = good fit to normality
- Large $D$ = poor fit, evidence against normality

**Note**: When $\mu, \sigma$ are estimated from data, test becomes conservative (may fail to reject even when data are non-normal)

```{r ks-test}
#| echo: true
#| code-fold: false

# Standardize residuals
std_residuals <- scale(residuals)
# KS test against standard normal
ks.test(std_residuals, "pnorm")
```

**Note**: $p$-value may be overly conservative when parameters are estimated.

---

## Anderson-Darling test

**Test**: $H_0$: Residuals $\sim \text{Normal}(\mu, \sigma^2)$

**Test statistic**: $$A^2 = -N - \frac{1}{N}\sum_{i=1}^N (2i-1)[\log \Phi(e_{(i)}) + \log(1-\Phi(e_{(N+1-i)}))]$$

where $e_{(i)}$ are ordered residuals.

**What A² measures**: Weighted squared distance between empirical and theoretical CDFs

- **Key advantage**: Gives more weight to deviations in the distribution tails than KS test
- Small $A^2$ = consistent with normality
- Large $A^2$ = evidence against normality (especially tail behavior)

**Why it's better**: More sensitive to departures that matter for inference (tail probabilities)

```{r ad-test}
#| echo: true
#| code-fold: false
#| warning: false

library(nortest)
ad.test(residuals)
```

---

# Homoscedasticity

## Constant variance assumption

**Assumption**: $\text{Var}(\epsilon_i \mid X_i = x) = \sigma^2$ for all $x$

**Violation** (heteroscedasticity) causes:
1. Inefficient estimates (higher variance)
2. Biased standard errors → invalid tests
3. Confidence intervals have wrong coverage

---

## Residuals vs. fitted plot

```{r residuals-fitted}
#| echo: true
#| code-fold: false
#| fig-width: 7
#| fig-height: 5

plot(fitted_values, residuals,
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs. Fitted Values")
abline(h = 0, col = "red", lwd = 2)
```

Random scatter around zero with constant spread indicates homoscedasticity. Funnel shapes indicate heteroscedasticity.

---

## Scale-location plot

```{r scale-location}
#| echo: true
#| code-fold: false
#| fig-width: 7
#| fig-height: 5

standardized_res <- rstandard(model)

plot(fitted_values, sqrt(abs(standardized_res)),
     xlab = "Fitted Values",
     ylab = expression(sqrt("|Standardized Residuals|")),
     main = "Scale-Location Plot")
lines(lowess(fitted_values, sqrt(abs(standardized_res))),
      col = "red", lwd = 2)
```

Horizontal line with equal spread indicates homoscedasticity.

---

## Breusch-Pagan test

**Test**: $H_0$: Homoscedasticity ($\text{Var}(\epsilon_i) = \sigma^2$ constant)

**Procedure**:
1. Fit original model, obtain residuals $\hat\epsilon_i$
2. Regress $\hat\epsilon_i^2$ on predictors $X_1, \ldots, X_p$
3. Compute $\text{BP} = N \times R^2$ from auxiliary regression
4. Under $H_0$: $\text{BP} \sim \chi^2_p$ (asymptotically)

**Decision**: Reject $H_0$ if $\text{BP} > \chi^2_{p,\alpha}$ critical value

```{r bp-test}
#| echo: true
#| code-fold: false
#| warning: false

library(lmtest)
bptest(model)
```

---

# Influential Observations

## Outliers and leverage

**Outlier**: Observation with large residual (unusual $Y$ value)

**High leverage**: Observation with extreme $X$ value

**Influential**: Observation that substantially changes regression when removed

---

## Cook's distance

**Measure**: Influence of observation $i$ on all fitted values

$$D_i = \frac{\sum_{j=1}^N (\hat{y}_j - \hat{y}_{j(i)})^2}{(p+1) \times \text{MSE}}$$

where $\hat{y}_{j(i)}$ is fitted value when observation $i$ is excluded.

**Rule of thumb**: $D_i > 1$ indicates high influence (or $D_i > 4/N$ for smaller threshold)

```{r cooks-distance}
#| echo: true
#| code-fold: false
#| fig-width: 7
#| fig-height: 5

cooks_d <- cooks.distance(model)

plot(cooks_d, type = "h",
     main = "Cook's Distance",
     ylab = "Cook's Distance",
     xlab = "Observation Index")
abline(h = 1, col = "red", lty = 2, lwd = 2)
```

---

## Residuals vs. leverage plot

```{r residuals-leverage}
#| echo: true
#| code-fold: false
#| fig-width: 7
#| fig-height: 5

plot(model, which = 5)
```

Points in upper/lower right corners are both outliers and high leverage (most influential).

---

## Identifying influential points

```{r identify-influential}
#| echo: true
#| code-fold: false

# Identify observations with high Cook's distance
influential <- which(cooks_d > 4/length(cooks_d))

cat("Number of influential observations:", length(influential), "\n")
if(length(influential) > 0 && length(influential) <= 10) {
  cat("Their Cook's distances:", head(cooks_d[influential], 10), "\n")
}
```

---

# Model Fit

## Coefficient of determination

$$R^2 = 1 - \frac{\text{RSS}}{\text{TSS}} = 1 - \frac{\sum_{i=1}^N (y_i - \hat{y}_i)^2}{\sum_{i=1}^N (y_i - \bar{y})^2}$$

- $R^2 \in [0,1]$: Proportion of variance in $Y$ explained by the model
- Higher values indicate better fit to the observed data
- **Important caveat**: $R^2$ always increases when adding more predictors
  - This doesn't necessarily mean a better model!
  - We need to assess for overfitting

```{r r-squared}
#| echo: true
#| code-fold: false

summary(model)$r.squared
```

**Note**: A later unit on model evaluation will develop methods for guarding against overfitting.

---

# Hypothesis Testing

## Connecting to Problem Set 2

**Recall Exercise 4.2**: You manually computed Welch's t-statistic for comparing vowel formants (IY vs. IH)

**What you did**:

1. Extracted F1 values for each vowel
2. Computed means, standard deviations, sample sizes
3. Calculated t-statistic: $$t = \frac{\bar{x}_{\text{IY}} - \bar{x}_{\text{IH}}}{\sqrt{s^2_{\text{IY}}/n_{\text{IY}} + s^2_{\text{IH}}/n_{\text{IH}}}}$$
4. Computed degrees of freedom using Welch-Satterthwaite formula
5. Found p-value using `pt(-abs(t), df)`

**Same idea for regression**: Test if a coefficient differs from 0

---

## Hypothesis test for regression coefficients

Testing whether a predictor has an effect on the response:

- **Null hypothesis** $H_0$: $\beta_1 = 0$ (no relationship between $X$ and $Y$)
- **Alternative hypothesis** $H_1$: $\beta_1 \neq 0$ (X affects Y)

**Test statistic**:

$$t = \frac{\hat\beta_1 - 0}{SE(\hat\beta_1)} = \frac{\hat\beta_1}{SE(\hat\beta_1)}$$

**Under $H_0$**: $t \sim t_{N-2}$ (for simple regression)

**Decision rule**: Reject $H_0$ if $|t| > t_{\alpha/2, N-2}$ or equivalently if $p < \alpha$

---

## Understanding the test statistic

The t-statistic measures how many standard errors the estimate is from the null value:

$$t = \frac{\hat\beta_1}{SE(\hat\beta_1)}$$

- **Numerator**: How far is $\hat\beta_1$ from 0?
- **Denominator**: How uncertain are we about $\hat\beta_1$?
- **Large $|t|$**: Strong evidence against $H_0$ (effect is many SEs away from 0)
- **Small $|t|$**: Weak evidence against $H_0$ (effect is close to 0 relative to uncertainty)

---

## Computing the test in R

```{r inference}
#| echo: true
#| code-fold: false

# Coefficient table with hypothesis tests
summary(model)$coefficients
```

Output columns:

- **Estimate**: $\hat\beta_j$ (point estimate)
- **Std. Error**: $SE(\hat\beta_j)$ (standard error)
- **t value**: $t = \text{Estimate} / \text{Std. Error}$
- **Pr(>|t|)**: $p$-value = $P(|T| \geq |t| \mid H_0 \text{ true})$

**Interpretation**: Small p-value (e.g., $p < 0.05$) $\Rightarrow$ strong evidence against $H_0$ $\Rightarrow$ conclude $\beta_j \neq 0$

---

## Diagnostic plots summary

```{r diagnostic-plots}
#| echo: true
#| code-fold: false
#| fig-width: 10
#| fig-height: 8

par(mfrow = c(2, 2))
plot(model)
```

Standard R diagnostic plots: (1) Residuals vs Fitted, (2) Q-Q, (3) Scale-Location, (4) Residuals vs Leverage

---

## Conclusion

We have:

1. Fit an OLS regression model (which is the MLE under normality)
2. Constructed confidence intervals for coefficients
3. Performed hypothesis tests for coefficients
4. Checked normality (Q-Q plot, Shapiro-Wilk, KS, AD tests)
5. Checked homoscedasticity (residual plots, Breusch-Pagan test)
6. Identified influential observations (Cook's distance, leverage)
7. Assessed model fit ($R^2$)

This diagnostic process ensures our assumptions hold and inference is valid.

---
