# Generalized Linear Mixed Effects Models

## From GLM to GLMM

So far, we've studied **linear mixed effects models** with Normal responses.

But what about:

- **Binary outcomes** (grammaticality judgments, choice data)?
- **Count data** (word frequencies, occurrence counts)?
- **Non-normal distributions** in general?

**Solution**: Generalized Linear Mixed Effects Models (**GLMMs**)

GLMMs extend the [GLM framework](generalized-linear-models.qmd) by adding random effects to the systematic component.

---

## The GLM framework (recap)

Recall that [generalized linear models](generalized-linear-models.qmd) have three components:

1. **Random component**: $Y_i \mid \mathbf{x}_i \sim f(\mu_i, \phi)$
   - Response follows a distribution from the exponential family
   - Mean $\mu_i$ depends on predictors

2. **Systematic component**: $\eta_i = \mathbf{x}_i^T \boldsymbol{\beta}$
   - Linear predictor combines fixed effects

3. **Link function**: $g(\mu_i) = \eta_i$
   - Connects mean to linear predictor
   - Ensures mean stays in valid range

**Examples**:

- Logistic regression: $\log\left(\frac{\pi_i}{1-\pi_i}\right) = \eta_i$
- Poisson regression: $\log(\lambda_i) = \eta_i$

---

## Extending GLMs: Adding random effects

GLMMs extend the **systematic component** by adding random effects:

$$\eta_{ij} = \mathbf{x}_{ij}^T \boldsymbol{\beta} + \mathbf{z}_{ij}^T \boldsymbol{\alpha}_i$$

where:

- $\mathbf{x}_{ij}^T \boldsymbol{\beta}$: Fixed effects (systematic component)
- $\mathbf{z}_{ij}^T \boldsymbol{\alpha}_i$: Random effects (varying by group $i$)
- $\boldsymbol{\alpha}_i \sim N(\mathbf{0}, \boldsymbol{\Sigma})$: Random effects distribution

**Three components remain**:

1. **Random**: $Y_{ij} \mid \mathbf{x}_{ij}, \boldsymbol{\alpha}_i \sim f(\mu_{ij}, \phi)$
2. **Systematic**: $\eta_{ij} = \mathbf{x}_{ij}^T \boldsymbol{\beta} + \mathbf{z}_{ij}^T \boldsymbol{\alpha}_i$
3. **Link**: $g(\mu_{ij}) = \eta_{ij}$

---

## GLM → GLMM diagram

```{r glm-glmm-diagram}
#| echo: false
#| fig-width: 14
#| fig-height: 7

library(ggplot2)
library(grid)
library(gridExtra)

# Create text annotations for comparison
glm_text <- data.frame(
    x = c(1, 1, 1),
    y = c(3, 2, 1),
    label = c(
        "1. Random: Y ~ f(μ, φ)",
        "2. Systematic: η = Xβ",
        "3. Link: g(μ) = η"
    ),
    hjust = 0
)

glmm_text <- data.frame(
    x = c(1, 1, 1),
    y = c(3, 2, 1),
    label = c(
        "1. Random: Y | α ~ f(μ, φ)",
        "2. Systematic: η = Xβ + Zα",
        "3. Link: g(μ) = η"
    ),
    hjust = 0
)

p1 <- ggplot(glm_text, aes(x = x, y = y, label = label)) +
    geom_text(size = 6, hjust = 0, fontface = "bold", color = "coral") +
    xlim(0.5, 4) +
    ylim(0.5, 3.5) +
    labs(title = "GLM (Generalized Linear Model)",
         subtitle = "Fixed effects only") +
    theme_void() +
    theme(plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
          plot.subtitle = element_text(size = 14, hjust = 0.5))

p2 <- ggplot(glmm_text, aes(x = x, y = y, label = label)) +
    geom_text(size = 6, hjust = 0, fontface = "bold", color = "steelblue") +
    xlim(0.5, 4) +
    ylim(0.5, 3.5) +
    labs(title = "GLMM (Generalized Linear Mixed Model)",
         subtitle = "Fixed effects + random effects") +
    theme_void() +
    theme(plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
          plot.subtitle = element_text(size = 14, hjust = 0.5))

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

**Key difference**: Systematic component includes random effects $\mathbf{z}_{ij}^T \boldsymbol{\alpha}_i$.

---

## Example: Logistic GLMM

**Scenario**: Grammaticality judgments (binary: 0 = unacceptable, 1 = acceptable) from multiple participants.

**GLM (logistic regression)**:

$$\log\left(\frac{\pi_{ij}}{1-\pi_{ij}}\right) = \beta_0 + \beta_1 x_{ij}$$

- Ignores participant variation
- Assumes all participants have same response pattern

**GLMM (logistic mixed model)**:

$$\log\left(\frac{\pi_{ij}}{1-\pi_{ij}}\right) = \beta_0 + \beta_1 x_{ij} + u_i$$

where $u_i \sim N(0, \sigma_u^2)$ is participant $i$'s random intercept.

- Participants differ in baseline acceptance rates
- Generalizes to new participants

---

## Example: Poisson GLMM

**Scenario**: Counts of verb-frame cooccurrences, with verbs nested within semantic classes.

**GLM (Poisson regression)**:

$$\log(\lambda_{ij}) = \beta_0 + \beta_1 x_{ij,\text{frame}} + \beta_2 x_{ij,\text{class}}$$

- Fixed effects for frame and class
- Ignores within-class variation among verbs

**GLMM (Poisson mixed model)**:

$$\log(\lambda_{ijk}) = \beta_0 + \beta_1 x_{ijk,\text{frame}} + \beta_2 x_{ijk,\text{class}} + \alpha_j$$

where $\alpha_j \sim N(0, \sigma_u^2)$ is verb $j$'s random intercept.

- Verbs within a class vary in frequency preferences
- Generalizes beyond observed verbs

---

## Conditional vs. marginal interpretation

GLMMs have two perspectives on fixed effects:

**Conditional (subject-specific)**:

- Fixed effects are interpreted **conditional on** random effects
- "For a given participant/item, how does X affect Y?"
- Example: "Within a specific verb, how does frame affect count?"

**Marginal (population-average)**:

- Fixed effects are interpreted **averaging over** random effects
- "On average across all participants/items, how does X affect Y?"
- Example: "On average across all verbs, how does frame affect count?"

**Important**: For non-linear link functions (logit, log), conditional ≠ marginal!

- Conditional effects are typically larger in magnitude
- Use GEE (Generalized Estimating Equations) for marginal effects

---

## Plot: Conditional vs. marginal effects

```{r conditional-marginal-plot}
#| echo: false
#| fig-width: 12
#| fig-height: 7

# Simulate logistic GLMM data
set.seed(42)
n_subjects <- 10
n_obs_per_subject <- 50

# Subject random intercepts
subject_intercepts <- rnorm(n_subjects, mean = 0, sd = 1.5)

# Simulate data
sim_data <- expand_grid(
    subject = 1:n_subjects,
    obs = 1:n_obs_per_subject
) %>%
    mutate(
        x = rnorm(n(), mean = 0, sd = 1),
        alpha_i = subject_intercepts[subject],
        # Conditional (subject-specific) logit
        logit_cond = -0.5 + 1.5 * x + alpha_i,
        prob_cond = plogis(logit_cond),
        y = rbinom(n(), size = 1, prob = prob_cond)
    )

# Fit models
library(lme4)
glmm_model <- glmer(y ~ x + (1|subject), data = sim_data, family = binomial)

# Subject-specific curves
p1 <- ggplot(sim_data, aes(x = x, y = prob_cond, group = subject)) +
    geom_line(alpha = 0.5, color = "steelblue", linewidth = 1) +
    labs(x = "Predictor X", y = "P(Y = 1)",
         title = "Conditional (Subject-Specific) Effects",
         subtitle = "Each line = one subject | Slope = 1.5 (conditional effect)") +
    theme_minimal(base_size = 14) +
    ylim(0, 1)

# Marginal curve (averaging over subjects)
marginal_pred <- sim_data %>%
    group_by(x_bin = cut(x, breaks = 20)) %>%
    summarize(
        x_mean = mean(x),
        prob_marginal = mean(y),
        .groups = "drop"
    )

p2 <- ggplot(marginal_pred, aes(x = x_mean, y = prob_marginal)) +
    geom_point(size = 3, color = "coral") +
    geom_smooth(method = "loess", se = TRUE, color = "darkred", fill = "pink", alpha = 0.3) +
    labs(x = "Predictor X", y = "P(Y = 1)",
         title = "Marginal (Population-Average) Effect",
         subtitle = "Averaged over all subjects | Slope < 1.5 (attenuated)") +
    theme_minimal(base_size = 14) +
    ylim(0, 1)

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

**Key point**: Non-linearity of link function causes marginal effect to be smaller than conditional.

---

## Estimation: Laplace approximation

GLMMs do not have closed-form solutions. Estimation requires numerical approximation.

**Likelihood**:

$$L(\boldsymbol{\beta}, \boldsymbol{\Sigma}) = \prod_{i=1}^N \int f(y_i \mid \boldsymbol{\alpha}_i, \boldsymbol{\beta}) \, f(\boldsymbol{\alpha}_i \mid \boldsymbol{\Sigma}) \, d\boldsymbol{\alpha}_i$$

- Integral over random effects (high-dimensional!)
- No closed form for non-normal responses

**Laplace approximation**:

- Approximates the integral using a second-order Taylor expansion
- Fast, but can be inaccurate for small sample sizes or extreme probabilities
- Used by `lme4::glmer()`

**Alternatives**:

- **Adaptive Gauss-Hermite quadrature**: More accurate but slower
- **MCMC (Bayesian)**: Gold standard, fully explores posterior (Stan, brms)

---

## Fitting GLMMs in R: lme4

```{r glmer-syntax}
#| echo: true
#| eval: false

library(lme4)

# Logistic GLMM
glmer(binary_outcome ~ predictor + (1|group),
      data = df, family = binomial)

# Poisson GLMM
glmer(count_outcome ~ predictor + (1|group),
      data = df, family = poisson)

# Negative binomial GLMM (overdispersion)
glmer.nb(count_outcome ~ predictor + (1|group),
         data = df)
```

**Family options**:

- `binomial`: Logistic regression
- `poisson`: Poisson regression
- Use `glmer.nb()` for negative binomial (handles overdispersion)

---

## Convergence issues

GLMMs are complex models. Common issues:

**Convergence warnings**:

- "Model failed to converge"
- "Model is nearly unidentifiable"

**Causes**:

- Too many random effects for available data
- Random effects variance near zero (singular fit)
- Optimization stuck in local optimum

**Solutions**:

1. Simplify random effects structure
2. Center/scale predictors
3. Increase iterations: `control = glmerControl(optimizer = "bobyqa")`
4. Try different optimizers
5. Use Bayesian methods (brms) for more robust estimation

---

## Overdispersion in GLMMs

Recall from [Poisson regression](poisson-regression.qmd): the Poisson model assumes $\text{Var}(Y) = \mathbb{E}[Y] = \lambda$.

**Overdispersion**: $\text{Var}(Y) > \lambda$ due to:

- Clustering (observations within groups are similar)
- Heterogeneity (unmeasured factors vary across groups)

**Random effects address overdispersion**:

- Model the clustering/heterogeneity explicitly
- Random effects "soak up" extra variance

**Result**: Adding random effects can reduce/eliminate overdispersion.

**Next section**: Demonstrate this with VALEX verb-frame data.

---

## Summary: GLMM theory

**GLMMs extend GLMs** by adding random effects to the systematic component:

$$\eta_{ij} = \mathbf{x}_{ij}^T \boldsymbol{\beta} + \mathbf{z}_{ij}^T \boldsymbol{\alpha}_i$$

**Three components**:

1. Random: Exponential family distribution
2. Systematic: Fixed effects + random effects
3. Link: Connects mean to linear predictor

**Interpretation**:

- Conditional (subject-specific) vs. marginal (population-average)
- Non-linear links cause these to differ

**Estimation**: Laplace approximation (lme4) or MCMC (Stan/brms)

**Next**: Poisson GLMM example with VALEX data

---
