# Multiple Regression

## Extension to multiple predictors

In simple regression, we modeled log RT as a function of surprisal alone.

But reading times depend on multiple factors:

- **Surprisal**: -log₂ p(word | context) from language model
- **Acceptability**: Human grammaticality judgments
- **Condition**: Experimental manipulation (early vs. late coercion)

Multiple linear regression lets us model their joint effects.

---

## Loading data with multiple predictors

```{r load-full-data}
#| echo: true
#| code-fold: false

# Load merged dataset with both surprisal and acceptability
spr_full <- read_csv("data/word_rt_surprisal_acceptability.csv",
                     show_col_types = FALSE)

# Filter to observations with acceptability ratings
spr_full <- spr_full %>%
  filter(!is.na(mean_acceptability))

# Convert condition to factor
spr_full <- spr_full %>%
  mutate(
    condition = factor(condition,
                      levels = c("early coercion", "late coercion"))
  )

# Preview
head(spr_full)
```

---

## Dataset structure

```{r data-structure}
#| echo: true
#| code-fold: false

# Dimensions
cat("Observations:", nrow(spr_full), "\n")

# Variables
cat("Predictors: surprisal, mean_acceptability, condition\n")
cat("Response: log_rt\n")
```

---

# Multicollinearity

## Checking for collinearity

Before fitting a model with both surprisal and acceptability, we should check if they're **correlated**.

Why? **Multicollinearity** (high correlation between predictors) causes:

1. Inflated standard errors → unstable estimates
2. Difficulty separating individual effects
3. Coefficients sensitive to small data changes

Let's check the correlation to see if this is a concern in our data.

---

## Correlation between predictors

```{r correlation-plot}
#| echo: true
#| code-fold: false
#| fig-width: 8
#| fig-height: 5

# Compute correlation
cor_value <- cor(spr_full$surprisal, spr_full$mean_acceptability,
                 use = "complete.obs")

cat("Correlation:", round(cor_value, 3), "\n")

# Scatterplot
ggplot(spr_full, aes(x = surprisal, y = mean_acceptability)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(x = "Surprisal (bits)",
       y = "Acceptability (z-score)",
       title = paste0("Correlation: ", round(cor_value, 3)))
```

---

## Interpreting the correlation

Looking at the correlation value and scatterplot:

- The correlation is **weak to negligible** (likely |r| < 0.3)
- There's no strong linear relationship between surprisal and acceptability
- Multicollinearity is **not a major concern** in this dataset

**Good news**: We can include both predictors without worrying about inflated standard errors.

**But let's verify** using the Variance Inflation Factor (VIF)...

---

## Variance Inflation Factor (VIF)

**Formal measure** of multicollinearity for predictor $j$:

$$\text{VIF}_j = \frac{1}{1 - R^2_j}$$

where $R^2_j$ is from regressing $X_j$ on all other predictors.

**Interpretation**:
- VIF = 1: No multicollinearity
- VIF > 5: Moderate multicollinearity
- VIF > 10: Severe multicollinearity (problematic)

---

## Computing VIF

```{r vif-check}
#| echo: true
#| code-fold: false
#| warning: false

library(car)

# Fit model with both predictors
model_both <- lm(log_rt ~ surprisal + mean_acceptability + condition,
                 data = spr_full)

# Compute VIF
vif(model_both)
```

VIF values close to 1 indicate **no multicollinearity**. All values well below 5 (the common threshold for concern).

---

## What if we had collinearity?

In this dataset, multicollinearity is not a problem. But it's important to know how to handle it when it does occur.

One approach is **residualization**: orthogonalizing predictors to remove their correlation.

Let's demonstrate this technique even though we don't strictly need it here—it's a valuable tool for your analytic toolkit.

---

## Residualization approach

**The idea**: Remove the correlation between predictors by residualizing one on the other.

**Procedure**:

1. Regress surprisal on acceptability: $\text{Surprisal} = \gamma_0 + \gamma_1 \text{Acceptability} + \eta$
2. Extract residuals: $\hat\eta = \text{Surprisal} - (\hat\gamma_0 + \hat\gamma_1 \text{Acceptability})$
3. Use $\hat\eta$ as predictor instead of raw surprisal

**Result**: $\hat\eta$ captures variation in surprisal **orthogonal to** (independent of) acceptability.

**Why this direction?** Surprisal from a language model likely reflects many factors, including acceptability/grammaticality. By residualizing surprisal on acceptability, we isolate the component of surprisal that's independent of acceptability ratings.

---

## Implementing residualization

```{r residualization}
#| echo: true
#| code-fold: false

# Residualize surprisal on acceptability
resid_model <- lm(surprisal ~ mean_acceptability, data = spr_full)
spr_full$surprisal_resid <- residuals(resid_model)

# Check correlation (should be ~0)
cor(spr_full$mean_acceptability, spr_full$surprisal_resid, use = "complete.obs")

# Now fit with residualized predictor
model_resid <- lm(log_rt ~ surprisal_resid + mean_acceptability + condition,
                  data = spr_full)
summary(model_resid)
```

Now `mean_acceptability` and `surprisal_resid` are uncorrelated by construction.

---

## Interpreting residualized coefficients

With residualization:

- $\beta_{\text{acceptability}}$: Effect of acceptability on log RT
- $\beta_{\text{surprisal\_resid}}$: Effect of surprisal **beyond what's explained by acceptability**

The second coefficient answers: "Does surprisal predict RT over and above acceptability?"

This isolates the unique contribution of surprisal that isn't already captured by acceptability judgments.

---

## Visualizing residualization

```{r residualization-visualization}
#| echo: false
#| fig-width: 12
#| fig-height: 5

library(patchwork)

# Before residualization
p1 <- ggplot(spr_full, aes(x = mean_acceptability, y = surprisal)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(x = "Acceptability (z-score)",
       y = "Surprisal (bits)",
       title = "Before Residualization",
       subtitle = paste0("r = ", round(cor(spr_full$surprisal,
                                           spr_full$mean_acceptability,
                                           use = "complete.obs"), 3))) +
  theme_minimal(base_size = 12)

# After residualization
p2 <- ggplot(spr_full, aes(x = mean_acceptability, y = surprisal_resid)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Acceptability (z-score)",
       y = "Residualized Surprisal (bits)",
       title = "After Residualization",
       subtitle = paste0("r = ", round(cor(spr_full$surprisal_resid,
                                           spr_full$mean_acceptability,
                                           use = "complete.obs"), 3))) +
  theme_minimal(base_size = 12)

p1 + p2
```

Residualization removes the linear relationship, making the predictors orthogonal (r ≈ 0).

---

## Understanding the residuals

```{r residual-interpretation}
#| echo: false
#| fig-width: 10
#| fig-height: 5

ggplot(spr_full, aes(x = surprisal, y = surprisal_resid)) +
  geom_point(alpha = 0.3, size = 1.5, color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Original Surprisal (bits)",
       y = "Residualized Surprisal (bits)",
       title = "Residualized vs. Original Surprisal",
       subtitle = "Positive residuals = higher surprisal than predicted by acceptability alone") +
  theme_minimal(base_size = 13)
```

Residuals represent the **unexplained component** of surprisal after accounting for acceptability.

---

## When is residualization really needed?

Residualization is most useful when predictors are **highly correlated** (r > 0.7).

**Example scenario**: Comparing surprisal from two different autoregressive language models:

- GPT-2 small and GPT-2 medium surprisal would likely be highly correlated (r > 0.8)
- Both measure similar aspects of predictability
- High VIF values would indicate multicollinearity
- Residualization (or choosing one model) would be necessary

Let's actually demonstrate this with real data...

---

## Simulating multicollinearity with language models

To demonstrate multicollinearity effects, we **simulate** a second set of surprisal values that are highly correlated with our real surprisal values (r ≈ 0.95).

This reflects what we would see if we computed surprisal from two similar Spanish language models.

```{r load-two-models}
#| echo: true
#| code-fold: false

# Load data with surprisals from both models
spr_two_models <- read_csv("data/word_rt_surprisal.csv", show_col_types = FALSE)

# Preview
head(spr_two_models %>% select(word, surprisal, surprisal_gpt2_small2))
```

---

## Correlation between language models

```{r lm-correlation}
#| echo: true
#| code-fold: false
#| fig-width: 8
#| fig-height: 5

# Compute correlation
cor_lms <- cor(spr_two_models$surprisal, spr_two_models$surprisal_gpt2_small2,
               use = "complete.obs")

cat("Correlation between models:", round(cor_lms, 3), "\n")

# Visualize
ggplot(spr_two_models, aes(x = surprisal, y = surprisal_gpt2_small2)) +
  geom_point(alpha = 0.2, size = 1) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(x = "Surprisal: Model 1 (bits)",
       y = "Surprisal: Model 2 (bits)",
       title = paste0("High Correlation: r = ", round(cor_lms, 3))) +
  theme_minimal(base_size = 13)
```

**This is true multicollinearity**—both models capture similar aspects of predictability.

---

## VIF with correlated language models

```{r vif-two-models}
#| echo: true
#| code-fold: false

library(car)

# Fit model with both language model surprisals
model_two_lms <- lm(log_rt ~ surprisal + surprisal_gpt2_small2 + condition,
                    data = spr_two_models)

# Compute VIF
vif(model_two_lms)
```

The VIF values for the two surprisal measures are **very high** (> 10), indicating severe multicollinearity.

---

## Residualization with correlated predictors

Now residualization is **necessary**, not just pedagogical!

```{r residualize-lms}
#| echo: true
#| code-fold: false

# Residualize medium model surprisal on small model surprisal
resid_lm_model <- lm(surprisal_gpt2_small2 ~ surprisal, data = spr_two_models)
spr_two_models$surprisal_medium_resid <- residuals(resid_lm_model)

# Check correlation (should be ~0)
cor(spr_two_models$surprisal, spr_two_models$surprisal_medium_resid,
    use = "complete.obs")

# Fit model with residualized predictor
model_resid_lms <- lm(log_rt ~ surprisal + surprisal_medium_resid + condition,
                      data = spr_two_models)

# Check VIF
vif(model_resid_lms)
```

VIF values should now be close to 1!

---

## Visualizing residualization with correlated predictors

```{r viz-resid-lms}
#| echo: false
#| fig-width: 12
#| fig-height: 5

library(patchwork)

# Before
p1 <- ggplot(spr_two_models, aes(x = surprisal, y = surprisal_gpt2_small2)) +
  geom_point(alpha = 0.2, size = 1) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(x = "Model 1 Surprisal",
       y = "Model 2 Surprisal (simulated)",
       title = "Before Residualization",
       subtitle = paste0("r = ", round(cor_lms, 3))) +
  theme_minimal(base_size = 11)

# After
p2 <- ggplot(spr_two_models, aes(x = surprisal, y = surprisal_medium_resid)) +
  geom_point(alpha = 0.2, size = 1) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
  labs(x = "Model 1 Surprisal",
       y = "Model 2 Surprisal (simulated, residualized)",
       title = "After Residualization",
       subtitle = paste0("r ≈ ", round(cor(spr_two_models$surprisal,
                                           spr_two_models$surprisal_medium_resid,
                                           use = "complete.obs"), 3))) +
  theme_minimal(base_size = 11)

p1 + p2
```

---

## Key takeaway: Always check first

**Workflow for handling potential multicollinearity**:

1. ✅ **Check correlation matrix** between all predictors
2. ✅ **Compute VIF** after fitting initial model
3. ❓ **If VIF > 5**: Consider residualization, dropping predictors, or regularization
4. ✅ **If VIF < 5**: Proceed with standard multiple regression

**In this dataset**: VIF values are low, so we can include both predictors without concern.

---

## Summary: Two demonstrations of residualization

We've seen residualization in two contexts:

**1. Surprisal and Acceptability (low correlation)**
- Demonstrated the *technique* of residualization
- Not actually needed (VIF < 5, weak correlation)
- Pedagogical example to learn the method

**2. Two Language Models (high correlation)**
- Actual multicollinearity problem (VIF > 5, r > 0.8)
- Residualization is *necessary* to get stable estimates
- Real-world scenario you'll encounter comparing similar measures

**Key lesson**: Always check VIF before deciding if residualization is needed!

---

# Multiple Regression Formalism

## Multiple linear regression model

With $p$ predictors, the model is:

$$Y_i = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \cdots + \beta_p X_{ip} + \epsilon_i$$

where $\epsilon_i \sim \text{Normal}(0, \sigma^2)$.

Each $\beta_j$ represents the partial effect of $X_j$ on $Y$, holding all other predictors constant.

---

## Matrix notation

Let $\mathbf{y} = (y_1, \ldots, y_N)^T$ be the vector of responses.

Let $\mathbf{X}$ be the $N \times (p+1)$ design matrix:

$$\mathbf{X} = \begin{pmatrix}
1 & x_{11} & x_{12} & \cdots & x_{1p}\\
1 & x_{21} & x_{22} & \cdots & x_{2p}\\
\vdots & \vdots & \vdots & \ddots & \vdots\\
1 & x_{N1} & x_{N2} & \cdots & x_{Np}
\end{pmatrix}$$

Let $\boldsymbol{\beta} = (\beta_0, \beta_1, \ldots, \beta_p)^T$.

---

## Matrix form

The model is:

$$\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}$$

where $\boldsymbol{\epsilon} \sim \text{Normal}(\mathbf{0}, \sigma^2 \mathbf{I})$.

The OLS estimate is:

$$\hat{\boldsymbol{\beta}} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}$$

assuming $\mathbf{X}^T\mathbf{X}$ is invertible.

---

## Categorical predictors

Categorical variables are encoded using dummy (indicator) variables.

For a categorical predictor with $K$ levels, we create $K-1$ dummy variables.

One level serves as the reference (baseline), and each dummy variable compares another level to the reference.

This is called **treatment coding** or **dummy coding**.

---

## Dummy coding example

For `condition` with two levels (early coercion, late coercion):

- **Reference**: early coercion (coded as 0)
- **Late coercion**: coded as 1

The model includes one dummy variable:

$$\text{LogRT}_i = \beta_0 + \beta_1 \mathbb{1}[\text{late}]_i + \cdots$$

---

## Interpreting dummy coefficients

$\beta_0$: Expected log RT in the reference condition (early coercion)

$\beta_1$: Difference in log RT between late coercion and early coercion

Each coefficient is a **contrast** between levels.

---

# Contrast Coding Schemes

## Reference coding vs. sum coding

There are multiple ways to encode categorical variables numerically. Two common schemes:

**Reference coding** (dummy coding, treatment coding):
- One level serves as baseline (coded 0)
- Other levels compared to baseline
- Coefficients = differences from reference level

**Sum coding** (effects coding, deviation coding):
- All levels sum to zero
- Coefficients = deviation from grand mean
- Tests main effects symmetrically

---

## Reference coding in detail

For a two-level factor (early coercion, late coercion):

| Condition | Reference coding |
|-----------|-----------------|
| early coercion | 0 |
| late coercion | 1 |

**Model**: $Y_i = \beta_0 + \beta_1 X_i + \epsilon_i$

**Interpretation**:
- $\beta_0$: Mean of reference level (early coercion)
- $\beta_1$: Difference (late coercion - early coercion)

---

## Sum coding in detail

For a two-level factor (early coercion, late coercion):

| Condition | Sum coding |
|-----------|-----------|
| early coercion | -0.5 |
| late coercion | 0.5 |

**Model**: $Y_i = \beta_0 + \beta_1 X_i + \epsilon_i$

**Interpretation**:
- $\beta_0$: Grand mean (average across both conditions)
- $\beta_1$: Difference from grand mean to late coercion (= half the difference between conditions)

Note: Sum coding weights sum to 0: $-0.5 + 0.5 = 0$

---

## Implementing coding schemes in R

```{r coding-schemes}
#| echo: true
#| code-fold: false

# Create copy of data for demonstration
spr_ref <- spr_full
spr_sum <- spr_full

# Reference coding (default in R)
contrasts(spr_ref$condition) <- contr.treatment(2)
contrasts(spr_ref$condition)

# Sum coding
contrasts(spr_sum$condition) <- contr.sum(2)
contrasts(spr_sum$condition)
```

---

## Comparing models with different coding

```{r compare-coding}
#| echo: true
#| code-fold: false

# Fit with reference coding
model_ref <- lm(log_rt ~ surprisal + mean_acceptability + condition,
                data = spr_ref)

# Fit with sum coding
model_sum <- lm(log_rt ~ surprisal + mean_acceptability + condition,
                data = spr_sum)

# Compare intercepts and condition coefficients
cat("Reference coding:\n")
cat("Intercept:", coef(model_ref)[1], "\n")
cat("Condition:", coef(model_ref)[4], "\n\n")

cat("Sum coding:\n")
cat("Intercept:", coef(model_sum)[1], "\n")
cat("Condition:", coef(model_sum)[4], "\n")
```

---

## Interpretation differences

**Reference coding**:
- Intercept = mean of early coercion condition
- Condition coefficient = (late coercion mean) - (early coercion mean)

**Sum coding**:
- Intercept = grand mean across both conditions
- Condition coefficient = (late coercion mean) - (grand mean)

**Note**: The condition coefficient under sum coding is **half** the difference between conditions.

**Model fit is identical**: Both achieve same predictions, $R^2$, residuals. Only parameterization differs.

---

## When to use each coding scheme

**Reference coding**:
- When one level is a natural baseline (e.g., control condition)
- When comparing treatments to control
- More intuitive for experimental contrasts

**Sum coding**:
- When no natural baseline exists
- When testing symmetrical main effects
- Standard in ANOVA-style analyses
- Useful for balanced designs

**Choice affects interpretation, not model fit.**

---

## Visualizing the data: condition effects

```{r plot-by-condition}
#| echo: false
#| fig-width: 10
#| fig-height: 5

ggplot(spr_full, aes(x = condition, y = log_rt, fill = condition)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.2, size = 1) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Condition",
       y = "Log Reaction Time (s)",
       title = "Log RT by Condition") +
  theme(legend.position = "none")
```

---

## Visualizing by condition

```{r plot-by-condition-new}
#| echo: false
#| fig-width: 10
#| fig-height: 5

ggplot(spr_full, aes(x = condition, y = log_rt, fill = condition)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.2, size = 1) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Condition",
       y = "Log Reaction Time (s)",
       title = "Log RT by Condition") +
  theme(legend.position = "none")
```

---

## Visualizing continuous predictors

```{r plot-continuous-predictors}
#| echo: false
#| fig-width: 12
#| fig-height: 5

library(patchwork)

p1 <- ggplot(spr_full, aes(x = surprisal, y = log_rt)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(x = "Surprisal (bits)",
       y = "Log RT",
       title = "Log RT vs. Surprisal")

p2 <- ggplot(spr_full, aes(x = mean_acceptability, y = log_rt)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(x = "Acceptability (z-score)",
       y = "Log RT",
       title = "Log RT vs. Acceptability")

p1 + p2
```

---

## Interaction visualization

```{r plot-interaction-viz}
#| echo: true
#| code-fold: false
#| fig-width: 10
#| fig-height: 5

ggplot(spr_full, aes(x = surprisal, y = log_rt, color = condition)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "Surprisal (bits)",
       y = "Log Reaction Time (s)",
       title = "Log RT vs. Surprisal by Condition",
       color = "Condition") +
  theme_minimal(base_size = 12)
```

Different slopes suggest an **interaction effect** between surprisal and condition.

---

# Additive Model

## Model specification

We start with an additive model (no interactions):

$$\text{LogRT}_i = \beta_0 + \beta_1 \text{Surprisal}_i + \beta_2 \text{Acceptability}_i + \beta_3 \mathbb{1}[\text{late}]_i + \epsilon_i$$

This assumes effects are **additive**—the effect of surprisal is the same across both conditions.

---

## Fitting the additive model

```{r fit-additive}
#| echo: true
#| code-fold: false

# Fit additive model
model_additive <- lm(log_rt ~ surprisal + mean_acceptability + condition,
                     data = spr_full)

# Summary
summary(model_additive)
```

---

## Interpreting coefficients

```{r coef-additive}
#| echo: true
#| code-fold: false

# Extract coefficients
coef(model_additive)
```

$\beta_1$ (surprisal): Effect of surprisal on log RT, controlling for acceptability and condition

$\beta_2$ (acceptability): Effect of acceptability on log RT, controlling for surprisal and condition

$\beta_3$ (late coercion): Difference from early coercion, controlling for surprisal and acceptability

---

## Visualizing additive model predictions

```{r plot-additive-predictions}
#| echo: false
#| fig-width: 10
#| fig-height: 5

# Create prediction grid
pred_grid <- expand.grid(
  surprisal = seq(min(spr_full$surprisal, na.rm = TRUE),
                  max(spr_full$surprisal, na.rm = TRUE),
                  length.out = 50),
  mean_acceptability = mean(spr_full$mean_acceptability, na.rm = TRUE),
  condition = levels(spr_full$condition)
)

# Predict
pred_grid$predicted <- predict(model_additive, newdata = pred_grid)

# Plot
ggplot() +
  geom_point(data = spr_full,
             aes(x = surprisal, y = log_rt, color = condition),
             alpha = 0.3, size = 1) +
  geom_line(data = pred_grid,
            aes(x = surprisal, y = predicted, color = condition),
            linewidth = 1) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "Surprisal (bits)",
       y = "Log Reaction Time (s)",
       title = "Additive Model: Parallel Lines (Same Slope for Surprisal)",
       color = "Condition") +
  theme_minimal(base_size = 12)
```

Note: Lines are parallel (same slope for surprisal across conditions).

---

# Interaction Effects

## What are interactions?

An interaction occurs when the effect of one predictor depends on the level of another predictor.

For example: Does the effect of acceptability differ across conditions?

We model this by including **interaction terms**: products of predictors.

---

## Interaction model specification

$$\begin{align*}
\text{LogRT}_i &= \beta_0 + \beta_1 \text{Surprisal}_i + \beta_2 \text{Acceptability}_i + \beta_3 \mathbb{1}[\text{late}]_i \\
&\quad + \beta_4 \text{Surprisal}_i \times \mathbb{1}[\text{late}]_i \\
&\quad + \beta_5 \text{Acceptability}_i \times \mathbb{1}[\text{late}]_i + \epsilon_i
\end{align*}$$

---

## Interpreting interaction coefficients

$\beta_4$: How much the effect of surprisal changes in late coercion vs. early coercion

$\beta_5$: How much the effect of acceptability changes in late coercion vs. early coercion

Non-zero interaction coefficients indicate the slopes differ across conditions.

---

# Interactions and Coding Schemes

## How coding affects interaction interpretation

Interaction interpretation depends on the coding scheme for categorical predictors.

With **reference coding**: Interaction tests whether slopes differ **between specific levels**.

With **sum coding**: Interaction tests whether slopes differ **from the average slope**.

Both test the same hypothesis (presence of interaction), but coefficients have different meanings.

---

## Interaction with reference coding

Model: $Y_i = \beta_0 + \beta_1 X_i + \beta_2 C_i + \beta_3 X_i \times C_i + \epsilon_i$

where $C_i = 0$ (early coercion) or $1$ (late coercion).

**For early coercion** ($C_i = 0$):
$$Y_i = \beta_0 + \beta_1 X_i$$

**For late coercion** ($C_i = 1$):
$$Y_i = (\beta_0 + \beta_2) + (\beta_1 + \beta_3) X_i$$

$\beta_3$ = (slope for late coercion) - (slope for early coercion)

---

## Interaction with sum coding

Model: $Y_i = \beta_0 + \beta_1 X_i + \beta_2 C_i + \beta_3 X_i \times C_i + \epsilon_i$

where $C_i = -0.5$ (early coercion) or $0.5$ (late coercion).

**For early coercion** ($C_i = -0.5$):
$$Y_i = (\beta_0 - 0.5\beta_2) + (\beta_1 - 0.5\beta_3) X_i$$

**For late coercion** ($C_i = 0.5$):
$$Y_i = (\beta_0 + 0.5\beta_2) + (\beta_1 + 0.5\beta_3) X_i$$

$\beta_1$ = average slope across conditions

$\beta_3$ = (slope for late coercion) - (slope for early coercion)

---

## Comparing interaction models

```{r interaction-coding-comparison}
#| echo: true
#| code-fold: false

# Interaction with reference coding
model_int_ref <- lm(log_rt ~ surprisal * condition +
                             mean_acceptability * condition,
                    data = spr_ref)

# Interaction with sum coding
model_int_sum <- lm(log_rt ~ surprisal * condition +
                             mean_acceptability * condition,
                    data = spr_sum)

# Extract surprisal coefficients
cat("Reference coding:\n")
cat("Main effect of surprisal:", coef(model_int_ref)["surprisal"], "\n")
cat("Interaction:", coef(model_int_ref)["surprisal:conditionlate coercion"], "\n\n")

cat("Sum coding:\n")
cat("Main effect of surprisal:", coef(model_int_sum)["surprisal"], "\n")
cat("Interaction:", coef(model_int_sum)["surprisal:condition1"], "\n")
```

---

## Interpretation comparison

**Reference coding**:
- Main effect = slope in reference condition (early coercion)
- Interaction = change in slope for late coercion

**Sum coding**:
- Main effect = average slope across conditions
- Interaction = half the difference in slopes

**Key difference**: Under sum coding, the main effect is the **average** effect. Under reference coding, it is the effect in the **baseline** condition.

---

## Computing condition-specific slopes

```{r compute-slopes-by-coding}
#| echo: true
#| code-fold: false

# Reference coding
beta_surp_ref <- coef(model_int_ref)["surprisal"]
beta_int_ref <- coef(model_int_ref)["surprisal:conditionlate coercion"]

slope_early_ref <- beta_surp_ref
slope_late_ref <- beta_surp_ref + beta_int_ref

cat("Reference coding:\n")
cat("Early coercion slope:", slope_early_ref, "\n")
cat("Late coercion slope:", slope_late_ref, "\n\n")

# Sum coding
beta_surp_sum <- coef(model_int_sum)["surprisal"]
beta_int_sum <- coef(model_int_sum)["surprisal:condition1"]

slope_early_sum <- beta_surp_sum - 0.5 * beta_int_sum
slope_late_sum <- beta_surp_sum + 0.5 * beta_int_sum

cat("Sum coding:\n")
cat("Early coercion slope:", slope_early_sum, "\n")
cat("Late coercion slope:", slope_late_sum, "\n")
```

Note: Condition-specific slopes are identical regardless of coding scheme.

---

## Which coding for interactions?

**Reference coding** advantages:
- Direct test of difference from baseline
- Intuitive when one condition is control
- Coefficients directly give condition-specific effects

**Sum coding** advantages:
- Main effect represents average across conditions
- Symmetrical treatment of conditions
- Matches ANOVA decomposition

**Both test the same interaction hypothesis.** Choice depends on research question and interpretation preferences.

---

## Fitting the interaction model

```{r fit-interaction}
#| echo: true
#| code-fold: false

# Fit interaction model
model_interaction <- lm(log_rt ~ surprisal * condition +
                                 mean_acceptability * condition,
                        data = spr_full)

# Summary
summary(model_interaction)
```

---

## Comparing models

```{r compare-models-anova}
#| echo: true
#| code-fold: false

# ANOVA to compare nested models
anova(model_additive, model_interaction)
```

The $F$-test tells us whether the interaction terms significantly improve model fit.

---

## Visualizing interaction model

```{r plot-interaction}
#| echo: false
#| fig-width: 10
#| fig-height: 5

# Prediction grid
pred_grid_int <- expand.grid(
  surprisal = seq(min(spr_full$surprisal, na.rm = TRUE),
                  max(spr_full$surprisal, na.rm = TRUE),
                  length.out = 50),
  mean_acceptability = mean(spr_full$mean_acceptability, na.rm = TRUE),
  condition = levels(spr_full$condition)
)

# Predict
pred_grid_int$predicted <- predict(model_interaction, newdata = pred_grid_int)

# Plot
ggplot() +
  geom_point(data = spr_full,
             aes(x = surprisal, y = log_rt, color = condition),
             alpha = 0.3, size = 1) +
  geom_line(data = pred_grid_int,
            aes(x = surprisal, y = predicted, color = condition),
            linewidth = 1.5) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "Surprisal (bits)",
       y = "Log Reaction Time (s)",
       title = "Interaction Model: Different Slopes by Condition",
       color = "Condition") +
  theme_minimal(base_size = 12)
```

Lines now have different slopes—the effect of surprisal differs by condition.

---

## Coefficient table

```{r coef-table}
#| echo: true
#| code-fold: false

# Full coefficient table
summary(model_interaction)$coefficients
```

---

# Diagnostic Checks for Multiple Regression

## Same assumptions apply

Multiple regression makes the same assumptions as simple regression:

1. **Linearity**: Linear relationship between predictors and response
2. **Normality**: Errors are normally distributed
3. **Homoscedasticity**: Constant error variance
4. **Independence**: Errors are independent

We check these using the same diagnostic tools.

---

## Residuals vs. fitted

```{r diag-resid-fitted}
#| echo: true
#| code-fold: false
#| fig-width: 7
#| fig-height: 5

plot(model_interaction, which = 1)
```

---

## Q-Q plot

```{r diag-qq}
#| echo: true
#| code-fold: false
#| fig-width: 7
#| fig-height: 5

plot(model_interaction, which = 2)
```

---

## Scale-location plot

```{r diag-scale-location}
#| echo: true
#| code-fold: false
#| fig-width: 7
#| fig-height: 5

plot(model_interaction, which = 3)
```

---

## Cook's distance

```{r diag-cooks}
#| echo: true
#| code-fold: false
#| fig-width: 7
#| fig-height: 5

plot(model_interaction, which = 4)
```

---

## Residuals vs. leverage

```{r diag-leverage}
#| echo: true
#| code-fold: false
#| fig-width: 7
#| fig-height: 5

plot(model_interaction, which = 5)
```

---

## Formal normality tests

```{r multi-normality-tests}
#| echo: true
#| code-fold: false

# Extract residuals
resid_multi <- residuals(model_interaction)

# Shapiro-Wilk test (max N = 5000)
if (length(resid_multi) <= 5000) {
  shapiro.test(resid_multi)
} else {
  set.seed(42)
  subsample <- sample(resid_multi, 5000)
  cat("Note: Using random subsample of 5000 observations (N =", length(resid_multi), ")\n")
  shapiro.test(subsample)
}

# Anderson-Darling test
nortest::ad.test(resid_multi)
```

---

## Breusch-Pagan test

```{r multi-bp-test}
#| echo: true
#| code-fold: false

# Test for heteroscedasticity
lmtest::bptest(model_interaction)
```

---

# Multicollinearity

## The multicollinearity problem

When predictors are highly correlated, coefficient estimates become unstable.

High correlations among predictors lead to:

1. Large standard errors
2. Unstable coefficient estimates
3. Difficulty interpreting individual effects

This is **multicollinearity**.

---

## Variance Inflation Factor (VIF)

VIF measures how much the variance of a coefficient is inflated due to collinearity.

For predictor $X_j$:

$$\text{VIF}_j = \frac{1}{1 - R^2_j}$$

where $R^2_j$ is from regressing $X_j$ on all other predictors.

**Rule of thumb**: VIF $> 10$ indicates problematic collinearity; VIF $> 5$ is concerning.

---

## Checking VIF

```{r vif-interaction}
#| echo: true
#| code-fold: false

library(car)

# Compute VIF
vif(model_interaction)
```

High VIF values indicate multicollinearity issues.

---

## Dealing with multicollinearity

Solutions:

1. **Remove correlated predictors**: Drop one of the highly correlated variables
2. **Center predictors**: Mean-center continuous variables (reduces collinearity in interaction models)
3. **Combine predictors**: Create composite variables via PCA
4. **Regularization**: Use ridge regression or lasso
5. **Bayesian priors**: Informative priors can stabilize estimates

---

# Advanced Visualizations

## Marginal effects plots

Marginal effects show the predicted values while varying one predictor and holding others at typical values.

This clarifies the effect of each predictor in the presence of interactions.

---

## Marginal effects: acceptability by condition

```{r marginal-condition}
#| echo: false
#| fig-width: 10
#| fig-height: 6

# Create grid: vary acceptability and condition, hold surprisal at median
marg_cond <- expand.grid(
  mean_acceptability = seq(min(spr_full$mean_acceptability, na.rm = TRUE),
                          max(spr_full$mean_acceptability, na.rm = TRUE),
                          length.out = 100),
  condition = levels(spr_full$condition),
  surprisal = median(spr_full$surprisal, na.rm = TRUE)
)

# Predictions with standard errors
marg_cond_pred <- predict(model_interaction, newdata = marg_cond, se.fit = TRUE)
marg_cond$fit <- marg_cond_pred$fit
marg_cond$se <- marg_cond_pred$se.fit
marg_cond$lower <- marg_cond$fit - 1.96 * marg_cond$se
marg_cond$upper <- marg_cond$fit + 1.96 * marg_cond$se

# Plot
ggplot(marg_cond, aes(x = mean_acceptability, y = fit, color = condition, fill = condition)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, color = NA) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = "Mean Acceptability (z-score)",
       y = "Predicted Log RT (s)",
       title = "Marginal Effect of Acceptability by Condition",
       color = "Condition",
       fill = "Condition") +
  theme_minimal(base_size = 13)
```

---

## Marginal effects: surprisal by condition

```{r marginal-surprisal}
#| echo: false
#| fig-width: 10
#| fig-height: 6

# Create grid: vary surprisal and condition, hold acceptability at mean
marg_surp <- expand.grid(
  surprisal = seq(min(spr_full$surprisal, na.rm = TRUE),
                  max(spr_full$surprisal, na.rm = TRUE),
                  length.out = 100),
  condition = levels(spr_full$condition),
  mean_acceptability = mean(spr_full$mean_acceptability, na.rm = TRUE)
)

# Predictions with standard errors
marg_surp_pred <- predict(model_interaction, newdata = marg_surp, se.fit = TRUE)
marg_surp$fit <- marg_surp_pred$fit
marg_surp$se <- marg_surp_pred$se.fit
marg_surp$lower <- marg_surp$fit - 1.96 * marg_surp$se
marg_surp$upper <- marg_surp$fit + 1.96 * marg_surp$se

# Plot
ggplot(marg_surp, aes(x = surprisal, y = fit, color = condition, fill = condition)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, color = NA) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = "Surprisal (bits)",
       y = "Predicted Log RT (s)",
       title = "Marginal Effect of Surprisal by Condition",
       color = "Condition",
       fill = "Condition") +
  theme_minimal(base_size = 13)
```

---

# Bayesian Multiple Regression

## Extending to multiple predictors

Bayesian multiple regression extends naturally from simple regression.

We specify priors for all coefficients:

$$\beta_j \sim \text{Normal}(0, \sigma_{\beta})$$

for $j = 0, 1, \ldots, p$.

The posterior is computed via MCMC.

---

## Fitting with brms

```{r brms-multiple}
#| echo: true
#| code-fold: false
#| message: false
#| warning: false
#| results: hide
#| eval: false

library(brms)

# Fit Bayesian interaction model
bayes_multi <- brm(
  log_rt ~ surprisal * condition + mean_acceptability * condition,
  data = spr_full,
  family = gaussian(),
  prior = c(
    prior(normal(0, 5), class = Intercept),
    prior(normal(0, 5), class = b),
    prior(student_t(3, 0, 2.5), class = sigma)
  ),
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 42,
  file = "stan_output/bayes_multi"
)
```

---

## Model summary

```{r brms-multi-summary}
#| echo: true
#| code-fold: false
#| eval: false

summary(bayes_multi)
```

---

## Posterior distributions for key coefficients

```{r posterior-distributions}
#| echo: true
#| code-fold: false
#| fig-width: 10
#| fig-height: 7
#| eval: false

plot(bayes_multi, variable = c("b_surprisal",
                                "b_mean_acceptability",
                                "b_conditionlatecoercion"))
```

---

## Posterior predictive check

```{r ppc-multi}
#| echo: true
#| code-fold: false
#| fig-width: 8
#| fig-height: 5
#| eval: false

pp_check(bayes_multi, ndraws = 100)
```

---

## Credible intervals for all coefficients

```{r credible-intervals-multi}
#| echo: true
#| code-fold: false
#| eval: false

# Extract posterior samples
posterior_multi <- as_draws_df(bayes_multi)

# Compute 95% credible intervals for all coefficients
coef_names <- grep("^b_", names(posterior_multi), value = TRUE)
coef_names <- coef_names[!grepl("Intercept", coef_names)]

for (coef_name in coef_names) {
  ci <- quantile(posterior_multi[[coef_name]], c(0.025, 0.975))
  cat(sprintf("%s: [%.3f, %.3f]\n", coef_name, ci[1], ci[2]))
}
```

---

## Hypothesis testing: directional effects

```{r hypothesis-directional}
#| echo: true
#| code-fold: false
#| eval: false

# Probability that acceptability has a negative effect
prob_neg_accept <- mean(posterior_multi$b_mean_acceptability < 0)
cat("P(β_acceptability < 0 | data) =", prob_neg_accept, "\n")

# Probability that late coercion differs from early coercion
prob_late_diff <- mean(posterior_multi$b_conditionlatecoercion != 0)
cat("P(β_late ≠ 0 | data) ≈", prob_late_diff, "\n")
```

---

## Visualizing Bayesian predictions

```{r bayes-predictions-plot}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| eval: false

# Prediction grid
bayes_pred_grid <- expand.grid(
  surprisal = seq(min(spr_full$surprisal, na.rm = TRUE),
                  max(spr_full$surprisal, na.rm = TRUE),
                  length.out = 50),
  mean_acceptability = mean(spr_full$mean_acceptability, na.rm = TRUE),
  condition = levels(spr_full$condition)
)

# Predict with uncertainty
bayes_preds <- predict(bayes_multi, newdata = bayes_pred_grid)
bayes_pred_grid <- cbind(bayes_pred_grid, bayes_preds)

# Plot
ggplot() +
  geom_point(data = spr_full,
             aes(x = surprisal, y = log_rt, color = condition),
             alpha = 0.3, size = 1) +
  geom_line(data = bayes_pred_grid,
            aes(x = surprisal, y = Estimate, color = condition),
            linewidth = 1.5) +
  geom_ribbon(data = bayes_pred_grid,
              aes(x = surprisal, ymin = Q2.5, ymax = Q97.5, fill = condition),
              alpha = 0.2) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = "Surprisal (bits)",
       y = "Log Reaction Time (s)",
       title = "Bayesian Predictions with 95% Credible Intervals",
       color = "Condition",
       fill = "Condition") +
  theme_minimal(base_size = 12)
```

---

# Interpreting the Full Model

## Putting it all together

Our final model includes:

- **Main effects**: Surprisal, acceptability, condition
- **Interactions**: Surprisal × condition, acceptability × condition

Key findings:

1. Higher surprisal predicts slower reading (positive slope)
2. Higher acceptability predicts faster reading (negative slope)
3. The effects of both surprisal and acceptability differ across conditions

---

## Conditional effects

The effect of surprisal and acceptability depends on condition.

For early coercion (reference level):

$$\text{Surprisal effect} = \beta_{\text{surprisal}}$$

$$\text{Acceptability effect} = \beta_{\text{acceptability}}$$

For late coercion:

$$\text{Surprisal effect} = \beta_{\text{surprisal}} + \beta_{\text{surprisal×late}}$$

$$\text{Acceptability effect} = \beta_{\text{acceptability}} + \beta_{\text{acceptability×late}}$$

---

## Visualizing conditional slopes

```{r conditional-slopes}
#| echo: false
#| fig-width: 10
#| fig-height: 6

# Extract interaction coefficients from frequentist model
coefs <- coef(model_interaction)

# Compute slopes for each predictor × condition combination
slopes_df <- data.frame(
  condition = rep(levels(spr_full$condition), times = 2),
  predictor = rep(c("Surprisal", "Acceptability"), each = 2),
  slope = c(
    coefs["surprisal"],  # surprisal, early coercion
    coefs["surprisal"] + coefs["surprisal:conditionlate coercion"],  # surprisal, late coercion
    coefs["mean_acceptability"],  # acceptability, early coercion
    coefs["mean_acceptability"] + coefs["mean_acceptability:conditionlate coercion"]  # acceptability, late coercion
  )
)

# Plot
ggplot(slopes_df, aes(x = condition, y = slope, color = predictor, group = predictor)) +
  geom_point(size = 4) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_brewer(palette = "Set1") +
  labs(x = "Condition",
       y = "Effect on Log RT",
       title = "Conditional Slopes: How Predictor Effects Vary by Condition",
       color = "Predictor") +
  theme_minimal(base_size = 13)
```

---

## Effect sizes

Beyond statistical significance, we want to know the **magnitude** of effects.

For continuous predictors, standardized coefficients indicate effect size:

$$\beta_{\text{std}} = \beta \times \frac{\text{SD}(X)}{\text{SD}(Y)}$$

For categorical predictors, Cohen's $d$ measures standardized mean differences.

---

## Computing effect sizes

```{r effect-sizes}
#| echo: true
#| code-fold: false

library(effectsize)

# Standardized coefficients
standardize_parameters(model_interaction)
```

---

# Summary

## What we've covered

1. **Multiple regression**: Modeling with multiple predictors
2. **Categorical encoding**: Dummy variables and interpretation
3. **Interactions**: When effects depend on other variables
4. **Diagnostics**: Same checks apply to multiple regression
5. **Multicollinearity**: VIF and how to address it
6. **Visualizations**: Faceting, marginal effects, interaction plots
7. **Bayesian approach**: brms for multiple regression

---

## Key takeaways

- Multiple regression models joint effects of several predictors
- Categorical predictors are encoded as dummy variables
- Interactions capture non-additive effects
- Diagnostic checks ensure valid inference
- Visualizations clarify complex relationships
- Bayesian methods provide full uncertainty quantification

---

## Next steps

The [mixed-effects models](structured-data.qmd) unit extends these models with **random effects**, allowing us to model hierarchical structure in linguistic data (e.g., variation across participants and items).

This leads to **mixed-effects models**, which are the workhorse of modern psycholinguistic analysis.

---
