# Negative Binomial Distribution {.center background-color="#2166ac"}

## More Flexible Distribution

One such distribution that gives us more flexibility in this respect is the **negative binomial distribution**, which is very useful for modeling token frequency in text.

::: {.fragment}
This distribution effectively generalizes the geometric by allowing us to control the exponent on $\pi$ with a new parameter $r$.
:::

::: {.fragment}
$$\text{NegBin}(k; \pi, r) = {k+r-1 \choose r-1}(1-\pi)^{k}\pi^{r}$$
:::

---

## Normalizing Constant

This added flexibility in turn requires us to add an additional term ${k+r-1 \choose r-1} = \frac{(k+r-1)!}{(r-1)!\,(k)!}$ that ensures that the series $\sum_{k=0}^\infty \text{NegBin}(k; \pi, r)$ converges to $1$.

::: {.fragment}
The pieces of this term that do not include the value we're computing the probability of–i.e. $\frac{1}{(r-1)!}$–are often called the **normalizing constant**. We will make extensive use of this concept as the course moves forward.
:::

---

## Relationship to Geometric

When $r = 1$, we of course just get the geometric distribution. As such, if we keep $r = 1$, manipulating $\pi$ will have the same effect we saw above.

```{r}
#| echo: true
#| fig-width: 10
#| fig-height: 6

# When r = 1, NegBin reduces to Geometric
p <- 0.5
r <- 1
k <- 0:9

negbin_r1_data <- tibble(
  k = k,
  probability = dnbinom(k, size = r, prob = p)
)

ggplot(negbin_r1_data, aes(x = k, y = probability)) +
  geom_point(size = 4, color = "#b2182b") +
  geom_segment(aes(xend = k, yend = 0), color = "#b2182b", size = 2) +
  scale_x_continuous(breaks = k) +
  labs(title = paste("PMF of NegBin(", p, ",", r, ") distribution"),
       x = "k", y = "Probability") +
  theme_minimal()
```

---

## Mass Shifting with Parameter r

As $r$ grows, though, we get very different behavior: $p(k)$ is no longer always greater than $p(k + 1)$. Another way of saying this is that we can use $r$ to shift the probability mass rightward.

```{r}
#| echo: true
#| fig-width: 12
#| fig-height: 8

# Compare different r values
p <- 0.5
r_values <- c(1, 5, 10)
k <- 0:15

negbin_comparison <- map_df(r_values, function(r_val) {
  tibble(
    k = k,
    probability = dnbinom(k, size = r_val, prob = p),
    r = paste("r =", r_val)
  )
})

ggplot(negbin_comparison, aes(x = k, y = probability, color = r)) +
  geom_point(size = 3) +
  geom_line(size = 1.2) +
  facet_wrap(~ r, scales = "free_y") +
  scale_color_manual(values = c("r = 1" = "#2166ac", 
                                "r = 5" = "#b2182b", 
                                "r = 10" = "#4daf4a")) +
  scale_x_continuous(breaks = seq(0, 15, 3)) +
  labs(title = "PMF of NegBin(π, r) distribution for different r values",
       subtitle = paste("π =", p),
       x = "k", y = "Probability") +
  theme_minimal() +
  guides(color = "none")
```

---

## Effect of π on Mass Shifting

The mass-shifting effect is modulated by $\pi$: it accelerates with small $\pi$...

```{r}
#| echo: true
#| fig-width: 12
#| fig-height: 6

# Effect of different pi values
r <- 5
pi_values <- c(0.1, 0.5, 0.9)
k <- 0:20

pi_comparison <- map_df(pi_values, function(p_val) {
  tibble(
    k = k,
    probability = dnbinom(k, size = r, prob = p_val),
    pi = paste("π =", p_val)
  )
})

ggplot(pi_comparison, aes(x = k, y = probability, color = pi)) +
  geom_point(size = 2) +
  geom_line(size = 1) +
  scale_color_manual(values = c("π = 0.1" = "#2166ac", 
                                "π = 0.5" = "#b2182b", 
                                "π = 0.9" = "#4daf4a")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  labs(title = "Mass-shifting effect modulated by π",
       subtitle = paste("r =", r),
       x = "k", y = "Probability") +
  theme_minimal()
```

---

## ...but decelerates with large π

```{r}
#| echo: true
#| fig-width: 12
#| fig-height: 6

# Large r with different pi values
r <- 40
pi_values <- c(0.5, 0.9)
k <- 0:60

large_r_comparison <- map_df(pi_values, function(p_val) {
  tibble(
    k = k,
    probability = dnbinom(k, size = r, prob = p_val),
    pi = paste("π =", p_val)
  )
})

ggplot(large_r_comparison, aes(x = k, y = probability, color = pi)) +
  geom_point(size = 1.5) +
  geom_line(size = 1) +
  scale_color_manual(values = c("π = 0.5" = "#b2182b", 
                                "π = 0.9" = "#4daf4a")) +
  scale_x_continuous(breaks = seq(0, 60, 10)) +
  labs(title = "Mass-shifting decelerates with large π",
       subtitle = paste("r =", r),
       x = "k", y = "Probability") +
  theme_minimal()
```

---

## Real Word Length Distribution

This point can be seen if we look at the distribution of word lengths at the type level in the CMU Pronouncing Dictionary, which contains phonemic transcriptions of English words.

```{python}
#| echo: true
#| python.reticulate: false
import pandas as pd
from urllib.request import urlopen

cmudict_url = "http://svn.code.sf.net/p/cmusphinx/code/trunk/cmudict/cmudict-0.7b"

with urlopen(cmudict_url) as cmudict:
    words = [
        line.split()[1:] for line in cmudict if line[0] != 59
    ]

# Extract word lengths
word_lengths = [len(w) for w in words]

# Save to CSV for R processing
word_data = pd.DataFrame({'word_length': word_lengths})
word_data.to_csv('data/cmu_word_lengths.csv', index=False)

print(f"Processed {len(word_lengths)} words")
print(f"Length range: {min(word_lengths)} to {max(word_lengths)} phonemes")
```

```{r}
#| echo: true
#| fig-width: 12
#| fig-height: 6

# Load the processed CMU data
cmu_data <- read_csv("data/cmu_word_lengths.csv")

ggplot(cmu_data, aes(x = word_length)) +
  geom_histogram(bins = 32, fill = "#2166ac", color = "white", alpha = 0.8) +
  labs(title = "Distribution of word length in phonemes",
       x = "Word length in phonemes",
       y = "Count") +
  theme_minimal()
```

---

## Application to Word Lengths

We won't talk about how to fit a distribution to some data until later, when we talk about parameter estimation; but the negative binomial distribution can provide a reasonably good description of the empirical distribution of word lengths.

::: {.fragment}
One way to visualize this is to compare the empirical CDF with the CDF of the best fitting negative binomial.
:::

---

## Fitting Negative Binomial to Real Data

```{r}
#| echo: true

# Fit negative binomial distribution to real CMU data
library(MASS)

word_lengths <- cmu_data$word_length

# Fit negative binomial using method of moments/MLE
nb_fit <- fitdistr(word_lengths, "negative binomial")

# Extract parameters
r_fitted <- nb_fit$estimate[["size"]]
p_fitted <- nb_fit$estimate[["size"]] / (nb_fit$estimate[["size"]] + nb_fit$estimate[["mu"]])

cat(sprintf("Fitted parameters: r = %.2f, p = %.2f\n", r_fitted, p_fitted))
cat(sprintf("Mean = %.2f, Variance = %.2f\n", 
            nb_fit$estimate[["mu"]], 
            nb_fit$estimate[["mu"]] * (1 + nb_fit$estimate[["mu"]]/nb_fit$estimate[["size"]])))
```

---

## Empirical vs. Fitted Comparison

```{r}
#| echo: true
#| fig-width: 12
#| fig-height: 7

# Create empirical vs fitted CDF comparison
k_seq <- 1:30
empirical_cdf <- ecdf(word_lengths)
fitted_cdf <- pnbinom(k_seq, size = r_fitted, prob = p_fitted)

cdf_data <- tibble(
  k = k_seq,
  empirical = empirical_cdf(k_seq),
  fitted = fitted_cdf
) %>%
  pivot_longer(cols = c(empirical, fitted), 
               names_to = "type", 
               values_to = "probability")

ggplot(cdf_data, aes(x = k, y = probability, color = type)) +
  geom_line(size = 1.5) +
  geom_point(size = 2) +
  scale_color_manual(
    values = c("empirical" = "#2166ac", "fitted" = "#b2182b"),
    labels = c("Empirical CDF", 
               paste0("Estimated CDF\nNegBin(", round(p_fitted, 2), ", ", round(r_fitted, 2), ")"))
  ) +
  labs(title = "CDF of word length in phonemes",
       x = "Word length in phonemes",
       y = "Cumulative relative frequency/probability",
       color = NULL) +
  theme_minimal()
```

---

## Poisson as Limiting Case

A limiting case of the negative binomial distribution that you may be familiar with is the **Poisson distribution**.

$$\text{Pois}(k; \lambda) = \frac{\lambda^k\exp(-\lambda)}{k!}$$

::: {.fragment}
The Poisson distribution arises as:
$$\text{Pois}(k; \lambda) = \lim_{r \rightarrow \infty} \text{NegBin}
\left(k; r, \frac{\lambda}{r + \lambda}\right)$$
:::

::: {.fragment}
**Note**: We return to the Poisson distribution in [Poisson regression](../regression-models/poisson-regression.qmd), where we use it for parameter estimation.
:::
