# Overdispersion and Random Effects

## Connecting count regression to mixed-effects models

**Count regression**: Overdispersion in count data

- Poisson assumes $\text{Var}(Y) = \mathbb{E}[Y] = \lambda$
- Real data often show $\text{Var}(Y) > \lambda$ (overdispersion)
- Solutions: quasi-Poisson, negative binomial

**Mixed-effects models**: Random effects as a solution to overdispersion

- Overdispersion often caused by **clustering** and **heterogeneity**
- Random effects **explicitly model** these sources of extra variation
- Random effects **quantify** how much overdispersion comes from grouping

**Key insight**: Overdispersion and random effects are two perspectives on the same phenomenon.

---

## Sources of overdispersion

**1. Clustering**: Observations within groups are correlated

- Example: Verbs within a VerbNet class are more similar to each other
- Creates positive within-group correlation
- Violates independence assumption

**2. Heterogeneity**: Unmeasured factors vary across groups

- Example: Individual verbs have idiosyncratic frequency preferences
- Some verbs highly frequent, others rare
- Not captured by fixed effects alone

**3. Dependence**: Repeated measures from same units

- Example: Multiple frames for the same verb
- Observations from same verb are not independent

**All three create extra-Poisson variation**: $\text{Var}(Y) > \mathbb{E}[Y]$

---

## Three approaches to overdispersion

**Approach 1: Quasi-Poisson** (see [Poisson regression](poisson-regression.qmd))

$$\text{Var}(Y_i) = \phi \lambda_i$$

- Scales standard errors by $\sqrt{\phi}$ (dispersion parameter)
- Doesn't model the **source** of overdispersion
- Just acknowledges it exists and adjusts inference

**Approach 2: Negative binomial** (see [Poisson regression](poisson-regression.qmd))

$$Y_i \sim \text{NegBin}(\mu_i, \theta)$$

- Adds extra variance parameter $\theta$
- Variance-mean relationship: $\text{Var}(Y_i) = \mu_i + \frac{\mu_i^2}{\theta}$
- Models overdispersion but doesn't identify its source

**Approach 3: Random effects**

$$\log(\lambda_{ij}) = \mathbf{x}_{ij}^T \boldsymbol{\beta} + u_i, \quad u_i \sim N(0, \sigma_u^2)$$

- **Explicitly models** the source: group-level variation
- **Quantifies** overdispersion via $\sigma_u^2$
- Allows **generalization** beyond observed groups

---

## Variance decomposition in Poisson GLMM

For a Poisson GLMM with random intercepts:

$$\log(\lambda_{ij}) = \beta_0 + \mathbf{x}_{ij}^T \boldsymbol{\beta} + u_i$$

where $u_i \sim N(0, \sigma_u^2)$.

**Total variance** has two components:

1. **Poisson variance**: $\text{Var}(Y_{ij} \mid u_i) = \lambda_{ij}$ (inherent randomness)
2. **Random effects variance**: $\text{Var}(u_i) = \sigma_u^2$ (group-level heterogeneity)

**Marginal variance** (unconditional on random effects):

$$\text{Var}(Y_{ij}) = \mathbb{E}[\text{Var}(Y_{ij} \mid u_i)] + \text{Var}[\mathbb{E}(Y_{ij} \mid u_i)]$$

The second term captures extra variation due to random effects → **explains overdispersion**.

---

## Plot 1: Variance decomposition

```{r variance-decomposition-overdispersion}
#| echo: false
#| fig-width: 12
#| fig-height: 7

# Create visualization of variance components
variance_components <- tibble(
    Model = c("Poisson\n(No REs)", "Poisson\n(No REs)",
              "Poisson GLMM\n(Verb REs)", "Poisson GLMM\n(Verb REs)"),
    Component = c("Assumed Variance\n(Poisson)", "Observed Variance",
                  "Poisson Variance", "Random Effect\nVariance"),
    Variance = c(1, 47.6, 1, 46.6),  # Stylized based on actual deviance/df ratios
    Type = c("Assumed", "Observed", "Poisson", "RE")
)

ggplot(variance_components, aes(x = Model, y = Variance, fill = Component)) +
    geom_bar(stat = "identity", position = "stack", alpha = 0.8) +
    scale_fill_manual(values = c(
        "Assumed Variance\n(Poisson)" = "gray80",
        "Observed Variance" = "coral",
        "Poisson Variance" = "gray80",
        "Random Effect\nVariance" = "steelblue"
    )) +
    geom_text(aes(label = sprintf("%.1f", Variance)),
              position = position_stack(vjust = 0.5), size = 6, fontface = "bold") +
    labs(x = "", y = "Variance",
         title = "Variance Decomposition: Poisson vs. Poisson GLMM",
         subtitle = "Random effects explicitly partition variance, explaining overdispersion",
         fill = "") +
    theme_minimal(base_size = 16) +
    theme(legend.position = "top")
```

**Left**: Poisson assumes variance = mean, but observed variance >> assumed.

**Right**: GLMM partitions variance into Poisson (inherent) + random effects (overdispersion).

---

## Comparing three approaches

Let's compare quasi-Poisson, negative binomial, and Poisson GLMM on the VALEX data:

```{r compare-approaches}
#| echo: true
#| code-fold: false

# Load models (V0 and V2 already fitted)
# V0: Poisson GLM
# V2: Poisson GLMM

# Fit quasi-Poisson
v1 <- glm(count ~ frame + verbnet_class,
          family = quasipoisson,
          data = valex_model)

# Fit negative binomial
library(MASS)
v_nb <- glm.nb(count ~ frame + verbnet_class,
               data = valex_model)

# Compare dispersion
cat("Dispersion comparison:\n")
cat(sprintf("Poisson (V0): deviance/df = %.2f\n", deviance(v0) / df.residual(v0)))
cat(sprintf("Quasi-Poisson (V1): φ = %.2f\n", summary(v1)$dispersion))
cat(sprintf("Negative binomial (V_NB): θ = %.2f\n", v_nb$theta))
cat(sprintf("Poisson GLMM (V2): deviance/df = %.2f\n", deviance(v2) / df.residual(v2)))
```

---

## Plot 2: Dispersion reduction across models

```{r dispersion-reduction}
#| echo: false
#| fig-width: 12
#| fig-height: 7

# Dispersion metrics
dispersion_data <- tibble(
    Model = c("Poisson\n(V0)", "Quasi-Poisson\n(V1)", "Negative Binomial\n(V_NB)", "Poisson GLMM\n(V2)"),
    Dispersion = c(
        deviance(v0) / df.residual(v0),
        summary(v1)$dispersion,
        deviance(v_nb) / df.residual(v_nb),
        deviance(v2) / df.residual(v2)
    ),
    Type = c("Overdispersed", "Adjusted SEs", "Extra Parameter", "Random Effects")
) %>%
    mutate(Model = factor(Model, levels = Model))

ggplot(dispersion_data, aes(x = Model, y = Dispersion, fill = Type)) +
    geom_bar(stat = "identity", alpha = 0.8) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "red", linewidth = 1) +
    geom_text(aes(label = sprintf("%.2f", Dispersion)),
              vjust = -0.5, size = 6, fontface = "bold") +
    annotate("text", x = 1.5, y = 1.5, label = "Target: deviance/df ≈ 1",
             color = "red", size = 5) +
    scale_fill_manual(values = c("Overdispersed" = "coral",
                                  "Adjusted SEs" = "orange",
                                  "Extra Parameter" = "gold",
                                  "Random Effects" = "steelblue")) +
    labs(x = "", y = "Dispersion (deviance / df)",
         title = "Overdispersion Across Model Types",
         subtitle = "Random effects dramatically reduce overdispersion by modeling its source",
         fill = "Approach") +
    theme_minimal(base_size = 16) +
    theme(legend.position = "top") +
    ylim(0, max(dispersion_data$Dispersion) * 1.1)
```

**Key result**: Poisson GLMM (V2) achieves best dispersion control by modeling verb-level variation.

---

## How much overdispersion do random effects explain?

**Quantifying overdispersion reduction**:

Overdispersion in V0 (no REs):

$$\phi_{\text{V0}} = \frac{\text{deviance}}{\text{df}} \approx 47.6$$

Dispersion after adding verb REs (V2):

$$\phi_{\text{V2}} = \frac{\text{deviance}}{\text{df}} \approx 2.3$$

**Proportion of overdispersion explained**:

$$\frac{\phi_{\text{V0}} - \phi_{\text{V2}}}{\phi_{\text{V0}} - 1} = \frac{47.6 - 2.3}{47.6 - 1} \approx 0.97$$

**Interpretation**: Random effects explain ~97% of overdispersion! Almost all extra variation is due to verb-level heterogeneity.

---

## Plot 3: Predicted vs. observed variance

```{r predicted-observed-variance}
#| echo: false
#| fig-width: 12
#| fig-height: 7

# Calculate predicted means and observed variances by frame
# Use valex_model (same data v0 was fitted on)
variance_by_frame <- valex_model %>%
    group_by(frame) %>%
    summarize(
        mean_obs = mean(count, na.rm = TRUE),
        var_obs = var(count, na.rm = TRUE),
        n = n(),
        .groups = "drop"
    ) %>%
    filter(n >= 2, !is.na(var_obs)) %>%  # Need at least 2 obs to calculate variance
    filter(var_obs < quantile(var_obs, 0.95, na.rm = TRUE))  # Remove extreme outliers for visualization

# Plot: Poisson prediction (var = mean) vs observed
ggplot(variance_by_frame, aes(x = mean_obs, y = var_obs)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "coral",
                linewidth = 1.5, alpha = 0.7) +
    geom_point(size = 3, color = "steelblue", alpha = 0.7) +
    annotate("text", x = 50, y = 50, label = "Poisson assumption:\nVar = Mean",
             color = "coral", size = 5, hjust = 0) +
    annotate("text", x = 100, y = 1500,
             label = "Observed: Var >> Mean\n(Overdispersion)",
             color = "steelblue", size = 5, hjust = 0) +
    labs(x = "Observed Mean Count", y = "Observed Variance",
         title = "Variance vs. Mean by Syntactic Frame",
         subtitle = "Poisson assumes Var = Mean (red line) | Reality: Var >> Mean (points above line)") +
    theme_minimal(base_size = 14)
```

**Observation**: Variance dramatically exceeds mean—clear evidence of overdispersion.

---

## Plot 4: Residuals comparison (all three approaches)

```{r residuals-three-approaches}
#| echo: false
#| fig-width: 14
#| fig-height: 6

# Calculate residuals
valex_model$resid_v0 <- residuals(v0, type = "deviance")
valex_model$resid_v1 <- residuals(v1, type = "deviance")
valex_model$resid_v2 <- residuals(v2, type = "deviance")

# Create combined data
resid_combined <- tibble(
    Model = rep(c("Poisson\n(V0)", "Quasi-Poisson\n(V1)", "Poisson GLMM\n(V2)"),
                each = nrow(valex_model)),
    Residuals = c(valex_model$resid_v0, valex_model$resid_v1, valex_model$resid_v2)
) %>%
    mutate(Model = factor(Model, levels = c("Poisson\n(V0)", "Quasi-Poisson\n(V1)", "Poisson GLMM\n(V2)")))

ggplot(resid_combined, aes(x = Model, y = Residuals, fill = Model)) +
    geom_violin(alpha = 0.7) +
    geom_boxplot(width = 0.2, fill = "white", outlier.alpha = 0.2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    scale_fill_manual(values = c("coral", "orange", "steelblue")) +
    labs(x = "", y = "Deviance Residuals",
         title = "Residual Distributions Across Three Approaches",
         subtitle = "Poisson GLMM achieves smallest, most symmetric residuals") +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none") +
    ylim(-10, 10)
```

**V0**: Extreme outliers (overdispersion unaccounted for)

**V1**: Same residuals as V0 (just adjusts SEs, doesn't improve fit)

**V2**: Much tighter, symmetric distribution (random effects improve fit)

---

## Plot 5: QQ plots comparison

```{r qq-plots-comparison}
#| echo: false
#| fig-width: 14
#| fig-height: 5

# Create QQ plots
p1 <- ggplot(valex_model, aes(sample = resid_v0)) +
    stat_qq() +
    stat_qq_line(color = "red") +
    labs(x = "Theoretical Quantiles", y = "Sample Quantiles",
         title = "Poisson (V0)",
         subtitle = "Heavy tails indicate overdispersion") +
    theme_minimal(base_size = 12)

p2 <- ggplot(valex_model, aes(sample = resid_v1)) +
    stat_qq() +
    stat_qq_line(color = "red") +
    labs(x = "Theoretical Quantiles", y = "Sample Quantiles",
         title = "Quasi-Poisson (V1)",
         subtitle = "Same as V0 (adjusts SEs only)") +
    theme_minimal(base_size = 12)

p3 <- ggplot(valex_model, aes(sample = resid_v2)) +
    stat_qq() +
    stat_qq_line(color = "red") +
    labs(x = "Theoretical Quantiles", y = "Sample Quantiles",
         title = "Poisson GLMM (V2)",
         subtitle = "Much closer to diagonal") +
    theme_minimal(base_size = 12)

gridExtra::grid.arrange(p1, p2, p3, ncol = 3)
```

**V2 QQ plot**: Residuals much closer to normal distribution—better model specification.

---

## When to use each approach

**Quasi-Poisson**: Quick fix

- Adjusts standard errors for overdispersion
- Doesn't model source of overdispersion
- Use when you don't care about group structure
- Cannot compare models via AIC/BIC

**Negative binomial**: Flexible distribution

- Extra variance parameter
- Models overdispersion without specifying source
- Use when overdispersion is intrinsic (not due to clustering)
- Good for count data with high variance

**Poisson GLMM**: Mechanistic modeling

- Explicitly models source of overdispersion (clustering/heterogeneity)
- Allows generalization beyond observed groups
- Quantifies group-level variation
- Use when you have grouping structure and want to generalize
- Preferred when theory suggests clustering

---

## Connections to Problem Set 3

**Problem Set 3**: Heteroscedasticity in MALD data

- Variance in RT decreased with word frequency
- Modeled via `varPower()` in `gls()`: $\text{Var}(\epsilon_i) = \sigma^2 |x_i|^{2\delta}$
- Variance as a function of a continuous predictor

**Today**: Random effects for clustered data

- Variance differs across groups (participants, items, verbs)
- Modeled via random intercepts: $u_i \sim N(0, \sigma_u^2)$
- Variance as a function of grouping factors

**Connection**: Both model heterogeneity in variance

- PS3: Conditional heteroscedasticity (variance depends on X)
- Today: Group-level heterogeneity (variance differs across groups)

**Unified framework**: Distributional regression (both mean and variance modeled)

---

## Summary: Overdispersion and random effects

**Key insights**:

1. Overdispersion often caused by **clustering** and **heterogeneity**
2. Random effects **explicitly model** these sources
3. Random effects **quantify** how much variance is due to grouping
4. Three approaches: quasi-Poisson (adjust SEs), negative binomial (extra parameter), Poisson GLMM (model source)

**VALEX results**:

- Poisson GLM: severe overdispersion (φ ≈ 48)
- Poisson GLMM: reduced overdispersion (φ ≈ 2.3)
- Random effects explain ~97% of overdispersion

**Recommendation**: Use Poisson GLMM when you have grouping structure—models source of overdispersion and allows generalization.

---
