# Random Effects: Theory

## What is a random effect?

A **random effect** is a parameter treated as a random variable drawn from a population distribution.

**Fixed effect** (β): A coefficient we estimate directly

- Example: $\beta_1$ = effect of log frequency on log RT
- We estimate $\beta_1$ = -0.05, meaning higher frequency → faster RT
- Inference: "The frequency effect is -0.05"

**Random effect** (α): A parameter drawn from a distribution

- Example: $u_i$ = participant i's deviation from population mean RT
- We estimate $u_i \sim N(0, \sigma_u^2)$ where $\sigma_u^2$ = 0.019
- Inference: "Participants vary with SD = 0.14 around the population mean"

**Key difference**: Fixed effects estimate specific values; random effects estimate variance parameters.

---

## Why random effects?

Random effects model **sources of variation** that arise from **sampling** larger populations.

**Participants are a random sample** from the population of speakers:

- We test 30 participants, but want to generalize to all native speakers
- Each participant has their own baseline RT (some fast, some slow)
- We don't care about estimating each individual's exact RT
- We care about **how much participants vary** (the variance σ²_subject)

**Items are a random sample** from the population of words:

- We test 200 words, but want to generalize to the lexicon
- Each word has its own baseline RT (controlling for frequency)
- Some words are intrinsically easier/harder to process
- We care about **how much items vary** (the variance σ²_item)

---

## Fixed vs. random: The generalization question

Ask: **Do we want to generalize beyond the observed levels?**

**Participants**: YES

- We sampled 30 people from a population
- We want inferences to apply to **new, unseen participants**
- Therefore: Random effect

**Items**: YES

- We sampled 200 words from the lexicon
- We want inferences to apply to **new, unseen words**
- Therefore: Random effect

**Log frequency**: NO

- Frequency is a continuous predictor, not a grouping factor
- We want to estimate the specific effect of frequency
- Therefore: Fixed effect

---

## The sampling process

```{r sampling-diagram}
#| echo: false
#| fig-width: 12
#| fig-height: 6

library(ggplot2)
library(tidyverse)

# Create visualization of sampling process
set.seed(42)

# Population
pop_data <- tibble(
    id = 1:1000,
    value = rnorm(1000, mean = 0, sd = 0.15),
    sampled = id %in% sample(1:1000, 30)
)

p1 <- ggplot(pop_data, aes(x = value, fill = sampled)) +
    geom_histogram(bins = 50, alpha = 0.7) +
    scale_fill_manual(values = c("gray70", "steelblue"),
                      labels = c("Population", "Sample"),
                      name = "") +
    labs(x = "Random Intercept α", y = "Count",
         title = "Population of Participants",
         subtitle = "We sample 30 from a population of thousands") +
    theme_minimal(base_size = 14) +
    theme(legend.position = "top")

# Sample distribution
sample_data <- pop_data %>% filter(sampled)

p2 <- ggplot(sample_data, aes(x = value)) +
    geom_histogram(bins = 15, fill = "steelblue", alpha = 0.7) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
    annotate("text", x = 0.05, y = 6, label = "Population mean = 0",
             color = "red", hjust = 0) +
    labs(x = "Random Intercept u", y = "Count",
         title = "Our Sample (n = 30)",
         subtitle = "Sample estimates population variance") +
    theme_minimal(base_size = 14)

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

We observe the sample, estimate $\sigma_u^2$, and make inferences about the population.

---

## Mathematical formulation

For participant $i$ responding to item $j$:

$$Y_{ij} = \beta_0 + \beta_1 x_{ij} + u_i + \epsilon_{ij}$$

where:

- $\beta_0$: Population mean (fixed intercept)
- $\beta_1$: Fixed effect of predictor $x_{ij}$
- $u_i$: Random intercept for participant $i$
- $\epsilon_{ij}$: Residual error for observation $ij$

**Distributional assumptions**:

$$u_i \sim N(0, \sigma_u^2)$$
$$\epsilon_{ij} \sim N(0, \sigma^2)$$

**Parameters to estimate**: $\beta_0, \beta_1, \sigma_u^2, \sigma^2$

---

## Extending the GLM framework

Recall the three components of [generalized linear models](generalized-linear-models.qmd):

1. **Random component**: Distribution of $Y_i$
2. **Systematic component**: $\eta_i = \mathbf{x}_i^T \boldsymbol{\beta}$
3. **Link function**: $g(\mu_i) = \eta_i$

Mixed effects models extend the **systematic component** by adding random effects:

$$\eta_{ij} = \mathbf{x}_{ij}^T \boldsymbol{\beta} + u_i$$

The **random component** now has **two sources of variation**:

- **Residual (within-group)**: $\epsilon_{ij} \sim N(0, \sigma^2)$
- **Random effects (between-group)**: $u_i \sim N(0, \sigma_u^2)$

The **link function** remains unchanged: $g(\mu_{ij}) = \eta_{ij}$

---

## Variance decomposition

Total variance in $Y_{ij}$ comes from multiple sources:

$$\text{Var}(Y_{ij}) = \sigma_u^2 + \sigma^2$$

- $\sigma_u^2$: Between-participant variance (how much participants differ)
- $\sigma^2$: Within-participant variance (residual noise)

```{r variance-decomposition}
#| echo: false
#| fig-width: 8
#| fig-height: 6

# Load actual variance components from M3
vc <- read_csv("data/model_results/m3_variance_components.csv", show_col_types = FALSE)

vc_plot <- vc %>%
    mutate(
        component = case_when(
            grp == "subject" ~ "Between-Subject",
            grp == "item" ~ "Between-Item",
            grp == "Residual" ~ "Within (Residual)"
        ),
        percentage = vcov / sum(vcov) * 100
    )

ggplot(vc_plot, aes(x = "", y = vcov, fill = component)) +
    geom_bar(stat = "identity", width = 1) +
    coord_polar(theta = "y") +
    scale_fill_manual(values = c("steelblue", "coral", "gray70")) +
    labs(title = "Variance Decomposition: MALD Model M3",
         subtitle = sprintf("Total variance partitioned into three components"),
         fill = "Component") +
    theme_void(base_size = 14) +
    theme(legend.position = "right") +
    geom_text(aes(label = sprintf("%.1f%%", percentage)),
              position = position_stack(vjust = 0.5), size = 5, color = "white")
```

**Intraclass correlation (ICC)**: Proportion of variance due to grouping

$$\text{ICC} = \frac{\sigma_u^2}{\sigma_u^2 + \sigma^2}$$

---

## Random effects as deviations

Each participant has a random intercept $u_i$ representing their deviation from the population mean.

```{r random-intercepts-histogram}
#| echo: false
#| fig-width: 10
#| fig-height: 6

# Load actual model
load("data/model_results/all_models.RData")

# Extract random intercepts for subjects from M3
library(lme4)
subject_REs <- ranef(m3)$subject %>%
    rownames_to_column("subject") %>%
    rename(intercept = `(Intercept)`)

ggplot(subject_REs, aes(x = intercept)) +
    geom_histogram(bins = 15, fill = "steelblue", alpha = 0.7, color = "white") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
    annotate("text", x = 0.05, y = 6,
             label = sprintf("Mean = 0\nSD = %.3f", sd(subject_REs$intercept)),
             hjust = 0, color = "red", size = 5) +
    labs(x = "Random Intercept (log RT units)", y = "Count",
         title = "Distribution of Subject Random Intercepts",
         subtitle = "Each participant's deviation from population mean (Model M3)") +
    theme_minimal(base_size = 14)
```

Positive intercepts: slower than average participants

Negative intercepts: faster than average participants

---

## Caterpillar plot: Ranking random effects

A **caterpillar plot** displays random effects with confidence intervals, ordered by magnitude.

```{r caterpillar-plot}
#| echo: false
#| fig-width: 10
#| fig-height: 8

# Get random effects with SEs
subject_REs_conf <- ranef(m3, condVar = TRUE)$subject %>%
    rownames_to_column("subject") %>%
    rename(intercept = `(Intercept)`) %>%
    mutate(
        se = sqrt(attr(ranef(m3, condVar = TRUE)$subject, "postVar")[1, 1, ]),
        lower = intercept - 1.96 * se,
        upper = intercept + 1.96 * se
    ) %>%
    arrange(intercept) %>%
    mutate(subject = factor(subject, levels = subject))

ggplot(subject_REs_conf, aes(x = intercept, y = subject)) +
    geom_point(size = 3, color = "steelblue") +
    geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.3, color = "steelblue") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
    labs(x = "Random Intercept (log RT units)", y = "Subject",
         title = "Caterpillar Plot: Subject Random Intercepts",
         subtitle = "95% confidence intervals | Ordered from slowest to fastest") +
    theme_minimal(base_size = 14) +
    theme(axis.text.y = element_text(size = 10))
```

Subjects at the top are consistently slower; those at the bottom are consistently faster.

---

## Why not fixed effects for participants?

Why not estimate 30 separate fixed effects for participants?

**Problems with fixed effects**:

1. **Loss of generalizability**: Inferences limited to these 30 people
2. **Parameter explosion**: 30 parameters instead of 1 variance parameter
3. **No pooling**: Each participant estimated independently (inefficient)
4. **Overfitting**: More parameters = better fit to training data but worse generalization

---

## Benefits of random effects over fixed effects

**Random effects provide**:

1. **Generalizability**: Infer about population, not just sample
2. **Parsimony**: 1 variance parameter ($\sigma_u^2$) captures all participant variation
3. **Shrinkage**: Extreme estimates pulled toward population mean (regularization)
4. **Efficiency**: Borrow strength across participants (partial pooling)

---

## Shrinkage and borrowing strength

Random effects models apply **shrinkage**: Extreme group estimates are pulled toward the population mean.

**Why?** Groups with few observations or extreme values are likely overestimates. Shrinkage regularizes by borrowing information from other groups.

**Fixed effects**: No pooling—each group estimated independently

**Random effects**: Partial pooling—groups inform each other through shared variance

**No pooling** (complete pooling): Ignore groups entirely

This is the **bias-variance tradeoff** in action: Shrinkage increases bias but reduces variance, improving prediction.

---

## Summary: Random effects

**Random effects model sources of variation from sampling**:

- Participants sampled from population of speakers → $u_i \sim N(0, \sigma_u^2)$
- Items sampled from population of stimuli → $v_j \sim N(0, \sigma_v^2)$
- Verbs sampled within semantic classes → $w_k \sim N(0, \sigma_w^2)$

**Key concepts**:

- **Variance components**: $\sigma_u^2, \sigma_v^2, \sigma_w^2$ capture between-group variation
- **Generalizability**: Inferences extend beyond observed groups
- **Shrinkage**: Regularization via partial pooling
- **Systematic + random**: Extends GLM framework by adding REs to linear predictor

---
