# Regression Models

## Modeling relationships

So far, we have focused on modeling distributions of single variables or collections of independent variables.

In linguistics, we often want to model the relationship between variables—to predict one variable from another.

---

## Example: sentence processing

In psycholinguistics, researchers measure reading times to study how we process sentences in real time.

**Self-paced reading** presents sentences word-by-word on a computer screen. Participants press a button to advance to the next word.

The time between button presses (reaction time) reflects processing difficulty: longer RTs indicate difficulty integrating the current word.

Today we'll use a computational measure—**surprisal**—to predict reading times.

---

## Running example: aspectual coercion

We'll analyze data from a study on Spanish progressive statives.

Stative verbs (know, believe) typically reject the progressive form: *está conociendo* ('is knowing').

But adding a dynamic adverbial makes them acceptable: *está conociendo poco a poco* ('is knowing little by little').

This is aspectual coercion—semantic reinterpretation triggered by context.

---

## Experimental design

The study tested three conditions:

1. **Dynamic Control**: Compatible verb in progressive
   *Pedro está **pensando** en Dios cada día más...* ('Pedro is thinking...')

2. **Early Coercion**: Stative verb + adverbial immediately after verb
   *Pedro está **creyendo** cada día más en Dios...* ('Pedro is believing more...')

3. **Late Coercion**: Stative verb + adverbial after object
   *Pedro está **creyendo** en Dios cada día más...* ('Pedro is believing in God more...')

Participants read sentences word-by-word and answered comprehension questions.

---

# Language Models

## What is a language model?

A **language model** is a probability distribution over strings.

Formally:

- Let $\Sigma$ be a vocabulary (the set of all Spanish words)
- Let $\Sigma^*$ be the set of all sentences (strings of words from $\Sigma$)

A language model assigns a probability to each sentence $s \in \Sigma^*$:

$$p(s) \quad \text{where} \quad s \in \Sigma^*$$

This is a probability measure on the sample space $\Sigma^*$.

---

## Formal definition of $\Sigma^*$

The set of all strings over vocabulary $\Sigma$ is defined as:

$$\Sigma^* = \bigcup_{n=0}^\infty \Sigma^n$$

where:

- $\Sigma^0 = \{\varepsilon\}$ (the empty string)
- $\Sigma^1 = \Sigma$ (all single words)
- $\Sigma^2 = \{w_1 w_2 : w_1, w_2 \in \Sigma\}$ (all 2-word sequences)
- $\Sigma^3$ (all 3-word sequences), etc.

---

## The autoregressive factorization

How do we compute $p(s)$ for a sentence $s = w_1 w_2 \ldots w_n$?

By the [**chain rule** of probability](../foundations/probability-concepts.qmd):

$$p(w_1, w_2, \ldots, w_n) = p(w_1) \prod_{i=2}^n p(w_i \mid w_1, \ldots, w_{i-1})$$

An **autoregressive language model** computes each conditional probability $p(w_i \mid w_1, \ldots, w_{i-1})$.

That is, it predicts each word given all previous words.

---

## Surprisal

The **surprisal** of word $w_i$ in context is defined as:

$$\text{Surprisal}(w_i) = -\log_2 p(w_i \mid w_1, \ldots, w_{i-1})$$

**Interpretation**: How unexpected is $w_i$ given the preceding context?

- Low probability → High surprisal → Unexpected
- High probability → Low surprisal → Expected

**Key hypothesis**: More surprising words take longer to read.

---

## Computing surprisal with GPT-2

We use a pretrained Spanish GPT-2 model to compute surprisal values.

For each word in each sentence:

1. Feed the previous words into the model
2. Get $p(w_i \mid w_1, \ldots, w_{i-1})$
3. Compute surprisal: $-\log_2 p(w_i \mid \text{context})$

This gives us a continuous predictor for each word's reading time.

---

## Loading the data

```{r load-data}
#| echo: true
#| code-fold: false

library(tidyverse)

# Load word-by-word reading times with surprisal values
word_data <- read_csv("data/word_rt_surprisal.csv",
                      show_col_types = FALSE)

# Preview the data
head(word_data, 10)
```

---

## Data structure

Each row is a single word presentation to a single participant:

- `participant`: Participant ID
- `sentence`: Full sentence text
- `condition`: Experimental condition (dynamic control, early coercion, late coercion)
- `verb`: Critical verb
- `word`: Current word shown
- `word_position`: Position in sentence (0-indexed)
- `rt`: Reaction time in seconds
- `log_rt`: Natural log of RT
- `surprisal`: $-\log_2 p(\text{word} \mid \text{context})$ from GPT-2

---

## Distribution of reaction times

```{r plot-rt-distribution}
#| echo: false
#| fig-width: 8
#| fig-height: 5

ggplot(word_data, aes(x = rt)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", alpha = 0.8) +
  labs(x = "Reaction Time (seconds)",
       y = "Count",
       title = "Distribution of Raw Reading Times") +
  theme_minimal(base_size = 14)
```

Raw RTs are right-skewed—most words are read quickly, but some take much longer.

---

## Distribution of log reaction times

```{r plot-log-rt-distribution}
#| echo: false
#| fig-width: 8
#| fig-height: 5

ggplot(word_data, aes(x = log_rt)) +
  geom_histogram(bins = 50, fill = "coral", color = "white", alpha = 0.8) +
  labs(x = "Log Reaction Time (log seconds)",
       y = "Count",
       title = "Distribution of Log Reading Times") +
  theme_minimal(base_size = 14)
```

Log transformation produces a more symmetric distribution, closer to normality.

**Why log?** Linear regression assumes normally distributed residuals—log RT is more appropriate.

---

## Visualizing the relationship

```{r plot-data}
#| echo: false
#| fig-width: 8
#| fig-height: 5

ggplot(word_data, aes(x = surprisal, y = log_rt)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(x = "Surprisal (bits)",
       y = "Log Reaction Time (log seconds)",
       title = "Surprisal vs. Reading Time")
```

There appears to be a positive relationship: higher surprisal predicts slower reading.

---

## The regression question

Given observations of surprisal $X$ and reaction time $Y$, we want to model:

1. How log RT changes as surprisal changes
2. The strength of the relationship
3. Uncertainty in our estimates

Linear regression provides a framework for answering these questions.

---
