---
title: "Problem Set 4: Commitment in naturally occurring discourse"
subtitle: "Statistical Methods in Linguistics"
format:
  html:
    toc: true
execute:
  enabled: false
---

## What this problem set asks

Suppose someone says, *I did not know that it did that*. The complement *it did
that* is not asserted. Yet a listener may still take the speaker to be committed to
its truth. When commitment survives an environment that normally removes an
entailment, the complement is said to [**project**](https://doi.org/10.18148/sub/2019.v23i2.601).

A traditional account begins with the embedding predicate. Predicates such as
*know* are classified as factive, while predicates such as *think* are classified as
nonfactive. If this classification fully determines projection, the complement of a
factive predicate should receive high commitment ratings across discourses. The
CommitmentBank shows why this prediction is too strong. Factive predicates tend to
receive higher ratings, but predicates within the same class differ, individual
discourses differ, and annotators use the scale differently.

The binary factive/nonfactive classification is thus too coarse to determine commitment ratings by itself. The assignment
asks how much that classification tells us, then asks whether the
embedding environment tells us more. A multilevel ordinal model lets us answer both
questions while retaining the seven ordered response categories and representing
the three sources of repeated observations.

The focal method is the [ordinal mixed model](../../regression-models/ordinal-mixed-models.qmd) introduced in the notes. Prior predictive
simulation, sampling diagnostics, posterior prediction, posterior predictive
checking, and grouped predictive comparison are parts of evaluating that one model
family. They do not introduce additional response models.

::: {.callout-note title="Reading"}
Read de Marneffe, Simons, and Tonhauser (2019), [“The CommitmentBank:
Investigating projection in naturally occurring
discourse”](https://doi.org/10.18148/sub/2019.v23i2.601), focusing on Sections 2,
3.1, 3.4, and 3.5. The reading is included because it defines the commitment task,
motivates the factivity and embedding predictors, documents the restriction to
epistemic modals, and shows that predicate identity leaves substantial variation
unexplained. We will reuse the data but replace the paper's separate factor tests
with one prespecified Bayesian model sequence and a fuller dependence structure.
:::

Before beginning, review [ordinal
regression](../../regression-models/ordinal-regression.qmd), [ordinal mixed
models](../../regression-models/ordinal-mixed-models.qmd), [prior
predictive distributions](../../statistical-inference/prior-predictive-distributions.qmd),
[Hamiltonian Monte Carlo](../../statistical-inference/hamiltonian-monte-carlo.qmd),
[R-hat](../../statistical-inference/rhat.qmd), [effective sample
size](../../statistical-inference/effective-sample-size.qmd), [posterior predictive
checks](../../statistical-inference/posterior-predictive-checks.qmd), and [Bayesian
grouped cross-validation](../../model-criticism-and-prediction/grouped-cross-validation.qmd).
These pages establish the ordinal link and its multilevel extension before this
problem set applies them to CommitmentBank.

## Begin with the measurement

The response scale has seven ordered categories:

| Rating | Interpretation |
|---:|---|
| `-3` | the speaker is certain that the complement is false |
| `0` | the speaker is uncertain whether the complement is true or false |
| `3` | the speaker is certain that the complement is true |

The intermediate categories express intermediate ordered judgments. The difference
between `-3` and `-2` need not equal the difference between `1` and `2`. Thus a normal
linear model would add a spacing assumption that the task did not establish.

Recall from [ordinal regression](../../regression-models/ordinal-regression.qmd) that an ordinal model asks whether a response falls at or below each boundary.
With seven categories, there are six boundaries. For boundary $k$, the cumulative
logit model is

$$
\operatorname{logit}\left[\Pr(Y_i \leq k)\right]
= \theta_k - \eta_i,
\qquad k \in \{-3,-2,-1,0,1,2\}.
$$

The ordered thresholds $\theta_k$ divide a latent commitment scale into the seven
observed categories. The linear predictor $\eta_i$ moves an observation along that
scale. Because $\eta_i$ is subtracted, a positive coefficient lowers the probability
of falling at or below every boundary and shifts probability toward higher ratings.

One coefficient is shared across all six cumulative logits. This is the
[proportional-odds assumption introduced earlier](../../regression-models/ordinal-regression.qmd#interpret-one-coefficient-at-both-boundaries). It says that a predictor shifts location on the
latent scale rather than having a separate effect at every response boundary. We
will retain this assumption throughout the assignment so that the model sequence
changes only the linguistic predictors.

## Put the repeated observations into the model

Let rating $i$ come from annotator $a[i]$, discourse $d[i]$, and predicate $v[i]$.
The linear predictor for the larger model is

$$
\eta_i
= \beta_F F_i
+ \boldsymbol{x}_{E_i}^{\mathsf T}\boldsymbol{\beta}_E
+ u_{a[i]} + w_{d[i]} + z_{v[i]},
$$

with

$$
u_a \sim \operatorname{Normal}(0,\sigma_a^2),
\qquad
w_d \sim \operatorname{Normal}(0,\sigma_d^2),
\qquad
z_v \sim \operatorname{Normal}(0,\sigma_v^2).
$$

The annotator intercept represents a person's general tendency to use higher or
lower categories. The discourse intercept represents properties of a particular
passage that every annotator saw. The predicate intercept represents lexical
differences that remain after the binary factivity classification.

Including factivity and predicate variation is not redundant. The factivity
coefficient estimates the average difference between the two lexical classes. The
predicate distribution asks how far individual predicates depart from that class
average. This distinction makes the inadequacy of a class-only account estimable.

## Prepare the data

The official repository contains corpus passages from three source corpora and does
not state a separate data license. The course repository thus supplies a
retrieval script rather than a copy of the data. From the repository root, run:

```sh
Rscript problem-sets/ps4/scripts/prepare-commitmentbank.R
```

The script downloads the official participant-level CSV, checks a fixed SHA-256
checksum, and creates a local analysis table without the passage or prompt text.

```{r}
library(brms)
library(loo)
library(posterior)
library(ggplot2)

ratings <- read.csv("data/commitmentbank-model.csv")

primary <- ratings[ratings$modal_type %in% c("none", "EP"), ]
primary$rating <- ordered(primary$rating, levels = -3:3)
primary$factive <- factor(primary$factive, levels = c("no", "yes"))
primary$embedding <- factor(
  primary$embedding,
  levels = c("conditional", "modal", "negation", "question")
)
primary$annotator <- factor(primary$annotator)
primary$discourse <- factor(primary$discourse)
primary$predicate <- factor(primary$predicate)
primary$cell <- interaction(
  primary$factive,
  primary$embedding,
  sep = ": "
)
primary <- droplevels(primary)

stopifnot(
  nrow(primary) == 9599L,
  nlevels(primary$discourse) == 982L,
  nlevels(primary$annotator) == 496L,
  nlevels(primary$predicate) == 45L
)
```

The filter follows the paper. Weak epistemic modals can diagnose projection because
they remove the matrix entailment without adding a non-epistemic interpretation.
Ability, circumstantial, and deontic modals do not provide the same diagnostic. We
will restore them only in the predeclared sensitivity analysis.

## Task 1: reconstruct the design (10 points) {#task-1-reconstruct-the-design}

Create one table of rating counts and a second table of discourse counts by
`factive` and `embedding`. Then answer the following questions.

1. Why is the rating an ordered response rather than a continuous response?
2. What is repeated within an annotator?
3. What is repeated within a discourse?
4. Why can the predicate effect not replace the discourse effect?
5. Which factivity by environment cells have the weakest discourse support?

State the target of the conditional predictions used later. They describe a rating
with the annotator, discourse, and predicate random intercepts set to zero. They are
not probabilities averaged over those random-effect distributions, and they do not
describe every English speaker or every clause-embedding predicate.

## Task 2: check the priors before fitting the data (15 points) {#task-2-check-the-priors-before-fitting-the-data}

Use the prespecified priors:

```{r}
prior_set <- c(
  prior(normal(0, 1.5), class = "Intercept"),
  prior(normal(0, 1), class = "b"),
  prior(exponential(1), class = "sd")
)
```

The `Intercept` class names the six ordered thresholds in this model. The `b` class
names the factivity and embedding coefficients. The `sd` class names the three group
standard deviations.

Define the larger formula now because a prior predictive check should use the model
whose posterior will later be interpreted.

```{r}
environment_formula <- bf(
  rating ~ factive + embedding +
    (1 | annotator) + (1 | discourse) + (1 | predicate),
  family = cumulative("logit")
)

# Fit with sample_prior = "only".
```

Draw 500 replicated datasets from the prior predictive distribution. Plot the
overall proportion in each category for every replicate, then place the observed
proportions on the same plot. Answer three questions.

1. Do the priors permit all seven response categories?
2. Do they routinely imply that nearly every response occupies one category?
3. Would you revise the priors before seeing the posterior? Explain the proposed
   revision or explain why no revision is needed.

Do not choose a prior because it recreates the observed distribution closely. Prior
predictive checking rules out implications that were unreasonable before the
outcomes were used for fitting.

## Task 3: fit the factivity model (20 points) {#task-3-fit-the-factivity-model}

The first model asks whether the traditional lexical classification
predicts the ordered commitment response while partially pooling the three grouping
units.

```{r}
factivity_formula <- bf(
  rating ~ factive +
    (1 | annotator) + (1 | discourse) + (1 | predicate),
  family = cumulative("logit")
)

# Fit the factivity-only model with four chains, 1,000 warmup iterations, and 1,000 retained
# iterations per chain. Use the prespecified priors, adapt_delta = 0.95, and a
# fixed seed.
```

Before interpreting the model, report:

1. the number of divergent transitions;
2. the largest split R-hat;
3. the smallest bulk effective sample size among reported population and group-scale
   parameters; and
4. the smallest tail effective sample size among those parameters.

Then report the median and 95 percent posterior interval for `factiveyes`. Explain
its sign using the cumulative probability equation. Finally, report the three group
standard deviations and inspect the predicate deviations for *know*, *realize*,
*accept*, and *tell*.

The factivity coefficient and predicate deviations answer different questions. A
positive class coefficient can coexist with substantial lexical departures. Explain
how that combination bears on the limits of a class-only account.

## Task 4: add embedding environment (20 points) {#task-4-add-embedding-environment}

Fit the model that adds embedding environment using `environment_formula`. Use
the same data, priors, random structure, chains, iterations, and seed as the factivity-only model. This is a
prespecified comparison: the only formula change is the addition of `embedding`.

Perform the same diagnostic audit used for the factivity-only model. Then summarize the three embedding
coefficients. Conditional antecedents are the reference level, so each coefficient
compares one environment with a conditional antecedent while holding factivity fixed
and setting the grouping deviations to their modeled values.

Environment is constant within a discourse. This fact determines how the two models
must be compared. If we leave out one rating, the remaining ratings of that discourse
still estimate its discourse intercept. The intercept can then carry information
that the environment predictor would otherwise supply. A rating-level comparison
would answer the wrong question.

The prespecified comparison instead uses five folds grouped by discourse and stratified by
factivity and environment. Every rating of a discourse enters the same fold. Each
held-out discourse receives one new discourse intercept drawn from the fitted group
distribution, and that intercept is shared across its ratings. The ten model fits across
five folds are slow, so you will audit the supplied discourse-level output instead of refitting them.

[Download the grouped comparison
results](data/commitmentbank-grouped-cv.csv), then read them into R.

```{r}
grouped_cv <- read.csv("data/commitmentbank-grouped-cv.csv")

stopifnot(
  nrow(grouped_cv) == 982L,
  length(unique(grouped_cv$case)) == 982L,
  setequal(grouped_cv$fold, 1:5)
)
```

First verify that each fold contains all eight factivity by environment cells and
between 194 and 200 discourses. Each value $\ell_{d,m}$ in the supplied file is the
log predictive density for all ratings in discourse $d$ under model $m$. This is the
[group-level pointwise quantity](../../model-criticism-and-prediction/grouped-cross-validation.qmd)
required by the new-discourse target. Then compute

$$
\Delta = \sum_d \left(\ell_{d,\mathrm{environment}} -
\ell_{d,\mathrm{factivity}}\right)
$$

and its discourse-level standard error,

$$
\operatorname{SE}(\Delta)
= \sqrt{D\operatorname{Var}\left(\ell_{d,\mathrm{environment}} -
\ell_{d,\mathrm{factivity}}\right)}.
$$

Report the difference, its standard error, and their ratio. Explain why one row is
one discourse rather than one rating. State whether the difference exceeds the
prespecified two-standard-error threshold. Finally, explain one limitation of the standard
error: the discourse is the pointwise unit, but the five training sets overlap.

## Task 5: return to response categories (20 points) {#task-5-return-to-response-categories}

Create the eight combinations of factivity and embedding environment. Use
`posterior_epred()` with `re_formula = NA` to obtain conditional category
probabilities from the model that includes embedding environment, with every random
intercept set to zero. Do not call these population-averaged probabilities, because
they do not integrate over the fitted random-effect distributions.

For each combination, report:

1. the posterior mean and 95 percent interval for each of the seven category
   probabilities;
2. the probability of a high commitment response, defined before fitting as
   $\Pr(Y \geq 2)$; and
3. the posterior expected rating, used only as a compact summary of the seven
   probabilities.

Plot the seven category probabilities rather than only the expected ratings. Then
answer three substantive questions.

1. Does factivity shift probability toward stronger commitment in every
   environment?
2. Which environment has the largest probability of negative commitment responses?
3. Does any environment make the factive and nonfactive distributions identical?

The additive model builds in a common latent factivity shift. It does not build in
identical response probabilities, because the inverse link and thresholds make the
probability differences depend on position along the scale.

## Task 6: criticize the model (10 points) {#task-6-criticize-the-model}

Generate posterior predictive datasets from the model that includes embedding environment. Compare observed and replicated
category proportions overall, by factivity, by embedding environment, and within the
eight factivity by environment cells. Identify one feature the model reproduces and
one feature it misses. A fitted coefficient is not a substitute for checking the
response distribution that the model generates.

Next perform the prespecified sensitivity analysis. Restore all modal types, rebuild the
factors in the same order, and refit the model that includes embedding environment without changing its formula or priors.
Compare the factivity and embedding summaries with the primary fit. Focus on the
modal prediction because every restored item contains a non-epistemic modal.

Explain whether the sensitivity result supports the paper's decision to restrict the
projection diagnostic to epistemic modals. Do not select whichever fit gives a more
dramatic coefficient. The two fits answer questions about different inclusion rules.

## Task 7: license the claim (5 points) {#task-7-license-the-claim}

Write a six-to-eight sentence results paragraph that does all of the following:

1. names the ordered response and the two-model sequence;
2. reports the factivity direction and uncertainty;
3. reports variation among annotators, discourses, and predicates;
4. reports the embedding model comparison with its predictive target;
5. describes the category-level factivity and negation pattern;
6. reports the modal-filter sensitivity; and
7. limits the claim to these naturally occurring corpus discourses and the recorded
   annotators.

Do not say that factivity determines projection. The intended conclusion is narrower:
factivity supplies a positive but coarse shift, while embedding environment and the
modeled grouping units account for additional structure in commitment judgments.

## Submission checklist

1. The data checksum passes and no corpus text is submitted.
2. The response remains an ordered factor with levels `-3` through `3`.
3. The factivity-only model and the model that adds embedding environment use identical priors and random structures.
4. Every fitted model receives a diagnostic audit before interpretation.
5. The grouped comparison uses discourse as its pointwise unit and reports the
   uncertainty of the difference.
6. Posterior predictions show category probabilities, not only latent coefficients.
7. The sensitivity analysis changes only the modal inclusion rule.
8. The final paragraph states what the corpus sampling does not license.
