---
title: "Bootstrap confidence intervals"
---

The [sampling distribution page](properties-of-estimators.qmd) described how an estimator varies across possible samples, and the [confidence interval page](confidence-intervals.qmd) used that variation to construct an interval. The bootstrap approximates the same sampling distribution by resampling. Suppose we observe median fundamental frequency for twelve speakers:

```{r}
#| label: f0-sample
#| echo: true
f0 <- c(182, 195, 201, 176, 188, 210, 193, 184, 205, 199, 180, 191)
median(f0)
```

The sample median is 192 Hz. An analytic sampling distribution for the median is less familiar than the sampling distribution of a mean. The [**nonparametric bootstrap**](https://openintro-ims.netlify.app/foundations-bootstrapping.html) approximates it by resampling the observed units.

The target is the population median of the speaker-level summaries. The
estimator is the median of the twelve observed summaries. We would like to know
how much that estimator might change if the sampling process produced another
set of twelve speakers from the same target population.

Now, you might wonder how we can approximate repeated sampling when the population distribution is precisely what we do not know. If the population distribution were known, we could repeatedly sample from it
and calculate a median after each sample. But the population distribution is
the object we do not know. The nonparametric bootstrap replaces it with the
[**empirical distribution**](https://openintro-ims.netlify.app/foundations-bootstrapping.html), which assigns probability $1/12$ to each observed
speaker summary. This replacement is the bootstrap's central approximation. It lets the observed units stand in for a population distribution, then applies the estimator anew to each resample.

## Drawing one bootstrap sample

Sample twelve values from the observed twelve, with replacement:

```{r}
#| label: one-bootstrap-sample
#| echo: true
set.seed(414)
f0_star <- sample(f0, size = length(f0), replace = TRUE)
f0_star
median(f0_star)
```

Because sampling is with replacement, some speakers appear more than once and others do not appear in this bootstrap sample.

For instance, one bootstrap sample might contain

$$
176,180,180,184,188,191,193,193,199,201,205,210.
$$

The value 180 occurs twice, the value 193 occurs twice, and several observed
values do not occur. Its median is 192. Another resample will generally contain
a different pattern of repetitions and omissions, so its median may differ.

Sampling with replacement is essential. If every bootstrap sample contained
each original speaker exactly once, every sample would merely reorder the same
twelve numbers. The median would always remain 192, and the resulting
distribution would contain no information about sampling variation.

## Repeating the procedure

For bootstrap replicate $b$:

1. draw a sample of size $n$ from the observed units with replacement;
2. calculate the estimator on that resample; and
3. store the resulting estimate $\widehat{\theta}^{*(b)}$.

Repeat these steps many times:

```{r}
#| label: bootstrap-medians
#| echo: true
set.seed(414)

bootstrap_medians <- replicate(
  10000,
  median(sample(f0, size = length(f0), replace = TRUE))
)

hist(bootstrap_medians, xlab = "bootstrap median", main = "")
```

The empirical distribution of bootstrap estimates approximates the estimator's sampling distribution, using the observed empirical distribution in place of the unknown population distribution.

The two distributions involved here should not be conflated. The empirical
distribution of the twelve $f_0$ values approximates the population distribution
of speaker summaries. The distribution of the 10,000 bootstrap medians
approximates the sampling distribution of the median estimator. We resample
observations from the first distribution to construct the second.

The following summary makes that separation visible:

```{r}
#| label: summarize-bootstrap-medians
#| echo: true
c(
  observed_median = median(f0),
  bootstrap_mean = mean(bootstrap_medians),
  bootstrap_standard_error = sd(bootstrap_medians)
)
```

The standard deviation of the bootstrap medians is a bootstrap estimate of the
standard error. It describes variation in the estimator across bootstrap
samples. It is not the standard deviation of speaker fundamental frequency.

## Forming a percentile interval

A 95% [**percentile bootstrap confidence interval**](https://openintro-ims.netlify.app/foundations-bootstrapping.html) takes the $.025$ and $.975$ quantiles of the bootstrap estimates:

```{r}
#| label: bootstrap-percentile-interval
#| echo: true
quantile(bootstrap_medians, c(.025, .975))
```

The calculation is simple, but the percentile interval is not always the most accurate bootstrap construction. Its purpose here is to make the resampling logic visible.

Suppose the resulting interval is $[182,201]$ Hz. The repeated-sampling claim is
about the procedure. If we could repeatedly sample twelve speakers from the
target population and construct an interval in this way, the method is intended
to cover the population median in about 95 percent of those repetitions under
conditions where the bootstrap approximation works well. The endpoints are not
probabilities for the fixed population median.

The percentile method uses the quantiles of the bootstrap estimates directly.
It tends to work best when the bootstrap distribution is reasonably centered on
the observed estimate and not strongly skewed. With only twelve observations,
the sample may give a poor approximation to the population, especially in its
tails. The interval can also have coarse endpoints because a median from a small
sample can take only a limited set of values.

## Choosing the resampling unit

The example has one summary per speaker, so resampling speakers matches the proposed independent unit.

If the raw table instead contained many vowel tokens per speaker, resampling individual rows would treat those tokens as exchangeable across speakers and destroy the within-speaker structure. A suitable bootstrap might resample speakers and retain each selected speaker's observations together.

Consider a table with 20 speakers and 30 vowel tokens per speaker. The table has
600 rows, but it does not necessarily contain 600 independent units. If the
scientific claim concerns speakers, one speaker should enter or leave a
bootstrap sample as a block. A selected speaker may appear twice, along with both
copies of that speaker's complete token set. An omitted speaker contributes no
tokens to that replicate.

The statistic must also be recomputed from the resampled data. If the analysis
first averages tokens within speaker and then compares vowel categories within
speaker, each bootstrap replicate must repeat those operations. Resampling a
finished list of token-level residuals would correspond to a different procedure.

The bootstrap reproduces the sampling assumptions encoded by the resampling design. It does not decide which units are independent.

## Checking simulation error

The earlier [random seed page](../random-variables-and-distributions/random-seeds.qmd) separated reproducibility from validity. Here the number of bootstrap replicates $B$ controls Monte Carlo precision. With a
small $B$, the estimated quantiles may move appreciably when the random seed
changes. We can check this source of variation by running the same procedure
with several values of $B$.

```{r}
#| label: bootstrap-replicate-sensitivity
#| echo: true
bootstrap_interval <- function(B, seed) {
  set.seed(seed)
  estimates <- replicate(
    B,
    median(sample(f0, size = length(f0), replace = TRUE))
  )
  quantile(estimates, c(.025, .975))
}

bootstrap_interval(500, 214)
bootstrap_interval(5000, 214)
bootstrap_interval(5000, 414)
```

If the 5,000-replicate intervals agree closely while the 500-replicate interval
moves, more simulation has reduced Monte Carlo error. If large bootstrap runs
still produce an unstable scientific conclusion, the issue may lie in the small
original sample or the resampling design. Increasing $B$ cannot create new
speakers.

::: {.callout-note title="Check one replicate"}
1. Why must a nonparametric bootstrap sample contain $n$ draws rather than every
   original case exactly once? Explain what would happen to the estimate if no
   cases were repeated or omitted.
2. Distinguish the empirical distribution of the observations from the bootstrap
   distribution of the estimates.
3. If a dataset contains repeated tokens from each speaker, state which rows
   should move together when speakers are the intended independent units.
4. Explain why increasing $B$ reduces simulation error but does not repair a
   sample containing too few speakers.
:::
