---
title: "Maximum likelihood for a Poisson mean"
---

The [maximum-likelihood page](maximum-likelihood.qmd) introduced estimation with
a Bernoulli response. We now repeat the same reasoning with the [Poisson distribution](../random-variables-and-distributions/poisson-distribution.qmd) for a count response. The question is whether maximizing a count likelihood leads to a familiar sample summary. Deriving the answer will also show which steps depend specifically on the Poisson model.

## Starting with a linguistic count

The Universal Dependencies English Web Treebank contains annotated dependency
relations. Consider a teaching extract of 2,594 wh expressions and the linear
distance from each wh expression to its annotated head. A relative pronoun such
as *who* may be near its head, as in *the officer who was working*, or separated
from it by several intervening tokens.

::: {.callout-note title="Data source"}
The extract comes from the [Universal Dependencies English Web
Treebank](https://universaldependencies.org/treebanks/en_ewt/index.html), derived
from the English Web Treebank described by [Silveira et al.
(2014)](https://aclanthology.org/L14-1067/). The variable used here is a distance
between a selected wh-form and its head in the basic dependency annotation. More specifically, it is the absolute difference between their token indices. The stored file retains legacy column names such as `gap_position`, but that position is the annotated head token, not a silent gap. The measure should not be treated as a complete representation of a filler-gap dependency or its structural distance.
:::

Let $X_i$ be this wh-to-head token distance for observation $i$. The observed distances are
positive integers, but the Poisson family also assigns probability to zero. We
will temporarily ignore that mismatch so that we can isolate the estimation
step. Our working model is

$$
X_i\mathrel{\overset{\text{IID}}{\sim}}\operatorname{Poisson}(\lambda).
$$

The model says that every observation shares one Poisson mean $\lambda$ and that the
observations are independent after we condition on that mean. Those claims are
strong. We will check one of their consequences after finding the estimate.

## Writing the likelihood one observation at a time

For one observed distance $x_i$, the Poisson probability mass is

$$
p(x_i\mid\lambda)
=\frac{e^{-\lambda}\lambda^{x_i}}{x_i!}.
$$

For $n$ independent observations, multiplication gives the joint probability

$$
p(\mathbf{x}\mid\lambda)
=\prod_{i=1}^{n}\frac{e^{-\lambda}\lambda^{x_i}}{x_i!}.
$$

Once the distances are observed, we hold $\mathbf{x}$ fixed and view the same
expression as a function of $\lambda$:

$$
L(\lambda;\mathbf{x})
=\prod_{i=1}^{n}\frac{e^{-\lambda}\lambda^{x_i}}{x_i!}.
$$

This is the likelihood function. It compares candidate values of the mean by
asking how much probability each candidate assigns to the observed collection
of distances.

## Taking logs before differentiating

The product contains one small probability for each observation. Taking a log
turns that product into a sum:

$$
\begin{aligned}
\ell(\lambda;\mathbf{x})
&=\log L(\lambda;\mathbf{x})\\
&=\sum_{i=1}^{n}
\left[x_i\log\lambda-\lambda-\log(x_i!)\right]\\
&=\left(\sum_{i=1}^{n}x_i\right)\log\lambda
  -n\lambda
  -\sum_{i=1}^{n}\log(x_i!).
\end{aligned}
$$

The final term depends on the observations but not on $\lambda$. It changes the
height of the log-likelihood curve, but it cannot change where that curve is
maximized.

## Finding the maximizing mean

Differentiate with respect to $\lambda$:

$$
\frac{\mathrm{d}}{\mathrm{d}\lambda}
\ell(\lambda;\mathbf{x})
=\frac{\sum_{i=1}^{n}x_i}{\lambda}-n.
$$

At an interior maximum, this derivative is zero. We now isolate $\widehat{\lambda}$ one algebraic move at a time:

$$
\begin{aligned}
\frac{\sum_{i=1}^{n}x_i}{\widehat{\lambda}}-n&=0,\\
\sum_{i=1}^{n}x_i&=n\widehat{\lambda},\\
\widehat{\lambda}_{\mathrm{MLE}}
&=\frac{1}{n}\sum_{i=1}^{n}x_i\\
&=\bar{x}.
\end{aligned}
$$

The maximum-likelihood estimate of a shared Poisson mean is the sample mean. This
equality is a consequence of the Poisson model and its likelihood. It is not a
general definition of maximum likelihood.

## Verifying the result on the dependency distances

The data are available in
[`wh_filler_gap.csv`](../problem-sets/ps1-2/data/wh_filler_gap.csv). We can compute
the analytic estimate and compare it with a numerical optimizer.

```{r}
#| label: poisson-mle-wh-distance
#| echo: true
wh <- read.csv("../problem-sets/ps1-2/data/wh_filler_gap.csv")
x <- wh$linear_distance

analytic_mle <- mean(x)

negative_log_likelihood <- function(lambda, observations) {
  -sum(dpois(observations, lambda = lambda, log = TRUE))
}

numerical_fit <- optimize(
  negative_log_likelihood,
  interval = c(.01, 20),
  observations = x
)

c(
  analytic = analytic_mle,
  numerical = numerical_fit$minimum
)
```

Both calculations return approximately $3.007$. The agreement is not a second
piece of evidence about the linguistic phenomenon. It is a check that the code
and the algebra solve the same optimization problem.

We can also inspect the log-likelihood near the maximum.

```{r}
#| label: poisson-log-likelihood-curve
#| echo: true
lambda_grid <- seq(2.5, 3.5, length.out = 300)
log_likelihood <- vapply(
  lambda_grid,
  function(lambda) sum(dpois(x, lambda = lambda, log = TRUE)),
  numeric(1)
)

plot(
  lambda_grid, log_likelihood,
  type = "l",
  xlab = expression(lambda),
  ylab = "log-likelihood"
)
abline(v = analytic_mle, lty = 2)
```

The curve is highest at the sample mean. Values near the maximum have similar
support from the observed data, while more distant values assign lower joint
probability to those observations.

## Separating estimation from model adequacy

The estimate answers a conditional question: if the distances were independent
draws from one Poisson distribution, which value of $\lambda$ would fit them
best? It does not establish that the Poisson family is a good description.

For this extract,

$$
\bar{x}\approx3.01
\qquad\text{and}\qquad
s^2\approx10.46.
$$

A Poisson variable has equal mean and variance. The fitted model must thus have
both mean and variance near 3.01, while the empirical variance is more than
three times as large. The distances also cannot be zero in this particular
extract, even though the fitted distribution assigns positive probability to
zero.

These discrepancies identify two distinct model failures. The support mismatch
concerns which values can occur. The variance mismatch concerns how probability
is distributed over the values that can occur. Maximum likelihood finds the
best parameter within the chosen family. It cannot repair a family whose
structural assumptions miss the response.

::: {.callout-note title="Work through the estimate"}
1. For the distances $1,2,2,3,7$, calculate
   $\widehat{\lambda}_{\mathrm{MLE}}$.
2. Explain why $\sum_i\log(x_i!)$ disappears when we differentiate with respect
   to $\lambda$.
3. State the conditional question answered by the estimate.
4. Give one reason that wh dependency distances from the same document might
   not be independent.
:::
