---
title: "Posterior predictive distributions"
---

The [prior predictive distribution](prior-predictive-distributions.qmd) averaged the observation model over the prior. A posterior predictive distribution performs the same averaging over the [posterior distribution](posterior-distributions.qmd).

Continue with the picture-matching task for an ambiguous prepositional phrase. Let a response of one mean that the participant selects the picture corresponding to verb-phrase attachment, and let $\Pi$ be this participant's response probability across the declared item set. Suppose the observed responses $\mathbf{x}$ produce the posterior

$$
\Pi\mid\mathbf{X}=\mathbf{x}
\sim
\operatorname{Beta}(18,6).
$$

The posterior describes uncertainty about the participant's response probability under this model. For compactness, later expressions use conditioning on $\mathbf{x}$ as shorthand for conditioning on the event $\mathbf{X}=\mathbf{x}$. But another block records responses, not parameter values. A [**posterior predictive distribution**](https://bruno.nicenboim.me/bayescogsci/ch-introBDA.html) describes those possible responses that have not yet been observed; the [Stan User's Guide](https://mc-stan.org/docs/stan-users-guide/posterior-predictive-checks.html#simulating-from-the-posterior-predictive-distribution) gives the corresponding integral and simulation procedure.

## Averaging over parameter uncertainty

Let $\widetilde{X}$ denote a future binary response. Conditional on a fixed parameter value,

$$
\widetilde{X}\mid\Pi=\pi
\sim
\operatorname{Bernoulli}(\pi).
$$

Because $\pi$ is not known, prediction averages this observation model over the posterior:

$$
\mathbb{P}(\widetilde{X}=\widetilde{x}\mid\mathbf{X}=\mathbf{x})
=
\int
\mathbb{P}(\widetilde{X}=\widetilde{x}\mid\Pi=\pi)
f_{\Pi\mid\mathbf{X}}(\pi\mid\mathbf{x})
\,\mathrm{d}\pi.
$$

The probability mass inside the integral describes a possible future response at a fixed $\pi$. The posterior density supplies the weight assigned to each value of $\pi$ after observing $\mathbf{x}$.

For one future success,

$$
\begin{aligned}
\mathbb{P}(\widetilde{X}=1\mid\mathbf{X}=\mathbf{x})
&=\int \pi f_{\Pi\mid\mathbf{X}}(\pi\mid\mathbf{x})\,\mathrm{d}\pi\\
&=\mathbb{E}[\Pi\mid\mathbf{X}=\mathbf{x}]\\
&=\frac{18}{18+6}\\
&=.75.
\end{aligned}
$$

Though the prediction equals the posterior mean for one Bernoulli response, this equality does not make the posterior and predictive distributions identical. The posterior is continuous on $(0,1)$, while the future response is either zero or one.

## Predicting a future block

Now let $\widetilde{K}$ be the number of verb-phrase attachment responses in a future block of $m=20$ matched items from the same participant. Conditional on $\pi$,

$$
\widetilde{K}\mid\Pi=\pi
\sim
\operatorname{Binomial}(20,\pi).
$$

The posterior predictive distribution is beta-binomial because it averages the binomial probability over a beta posterior. We can sample from it in two steps:

1. Draw $\pi^{(s)}\sim\operatorname{Beta}(18,6)$.
2. Draw $\widetilde{k}^{(s)}\sim\operatorname{Binomial}(20,\pi^{(s)})$.

```{r}
#| label: posterior-predictive-parse-count
#| echo: true

set.seed(722)
draw_count <- 100000
pi_draw <- rbeta(draw_count, shape1 = 18, shape2 = 6)
future_count <- rbinom(draw_count, size = 20, prob = pi_draw)

predictive_mean <- 20 * 18 / (18 + 6)
simulated_mean <- mean(future_count)

stopifnot(abs(simulated_mean - predictive_mean) < .05)
c(
  analytic_mean = predictive_mean,
  simulated_mean = simulated_mean,
  quantile(future_count, c(.025, .5, .975))
)
```

The analytic predictive mean is

$$
\mathbb{E}[\widetilde{K}\mid\mathbf{x}]
=20(.75)
=15.
$$

The posterior predictive distribution is more dispersed than a binomial distribution that fixes $\pi=.75$ because it mixes over posterior uncertainty about $\Pi$.

For this beta-binomial distribution, the predictive variance is

$$
\operatorname{Var}(\widetilde{K}\mid\mathbf{x})
=
20
\frac{18(6)}{(18+6)^2}
\frac{18+6+20}{18+6+1}
=6.60.
$$

If we fix $\pi=.75$, the binomial variance is only

$$
20(.75)(.25)=3.75.
$$

The difference between these two variances is $2.85$. To understand that difference correctly, we next separate the posterior predictive variance into its two components.

```{r}
#| label: posterior-predictive-variance
#| echo: true

analytic_variance <- 20 * 18 * 6 * (18 + 6 + 20) /
  ((18 + 6)^2 * (18 + 6 + 1))
plug_in_variance <- 20 * .75 * .25
posterior_variance_pi <- 18 * 6 / ((18 + 6)^2 * (18 + 6 + 1))
mean_conditional_variance <-
  20 * (.75 * .25 - posterior_variance_pi)
variance_conditional_mean <- 20^2 * posterior_variance_pi

stopifnot(abs(analytic_variance - 6.60) < 1e-12)
stopifnot(abs(var(future_count) - analytic_variance) < .10)
stopifnot(abs(mean_conditional_variance - 3.60) < 1e-12)
stopifnot(abs(variance_conditional_mean - 3.00) < 1e-12)
c(full_prediction = analytic_variance,
  fixed_parameter = plug_in_variance,
  mean_conditional_variance = mean_conditional_variance,
  variance_conditional_mean = variance_conditional_mean)
```

## Separating two sources of predictive spread

The law of total variance gives

$$
\operatorname{Var}(\widetilde{K}\mid\mathbf{x})
=
\mathbb{E}\left[
\operatorname{Var}(\widetilde{K}\mid\Pi,\mathbf{x})
\mid\mathbf{x}
\right]
+
\operatorname{Var}\left(
\mathbb{E}[\widetilde{K}\mid\Pi,\mathbf{x}]
\mid\mathbf{x}
\right).
$$

The first term averages future response variation over the posterior, while the second measures how much the conditional mean $20\Pi$ varies across the posterior. In the current example, the two terms are

$$
\begin{aligned}
\mathbb{E}\left[20\Pi(1-\Pi)\mid\mathbf{x}\right]
&=3.60,\\
\operatorname{Var}(20\Pi\mid\mathbf{x})
&=3.00.
\end{aligned}
$$

Their sum is $6.60$. The plug-in variance $20(.75)(.25)=3.75$ is not exactly the first component: evaluating the concave function $20\pi(1-\pi)$ at the posterior mean gives a slightly larger value than averaging it over the posterior. More importantly, the plug-in distribution omits the second component entirely. It is thus too narrow in this example.

## Keeping prediction and checking separate

A posterior predictive distribution says what the fitted model predicts after observing the data. It does not establish that those predictions resemble relevant features of the observed data. [Posterior predictive checking](posterior-predictive-checks.qmd) later compares simulated and observed summaries to assess model adequacy.

A future observation must also match the population and measurement process represented by the observation model. Prediction for a new trial from the same participant differs from prediction for a new participant when participants vary. The symbol $\widetilde{X}$ should be defined with the same care as the original response.

We use a tilde for a genuinely future response. Later, $\mathbf{X}^{\mathrm{rep}}$ will denote a replicated dataset generated under the design of the observed $\mathbf{x}$ for a posterior predictive check. Both use the posterior predictive distribution, but they answer different questions.

Monte Carlo error is separate again: using more simulated draws makes a computed predictive probability more stable, but it does not reduce posterior uncertainty or future response variation.

## Check your understanding

1. What is the support of $\Pi\mid\mathbf{x}$?
2. What is the support of $\widetilde{K}$ for a block of 20 responses?
3. Which source of predictive variance is lost by setting $\pi=.75$ for every future block?
4. Why does increasing the number of simulation draws not make future responses less variable?

This beta-binomial prediction is available algebraically because the posterior is conjugate. The [next page](beyond-conjugacy.qmd) asks what remains possible when the posterior does not belong to a familiar family.
