---
title: "Penalizing in-sample fit"
---

The [likelihood-ratio page](likelihood-ratio-tests.qmd) compared models connected by a specific parameter restriction. Information criteria address a different comparison by combining fitted likelihood with a penalty for estimated flexibility.

Suppose two models predict log reading time. One uses word length and lexical frequency. The other adds several interactions. The larger model will usually achieve at least as much maximized likelihood on the observations used to fit it, even when the added terms mostly follow sample-specific variation.

This difference is the [**optimism of training error**](https://www.statlearning.com/s/ISLRSeventhPrinting.pdf): performance on the observations used to fit a procedure tends to be better than its expected performance on new observations. An [**information criterion**](https://www.statlearning.com/s/ISLRSeventhPrinting.pdf) estimates a correction by combining fitted likelihood with a complexity penalty. The resulting score estimates a model's relative predictive discrepancy under a particular approximation. It is not a significance test, and its absolute value is not a measure with a standalone linguistic interpretation.

The [**Akaike information criterion (AIC)**](https://www.statlearning.com/s/ISLRSeventhPrinting.pdf) provides the principal case below; a later comparison distinguishes it from the [**Bayesian information criterion (BIC)**](https://www.statlearning.com/s/ISLRSeventhPrinting.pdf).

## Separating fit from penalty

Let $\ell$ be a model's maximized log likelihood and let $k$ be its number of estimated parameters. AIC is

$$
\operatorname{AIC}=2k-2\ell.
$$

The expression has two parts:

$$
\underbrace{-2\ell}_{\text{lack of fitted likelihood}}
+
\underbrace{2k}_{\text{complexity penalty}}.
$$

A larger maximized log likelihood makes $-2\ell$ smaller. Adding parameters makes $2k$ larger. Lower AIC is preferred because the criterion places good fitted likelihood and lower estimated optimism on the same direction of the scale.

The coefficient 2 in the penalty is not an arbitrary classroom convention. Under the assumptions motivating AIC, $2k$ approximates how optimistic the in-sample fit is as an estimate of performance on new observations from the represented process. The approximation concerns the complete candidate model and its maximum-likelihood fit.

## Working through two candidate models

Suppose model A has

$$
k_A=3
\qquad\text{and}\qquad
\ell_A=-210.
$$

Its AIC is

$$
\begin{aligned}
\operatorname{AIC}_A
&=2(3)-2(-210)\\
&=6+420\\
&=426.
\end{aligned}
$$

Now suppose model B has five parameters and a maximized log likelihood of $-207$:

$$
\begin{aligned}
\operatorname{AIC}_B
&=2(5)-2(-207)\\
&=10+414\\
&=424.
\end{aligned}
$$

The additional parameters improve log likelihood by 3. That improvement reduces the lack-of-fit term by 6, while the two additional parameters increase the penalty by 4. The improvement is thus larger than the added penalty, and model B has the lower AIC by 2.

Only this difference matters for the comparison:

$$
\Delta_A
=
\operatorname{AIC}_A-\min_m\operatorname{AIC}_m
=
426-424
=
2.
$$

The absolute score 424 is not an error measured in log milliseconds or words. It becomes informative only relative to AIC values computed from compatible candidate models.

## Calculating the score in R

If the maximized log likelihood and parameter count are already available, the formula is direct.

```{r}
#| eval: false

candidate <- data.frame(
  model = c("A", "B"),
  parameters = c(3, 5),
  log_likelihood = c(-210, -207)
)

candidate$aic <-
  2 * candidate$parameters -
  2 * candidate$log_likelihood

candidate$delta_aic <- candidate$aic - min(candidate$aic)
candidate
```

For fitted R models, `AIC()` extracts the same ingredients from the model objects.

```{r}
#| eval: false

AIC(fit_length, fit_surprisal)
```

This convenient call does not check the scientific comparability of the models. It is the analyst's responsibility to verify that both objects represent the same response observations and compatible likelihoods.

## Keeping the response and observations fixed

Information criteria compare models only when their likelihoods refer to the same observed responses. Suppose an ordinary linear regression is fit to raw reading time and another is fit to log reading time. The two likelihoods assign density to different response variables. Their AIC values are not directly comparable without accounting for the transformation.

The same problem arises when candidate models use different rows. If the richer model contains a predictor with missing values, its fitting function may silently drop more observations. Its log likelihood then summarizes a different dataset. An AIC difference would mix model specification with a change in observations.

A useful check is to compare the row identifiers used in every fit, not merely their reported sample sizes. Equal counts do not guarantee that the same tokens were included.

```{r}
#| eval: false

identical(
  rownames(model.frame(fit_a)),
  rownames(model.frame(fit_b))
)
```

This is a mechanical check for one necessary condition. It does not establish that the likelihood families are compatible.

## Comparing AIC with BIC

The BIC uses

$$
\operatorname{BIC}=k\log n-2\ell,
$$

where $n$ is the sample size supplied to the approximation. Its parameter penalty grows with $\log n$. Thus, AIC and BIC may prefer different models, particularly in larger datasets.

BIC is not merely AIC with a stricter penalty. The criteria arise from different motivations and approximations. AIC is motivated by relative expected predictive discrepancy. BIC is motivated by an approximation connected to model evidence under additional assumptions.

Repeated linguistic data create a further difficulty for BIC because the meaning of $n$ may be delicate. Ten thousand trials from 50 speakers do not supply 10,000 independent speakers. Treating every trial count as an unproblematic sample size can make the penalty look more decisive than the design warrants.

For current purposes, the main lesson is modest: report which criterion was used and why it matches the comparison. Do not describe BIC as a universally improved version of AIC.

## Understanding what AIC does not encode

AIC approximates relative expected performance for the observation-level likelihood under its regularity conditions. It does not create held-out predictions. It also does not know which unit a linguistic claim requires to be new.

Suppose a reading-time model contains repeated observations from the same passages. AIC cannot distinguish another position in an observed passage from a position in a passage omitted during fitting. Grouped cross-validation can encode that distinction by changing the separation between training and test data.

This difference does not make AIC useless. It makes the scope of the score narrower. AIC can compare a fixed collection of compatible candidate likelihood models without repeatedly refitting them. Direct validation is preferable when prediction for a group absent during fitting is central to the scientific claim.

## Avoiding threshold interpretation

An AIC difference of 2 is not a conventional significance threshold. The difference says that, among these candidates and under the AIC approximation, model B has a criterion value 2 units lower than model A. It does not supply a null tail probability.

The magnitude must also be interpreted alongside the candidate set and stability of the analysis. A small difference may change after a reasonable preprocessing decision. Reporting the component log likelihoods, parameter counts, and criterion differences makes the source of the comparison visible.

Thus, AIC ranks the specified candidates. It does not certify that the best-ranked candidate is an adequate account of the response process.

## Checking the penalized comparison

1. Why can maximized likelihood alone not select among models of different sizes?
2. In the worked comparison, how much does model B improve log likelihood, and how much additional AIC penalty does it receive?
3. What does an AIC difference of 2 describe, and what does it not describe?
4. Why should models fit to raw and log reading time not be compared by their printed AIC values?
5. Why can AIC not distinguish prediction for omitted items from prediction for omitted participants?
6. What additional difficulty does repeated linguistic data create for the $n$ in BIC?

AIC addresses optimism of training error by adding a parameter-count penalty to maximized likelihood. The score is relative, requires compatible responses and observations, and does not encode which groups should be absent during fitting. The [next page](psis-loo.qmd) develops a Bayesian leave-one-out approximation that averages over posterior uncertainty and retains pointwise contributions for diagnostic inspection.
