---
title: "Correlated predictors"
---

The [multiple-regression page](multiple-regression.qmd) interpreted each coefficient through a comparison that holds the other predictors fixed. Word length and lexical frequency often vary together. Very frequent words tend to be short. If both enter a reading-time regression, the model must learn their separate coefficients from words that break this usual association.

The problem is easiest to see with an extreme example.

## Beginning with perfect overlap

Suppose every word in a dataset satisfies

$$
\text{length}=10-\text{frequency}.
$$

The fitted mean is

$$
m
=\beta_0+\beta_F\text{frequency}
+\beta_L\text{length}.
$$

Substitute the exact relation:

$$
\begin{aligned}
m
&=\beta_0+\beta_F\text{frequency}
+\beta_L(10-\text{frequency})\\
&=(\beta_0+10\beta_L)
+(\beta_F-\beta_L)\text{frequency}.
\end{aligned}
$$

Only the combinations $\beta_0+10\beta_L$ and $\beta_F-\beta_L$ are identified. Many separate coefficient triples produce the same fitted means.

With perfect collinearity, unique partial coefficients do not exist.

## Moving to strong but imperfect overlap

Real predictors are rarely exact transformations of one another. But when they are strongly correlated, the data contain few comparisons between words of different length at similar frequency.

The coefficients are then learned from a narrow and possibly unusual subset of predictor combinations. Their sampling uncertainty grows, and small changes in the sample can move the separate estimates substantially.

This predictor overlap is [**collinearity**](https://appliedstatisticsforlinguists.org/bwinter_stats_proofs.pdf#page=119).

Strong predictor association also limits the directly observed comparisons. A partial frequency coefficient requires words that differ in
frequency while remaining similar in length. If almost every frequent word is
short and almost every rare word is long, the sample supplies few such
comparisons.

## Constructing a nearly collinear sample

Consider ten constructed word observations with centered log frequency and
centered length. The two predictors are almost perfectly negatively correlated.

```{r}
#| label: construct-correlated-lexical-predictors
#| echo: true
frequency <- seq(-2, 2, length.out = 10)
centered_length <-
  -1.8 * frequency + c(.2, -.1, .1, -.2, .15, -.15, .2, -.1, .1, -.2)

log_rt <-
  6 - .08 * frequency + .02 * centered_length +
  c(.01, -.005, .004, -.008, .006, -.004, .008, -.006, .005, -.01)

lexical_sample <- data.frame(
  frequency = frequency,
  centered_length = centered_length,
  log_rt = log_rt
)

cor(lexical_sample$frequency, lexical_sample$centered_length)
```

The correlation is about $-.998$. The sample contains almost no independent
movement in frequency and length.

Fit the additive model:

```{r}
#| label: fit-nearly-collinear-model
#| echo: true
fit_collinear <- lm(
  log_rt ~ frequency + centered_length,
  data = lexical_sample
)

coef(fit_collinear)
confint(fit_collinear)
```

The separate slope estimates need not resemble the coefficients used to
construct the responses. Many pairs of frequency and length slopes produce
nearly the same fitted values over this narrow diagonal band of predictor space.

## Perturbing one response

Now increase the final log response by $.03$ and refit:

```{r}
#| label: perturb-collinear-response
#| echo: true
perturbed_sample <- lexical_sample
perturbed_sample$log_rt[10] <- perturbed_sample$log_rt[10] + .03

fit_perturbed <- lm(
  log_rt ~ frequency + centered_length,
  data = perturbed_sample
)

rbind(
  original = coef(fit_collinear),
  perturbed = coef(fit_perturbed)
)
```

The frequency slope changes from about $-.003$ to $-.032$, and the length slope
changes from about $.063$ to $.045$. The one-response perturbation is small on
the response scale, but it moves the allocation of association across the two
predictors.

The fitted values move much less:

```{r}
#| label: compare-collinear-fitted-values
#| echo: true
max(abs(fitted(fit_collinear) - fitted(fit_perturbed)))
```

The largest change is about $.013$. This is the distinction between unstable
attribution and relatively stable interpolation.

## The comparison rather than a threshold

A correlation of $.70$ is not automatically acceptable or unacceptable. Ask which observations identify each theoretical contrast.

If a theory distinguishes exposure from form complexity, we may need both frequency and length even when their separate estimates are uncertain. That uncertainty may be the correct reflection of the data.

Removing length would make the frequency coefficient describe the joint pattern
in which frequency and length vary together. Keeping both asks for a partial
frequency comparison at fixed modeled length. These are different estimands.
Neither choice repairs missing comparison support.

If two columns are alternate noisy measurements of one construct, a measurement model or a prespecified composite may better represent the theory.

Dropping or residualizing a predictor changes the coefficient target. Regularization may stabilize predictions, but it cannot create predictor combinations that were not observed.

## Prediction can remain stable

Different coefficient combinations may make similar predictions within the observed predictor region. Thus collinearity can strongly affect attribution to separate coefficients while having less effect on interpolation.

Extrapolation remains risky because the fitted model has not learned what happens when the correlated predictors take novel combinations.

For instance, the constructed sample contains no word that is simultaneously
high in centered frequency and high in centered length. A prediction in that
corner depends on how the model divided the diagonal association between its two
slopes. Since multiple divisions fit the observed band similarly, the novel
combination can receive very different predictions.

## Reporting what the data distinguish

A useful analysis states the observed predictor relation, shows the support for
the theoretically relevant comparison, and reports uncertainty in the partial
coefficients. A correlation threshold alone cannot perform this argument.

The [next page](variance-inflation-factors.qmd) summarizes the precision loss
with a variance inflation factor. It does not replace the analysis of predictor support because
the numerical factor cannot say whether the scarce comparisons answer the
linguistic question.

::: {.callout-note title="Find the supporting observations"}
1. If length and frequency are negatively correlated, which words are most
   informative about a length coefficient holding frequency fixed?
2. Why can the separate slopes change while fitted values remain similar in the
   observed predictor band?
3. Explain how strong predictor correlation limits the comparisons that directly support a partial coefficient.
4. Why is prediction for a frequent long word especially uncertain in the
   constructed sample?
:::
