---
title: "Design matrices"
---

The [multiple-regression page](multiple-regression.qmd) wrote one linear predictor for each observation. A design matrix collects those observation-specific equations into one matrix equation.

A regression formula is converted into a numeric matrix before the model is fit.
This [**design matrix**](https://appliedstatisticsforlinguists.org/bwinter_stats_proofs.pdf#page=133) places one observation on each row and one coefficient
contribution on each column.

The model matrix is the numeric representation produced from the formula. Inspecting this representation is useful because coefficient meaning comes from the
numbers in the matrix, not from the surface form of a variable name.

Suppose three words have centered log frequencies $-1.2$, $.4$, and $.8$, along
with centered neighborhood densities 3, $-2$, and 1. The additive model is

```{r}
log_rt ~ log_frequency + neighborhood_density
```

Its design matrix is

$$
\mathbf{X}=
\begin{bmatrix}
1 & -1.2 & 3\\
1 & 0.4 & -2\\
1 & 0.8 & 1
\end{bmatrix}.
$$

The first column represents the intercept. The second contains centered frequency
values. The third contains centered neighborhood-density values. All three rows
contain 1 in the first column because every observation receives the intercept.

## The linear predictor is a matrix product

With coefficient vector

$$
\boldsymbol\beta=
\begin{bmatrix}
\beta_0\\\beta_1\\\beta_2
\end{bmatrix},
$$

all linear predictors are

$$
\boldsymbol\eta=\mathbf{X}\boldsymbol\beta.
$$

For the second row,

$$
\eta_2=1\beta_0+.4\beta_1-2\beta_2.
$$

The row shows exactly which coefficients contribute to that observation's fitted value.

Suppose

$$
\boldsymbol\beta
=
\begin{bmatrix}
6.2\\
-.08\\
.015
\end{bmatrix}.
$$

Then the second fitted conditional mean is

$$
\begin{aligned}
\eta_2
&=1(6.2)+.4(-.08)-2(.015)\\
&=6.138.
\end{aligned}
$$

The matrix product repeats this calculation for all three observations at once.

## Multiplying the matrix by hand

Matrix multiplication takes the dot product of each design row with the
coefficient vector:

$$
\begin{aligned}
\eta_1&=6.2+(-1.2)(-.08)+3(.015)=6.341,\\
\eta_2&=6.2+.4(-.08)-2(.015)=6.138,\\
\eta_3&=6.2+.8(-.08)+1(.015)=6.151.
\end{aligned}
$$

Thus

$$
\boldsymbol\eta
=
\begin{bmatrix}
6.341\\
6.138\\
6.151
\end{bmatrix}.
$$

These are three conditional means. The design matrix does not contain the
responses, and the multiplication does not add residual variation.

## Constructing the matrix in R

```{r}
#| label: construct-numeric-design-matrix
#| echo: true
lexical_rows <- data.frame(
  log_frequency = c(-1.2, .4, .8),
  neighborhood_density = c(3, -2, 1)
)

X <- model.matrix(
  ~ log_frequency + neighborhood_density,
  data = lexical_rows
)

beta <- c(6.2, -.08, .015)

X
as.vector(X %*% beta)
```

The first command exposes the model matrix. The second verifies the three hand
calculations. `%*%` is R's matrix-multiplication operator.

The column names provide another check:

```{r}
#| label: inspect-design-column-names
#| echo: true
colnames(X)
```

Their order must match the order of the coefficients in `beta`. Swapping the two
slope values without swapping the columns would calculate a different model.

## Interpreting columns through row changes

To read a coefficient, compare two hypothetical rows that differ in one column.
For instance,

$$
\begin{bmatrix}1&0&0\end{bmatrix}
$$

and

$$
\begin{bmatrix}1&1&0\end{bmatrix}
$$

differ only by one frequency unit. Their fitted values differ by $\beta_1$. The
same logic gives $\beta_2$ when only the third column changes by one unit.

This row-comparison method generalizes beyond numeric predictors. Later pages
show how categorical comparisons and products become columns, but their
coefficients are still read by comparing design rows.

## Keeping rows and variables separate

The matrix has one row per modeled observation, not one row per distinct word
type unless the dataset itself has one observation per word type. Repeated
tokens generate repeated rows, possibly with identical predictor values but
different responses.

Similarly, one original variable need not always become one matrix column.
The formula and contrast coding determine the actual columns. In this example, each numeric predictor supplies
one column, so the correspondence happens to be one-to-one.

One formula variable does not always produce one model-matrix column. That correspondence holds for these two numeric predictors
but will fail for categorical predictors. Inspecting `model.matrix()` prevents
the assumption from being carried into settings where it is false.

::: {.callout-note title="Reconstruct the matrix product"}
1. Calculate the fitted value for the first row using the displayed coefficient
   vector.
2. What changes in a design row when centered frequency increases by one unit
   and centered neighborhood density stays fixed?
3. Why does the design matrix not contain the response?
4. Explain why one formula variable does not always produce one model-matrix column and why that correspondence holds only for the
   current example.
:::
