---
title: "Deviance residuals"
---

The [deviance page](deviance.qmd) defined total deviance relative to a saturated model, while the earlier [residual page](residuals.qmd) defined observation-level discrepancies. [**Deviance residuals**](https://appliedstatisticsforlinguists.org/bwinter_stats_proofs.pdf#page=214) divide total deviance into signed observation-level contributions.

Let $d_i$ be observation $i$'s contribution to the deviance. The deviance residual is

$$
r_i^{D}
=
\operatorname{sign}(y_i-\widehat\mu_i)
\sqrt{d_i}.
$$

The sign tells us whether the observation lies above or below its fitted mean. The magnitude tells us how much it contributes to the total deviance.

This residual is scaled by the fitted response family because $d_i$ is calculated from the likelihood for
the declared response family.

## Why raw residuals are not enough

A raw residual is

$$
y_i-\widehat\mu_i.
$$

Under a Poisson model, a difference of three counts has a different meaning when the fitted count is one than when it is one hundred. Deviance residuals account for the response distribution when measuring mismatch.

For a Poisson response, the contribution is

$$
d_i
=2\left[
y_i\log\frac{y_i}{\widehat\mu_i}
-(y_i-\widehat\mu_i)
\right],
$$

with the first term treated as zero when $y_i=0$.

Suppose $y_i=4$ and $\widehat\mu_i=1$. The raw residual is 3, and

$$
d_i
=2[4\log4-3]
\approx5.090.
$$

The deviance residual is

$$
r_i^D=\sqrt{5.090}\approx2.256.
$$

Now suppose $y_i=103$ and $\widehat\mu_i=100$. The raw residual is again 3, but

$$
d_i
=2\left[103\log\frac{103}{100}-3\right]
\approx.089.
$$

The deviance residual is only about $.299$. A difference of three counts is much
more surprising relative to a fitted mean of 1 than a fitted mean of 100 under
the Poisson mean-variance relation.

```{r}
#| label: compare-poisson-deviance-residuals
#| echo: true
y <- c(4, 103)
mu <- c(1, 100)

contribution <- 2 * (y * log(y / mu) - (y - mu))
deviance_residual <- sign(y - mu) * sqrt(contribution)

data.frame(
  y = y,
  fitted_mean = mu,
  raw_residual = y - mu,
  deviance_residual = deviance_residual
)
```

## Use residuals to locate model failure

Plot deviance residuals against fitted values and theoretically relevant predictors. Also summarize them by speaker, item, lexical class, or other repeated units.

```{r}
#| label: plot-constructed-deviance-residuals
#| echo: true
count_data <- data.frame(
  predictor = 0:5,
  count = c(0, 1, 1, 3, 7, 11)
)

fit <- glm(count ~ predictor, family = poisson(), data = count_data)
dev_resid <- residuals(fit, type = "deviance")

plot(fitted(fit), dev_resid)
abline(h = 0, lty = 2)
```

A curved center suggests a missing mean relation. A changing spread suggests that the family variance relation is inadequate. A cluster of large residuals in one group suggests omitted group structure.

The plot should be connected to a specific discrepancy. If large positive FSRs
occur mainly in phrase-final positions, the model may underpredict counts there.
If their spread grows with the fitted mean, the Poisson variance relation may be
too restrictive.

## Do not expect perfect normality

Deviance residuals are constructed to be more symmetric than raw residuals, but they are not automatically independent standard normal observations. This is especially visible with binary data, where each response can take only two values.

Thus, a normality test on deviance residuals is not a general goodness of fit test for a generalized linear model. Use checks that reproduce the actual response and the discrepancy relevant to the scientific claim.

Using residual normality as a universal GLM approval criterion imports a familiar ordinary-linear-regression diagnostic
without asking whether it follows from the fitted discrete response model.

## Connect local and global checks

The squared deviance residuals sum to the model deviance:

$$
\sum_i (r_i^D)^2=D.
$$

The total says how much discrepancy there is under this measure. The residual plot says where that discrepancy occurs. Both are needed for a useful diagnosis.

Squaring removes the signs. Two models can thus have similar total
deviance while placing their positive and negative discrepancies in different
linguistic conditions. The observation-level plot retains information that the
sum discards.

The signed square root of a deviance contribution remains a model diagnostic, not a deletion rule. A large residual may
identify miscoding, an omitted predictor, a rare but valid observation, or an
inadequate family. Its source must be investigated.

::: {.callout-note title="Interpret a deviance residual"}
1. What determines the sign of a deviance residual?
2. Why can the same raw count residual have different importance at different fitted means?
3. Why is a normality test on deviance residuals not a general model check?
4. Which information is lost when signed residuals are squared and summed?
5. Explain why residual normality is not a universal approval criterion for generalized linear models.
:::
