---
title: "Deviance"
---

The earlier [maximum-likelihood page](../statistical-inference/maximum-likelihood.qmd) defined the likelihood as a function of the parameter with the observed data held fixed. For ordinary linear regression, residual sums of squares provide a natural measure of mismatch. Generalized linear models use [**deviance**](https://appliedstatisticsforlinguists.org/bwinter_stats_proofs.pdf#page=214) to compare a fitted model with a model that reproduces the observed responses as closely as the response family allows.

This likelihood comparison defines the model deviance. The saturated model is not a theoretically plausible model. It is a likelihood benchmark
that gives each observation enough freedom to attain the best family-specific
fit.

## The saturated model

A saturated model has enough parameters to fit every observation's response. Its likelihood is $L_{\mathrm{sat}}$. The fitted candidate model has likelihood $L_{\mathrm{fit}}$.

The deviance is

$$
D
=
2\left[
\log L_{\mathrm{sat}}
-
\log L_{\mathrm{fit}}
\right].
$$

The saturated model cannot fit worse than the candidate. The deviance is thus nonnegative. Smaller values indicate that the candidate is closer to the saturated fit.

The factor 2 places likelihood differences on the conventional deviance scale.
A deviance of zero means that the candidate attains the saturated likelihood. It
does not mean that the candidate is the true data-generating model.

## Calculating Bernoulli deviance

Suppose five binary responses contain three successes and two failures, and an
intercept-only model fits probability

$$
\widehat\pi=.6.
$$

The fitted log likelihood is

$$
\begin{aligned}
\log L_{\text{fit}}
&=3\log(.6)+2\log(.4)\\
&\approx-3.365.
\end{aligned}
$$

The saturated Bernoulli model assigns fitted probability 1 to each observed
success and 0 to each observed failure. Its log likelihood is 0 because every
observed response receives probability 1. Thus

$$
\begin{aligned}
D
&=2[0-(-3.365)]\\
&\approx6.730.
\end{aligned}
$$

```{r}
#| label: calculate-intercept-only-bernoulli-deviance
#| echo: true
y <- c(1, 1, 0, 1, 0)
pi_hat <- mean(y)
log_lik_fit <- sum(dbinom(y, size = 1, prob = pi_hat, log = TRUE))
deviance_by_hand <- 2 * (0 - log_lik_fit)

fit <- glm(y ~ 1, family = binomial())

c(
  by_hand = deviance_by_hand,
  from_model = deviance(fit)
)
```

The two values agree. The calculation compares the fitted common probability
with observation-specific saturated probabilities.

## Residual deviance is not meaningful alone

A deviance value has no universal good cutoff. Its scale depends on the family, sample size, and observed responses. Use it through comparisons, dispersion checks, and residual contributions.

For instance, doubling a dataset by duplicating every observation can roughly
double a deviance while leaving fitted response proportions unchanged. A raw
value cannot be interpreted independently of the number and form of
observations.

One decontextualized cutoff cannot interpret every deviance value. Deviance becomes informative through a declared model comparison or a family-specific diagnostic.

## Comparing nested models

Suppose model 0 is nested within model 1. The difference in deviance is

$$
D_0-D_1
=
2\left[
\log L_1-\log L_0
\right].
$$

This is the likelihood ratio statistic. Under regularity conditions and a null hypothesis that the smaller model is correct, it has an approximate chi squared distribution with degrees of freedom equal to the difference in parameter counts.

The approximation can fail at parameter boundaries, with small samples, or for irregular mixture models. The test also answers a fit question. It does not directly estimate held out prediction.

The models must be nested for this subtraction to have the usual likelihood
ratio interpretation. Model 0 is nested in model 1 when a constraint on the
larger model's parameters reproduces the smaller model. Setting one added slope
to zero is a common case.

If the two models use different response families or different observations,
their saturated references may not be comparable. Subtracting the displayed
deviances can then answer no clean nested-model question.

## Deviance and overdispersion

For a Poisson model, a residual deviance much larger than its residual degrees of freedom can be evidence that the conditional variation exceeds the Poisson mean variance relation. This ratio is a diagnostic, not a proof of one particular cause. Omitted grouping, excess zeros, and a negative binomial response can all produce related patterns.

The numerator and denominator should not be treated as an exact test without
checking the setting. A large ratio says that the fitted Poisson model leaves
more likelihood discrepancy than its parameter count would lead us to expect
under a rough reference. It does not identify which alternative model will fix
the discrepancy.

## In R

```{r}
#| label: retrieve-deviance
#| echo: true
deviance(fit)
df.residual(fit)

# For two fitted nested models on identical rows:
# anova(smaller_fit, larger_fit, test = "Chisq")
```

The models must be fit to the same observations and with comparable likelihoods. A difference caused by different missing value deletion is not a clean nested model comparison.

Thus a comparison should begin with a row check. The number and identity of
modeled observations must agree. The commented code records the operation
without assuming that two comparison models exist in this example.

::: {.callout-note title="Interpret deviance relative to the saturated model"}
1. What does the saturated model do?
2. How is a nested model deviance difference related to a likelihood ratio statistic?
3. Why is a large Poisson deviance ratio a clue rather than a diagnosis?
4. Recalculate the Bernoulli deviance for four successes and one failure under
   fitted probability $.8$.
5. Explain why one decontextualized deviance cutoff cannot apply to every model.
:::
