---
title: "Exposure offsets"
---

The [Poisson-regression page](poisson-regression.qmd) interpreted coefficients for counts observed over a common window. When observation windows differ, an exposure offset separates the expected rate from the number of opportunities.

Two corpus documents each contain 10 relative clauses. The first has 1,000 words and the second has 10,000 words.

Their counts are equal, but their rates differ:

$$
10/1000=.010
$$

and


$$
10/10000=.001.
$$

The longer document supplied ten times as many opportunities for a construction to occur.

A count combines how often an event occurs per opportunity with how many
opportunities were observed. A rate separates those quantities.

## Separate rate from exposure

Let $t_i>0$ be the exposure for observation $i$. Let $r_i$ be the expected event rate per exposure unit. Then the expected count is

$$
\lambda_i=t_ir_i.
$$

Take logarithms:

$$
\log\lambda_i
=\log t_i+\log r_i.
$$

Model the log rate with predictors:

$$
\log r_i
=\beta_0+\beta_1x_i.
$$

Substitution gives

$$
\log\lambda_i
=\log t_i+\beta_0+\beta_1x_i.
$$

The term $\log t_i$ is the [**exposure offset**](https://appliedstatisticsforlinguists.org/bwinter_stats_proofs.pdf#page=234).

The offset is known from the observation design. It enters the linear predictor
without an estimated coefficient. The remaining coefficients describe expected
rates after exposure has been accounted for.

## Reconstruct two expected counts

Suppose the expected relative-clause rate is $.002$ per word. For a 1,000-word
document,

$$
\lambda_1=1000(.002)=2.
$$

For a 10,000-word document,

$$
\lambda_2=10000(.002)=20.
$$

The expected counts differ by a factor of 10 even though the expected rates are
identical. The offset lets the Poisson mean represent that difference.

On the log scale,

$$
\log\lambda_2-\log\lambda_1
=\log10000-\log1000
=\log10.
$$

Exponentiating returns the expected-count ratio 10.

## Why its coefficient is fixed at one

The equation assumes that doubling exposure doubles the expected count while leaving the expected rate unchanged:

$$
\lambda_i
=t_i\exp(\beta_0+\beta_1x_i).
$$

The coefficient on $\log t_i$ is fixed at 1 because exposure defines the rate denominator.

If we estimated a free coefficient instead, we would ask whether counts scale proportionally with exposure. That is a different model and a different scientific question.

The fixed coefficient does not say that every document with twice the word count
must contain exactly twice the observed clauses. It says that the expected count
doubles when the modeled rate is held fixed. Poisson variation still produces
different observed counts around that expectation.

Interpreting proportional expected counts as deterministic observed counts is an
error.

## Fit an offset in R

```{r}
#| label: fit-relative-clause-rate-model
#| echo: true
documents <- data.frame(
  genre = factor(c("conversation", "conversation",
                   "news", "news")),
  word_count = c(500, 2000, 1000, 4000),
  relative_clause_count = c(2, 5, 7, 21)
)

fit_rate <- glm(
  relative_clause_count ~ genre + offset(log(word_count)),
  family = poisson(link = "log"),
  data = documents
)

coef(fit_rate)
```

The genre coefficient now compares expected relative-clause rates per word, not raw document counts.

To obtain rates, predict for an exposure of one word:

```{r}
#| label: predict-relative-clause-rates
#| echo: true
genre_rates <- data.frame(
  genre = levels(documents$genre),
  word_count = 1
)

genre_rates$rate_per_word <- predict(
  fit_rate,
  newdata = genre_rates,
  type = "response"
)

genre_rates$rate_per_1000_words <-
  1000 * genre_rates$rate_per_word

genre_rates
```

Setting exposure to one returns the fitted rate per exposure unit. Multiplying
by 1,000 changes only the reporting unit.

## Define the opportunity carefully

Word count is one possible exposure for document-level construction counts. It
assumes every word contributes proportionally to the opportunity for a relative
clause. A theory may instead define opportunities as noun phrases, clauses, or
positions that license a particular dependency.

These denominators answer different questions. A rate per word describes corpus
density. A rate per noun phrase may describe the probability that a noun phrase
is modified by a relative clause. The most convenient denominator is not
automatically the theoretically relevant one.

Substituting an available denominator for the theoretically intended opportunity can change the target rate. The offset can be numerically correct
and still answer the wrong rate question.

## Distinguish an offset from a predictor

A predictor receives an estimated coefficient because the analysis asks how the
rate changes with that variable. An offset receives a fixed coefficient because
it defines how the count scales with known opportunity.

Adding `log(word_count)` as an ordinary predictor would estimate a scaling
exponent:

$$
\log\lambda_i
=\beta_0+\gamma\log t_i+\beta_1x_i.
$$

The offset model fixes $\gamma=1$. Estimating $\gamma$ asks whether expected
counts scale sublinearly or superlinearly with document length. That may be a
useful hypothesis, but it no longer defines a simple rate per word.

Exposure units must be meaningful and reported. Rates per word, minute, speaker turn, or opportunity defined by a syntactic context answer different questions.

Zero exposure is not allowed because $\log0$ is undefined and no event can occur
without an opportunity. Such rows usually indicate that the observation unit or
denominator needs to be reconsidered.

::: {.callout-note title="Check proportional exposure"}
1. If two observations have the same predictors but one has three times the
   exposure, what ratio of expected counts does the offset model impose?
2. Why does this claim concern expected counts rather than observed counts?
3. Distinguish an exposure offset from an estimated exposure predictor.
4. Explain how an available denominator can define the wrong target rate.
:::
