---
title: "Fitting a linear regression"
---

The [straight-line mean](simple-linear-regression.qmd) specifies what the coefficients mean, while the earlier derivation of [maximum likelihood](../statistical-inference/maximum-likelihood.qmd#finding-the-maximum) supplies a general fitting principle. With normally distributed errors of constant variance, [**least squares**](https://appliedstatisticsforlinguists.org/bwinter_stats_proofs.pdf#page=85) and maximum likelihood select the same coefficient estimates.

Consider six constructed word observations:

| surprisal | log reading time |
|---:|---:|
| 2 | 5.78 |
| 3 | 5.82 |
| 4 | 5.93 |
| 5 | 5.99 |
| 6 | 6.05 |
| 7 | 6.13 |

We want values of $\beta_0$ and $\beta_1$ that place a line near these points.
The mean function tells us what a candidate line means, but it does not yet tell
us which candidate to choose. We need a rule that compares whole lines against
the observed responses.

## Defining each fitted value

For observation $i$, a candidate line gives

$$
\widehat{y}_i=\beta_0+\beta_1x_i.
$$

The vertical discrepancy between the observed and fitted response is

$$
y_i-\widehat{y}_i
=y_i-(\beta_0+\beta_1x_i).
$$

## Adding squared discrepancies

Least squares chooses the line that minimizes

$$
\operatorname{SSE}(\beta_0,\beta_1)
=\sum_{i=1}^{n}
\left[y_i-(\beta_0+\beta_1x_i)\right]^2.
$$

Squaring prevents positive and negative discrepancies from canceling and gives greater weight to larger misses.

For instance, residuals of $.2$ and $-.2$ sum to zero even though neither
observation lies on the line. Their squared values sum to $.08$. The squared
error objective thus records both misses without allowing one direction to erase
the other.

The objective evaluates a complete candidate line. Moving the intercept or slope
changes several fitted values at once, so the coefficient estimates must be
chosen jointly.

When errors are normally distributed with constant variance, the coefficient values that minimize this sum also maximize the likelihood.

To see the connection, the normal log likelihood for fixed $\sigma$ can be
written as

$$
\ell(\beta_0,\beta_1,\sigma)
=C-n\log\sigma
-\frac{1}{2\sigma^2}
\sum_{i=1}^{n}
\left[y_i-(\beta_0+\beta_1x_i)\right]^2,
$$

where $C$ does not depend on the coefficients. For a fixed positive value of
$\sigma$, increasing the squared-error sum lowers the log likelihood. Thus the
coefficient values with the smallest sum also have the largest normal
likelihood.

This equivalence does not say that squared error is a universally appropriate
fitting rule. It follows from the normal likelihood. A binary response or
a count response requires a different probability model, and later pages derive
the fitting objectives implied by those models.

## Deriving the two coefficient estimates

The least-squares slope for one predictor is

$$
\widehat\beta_1
=\frac{
\sum_{i=1}^{n}(x_i-\bar x)(y_i-\bar y)
}{
\sum_{i=1}^{n}(x_i-\bar x)^2
}.
$$

The numerator measures how predictor and response deviations move together. The
denominator measures the spread of the predictor. Their ratio gives response
change per predictor unit.

This expression also explains why the slope is not obtained by averaging the
ratios $y_i/x_i$. Both variables must first be expressed relative to their own
means. Otherwise the calculation confounds the line's vertical location with
its change across the predictor.

For the six observations,

$$
\bar x=4.5,
\qquad
\bar y=5.95,
$$

$$
\sum_i(x_i-\bar x)(y_i-\bar y)=1.25,
$$

and

$$
\sum_i(x_i-\bar x)^2=17.5.
$$

Thus

$$
\widehat\beta_1
=\frac{1.25}{17.5}
\approx.0714.
$$

The fitted line passes through the point $(\bar x,\bar y)$, so the intercept is

$$
\begin{aligned}
\widehat\beta_0
&=\bar y-\widehat\beta_1\bar x\\
&=5.95-.0714(4.5)\\
&\approx5.629.
\end{aligned}
$$

Thus the fitted conditional mean is

$$
\widehat m(x)=5.629+.0714x.
$$

## Fitting the line in R

```{r}
#| label: fit-constructed-reading-times
#| echo: true
reading_times <- data.frame(
  surprisal = 2:7,
  log_rt = c(5.78, 5.82, 5.93, 5.99, 6.05, 6.13)
)

fit <- lm(log_rt ~ surprisal, data = reading_times)
coef(fit)
```

The formula has three parts:

| syntax | role |
|---|---|
| `log_rt` | response |
| `~` | “is modeled as a function of” |
| `surprisal` | predictor |

The fitted coefficients define

$$
\widehat{m}(x)
=\widehat{\beta}_0+\widehat{\beta}_1x.
$$

For this fit, the output should return an intercept near $5.629$ and a slope near
$.0714$. The slope says that the fitted conditional mean increases by about
$.0714$ log-time units for each one-unit increase in surprisal. It does not say
that every observed response increases by that amount, since individual
responses vary around the line.

The software result is useful as a check on the hand derivation. But the hand
derivation supplies the interpretation: `lm()` has selected the same line that
minimizes the squared residuals.

## Obtaining fitted conditional means

```{r}
#| label: fitted-reading-time-means
#| echo: true
reading_times$fitted <- fitted(fit)
reading_times
```

Each fitted value is the model's estimated conditional mean at that observation's surprisal. It is not a claim that the individual response must equal the line.

## Calculating the residuals

The residual for observation $i$ is the observed response minus its fitted
conditional mean:

$$
e_i=y_i-\widehat m(x_i).
$$

The first word has fitted mean

$$
5.629+.0714(2)\approx5.7718,
$$

so its residual is approximately

$$
5.78-5.7718=.0082.
$$

A positive residual places the observed response above the line. A negative
residual places it below the line. With an intercept in the fitted model, the
residuals sum to zero up to numerical precision. This zero sum is a consequence
of the fitting rule, not evidence that the model is adequate.

```{r}
#| label: check-linear-residual-sum
#| echo: true
sum(residuals(fit))
sum(residuals(fit)^2)
```

The second value is the minimized squared-error sum. Any other intercept and
slope produce a sum that is at least as large for these observations.

## Separating fitting from evaluation

The fitted line is the best line under the declared squared-error rule. This is
a narrow claim. It does not establish that a line is a good representation of
the conditional mean, that the conditional response variance is constant, or that the
observations are conditionally independent.

The best-fitting line within a model family is not necessarily an adequate model. Optimization answers which member of the chosen model
family fits best. Adequacy asks whether that family captures the patterns needed
for the linguistic question. Residual checks later provide evidence about the
second question.

The coefficient estimates summarize this one sample. The [next page](linear-regression-uncertainty.qmd) studies how such estimates vary across possible samples.

::: {.callout-note title="Check the objective"}
1. If a candidate line produces residuals $-.1$, $.2$, and $-.3$, calculate its
   sum of squared errors for those three observations.
2. Why would the unsquared residual sum be a poor fitting objective?
3. Use the fitted equation to calculate the conditional mean at 6 bits of
   surprisal.
4. Explain the optimization-adequacy substitution in this example.
:::
