---
title: "Changing conditional variance"
---

The [linear-regression checks](checking-linear-regression.qmd) used changes in residual spread to diagnose a failure of the constant-variance assumption. The standard statistical term for such a change in conditional variance is [**heteroscedasticity**](https://appliedstatisticsforlinguists.org/bwinter_stats_proofs.pdf#page=85).

Suppose a phonetic model predicts vowel duration from local articulation rate.
Durations at slow rates may cluster tightly, while durations at fast rates may
vary more because speakers use different reduction strategies.

The ordinary linear regression model presented earlier assumes

$$
Y_i\mid x_i
\sim
\mathcal{N}(\mu_i,\sigma^2).
$$

The conditional mean $\mu_i$ may change with articulation rate, but the conditional standard deviation $\sigma$ is fixed.

Heteroscedasticity means that the conditional variance changes across observations:

$$
\operatorname{Var}(Y_i\mid x_i)=\sigma_i^2.
$$

The model is homoscedastic because one $\sigma$ describes every predictor value. Heteroscedasticity is a mismatch between
that assumption and a response whose conditional spread changes.

## Separating a mean pattern from a variance pattern

Consider two articulation-rate bands.

| rate band | fitted mean duration | residual SD |
|---|---:|---:|
| slow | 100 ms | 10 ms |
| fast | 70 ms | 30 ms |

The difference between 100 and 70 milliseconds is a mean pattern. The
difference between 10 and 30 milliseconds is a variance pattern. A model can
represent the first while missing the second.

The two distinctions can be reconstructed from six durations. Suppose the slow
group contains 90, 100, and 110 milliseconds, while the fast group contains 40,
70, and 100 milliseconds. Their means are 100 and 70, and their sample standard
deviations are 10 and 30.

```{r}
#| label: separate-vowel-means-and-spreads
#| echo: true
vowels <- data.frame(
  rate = rep(c("slow", "fast"), each = 3),
  duration = c(90, 100, 110, 40, 70, 100)
)

aggregate(
  duration ~ rate,
  data = vowels,
  FUN = function(x) c(mean = mean(x), sd = sd(x))
)
```

Pooling all six responses into one standard deviation would conceal the fact
that the fast group is three times as variable on this scale.

## Inspecting residual spread directly

A residual-versus-fitted plot should be read vertically. At each fitted value, ask how wide the residual cloud is.

```{r}
#| label: plot-vowel-residual-spread
#| echo: true
fit <- lm(duration ~ rate, data = vowels)

plot(
  fitted(fit),
  residuals(fit),
  xlab = "fitted mean",
  ylab = "residual"
)
abline(h = 0, lty = 2)
```

A fan shape suggests that conditional variance changes with the fitted mean. But
fitted values can combine several predictors, so residuals should also be plotted
against theoretically relevant predictors such as articulation rate, boundary
position, or speaker group.

Group summaries provide a second view:

```{r}
#| label: summarize-vowel-residual-spread
#| echo: true
vowels$residual <- residuals(fit)
aggregate(residual ~ rate, data = vowels, FUN = sd)
```

The groups are descriptive. They help locate the pattern but do not define the final variance model.

The grouping choice can create or hide a pattern. Very wide bins mix predictor
values whose spreads differ. Very narrow bins contain too few observations to
estimate spread reliably. A residual plot retains the continuous predictor,
while grouped summaries provide an interpretable second view.

## Understanding the inferential consequence

Ordinary least-squares coefficient estimates can remain centered on the intended linear projection under some forms of heteroscedasticity. Their usual standard errors, however, rely on the constant-variance model and can be wrong.

The direction of the error depends on where response spread and predictor
information occur. If the most extreme predictor values also have the widest
residual distributions, the slope can be less precise than the homoscedastic calculation
suggests. There is no rule that heteroscedasticity always makes every standard
error too small.

The mismatch also matters for prediction. A model with one $\sigma$ may produce intervals that are too wide for common words and too narrow for rare words.

In the vowel example, one pooled $\sigma$ may produce intervals that are too wide
for slow-rate durations and too narrow for fast-rate durations. The conditional
means can be approximately correct while the predictive intervals fail in
opposite directions.

An adequate mean fit does not establish an adequate response distribution. Residual checks must examine both the
center and vertical spread.

Three repairs answer different questions:

1. heteroscedasticity-consistent standard errors adjust coefficient uncertainty while leaving the variance pattern undescribed;
2. a response transformation changes the scale on which variance is modeled;
3. distributional regression models $\sigma_i$ directly.

The next page takes the third approach.

The choice depends on the scientific target. If spread is a nuisance for slope
uncertainty, an uncertainty correction may be enough. If speaker variability at
fast rates is theoretically meaningful, a model for $\sigma_i$ answers the more
direct question.

::: {.callout-note title="Separate center from spread"}
1. Which table column describes a mean pattern, and which describes a variance pattern?
2. Why can one residual plot conceal a speaker-specific variance pattern?
3. Which repair would you choose if variation itself bears on the linguistic claim?
4. Why does heteroscedasticity not always move every standard error in the same
   direction?
5. Explain why an adequate conditional mean does not establish an adequate response distribution.
:::
