---
title: "Influence"
---

The [preceding page](leverage-and-influence.qmd) defined leverage from predictor position alone. But an unusual
position need not change the fitted coefficients. [**Influence**](https://appliedstatisticsforlinguists.org/bwinter_stats_proofs.pdf#page=85) measures the
extent to which a fitted result depends on a particular observation.

Influence asks a case-deletion question: how much would the
fit change if observation $i$ were not used? This question combines the
observation's predictor position with the mismatch between its response and the
fitted conditional mean.

## Comparing two edge observations

Consider predictor values

$$
\mathbf{x}=(2,3,4,5,11).
$$

The point at $x=11$ has high leverage. Now consider two possible responses for
that point.

In dataset A, the responses are

$$
\mathbf{y}_A=(6.1,6.4,6.5,6.8,8.0).
$$

In dataset B, only the last response changes:

$$
\mathbf{y}_B=(6.1,6.4,6.5,6.8,10.5).
$$

The last observation has the same leverage in both datasets because the
predictor values are identical. But it should affect the fitted slope much more
in dataset B because its response lies farther from the relation supported by
the four central observations.

## Asking the deletion question directly

We can fit each dataset with and without the edge observation:

```{r}
#| label: compare-case-deleted-slopes
#| echo: true
dataset_a <- data.frame(
  x = c(2, 3, 4, 5, 11),
  y = c(6.1, 6.4, 6.5, 6.8, 8.0)
)

dataset_b <- transform(
  dataset_a,
  y = c(6.1, 6.4, 6.5, 6.8, 10.5)
)

fit_a <- lm(y ~ x, data = dataset_a)
fit_b <- lm(y ~ x, data = dataset_b)

fit_a_without_edge <- lm(y ~ x, data = dataset_a[-5, ])
fit_b_without_edge <- lm(y ~ x, data = dataset_b[-5, ])

rbind(
  a_full = coef(fit_a),
  a_without_edge = coef(fit_a_without_edge),
  b_full = coef(fit_b),
  b_without_edge = coef(fit_b_without_edge)
)
```

Removing the edge observation changes the dataset B slope much more. This is a
direct answer to the case-deletion question for one coefficient.

## Summarizing case deletion with Cook's distance

Refitting a model after deleting every observation can produce many coefficient
comparisons. Cook's distance summarizes the overall change in fitted values when
observation $i$ is deleted:

$$
D_i
=\frac{e_i^2}{p\widehat\sigma^2}
\frac{h_{ii}}{(1-h_{ii})^2},
$$

where $e_i$ is the residual, $h_{ii}$ is the leverage,
$\widehat\sigma^2$ is the fitted residual variance, and $p$ is the number of
estimated coefficients.

The formula exposes the two ingredients. A larger residual increases influence,
and larger leverage increases influence. Neither ingredient alone guarantees a
large Cook's distance.

```{r}
#| label: compare-cooks-distance
#| echo: true
cbind(
  x = dataset_a$x,
  leverage = hatvalues(fit_a),
  cooks_a = cooks.distance(fit_a),
  cooks_b = cooks.distance(fit_b)
)
```

The last observation has the same leverage in both columns but a different
Cook's distance. Its response determines whether the unusual predictor position
substantially changes the fit.

## Treating a large value as a question

A large Cook's distance identifies an observation that deserves investigation.
It does not decide that the observation is invalid. The useful sequence is:

1. verify the measurement, transcription, and coding;
2. check whether the observation belongs to the population targeted by the study;
3. inspect which coefficient or fitted region changes under deletion; and
4. report whether the theoretical conclusion depends on the observation.

These steps answer different questions. A miscoded value can be corrected. A
valid but unusual token cannot be deleted merely because it changes the result.

Deleting an observation only because it produces an inconvenient effect estimate changes the sample using information about the desired
conclusion. It can make the reported uncertainty much too small.

## Connecting influence to linguistic sampling

Suppose a study of a rare syntactic construction contains one verb with many
tokens in an unusual frame. Row-level Cook's distances ask about individual
tokens. But the theoretical result may depend on the verb as a group. Deleting
one token at a time may then make every value look modest even though deleting
the verb changes the conclusion.

The diagnostic unit should match the sampling claim. If the intended
generalization concerns verbs, a verb-level deletion analysis is informative. If
it concerns speakers, a speaker-level analysis is informative. This principle
does not require a multilevel model, though later models represent these grouped
sources of variation directly.

## Reporting the sensitivity

An influence analysis is most useful when it reports the quantity that changes.
For instance: "Deleting verb V changed the frame coefficient from .42 to .17,
while deleting any other verb changed it by less than .04." This statement lets
the reader evaluate the concentration of evidence.

A statement that "no influential observations were found" is weaker unless the
analysis defines the diagnostic, the unit of deletion, and the consequence that
was examined.

::: {.callout-note title="Answer the case-deletion question"}
1. Why can two observations with the same leverage have different influence?
2. Which two quantities enter Cook's distance?
3. Why does a large Cook's distance not by itself license deletion?
4. If the scientific generalization concerns verbs, why might deleting one row
   at a time answer the wrong question?
:::
