---
title: "Interaction terms"
---

The [multiple-regression page](multiple-regression.qmd) held each predictor association constant across the values of the other predictors. An [**interaction term**](https://appliedstatisticsforlinguists.org/bwinter_stats_proofs.pdf#page=149) relaxes that restriction by allowing one modeled association to depend on another predictor.

Suppose a reading experiment manipulates an aspectual mismatch and measures verb surprisal. The theory predicts that surprisal has a larger reading-time association in mismatching contexts than in matched contexts.

An additive model cannot express this claim because it gives both conditions the same surprisal slope.

The theoretical claim concerns a difference between associations. The interaction term lets the
conditional-mean model estimate that slope difference directly.

## Begin with the additive model

Let $C_i=0$ mark the matched condition and $C_i=1$ the mismatch condition:

$$
\eta_i
=\beta_0+\beta_1X_i+\beta_2C_i.
$$

For $C_i=0$, the surprisal slope is $\beta_1$. For $C_i=1$, the coefficient on $X_i$ is still $\beta_1$.

The condition changes the intercept only.

For instance, suppose $\beta_1=.04$ and $\beta_2=.10$. Increasing surprisal by
one bit raises the fitted mean by $.04$ in both conditions. The mismatch line is
shifted upward by $.10$, but the two lines remain parallel. A vertical shift is
not a slope difference.

## Add a product term

The interaction model is

$$
\eta_i
=\beta_0+\beta_1X_i+\beta_2C_i
+\beta_3X_iC_i.
$$

Substitute $C_i=0$:

$$
\eta_i=\beta_0+\beta_1X_i.
$$

The matched-condition slope is $\beta_1$.

Substitute $C_i=1$:

$$
\eta_i
=(\beta_0+\beta_2)
+(\beta_1+\beta_3)X_i.
$$

The mismatch-condition slope is $\beta_1+\beta_3$.

Thus $\beta_3$ is the difference between the two slopes on the linear-predictor scale.

More explicitly,

$$
\begin{aligned}
\text{slope}_{\text{mismatch}}
-\text{slope}_{\text{matched}}
&=(\beta_1+\beta_3)-\beta_1\\
&=\beta_3.
\end{aligned}
$$

If $\beta_1=.04$ and $\beta_3=.03$, the matched slope is $.04$ and the mismatch
slope is $.07$. The interaction coefficient is $.03$, not $.07$.

## Read the condition difference at two surprisal values

The mismatch-minus-matched difference at surprisal $x$ is

$$
\begin{aligned}
\Delta(x)
&=\eta(x,C=1)-\eta(x,C=0)\\
&=\beta_2+\beta_3x.
\end{aligned}
$$

Thus the condition difference itself changes with surprisal. At $x=0$, the
difference is $\beta_2$. At $x=4$, it is $\beta_2+4\beta_3$.

This makes the reference point for $x$ important. If surprisal is centered at 5
bits, $x=0$ means 5 bits on the original scale. The coefficient $\beta_2$ is then
the condition difference at 5 bits, not at zero original bits.

## Derive the difference in differences

An interaction can also be written as a difference in differences. Compare each
condition at surprisal values $x_a$ and $x_b$:

$$
\begin{aligned}
&[\eta(x_b,1)-\eta(x_a,1)]
-[\eta(x_b,0)-\eta(x_a,0)]\\
&\qquad=\beta_3(x_b-x_a).
\end{aligned}
$$

For a one-bit contrast, this quantity is $\beta_3$. For a four-bit contrast, it
is $4\beta_3$. The comparison should state the predictor interval because the
interaction coefficient is expressed per predictor unit.

## Fit the product in R

```{r}
#| label: fit-constructed-linear-interaction
#| echo: true
reading_times <- expand.grid(
  surprisal_centered = c(-2, 0, 2),
  condition = factor(c("matched", "mismatch"),
                     levels = c("matched", "mismatch"))
)

reading_times$log_rt <- with(
  reading_times,
  6.00 +
    .04 * surprisal_centered +
    .10 * (condition == "mismatch") +
    .03 * surprisal_centered * (condition == "mismatch")
)

fit_interaction <- lm(
  log_rt ~ surprisal_centered * condition,
  data = reading_times
)

coef(fit_interaction)
```

The `*` expands to the two lower-order terms and their product:

```{r}
#| label: inspect-interaction-formula-expansion
#| echo: true
terms(log_rt ~ surprisal_centered * condition)
```

Keeping the lower-order terms makes the coefficients correspond to the lines derived above.

The constructed responses follow the interaction equation exactly, so the fit
recovers coefficients 6.00, $.04$, $.10$, and $.03$ up to numerical precision.
Real observations would vary around these conditional means.

## Calculate response-scale predictions

The fitted lines are easier to check in a prediction table:

```{r}
#| label: predict-two-interacting-lines
#| echo: true
prediction_grid <- expand.grid(
  surprisal_centered = c(-2, 0, 2),
  condition = levels(reading_times$condition)
)

prediction_grid$predicted_log_rt <- predict(
  fit_interaction,
  newdata = prediction_grid
)

prediction_grid
```

At centered surprisal $-2$, the matched prediction is 5.92 and the mismatch
prediction is 5.96, a difference of $.04$. At centered surprisal 2, the matched
prediction is 6.08 and the mismatch prediction is 6.24, a difference of $.16$.
The condition contrast has increased by $.12$ across four surprisal units, which
equals $4(.03)$.

## Keep the lower-order terms

Removing `surprisal_centered` while retaining its product with condition would
force the matched-condition slope to zero. Removing `condition` would force the
two lines to have the same intercept at centered surprisal zero. These are
substantive constraints, not cosmetic simplifications.

Lower-order terms should not be removed merely because their coefficient tests
are not of interest. The product
coefficient is interpreted relative to the two lower-order terms. Removing them
changes the lines and the slope difference being estimated.

## Separate interaction from causation

An interaction states that a modeled association differs across conditions. It
does not by itself show that the condition causes the slope difference. If
condition was experimentally assigned while surprisal was measured, the two
parts of the claim may have different identification support. The coefficient
equation does not supply that design argument.

Thus the interaction should be reported as the comparison it answers: how much
the fitted surprisal slope differs between mismatch and matched contexts, on the
log-reading-time scale, over a declared predictor interval.

::: {.callout-note title="Check the second slope"}
1. If $\beta_1=.04$ and $\beta_3=.03$, what is the surprisal slope in each
   condition? Which coefficient is the slope difference?
2. If $\beta_2=.10$, calculate the mismatch-minus-matched difference at
   centered surprisal values 0 and 3.
3. Express the interaction over a five-bit surprisal contrast as a difference
   in differences.
4. Explain why lower-order terms should ordinarily remain when their interaction is included.
:::
