---
title: "Leverage"
---

The [residual diagnostics](checking-linear-regression.qmd) ask whether responses depart from fitted means in systematic ways. Leverage instead concerns where each observation falls among the predictor values used to fit the line.

Residual size tells us whether a response lies far from its fitted conditional
mean. It does not tell us whether the observation occupies an unusual position
on the predictor. [**Leverage**](https://appliedstatisticsforlinguists.org/bwinter_stats_proofs.pdf#page=85) measures the second property.

This distinction matters because an observation at the edge of the predictor
range can help determine the fitted slope even when its response lies close to
the fitted line. Leverage quantifies this unusual predictor position.

## Comparing positions on one predictor

Consider five words with predictor values

$$
\mathbf{x}=(2,3,4,5,11).
$$

The sample mean is

$$
\bar x=5.
$$

The word at $x=11$ is six units above the mean, while the word at $x=4$ is one
unit below it. The first word occupies a more unusual design position, regardless
of its observed response.

For simple linear regression with an intercept, observation $i$ has leverage

$$
h_{ii}
=\frac{1}{n}
+\frac{(x_i-\bar x)^2}
{\sum_{j=1}^n(x_j-\bar x)^2}.
$$

The first term gives every observation the same baseline leverage. The second
term increases with squared distance from the predictor mean.

For the five predictor values,

$$
\sum_{j=1}^5(x_j-\bar x)^2
=(-3)^2+(-2)^2+(-1)^2+0^2+6^2
=50.
$$

The leverage at $x=4$ is

$$
h
=\frac15+\frac{(-1)^2}{50}
=.22.
$$

The leverage at $x=11$ is

$$
h
=\frac15+\frac{6^2}{50}
=.92.
$$

The edge observation has very high leverage because its predictor value is far
from the center of this design.

## Calculating every leverage value

```{r}
#| label: calculate-simple-leverage
#| echo: true
x <- c(2, 3, 4, 5, 11)
n <- length(x)

leverage_by_hand <-
  1 / n + (x - mean(x))^2 / sum((x - mean(x))^2)

data.frame(x = x, leverage = leverage_by_hand)
sum(leverage_by_hand)
```

With an intercept and one predictor, the leverage values sum to 2. This is the
number of fitted coefficients. Their average is thus $2/n$, which gives a
useful reference for identifying observations far above the design's usual
leverage.

The reference is descriptive, not an automatic deletion rule. A high-leverage
observation may be exactly the observation needed to study the edge of a
linguistically relevant range.

## Retrieving leverage from a fitted model

Leverage depends only on the predictor values used to fit the model. We can
attach any response values to verify the software calculation:

```{r}
#| label: retrieve-fitted-leverage
#| echo: true
design_example <- data.frame(
  x = x,
  y = c(6.1, 6.4, 6.5, 6.8, 8.0)
)

fit <- lm(y ~ x, data = design_example)

cbind(
  by_hand = leverage_by_hand,
  from_model = hatvalues(fit)
)
```

The columns agree. If we changed all five responses but kept the predictor
values fixed, the leverage values would remain unchanged.

## Separating leverage from residual size

An observation may have high leverage and a small residual. The point at $x=11$
could lie almost exactly on the fitted line while still anchoring its right end.
In contrast, a point near $x=5$ could have a very unusual response and a large
residual while occupying an ordinary predictor position.

This gives two separate questions:

1. Is the response far from its fitted conditional mean?
2. Is the predictor value far from the center of the observed design?

Residuals answer the first. Leverage answers the second. A large residual and high leverage are thus different properties of an observation.

## Interpreting high leverage in a linguistic design

Suppose the predictor is corpus frequency. A genuinely rare word may have high
leverage after log frequency is entered into a small sample. Its position is not
necessarily a recording error. It may instead reveal that the sample contains
little information near that end of the frequency range.

The first response should be to inspect the design and the observation's
provenance. Is the value coded correctly? Does it belong to the target
population? Is the edge of the range theoretically relevant? Only then should
we ask how strongly it affects the fitted coefficients.

That final question is influence, which the [next page](influence.qmd) introduces
using case deletion.

::: {.callout-note title="Locate the unusual design position"}
1. Calculate the leverage for $x=5$ in the five-value example.
2. Why can leverage be calculated without looking at the responses?
3. Construct a case with high leverage and a small residual.
4. Explain why residual magnitude and leverage answer different questions.
:::
