---
title: "Verb type and temporal preposition choice"
---

[Logistic regression](logistic-regression.qmd) combines a Bernoulli response, logit link, and linear predictor. We now apply that model to a contrast represented through [categorical-predictor coding](coding-categorical-predictors.qmd).

The SuperTelicity cloze experiment asked participants to complete event descriptions with temporal phrases. We will focus on whether a completion used *in* or *for* and compare manner verbs with result verbs.

The linguistic question is whether the two verb classes differ in the probability of an *in* completion.

The comparison bears on event structure. An *in* phrase tends to describe time
to a culmination, while a *for* phrase tends to describe an event's duration.
Result verbs and manner verbs differ in how readily their lexical content
supports a culmination. The model estimates the completion difference between the two verb classes.

::: {.callout-note title="Dataset: SuperTelicity cloze responses"}
The [SuperTelicity repository](https://github.com/SuperMereo/supertelicity) contains the cloze responses and records the dataset authorship and Creative Commons Attribution ShareAlike 4.0 license.
This dataset is used here because verb class is a theoretically interpretable
predictor of temporal preposition choice, and the two-category subset supplies a
direct Bernoulli response.
:::

## Loading and representing the response

```{r}
#| label: load-supertelicity-cloze
#| echo: true
cloze <- read.csv("../slides/lecture11/supertelicity/data/data-cloze.csv")
cloze <- subset(cloze, verb_type %in% c("result", "manner") &
                        preposition %in% c("in", "for"))

cloze$in_response <- as.integer(cloze$preposition == "in")
cloze$verb_type <- relevel(factor(cloze$verb_type), ref = "result")
```

One observation represents one participant's cloze response to one sentence. The binary response is 1 for *in* and 0 for *for*.

The filter yields 9,160 responses. The model treats the two selected
prepositions as the complete outcome space for current purposes. Responses with
other prepositions are outside this particular binary question rather than
being recoded as *for*.

## Inspecting the observed response table

```{r}
#| label: summarize-supertelicity-responses
#| echo: true
with(cloze, table(verb_type, in_response))
aggregate(in_response ~ verb_type, data = cloze, FUN = mean)
```

The data contain 3,434 *in* responses and 1,446 *for* responses for result verbs.
For manner verbs, they contain 1,442 *in* responses and 2,838 *for* responses.
The observed *in* proportions are about $.704$ and $.337$.

These proportions show a large verb-class difference before regression. The model expresses that
difference on the log odds scale and supplies a framework that can later be
extended to repeated units.

## Stating the model

Let $X_i=1$ mark a manner verb and $X_i=0$ a result verb:

$$
Y_i\sim\operatorname{Bernoulli}(\pi_i)
$$

and


$$
\operatorname{logit}(\pi_i)
=\beta_0+\beta_1X_i.
$$

The intercept is the log odds of an *in* completion for result verbs. The slope is the manner-minus-result log-odds difference.

## Fitting the model

```{r}
#| label: fit-supertelicity-logistic
#| echo: true
fit_cloze <- glm(
  in_response ~ verb_type,
  family = binomial(link = "logit"),
  data = cloze
)

coef(fit_cloze)
```

The fitted intercept is about $.865$. Applying the inverse logit gives

$$
\operatorname{logit}^{-1}(.865)\approx.704,
$$

which is the fitted *in* probability for result verbs.

The manner coefficient is about $-1.542$. It is the manner-minus-result log odds
difference. Exponentiating gives

$$
e^{-1.542}\approx.214.
$$

Thus the fitted odds of an *in* completion for manner verbs are about $.214$
times the fitted odds for result verbs. This is an odds ratio, not a probability
ratio.

## Returning to probabilities

Create one row for each verb class:

```{r}
#| label: predict-supertelicity-probabilities
#| echo: true
new_cloze <- data.frame(
  verb_type = factor(c("result", "manner"),
                     levels = levels(cloze$verb_type))
)

new_cloze$in_probability <- predict(
  fit_cloze,
  newdata = new_cloze,
  type = "response"
)

new_cloze
```

These fitted probabilities answer the modeled comparison more directly than the raw log-odds coefficients.

The two probabilities should be about $.704$ for result verbs and $.337$ for
manner verbs. Their difference is

$$
.337-.704=-.367.
$$

The one-predictor model is saturated with respect to the two verb-class
proportions, so it reproduces the observed proportion in each class. The
coefficient and probability summaries are two parameterizations of the same
group comparison.

## Stating the licensed linguistic result

Within the selected *in* and *for* responses, result verbs have a substantially
higher observed and fitted probability of an *in* completion than manner verbs.
This pattern is consistent with the verb-class difference motivated by culmination structure.

The analysis does not show that lexical verb class alone caused each completion.
Individual verbs differ within a class, participants differ in their completion
preferences, and items differ in their temporal contexts. The present fit
deliberately ignores those repeated sources to isolate one-predictor logistic
regression.

The analysis is still incomplete as a final experiment model because participants and verbs repeat. For now, we isolate one-predictor logistic regression. The mixed-effects sequence will return to repeated units after their structure has been introduced.

The row-level standard errors incorrectly treat the 9,160 responses as independent. The point estimates
reproduce the two class proportions, but the displayed model understates the
sampling structure. This limitation is part of the result, not a reason to hide
the descriptive verb-class difference.

::: {.callout-note title="Check the reference"}
1. If the reference is changed from result to manner, which coefficient signs
   change? Why do the two fitted probabilities remain the same?
2. Convert the fitted manner coefficient into an odds ratio.
3. Calculate the manner-minus-result probability difference.
4. What unsupported independence claim is added when the row-level standard errors treat all responses as independent?
:::
