---
title: "Adding a second predictor"
---

The [simple linear model](simple-linear-regression.qmd) conditions the response mean on one predictor. Multiple regression extends the same conditional mean to several predictors, so each coefficient must be read relative to the others in the model.

Suppose a word-recognition study predicts log response time from two lexical properties:

1. log lexical frequency, which represents exposure to the word; and
2. phonological neighborhood density, which represents competition from similar word forms.

The theoretical question is whether each source of lexical structure is associated with response time when the other is held at the same modeled value.

## Extend the mean function

With one predictor,

$$
m(x_1)=\beta_0+\beta_1x_1.
$$

Add the second predictor:

$$
m(x_1,x_2)
=\beta_0+\beta_1x_1+\beta_2x_2.
$$

The ordinary multiple linear regression model is

$$
Y_i\mid x_{i1},x_{i2}
\sim
\mathcal{N}
\left(
\beta_0+\beta_1x_{i1}+\beta_2x_{i2},
\sigma^2
\right).
$$

## Compare observations at the same $x_2$

Hold neighborhood density at $x_2$. Increase log frequency by one unit:

$$
\begin{aligned}
m(x_1+1,x_2)-m(x_1,x_2)
&=\beta_1.
\end{aligned}
$$

Thus $\beta_1$ is the frequency association among observations compared at the same modeled neighborhood density.

Likewise, $\beta_2$ is the neighborhood-density association among observations at the same modeled log frequency.

The coefficients are [**partial regression coefficients**](https://appliedstatisticsforlinguists.org/bwinter_stats_proofs.pdf#page=119). Each describes the modeled association for one predictor while the other predictors are held fixed. It can differ from the marginal association between that predictor and the response.

The word *partial* indicates that other predictors are held constant. The frequency coefficient compares
model predictions that differ by one frequency unit while neighborhood density
is fixed. It need not equal the association obtained when neighborhood density
is allowed to vary along with frequency.

Adding a predictor changes the conditioning set for the other coefficients. It does not merely improve an existing coefficient. It
changes the modeled comparison that the coefficient represents.

## Calculate one prediction

Suppose

$$
\beta_0=6.2,
\qquad
\beta_1=-.08,
\qquad
\beta_2=.015.
$$

For a word with centered log frequency 1 and centered neighborhood density 4,

$$
m(1,4)
=6.2-.08(1)+.015(4)
=6.18.
$$

Both predictor contributions are added to the same conditional mean.

The same equation gives a direct comparison. Consider two words with centered
neighborhood density 4 and centered log frequencies 1 and 2. Their expected
responses differ by

$$
m(2,4)-m(1,4)
=\beta_1
=-.08.
$$

Now keep centered log frequency at 1 and compare neighborhood densities 4 and 6:

$$
m(1,6)-m(1,4)
=2\beta_2
=.03.
$$

Each coefficient controls a family of comparisons in which the other predictor
is held fixed in the model.

## Fit the additive model

```{r}
#| label: fit-constructed-multiple-regression
#| echo: true
lexical_decision <- data.frame(
  log_frequency = c(-1.5, -1, -.5, 0, .5, 1, 1.5, 2),
  neighborhood_density = c(14, 12, 13, 9, 10, 6, 7, 3),
  log_rt = c(6.540, 6.450, 6.440, 6.330, 6.315, 6.205, 6.195, 6.075)
)

fit_two <- lm(
  log_rt ~ log_frequency + neighborhood_density,
  data = lexical_decision
)

coef(fit_two)
```

This model assumes that the frequency slope is the same at every neighborhood density and vice versa. An interaction will later relax that assumption.

The fitted frequency coefficient is about $-.066$. It is the estimated
log-frequency association among model comparisons with neighborhood density held
fixed. The fitted neighborhood coefficient is about $.020$ at a fixed modeled
log frequency.

The intercept is the expected response when both centered predictors equal zero.
This interpretation is useful only if the centering constants have been stated.

## Compare the marginal and partial slopes

In the constructed observations, higher-frequency words tend to have fewer
phonological neighbors:

```{r}
#| label: inspect-predictor-association
#| echo: true
cor(
  lexical_decision$log_frequency,
  lexical_decision$neighborhood_density
)
```

The correlation is about $-.94$. Thus the one-predictor frequency model
compares words whose neighborhood densities also tend to differ:

```{r}
#| label: compare-marginal-and-partial-frequency-slopes
#| echo: true
fit_one <- lm(log_rt ~ log_frequency, data = lexical_decision)

c(
  marginal_frequency_slope = coef(fit_one)["log_frequency"],
  partial_frequency_slope = coef(fit_two)["log_frequency"]
)
```

The one-predictor slope is $-.125$, while the two-predictor slope is about
$-.066$. Neither is the uniquely correct description independently of a
question. The first describes the frequency association as neighborhood density
varies in this sample. The second describes the frequency association under a
model comparison that fixes neighborhood density.

This difference follows from the change in conditioning set. It should not be narrated as the second model
"correcting" the first unless the design and theory establish that the partial
comparison is the intended target.

## Adding a variable changes the coefficient target

The frequency coefficient in the two-predictor model is not merely a more accurate version of the one-predictor coefficient. It answers a different question because it conditions on neighborhood density.

Adding every available column does not automatically isolate a causal or theoretical effect. The predictor set should follow from the comparisons the theory requires and from the design that supports them.

For instance, controlling for a variable caused by both the predictor and
response can create a misleading partial association. Controlling for a noisy
proxy can also leave substantial variation in the property the analysis meant
to hold fixed. Regression performs the declared conditional comparison; it does
not determine whether that comparison supports a causal claim.

## Read predictions before coefficient labels

A reliable interpretation method is to choose two predictor rows, calculate
their fitted means, and subtract. If only frequency changes by one unit, the
difference is $\beta_1$. If both predictors change, the difference is

$$
\beta_1\Delta x_1+\beta_2\Delta x_2.
$$

This calculation prevents one coefficient from being used to describe a
contrast that also changes other predictors. The model's prediction for that
contrast may involve several coefficients.

::: {.callout-note title="Check the conditioning set"}
1. In words, distinguish the frequency slope in `log_rt ~ log_frequency` from
   the frequency slope in `log_rt ~ log_frequency + neighborhood_density`.
2. Using the displayed coefficients, compare predictions for two words that
   differ by two frequency units and three neighborhood-density units.
3. Explain how adding a predictor changes the comparison represented by the other coefficient.
4. Why does adding a predictor not by itself license a causal interpretation?
:::
