---
title: "Residuals"
---

The [fitting page](frequentist-linear-regression.qmd) used observed minus fitted discrepancies in the least-squares objective. We now inspect those discrepancies observation by observation.

For observation $i$, the fitted line supplies an estimated conditional mean

$$
\widehat{y}_i
=\widehat{\beta}_0+\widehat{\beta}_1x_i.
$$

The [**residual**](https://appliedstatisticsforlinguists.org/bwinter_stats_proofs.pdf#page=85) is the observed response minus this fitted value:

$$
e_i=y_i-\widehat{y}_i.
$$

The residual is an observation-level discrepancy. It records what remains after
subtracting the model's prediction for that observation.

The direction of the subtraction matters. We subtract the fitted value from the
observed response, not the observed response from the fitted value. This
convention makes a positive residual an observation above the fitted line.

## Calculate three residuals

For the six constructed observations, the fitted equation is

$$
\widehat y_i=5.629+.0714x_i.
$$

The first three residuals are approximately:

| observation | observed $y_i$ | fitted $\widehat{y}_i$ | residual $e_i$ |
|---:|---:|---:|---:|
| 1 | 5.7800 | 5.7714 | $.0086$ |
| 2 | 5.8200 | 5.8429 | $-.0229$ |
| 3 | 5.9300 | 5.9143 | $.0157$ |

A positive residual places the observation above its fitted mean. A negative residual places it below.

The magnitude records how poorly the conditional mean predicts that observation on the response scale.

For observation 2,

$$
e_2
=5.8200-5.8429
\approx-.0229.
$$

The fitted conditional mean is about $.0229$ log-time units above the observed
response. The residual does not say that the fitted value is wrong by the same
amount for all words with 3 bits of surprisal. It describes this observation.

## Compute them from the fitted model

```{r}
#| label: calculate-reading-time-residuals
#| echo: true
reading_times <- data.frame(
  surprisal = 2:7,
  log_rt = c(5.78, 5.82, 5.93, 5.99, 6.05, 6.13)
)

fit <- lm(log_rt ~ surprisal, data = reading_times)
reading_times$residual <- residuals(fit)
reading_times
```

Equivalently,

```{r}
#| label: verify-reading-time-residuals
#| echo: true
reading_times$log_rt - fitted(fit)
```

The equality of the two vectors verifies that residuals are observed responses minus fitted values. We can also attach fitted values
to see all three quantities together:

```{r}
#| label: inspect-fitted-values-and-residuals
#| echo: true
transform(
  reading_times,
  fitted = fitted(fit),
  reconstructed_response = fitted(fit) + residuals(fit)
)
```

The reconstructed response equals the observed response in every row because

$$
y_i=\widehat y_i+e_i.
$$

This identity holds by the definition of a residual. It does not depend on the
normality assumption.

## Residuals are not error terms

The ordinary linear regression model contains unobserved errors

$$
\varepsilon_i=Y_i-\mathbb{E}[Y_i\mid X_i].
$$

These errors are defined relative to the unknown true conditional mean. Residuals are defined relative to the fitted mean estimated from the same sample.

Thus $e_i$ is observable after fitting, while $\varepsilon_i$ is a model quantity we do not directly observe. Residuals provide evidence about the error assumptions, but they are not identical to the errors.

Suppose the true conditional mean at $x_i$ were $5.90$, the fitted mean were
$5.88$, and the observed response were $5.94$. The unobserved error would be

$$
\varepsilon_i=5.94-5.90=.04,
$$

while the residual would be

$$
e_i=5.94-5.88=.06.
$$

The two values differ because the fitted mean differs from the true mean. With
more informative data and an adequate model, residual patterns may approximate
error patterns more closely. They never become the same object by definition.

## A zero sum is not a fit check

In ordinary least squares with an intercept,

$$
\sum_i e_i=0.
$$

This equality follows from the fitting procedure. It can hold even when residuals trace a strong curve or widen systematically with fitted values.

For the constructed fit:

```{r}
#| label: verify-zero-residual-sum
#| echo: true
sum(residuals(fit))
```

The output may be a tiny nonzero value because computers approximate real-number
arithmetic. A value near $10^{-16}$ is numerical noise, not a meaningful failure
of the zero-sum property.

The zero sum also does not imply that positive and negative residuals are small.
Residuals of $1$, $-1$, $10$, and $-10$ sum to zero while containing large
observation-level discrepancies. The sum loses the sizes and their ordering
across predictors.

## Preserve the response scale

A raw residual has the response's units. Here it is measured in log-time units.
If the response were vowel duration in milliseconds, a raw residual would be in
milliseconds. This makes residual size interpretable only relative to the
response scale and the scientific setting.

Comparing raw residual magnitudes across models with different response
transformations can thus be misleading. The numerical scale has changed. On
this page, all interpretations remain on the fitted log-reading-time scale.

The [next page](checking-linear-regression.qmd) plots residuals to reveal patterns that their sum conceals.

::: {.callout-note title="Check the sign"}
1. An observed log reading time is 6.20 and its fitted mean is 6.05. Calculate
   and interpret the residual.
2. If the true conditional mean were 6.00, what would the corresponding error
   be? Why does it differ from the residual?
3. Explain why a zero residual sum cannot establish that all residuals are
   small.
4. What units does a raw residual inherit?
:::
