---
title: "Residualizing a predictor"
---

The [multiple-regression page](multiple-regression.qmd) introduced partial coefficients, and the [residual page](residuals.qmd) defined an observed minus fitted discrepancy. The process of [**residualization**](https://doi.org/10.1016/j.jml.2013.12.003) combines these two ideas, but it does not create a new directly observed linguistic variable.

Suppose surprisal and acceptability are correlated. One proposed response is to regress surprisal on acceptability and use the residual as a predictor of reading time.

The first regression is

$$
S_i
=
\alpha_0+\alpha_1A_i+R_i,
$$

where $S_i$ is surprisal, $A_i$ is acceptability, and $R_i$ is the surprisal residual.

The residual satisfies

$$
R_i
=
S_i-(\widehat{\alpha}_0+\widehat{\alpha}_1A_i).
$$

It is the part of observed surprisal not linearly predicted by acceptability in the fitted sample.

The final phrase is essential. $R_i$ is defined relative to one sample and one
linear residualizer. A different sample, a nonlinear residualizer, or a different
set of predictors would produce a different residualized variable.

The residual thus depends on the model used to residualize the predictor. It is
a statistical construction, not a directly observed linguistic property.

## Interpreting the new predictor

The second regression might be

$$
Y_i
=
\beta_0+\beta_RR_i+\beta_AA_i+\epsilon_i.
$$

The coefficient $\beta_R$ describes a change in reading time associated with residual surprisal while acceptability is held fixed. This is not the marginal association of the original surprisal variable.

In an ordinary linear regression containing the same acceptability term, the coefficient on $R_i$ matches the partial coefficient on $S_i$. Residualization changes the parameterization, not the underlying fitted subspace.

This equivalence can be derived. Since

$$
S_i=\widehat\alpha_0+\widehat\alpha_1A_i+R_i,
$$

substitute this expression into the original response model:

$$
\begin{aligned}
Y_i
&=\beta_0+\beta_SS_i+\beta_AA_i+\epsilon_i\\
&=\beta_0+\beta_S
(\widehat\alpha_0+\widehat\alpha_1A_i+R_i)
+\beta_AA_i+\epsilon_i\\
&=(\beta_0+\beta_S\widehat\alpha_0)
+(\beta_A+\beta_S\widehat\alpha_1)A_i
+\beta_SR_i+\epsilon_i.
\end{aligned}
$$

The coefficient on $R_i$ is the original partial surprisal coefficient
$\beta_S$. The intercept and acceptability coefficient change because the
coordinate system has changed.

## Verifying the equivalence

```{r}
#| label: verify-residualization-equivalence
#| echo: true
dat <- data.frame(
  acceptability = c(2.1, 2.8, 3.4, 4.2, 4.8, 5.5, 6.1, 6.7),
  surprisal = c(8.2, 7.4, 7.1, 6.0, 5.8, 4.9, 4.5, 3.7),
  log_rt = c(6.775, 6.712, 6.687, 6.609, 6.596,
             6.523, 6.501, 6.443)
)

fit_residualizer <- lm(surprisal ~ acceptability, data = dat)
dat$residual_surprisal <- residuals(fit_residualizer)

fit_original <- lm(log_rt ~ surprisal + acceptability, data = dat)
fit_residual <- lm(log_rt ~ residual_surprisal + acceptability, data = dat)

c(
  original_surprisal = coef(fit_original)["surprisal"],
  residual_surprisal = coef(fit_residual)["residual_surprisal"]
)
```

The coefficient on `surprisal` in the first response model equals the coefficient on `residual_surprisal` in the second, up to numerical precision.

The other coefficients are not equal. After residualization, the acceptability
coefficient absorbs both its original response association and the part of
surprisal that the residualizer assigned to acceptability. This is why the
residualized fit must be interpreted as a reparameterized system, not as the
original coefficient table with one purified predictor substituted.

## Residualizing both variables to expose the partial slope

The same partial slope can be obtained by residualizing the response with respect
to acceptability and regressing those response residuals on residual surprisal:

```{r}
#| label: residualize-response-for-partial-slope
#| echo: true
outcome_residualizer <- lm(log_rt ~ acceptability, data = dat)
dat$residual_log_rt <- residuals(outcome_residualizer)

fit_residual_both <- lm(
  residual_log_rt ~ residual_surprisal,
  data = dat
)

coef(fit_residual_both)["residual_surprisal"]
```

This calculation reveals the comparison behind a partial coefficient. First
remove the linear association with acceptability from both surprisal and the
response. Then relate the two remaining quantities.

## Residuals and construct interpretation

Residual surprisal is not “pure processing difficulty.” It depends on the sample, the residualizer formula, measurement error in both variables, and the chosen functional form.

It is also directional. Residualizing surprisal against acceptability does not
produce the same variable as residualizing acceptability against surprisal. The
first preserves surprisal units and defines deviations from predicted
surprisal. The second preserves acceptability units and defines deviations from
predicted acceptability.

If acceptability is measured with error, subtracting its fitted association does not remove the underlying construct cleanly. It can also induce difficult interpretations when the residualizer is trained outside the predictive evaluation loop.

Interpreting that residual as a purified theoretical construct is not licensed. The linear algebra establishes coefficient
equivalence. It does not establish that the residual corresponds to one causal
or cognitive mechanism.

## Keeping the residualizer inside cross-validation

For held-out prediction, estimate $\widehat{\alpha}_0$ and $\widehat{\alpha}_1$ using only the training fold. Then compute residual surprisal for the test fold from those training estimates.

Residualizing once on the complete dataset allows test-fold predictor relationships to influence the fitted transformation. That is leakage.

The same rule applies to centering constants, scaling constants, and any selected
functional form when these are learned from data. The test observations may be
transformed using training estimates, but they may not determine those estimates.

Residualization is often unnecessary for an explanatory multiple regression
because the original two-predictor model already estimates the same partial
surprisal coefficient. The explicit residualizer can still be pedagogically
useful because it displays which surprisal variation identifies the coefficient.

::: {.callout-note title="Interpret a residualized predictor"}
1. What does $R_i$ measure relative to the residualizer model?
2. Does residualization eliminate uncertainty due to correlated constructs?
3. Where must the residualizer be fitted during cross-validation?
4. Which coefficients remain equal between the original and residualized response
   models, and which may change?
5. Explain why residualizing a predictor does not isolate a pure theoretical construct.
:::
