---
title: "Sum coding"
---

The [categorical-coding page](coding-categorical-predictors.qmd) used treatment coding to make the intercept equal the fitted mean of a reference category. Sometimes the relevant reference is the unweighted average of the modeled category means. [**Sum coding**](https://appliedstatisticsforlinguists.org/bwinter_stats_proofs.pdf#page=133) uses columns whose values sum to zero so that the intercept represents that average.

This choice answers a different intercept question without changing the fitted
category means. The new reference is the category-weighted grand mean. It is
the arithmetic mean of the model's category-specific predictions, giving each
category one equal contribution.

Consider three clause types: declarative, interrogative, and infinitival. One sum-coding matrix is

| clause type | $C_1$ | $C_2$ |
|---|---:|---:|
| declarative | 1 | 0 |
| interrogative | 0 | 1 |
| infinitival | $-1$ | $-1$ |

The model is

$$
m_i
=
\beta_0+\beta_1C_{1i}+\beta_2C_{2i}.
$$

## Substitute each category

For declaratives,

$$
m_D=\beta_0+\beta_1.
$$

For interrogatives,

$$
m_Q=\beta_0+\beta_2.
$$

For infinitivals,

$$
m_I=\beta_0-\beta_1-\beta_2.
$$

Add the three fitted means:

$$
m_D+m_Q+m_I=3\beta_0.
$$

Thus

$$
\beta_0=\frac{m_D+m_Q+m_I}{3}.
$$

The intercept is the unweighted average of the three modeled category means.

Suppose the fitted category means are

$$
m_D=5.4,
\qquad
m_Q=6.0,
\qquad
m_I=5.7.
$$

Then

$$
\beta_0
=\frac{5.4+6.0+5.7}{3}
=5.7.
$$

The intercept is not tied to one clause type. It is the category-weighted grand mean across the three
types.

## Interpret the coefficients carefully

The coefficient $\beta_1$ is the declarative mean minus the grand mean. The coefficient $\beta_2$ is the interrogative mean minus the grand mean. The infinitival deviation is determined by the zero-sum constraint:

$$
-(\beta_1+\beta_2).
$$

These coefficients are not pairwise comparisons with the infinitival category. Treatment coding would provide those comparisons directly.

For the three means above,

$$
\beta_1=5.4-5.7=-.3
$$

and

$$
\beta_2=6.0-5.7=.3.
$$

The infinitival deviation is

$$
-(\beta_1+\beta_2)=0,
$$

which recovers $m_I=5.7$. The three deviations sum to zero, as the name *sum
coding* suggests.

## Distinguish the grand mean from the sample mean

Suppose a dataset contains 100 declaratives, 20 interrogatives, and 10
infinitivals. The raw sample mean gives the declarative category much more
weight because it contributes more observations. The category-weighted grand mean gives each of the
three fitted category means weight $1/3$.

The two quantities coincide when the category counts are equal or when the
category means happen to be equal. They need not coincide in an unbalanced
sample. Calling $\beta_0$ simply "the overall sample mean" can thus be wrong.

A report should thus say whether a mean is
observation-weighted or category-weighted. Sum coding supplies the latter
reference in the displayed factor-only model.

## Set the contrasts in R

```{r}
#| label: inspect-sum-contrast-matrix
#| echo: true
dat <- data.frame(
  clause_type = factor(
    c("declarative", "interrogative", "infinitival"),
    levels = c("declarative", "interrogative", "infinitival")
  ),
  rating = c(5.4, 6.0, 5.7)
)

contrasts(dat$clause_type) <- contr.sum(3)

fit_sum <- lm(
  rating ~ clause_type,
  data = dat
)
```

Always inspect the actual contrast matrix:

```{r}
#| label: print-sum-contrast-matrix
#| echo: true
contrasts(dat$clause_type)
coef(fit_sum)
```

Factor level order determines which row receives which coding values.

The coefficient output should reproduce the hand calculation: an intercept of
5.7, a first deviation of $-.3$, and a second deviation of $.3$.

## Coding does not change fitted category means

Treatment and sum coding span the same two-dimensional contrast space for a three-level predictor. They produce the same fitted values, residuals, likelihood, and omnibus comparison when the model otherwise remains the same.

They change the coefficient meanings. Choose the coding whose coefficients express the comparisons required by the linguistic question.

For instance, a theory may predict that interrogative complements differ from
the average selectional profile across clause types. Sum coding makes that
deviation direct. A theory that instead predicts a difference between
interrogatives and declaratives would be more directly represented by a coding
whose column encodes that pairwise comparison.

The correct coding is thus not the one with the most symmetric matrix. It is the
one whose rows and columns represent the comparisons required by the theory.

::: {.callout-note title="Reconstruct the coded means"}
1. What does $\beta_0$ represent under the displayed coding?
2. Which coefficient gives the interrogative deviation from that mean?
3. Why do treatment and sum coding have identical fitted values?
4. If the three fitted means are 4, 7, and 10, calculate the category-weighted grand mean and the two
   displayed sum-coded coefficients.
5. Distinguish the category-weighted grand mean from the observation-weighted sample mean in an unbalanced sample.
:::
