---
title: "Variance inflation from correlated predictors"
---

The [correlated-predictor page](correlated-predictors.qmd) explained why a coefficient becomes uncertain
when few observations separate two predictors. A [**variance inflation factor (VIF)**](https://appliedstatisticsforlinguists.org/bwinter_stats_proofs.pdf#page=119) measures that uncertainty increase in an ordinary linear regression.

As on the earlier [standard-error page](../statistical-inference/standard-errors.qmd), the relevant uncertainty is repeated-sample variation in an estimator. The diagnostic asks a counterfactual precision question: how much larger is the
sampling variance of coefficient $j$ because its predictor overlaps with the
other predictors in this design?

## Defining the auxiliary regression

For predictor $X_j$, regress it on all other predictors and record $R_j^2$.
Then

$$
\operatorname{VIF}_j
=
\frac{1}{1-R_j^2}.
$$

If the other predictors do not explain $X_j$, then $R_j^2=0$ and
$\operatorname{VIF}_j=1$. The sampling variance is not inflated relative to the
no-overlap reference.

Suppose word length is regressed on log frequency and neighborhood density,
producing

$$
R_{\text{length}}^2=.75.
$$

Then

$$
\operatorname{VIF}_{\text{length}}
=
\frac{1}{1-.75}=4.
$$

Under the usual linear-model conditions, the sampling variance of the length
coefficient is four times what it would be if length had the same marginal
variation but were uncorrelated with the other predictors. Its standard error
is multiplied by

$$
\sqrt{4}=2.
$$

Variance and standard error are not inflated by the same factor because standard
error is the square root of variance. If VIF is 9, coefficient variance is nine
times the reference and standard error is three times the reference.

## Connecting VIF to residual predictor variation

The auxiliary regression predicts $X_j$ from the other predictors. When
$R_j^2$ is large, little residual variation in $X_j$ remains. The response model
must estimate coefficient $j$ from this remaining variation.

Since

$$
1-R_j^2
$$

is the proportion of predictor variation left by the auxiliary regression, VIF
is its reciprocal. If 25 percent remains, VIF is $1/.25=4$. If 10 percent
remains, VIF is $1/.10=10$.

For length and frequency, the informative observations are unusually long
frequent words and unusually short rare words. A large VIF says that such
comparisons are scarce. It does not say that they are linguistically irrelevant.

## Calculating VIF from a predictor table

Suppose a morphological study uses stem length and affix count as predictors of
production duration:

```{r}
#| label: calculate-vif-from-auxiliary-regression
#| echo: true
morphology_design <- data.frame(
  stem_length = c(2, 3, 4, 4, 5, 6, 7, 8),
  affix_count = c(0, 0, 1, 0, 1, 2, 2, 3),
  duration = c(310, 335, 390, 365, 430, 505, 540, 620)
)

auxiliary <- lm(stem_length ~ affix_count, data = morphology_design)
r_squared_j <- summary(auxiliary)$r.squared
vif_j <- 1 / (1 - r_squared_j)

c(r_squared = r_squared_j, vif = vif_j, se_multiplier = sqrt(vif_j))
```

The response `duration` does not enter the auxiliary regression. VIF is a
property of the predictor design. Changing every duration while leaving the two
predictors fixed would leave the reported VIF unchanged.

For a model with only two predictors, the auxiliary $R^2$ equals their squared
correlation. With more predictors, $R_j^2$ measures how well their combination
predicts $X_j$.

## Avoiding a mechanical threshold

Rules such as VIF greater than 5 or 10 are conventions. The same value can have
different consequences depending on sample size, predictor range, and the
precision required for the scientific claim.

For instance, doubling a coefficient standard error may still leave enough
precision to distinguish two substantively different hypotheses in one study.
The same doubling may erase the distinction in another. The VIF quantifies a
design consequence; the theory determines whether that consequence is tolerable.

Inspect coefficient intervals, predictor support, and sensitivity to reasonable
specifications. Do not delete a theory-bearing predictor solely to cross a
threshold.

If the `car` package is installed, it can retrieve VIFs from a fitted model:

```{r}
#| label: retrieve-vif-from-fitted-model
#| echo: true
fit_morphology <- lm(
  duration ~ stem_length + affix_count,
  data = morphology_design
)

# car::vif(fit_morphology)
```

For categorical predictors represented by several columns, software may report
a generalized VIF. Its definition and degrees-of-freedom adjustment must be
checked before comparing it with a one-column VIF.

## Keeping prediction and attribution separate

Collinear models can make stable predictions within the observed predictor
region while assigning unstable coefficients to correlated predictors. VIF
diagnoses coefficient attribution, not predictive accuracy on new responses.

VIF also does not say which predictor should be removed. Deleting stem length
sets its partial coefficient to zero and changes the affix-count comparison. If
both predictors express distinct hypotheses, the enlarged uncertainty may be
the honest result.

Deleting a predictor based only on a VIF cutoff treats a diagnostic as a model-selection command. A VIF is more useful when paired with the support plot and the coefficient contrast the study needs.

::: {.callout-note title="Read the variance inflation factor"}
1. What VIF results when $R_j^2=.50$?
2. What happens to the coefficient standard error when VIF equals 9?
3. Why is a high VIF not sufficient reason to remove a theoretical predictor?
4. Why can VIF be calculated without the response?
5. Explain why a VIF cutoff does not by itself determine whether a predictor should be deleted.
:::
