---
title: "The chi-squared distribution"
---

The [normal distribution page](normal-distribution.qmd) introduced standard normal variables. Suppose three independent annotation discrepancies have been standardized relative to a reference model. Their values are

$$
Z_1=1,
\qquad
Z_2=-.5,
\qquad
Z_3=2.
$$

If discrepancy direction is not relevant, we can square each value and add:

$$
Q=Z_1^2+Z_2^2+Z_3^2.
$$

When $Z_1,\ldots,Z_\nu$ are independent standard normal variables, this sum has a [**chi-squared distribution**](https://openstax.org/books/introductory-statistics-2e/pages/11-1-facts-about-the-chi-square-distribution) with $\nu$ degrees of freedom:

$$
Q\sim\chi^2_\nu.
$$

The support is $[0,\infty)$ because a sum of squares cannot be negative.

## Calculate one squared discrepancy

For the three constructed values,

$$
\begin{aligned}
Q
&=1^2+(-.5)^2+2^2\\
&=1+.25+4\\
&=5.25.
\end{aligned}
$$

The value 2 makes the largest contribution because squaring gives greater weight to larger standardized departures.

If these were three independent standard normal discrepancies, the reference distribution would have $\nu=3$.

## Read center and spread from degrees of freedom

For $Q\sim\chi^2_\nu$,

$$
\mathbb{E}[Q]=\nu
$$

and

$$
\operatorname{Var}(Q)=2\nu.
$$

With three degrees of freedom, the mean is 3 and the variance is 6. The observed value 5.25 lies above the reference mean, but a tail probability is needed to state how much reference mass lies at least that far right.

## Calculate the upper tail

The upper tail probability is

$$
\mathbb{P}(Q\geq5.25).
$$

In R,

```{r}
#| eval: false
pchisq(5.25, df = 3, lower.tail = FALSE)
```

The result is approximately $.154$. Under the reference assumptions, about $.154$ of the probability lies on values at least as large as 5.25.

This is a statement about the statistic's reference distribution. It is not the probability that a null hypothesis is true.

## See what changing degrees of freedom does

With one degree of freedom, the distribution is strongly right skewed because one squared value often lies near zero but can occasionally be large. Adding more independent squared terms moves the center right and makes the shape less sharply skewed.

```{r}
#| eval: false
q <- seq(0, 20, length.out = 300)

plot(q, dchisq(q, df = 1), type = "l",
     xlab = "squared discrepancy", ylab = "density")
lines(q, dchisq(q, df = 4), lty = 2)
lines(q, dchisq(q, df = 10), lty = 3)
```

Degrees of freedom determine both center and spread for this family.

## Degrees of freedom count unconstrained variation

The symbol $\nu$ does not always count observations directly. It counts independent pieces of variation remaining after constraints.

In an $r$ by $c$ contingency table with row and column totals treated as fixed, the final row and column values are determined by the earlier cells. The number of free cell comparisons is

$$
\nu=(r-1)(c-1).
$$

The [chi-squared test page](../statistical-inference/chi-squared-test.qmd) later derives this count for a small linguistic table.

## A tail probability is not a hypothesis probability

An upper tail probability is not the probability that a hypothesis is true. The tail calculation conditions on a reference model and asks how often the statistic would be at least as large.

It does not reverse that conditional statement. Interpreting hypotheses requires a separate inferential framework developed later.

## Check your understanding

1. Why can a chi-squared value not be negative?
2. If $\nu=8$, what are the mean and variance?
3. Which constructed discrepancy contributes most to $Q=5.25$, and why?
4. State the upper tail probability in words without treating it as a hypothesis probability.

The family pages define target distributions. The [next page](inverse-transform-sampling.qmd) gives a general procedure for turning cumulative probability into simulated values.
