---
title: "The geometric distribution"
execute:
  enabled: true
  echo: true
  warning: false
  message: false
---

The original notes introduce the geometric family from the geometric series

$$
\sum_{k=0}^{\infty}\frac{1}{2^{k+1}}=1.
$$

Thus $p_K(k)\equiv2^{-(k+1)}$, for $k=0,1,2,\ldots$, is a probability mass function (PMF) with countably infinite support.

```{r}
k <- 0:9
mass <- 2^(-(k + 1))
plot(k, mass, type = "h", lwd = 5, lend = 1, col = "#B2182B",
     xlab = "k", ylab = "Probability",
     main = "PMF defined by a geometric series", bty = "l")
points(k, mass, pch = 19, col = "#B2182B")
```

## Definition and convention

Let $K$ count failures before the first success in independent Bernoulli trials with common success probability $\pi\in(0,1]$. We define

$$
K\sim\operatorname{Geometric}(\pi)
$$

by

$$
p_K(k)
\equiv\mathbb P(K=k)
=(1-\pi)^k\pi,
\qquad k=0,1,2,\ldots.
$$

This is the convention used by `dgeom` in R. If $T$ counts total trials through the first success, then $T\equiv K+1$ and

$$
\mathbb P(T=t)=(1-\pi)^{t-1}\pi,
\qquad t=1,2,\ldots.
$$

## Constructing the PMF from trial sequences

Suppose we sample word tokens until the first loanword, with constant loanword probability $\pi=.20$. The event $K=k$ requires $k$ nonloanwords followed by one loanword. Independence gives

$$
\mathbb P(K=k)=(.80)^k(.20).
$$

Thus

$$
\begin{aligned}
p_K(0)&=.20,\\
p_K(1)&=.80(.20)=.16,\\
p_K(2)&=.80^2(.20)=.128,\\
p_K(3)&=.80^3(.20)=.1024.
\end{aligned}
$$

The event $K\geq4$ requires four initial failures, so

$$
\mathbb P(K\geq4)=.80^4=.4096
$$

and $\mathbb P(T\leq4)=1-.4096=.5904$.

## The original parameter sequence

The following graphs reproduce the $\pi=.5$, $.1$, and $.9$ sequence from the original notes.

```{r}
plot_geometric <- function(probability) {
  k <- 0:9
  mass <- dgeom(k, prob = probability)
  plot(k, mass, type = "h", lwd = 5, lend = 1, col = "#B2182B",
       xlab = "Failures before the first success", ylab = "Probability",
       main = sprintf("PMF of Geometric(%.1f)", probability), bty = "l")
  points(k, mass, pch = 19, col = "#B2182B")
}

plot_geometric(.5)
plot_geometric(.1)
plot_geometric(.9)
```

For every $k\geq0$,

$$
\frac{p_K(k+1)}{p_K(k)}
=1-\pi<1.
$$

Thus every nondegenerate geometric PMF is strictly decreasing. Its mode is $K=0$. If $L\equiv K+1$ represents positive word length, its mode is necessarily $L=1$. The [negative-binomial page](negative-binomial-distribution.qmd) compares this restriction with the empirical distribution of phoneme counts in CMUdict.

Under this convention,

$$
\mathbb E[K]=\frac{1-\pi}{\pi},
\qquad
\operatorname{Var}(K)=\frac{1-\pi}{\pi^2}.
$$
