---
title: "The geometric distribution"
execute:
  enabled: true
  echo: true
  warning: false
  message: false
---

Suppose we sample word tokens until the first loanword appears. The response is not a fixed number of trials. It is the number of nonloanwords we see before the target. What distribution represents this waiting count?

This is a methodological example. A corpus application would need an explicit operational definition of *loanword* and a token-sampling order; the geometric model does not decide disputed etymological or borrowing classifications.

To begin with a particular case, suppose the loanword probability is $1/2$ on each [independent](../foundations/independence.qmd) draw. Zero failures occurs with probability $1/2$, one failure followed by a target with probability $1/4$, and two failures followed by a target with probability $1/8$. These probabilities form the geometric series

$$
\sum_{k=0}^{\infty}\frac{1}{2^{k+1}}=1.
$$

Thus $p_K(k)\equiv2^{-(k+1)}$, for $k=0,1,2,\ldots$, is a probability mass function (PMF) with countably infinite support. The index $k$ counts failures, so the first term corresponds to an immediate target.

```{r}
k <- 0:9
mass <- 2^(-(k + 1))
plot(k, mass, type = "h", lwd = 5, lend = 1, col = "#B2182B",
     xlab = "k", ylab = "Probability",
     main = "PMF defined by a geometric series", bty = "l")
points(k, mass, pch = 19, col = "#B2182B")
```

## Definition and convention

Basically, the geometric distribution extends this example from target probability $1/2$ to any fixed target probability. More specifically, the [**geometric distribution**](https://openstax.org/books/introductory-statistics-2e/pages/4-4-geometric-distribution) models a waiting count. Let $K$ count failures before the first success in independent Bernoulli trials with common success probability $\pi\in(0,1]$. We define

$$
K\sim\operatorname{Geometric}(\pi)
$$

by

$$
p_K(k)
\equiv\mathbb P(K=k)
=(1-\pi)^k\pi,
\qquad k=0,1,2,\ldots.
$$

This is the convention used by `dgeom` in R. If $T$ counts total trials through the first success, then $T\equiv K+1$ and

$$
\mathbb P(T=t)=(1-\pi)^{t-1}\pi,
\qquad t=1,2,\ldots.
$$

## Constructing the PMF from trial sequences

Suppose we sample word tokens until the first loanword, with constant loanword probability $\pi=.20$. The event $K=k$ requires $k$ nonloanwords followed by one loanword. Independence gives

$$
\mathbb P(K=k)=(.80)^k(.20).
$$

Thus

$$
\begin{aligned}
p_K(0)&=.20,\\
p_K(1)&=.80(.20)=.16,\\
p_K(2)&=.80^2(.20)=.128,\\
p_K(3)&=.80^3(.20)=.1024.
\end{aligned}
$$

The event $K\geq4$ requires four initial failures, so

$$
\mathbb P(K\geq4)=.80^4=.4096
$$

and $\mathbb P(T\leq4)=1-.4096=.5904$.

## Comparing the success probability

The following graphs compare $\pi=.5$, $.1$, and $.9$.

```{r}
plot_geometric <- function(probability) {
  k <- 0:9
  mass <- dgeom(k, prob = probability)
  plot(k, mass, type = "h", lwd = 5, lend = 1, col = "#B2182B",
       xlab = "Failures before the first success", ylab = "Probability",
       main = sprintf("PMF of Geometric(%.1f)", probability), bty = "l")
  points(k, mass, pch = 19, col = "#B2182B")
}

plot_geometric(.5)
plot_geometric(.1)
plot_geometric(.9)
```

For every $k\geq0$,

$$
\frac{p_K(k+1)}{p_K(k)}
=1-\pi<1.
$$

Thus every nondegenerate geometric PMF is strictly decreasing. Its mode is $K=0$. If $L\equiv K+1$ represents positive word length, its mode is necessarily $L=1$. The [negative-binomial page](negative-binomial-distribution.qmd) compares this restriction with the empirical distribution of phoneme counts in CMUdict.

## Deriving the expectation

Let $q\equiv1-\pi$. Differentiating the geometric series gives

$$
\sum_{k=0}^{\infty}q^k=\frac{1}{1-q}
\quad\Longrightarrow\quad
\sum_{k=1}^{\infty}kq^{k-1}=\frac{1}{(1-q)^2}.
$$

Thus

$$
\begin{aligned}
\mathbb E[K]
&=\sum_{k=0}^{\infty}kq^k\pi\\
&=\pi q\sum_{k=1}^{\infty}kq^{k-1}\\
&=\frac{1-\pi}{\pi}.
\end{aligned}
$$

Differentiating the geometric series twice also gives

$$
\mathbb E[K(K-1)]
=\frac{2(1-\pi)^2}{\pi^2}.
$$

Since $K^2=K(K-1)+K$,

$$
\begin{aligned}
\operatorname{Var}(K)
&=\mathbb E[K^2]-\mathbb E[K]^2\\
&=\mathbb E[K(K-1)]+\mathbb E[K]-\mathbb E[K]^2\\
&=\frac{1-\pi}{\pi^2}.
\end{aligned}
$$

The algebra matches the waiting-time interpretation. A smaller target probability increases both the expected number of failures and the uncertainty about how long the wait will be.

## Check your understanding

1. With $\pi=.25$, compute the probability of exactly three failures before the first target.
2. Under the same model, what event does $K\geq3$ describe in terms of the first three draws?
3. If $T$ counts draws through the first target, why is $T=K+1$?
4. Explain why the geometric PMF must decrease as $k$ increases.

The geometric family waits for one target. The [negative-binomial page](negative-binomial-distribution.qmd) keeps the same trial assumptions but waits for a declared number of targets.
