---
title: "The Poisson distribution"
---

Suppose we divide a conversation into one-minute intervals and count filled pauses in each interval. One observation may contain 0 filled pauses, another 2, and another 5. Which distribution can represent an unbounded count when the opportunity for events is a declared amount of exposure?

Here a filled pause means an event identified by a transcription convention, such as the `[uh]`, `[um]`, `[er]`, `[ah]`, and `[mm]` labels in the [Switchboard transcription specification](https://catalog.ldc.upenn.edu/docs/LDC94S19/sro_spec.html). The model can describe annotated counts per minute; it cannot by itself identify the cognitive source of those events.

The response is a nonnegative count:

$$
X\in\{0,1,2,3,\ldots\}.
$$

A common baseline model for such counts is the [**Poisson distribution**](https://appliedstatisticsforlinguists.org/bwinter_stats_proofs.pdf#page=234). We write

$$
X\sim\operatorname{Poisson}(\lambda),
$$

where $\lambda>0$ is the expected count in one specified unit of exposure.

The [probability mass function](probability-mass-functions.qmd) of a Poisson random variable specifies the distribution of one count. Now, you might wonder whether writing this distribution commits us to a process in continuous time. It does not. A homogeneous Poisson process provides one possible generative account under which events occur at a constant rate and increments in nonoverlapping intervals are independent. The Poisson count distribution can also be used as an observation model without claiming that the full event sequence follows that process.

## Exposure is part of the definition

The statement $\lambda=2$ is incomplete until the opportunity for events to occur is stated. It might mean two filled pauses per minute, per hundred words, or per speaker turn.

Here one observation is a one-minute interval, so

$$
\mathbb{E}[X]=\lambda=2
$$

means two filled pauses per minute on average under the model.

The expected count scales with exposure when the modeled event rate stays fixed. A thirty-second interval has half the exposure of a one-minute interval, so its expected count is

$$
2\left(\frac{30}{60}\right)=1.
$$

A five-minute interval has five times the exposure, so its expected count is

$$
2(5)=10.
$$

This scaling does not claim that every five-minute stretch contains exactly ten filled pauses. It changes the Poisson mean for the larger opportunity window.

## Calculating point probabilities

The Poisson PMF is

$$
p_X(x)
\equiv\mathbb{P}(X=x)
=\frac{e^{-\lambda}\lambda^x}{x!},
\qquad x\in\{0,1,2,\ldots\}.
$$

For $\lambda=2$, the probability of no filled pauses is

$$
\begin{aligned}
p_X(0)
&=\frac{e^{-2}2^0}{0!}\\
&=e^{-2}\\
&\approx.135.
\end{aligned}
$$

The probability of exactly two is

$$
\begin{aligned}
p_X(2)
&=\frac{e^{-2}2^2}{2!}\\
&=2e^{-2}\\
&\approx.271.
\end{aligned}
$$

These are probabilities of individual counts. We can also ask for the
probability of a range of counts. The probability of at most two filled pauses
is

$$
\begin{aligned}
F_X(2)
&=p_X(0)+p_X(1)+p_X(2)\\
&\approx .135+.271+.271\\
&\approx .677.
\end{aligned}
$$

The distinction between exactly two and at most two is the distinction between
a probability mass function and a cumulative distribution function. The first
selects one point in the support. The second adds the mass at every point up to
and including the requested count.

## Asking four kinds of question in R

R uses the same four prefixes for every named distribution family. For the
Poisson family, they are `dpois`, `ppois`, `qpois`, and `rpois`.

| function | statistical question | result when $\lambda=2$ |
|---|---|---|
| `dpois(2, 2)` | What is $p_X(2)$? | approximately $.271$ |
| `ppois(2, 2)` | What is $F_X(2)$? | approximately $.677$ |
| `qpois(.90, 2)` | What is the smallest $x$ with $F_X(x)\geq.90$? | $4$ |
| `rpois(6, 2)` | What might six new observations look like? | six simulated counts |

The first three calls are deterministic. They return the same answer whenever
their arguments are unchanged. The last call is a simulation. Its result may
change from one call to the next because it draws new values from the specified
distribution.

```{r}
#| label: four-poisson-questions
#| echo: true
dpois(2, lambda = 2)
ppois(2, lambda = 2)
qpois(.90, lambda = 2)
rpois(6, lambda = 2)
```

The quantile call deserves a slower reading. Counts are discrete, so there may
be no count whose cumulative probability is exactly $.90$. R returns the
smallest count at which the cumulative probability reaches or passes $.90$.
For $\lambda=2$, the cumulative probability through 3 is about $.857$, while
the cumulative probability through 4 is about $.947$. Thus the requested
quantile is 4.

## Drawing the whole distribution

A probability mass function can be interpreted more readily when we examine all of its
values together. The following code evaluates the PMF from 0 through 10.

```{r}
#| label: poisson-pmf
#| echo: true
counts <- 0:10
mass <- dpois(counts, lambda = 2)

plot(
  counts, mass,
  type = "h", lwd = 4,
  xlab = "filled pauses in one minute",
  ylab = "probability"
)
points(counts, mass, pch = 16)
```

The distribution places its greatest mass near 2, which is also its expected
value. But an expected value is not a promise about the next observation. Zero,
one, three, and larger counts remain possible. The parameter describes a
distribution over repeated observations, not the value that each observation
must take.

## One parameter controls center and spread

For a Poisson variable,

$$
\mathbb{E}[X]=\lambda
$$

and


$$
\operatorname{Var}(X)=\lambda.
$$

Thus a Poisson distribution with mean 2 must also have variance 2. The model cannot hold the mean fixed while increasing the variance.

This restriction may be too strong for linguistic counts. If speakers, documents, genres, or lexical items have different rates, the observed variance may be much larger than the mean. The [negative-binomial page](negative-binomial-distribution.qmd) gives one family whose variance can exceed its mean, though the scientific reason for extra variation still needs to be modeled.

## Diagnosing the first failure mode

Suppose 100 one-minute intervals have a sample mean of 2.1 filled pauses and a
sample variance of 9.4. A Poisson model fitted to the center would use a value of
$\lambda$ near 2.1. It would then require a variance near 2.1 as well. The gap
between 2.1 and 9.4 is evidence that the one-parameter family misses some
structure in the observations.

This pattern is called [**overdispersion**](https://appliedstatisticsforlinguists.org/bwinter_stats_proofs.pdf#page=234) relative to the Poisson model. It does
not by itself tell us why the extra variation occurs. Speakers may have
different pause rates. Speech tasks may differ. Pauses may cluster after moments
of planning difficulty. Each possibility suggests a different analysis. The
mean to variance comparison is thus a diagnostic, not an explanation.

::: {.callout-note title="Check the exposure"}
1. Two documents contain 3 target constructions each, but one has 500 words and
   the other 5,000. Explain why the raw counts do not represent the same rate.
2. With $X\sim\operatorname{Poisson}(2)$, explain the difference between
   `dpois(2, 2)` and `ppois(2, 2)` without referring to the R function names.
3. A corpus count has a mean of 4.3 and a variance of 18.7. State the Poisson
   restriction that this summary appears to violate. Then identify one linguistic
   source of the extra variation.
:::

The exact upshot is that a Poisson parameter is an expected count for a stated exposure, and that same parameter fixes the variance. Both pieces must be plausible before the family is a useful count model.
