---
title: "Reproducing a simulation"
---

The [preceding page](inverse-transform-sampling.qmd) introduced simulation from a target distribution. Suppose we simulate ten choices between two word orders under a target probability of $.60$ for order A. Running the simulation twice usually gives different sequences because the pseudorandom number generator advances after each draw.

A [**random seed**](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Random.html) sets that generator to a known initial state. It makes the computational sequence reproducible.

```{r}
#| eval: false
set.seed(214)
sample(c("A", "B"), size = 10, replace = TRUE,
       prob = c(.60, .40))

set.seed(214)
sample(c("A", "B"), size = 10, replace = TRUE,
       prob = c(.60, .40))
```

The two simulated vectors are identical. The seed does not change the target probabilities. It restarts the generator from the same state.

## Separate the procedure from one realization

The sampling procedure is the instruction to draw ten independent labels with probabilities $.60$ and $.40$. A seed determines which particular ten labels appear in one computational run.

Another seed gives another valid realization from the same procedure. Neither realization must contain exactly six A outcomes. The target proportion concerns repeated simulation, not a requirement imposed on each small sample.

This distinction is useful when a simulated summary enters an analysis. A stable result should not depend materially on the accidental choice of seed once enough simulation draws have been used.

## Set the seed once before related draws

Consider this incorrect loop:

```{r}
#| eval: false
simulated_count <- numeric(100)

for (b in 1:100) {
  set.seed(214)
  simulated_count[b] <- sum(runif(10) <= .60)
}
```

Every iteration resets the generator. The loop repeats the same simulated count 100 times.

Set the seed once instead:

```{r}
#| eval: false
set.seed(214)
simulated_count <- numeric(100)

for (b in 1:100) {
  simulated_count[b] <- sum(runif(10) <= .60)
}
```

Now the complete sequence is reproducible while individual iterations differ.

## Check simulation stability during development

Suppose we estimate the expected A count by averaging many simulated counts. We can compare several seeds:

```{r}
#| eval: false
estimate_mean_count <- function(seed, draws) {
  set.seed(seed)
  mean(replicate(draws, sum(runif(10) <= .60)))
}

sapply(c(11, 29, 47), estimate_mean_count, draws = 10000)
```

Each result should be near the theoretical expectation $10(.60)=6$. If conclusions change across seeds, the simulation may use too few draws or contain an implementation error.

Increasing draws reduces Monte Carlo variation. It does not repair a wrong target distribution.

## Place the seed at the right scope

One analysis may contain several related random operations. Setting one seed before the complete block makes the whole sequence reproducible. Resetting before every individual call can create unintended repetition or dependence.

The seed should be recorded with the code and set before the first random draw that affects the reported result.

## Reproducibility is not validity

A reproducible simulation is not necessarily statistically valid. A fixed seed cannot make dependent tokens independent, correct an incorrect PMF, or make a convenience sample representative.

Reproducibility answers whether another analyst can regenerate the same computation. Adequacy answers whether the computation represents the intended linguistic process. These questions require separate checks.

## Check your understanding

1. What object does `set.seed()` hold fixed?
2. Why does resetting the same seed inside a loop repeat outcomes?
3. If a simulated estimate changes substantially across seeds, what should be checked?
4. Why can a perfectly reproducible simulation still use an inadequate linguistic model?

Simulation draws one variable at a time or several together. The [next page](joint-probability.qmd) begins the multivariable sequence by assigning probability to value combinations.
