---
title: "Student's t distribution"
---

The [normal distribution page](normal-distribution.qmd) introduced standardization when the population scale is fixed. Let $X_1,\ldots,X_n$ be independent variables with common distribution $\mathcal N(\mu,\sigma^2)$, and define

$$
\overline X\equiv\frac1n\sum_{i=1}^nX_i
$$

and

$$
S^2\equiv\frac{1}{n-1}\sum_{i=1}^n(X_i-\overline X)^2.
$$

Then

$$
\frac{\overline X-\mu}{S/\sqrt n}\sim t_{n-1}.
$$

[**Student's t distribution**](https://online.stat.psu.edu/stat414/Lesson26) can be defined from two independent random variables. Let $Z\sim\mathcal N(0,1)$ and $U\sim\chi^2_\nu$, with $Z\perp U$, and define

$$
T\equiv\frac{Z}{\sqrt{U/\nu}}.
$$

Then

$$
T\sim t_\nu,
\qquad \nu>0.
$$

The parameter $\nu$ is the degrees of freedom. Because $U/\nu$ can be close to zero, the $t_\nu$ distribution assigns more probability to extreme values than the standard normal distribution when $\nu$ is finite.

The distribution is symmetric about zero:

$$
f_T(-t)=f_T(t).
$$

Its mean is zero only when $\nu>1$, and its variance is finite only when $\nu>2$.

## Comparing central intervals

The middle 95 percent of a standard normal distribution lies between approximately $-1.96$ and $1.96$. For $t_5$, the middle 95 percent lies between approximately

$$
-2.57\text{ and }2.57.
$$

The wider interval reflects heavier tails. A value of 2.2 is outside the central normal interval but inside the central $t_5$ interval.

As the degrees of freedom increase, the t family approaches the standard normal family.

| reference | lower $.025$ quantile | upper $.975$ quantile |
|---|---:|---:|
| $t_5$ | about $-2.57$ | about $2.57$ |
| $t_{10}$ | about $-2.23$ | about $2.23$ |
| $t_{30}$ | about $-2.04$ | about $2.04$ |
| standard normal | about $-1.96$ | about $1.96$ |

The table makes the limiting pattern visible without treating degrees of freedom as a generic scale parameter.

## Calculating a two-sided tail probability

Suppose the observed standardized value is $|t|=2.5$ with 5 degrees of freedom. Symmetry gives

$$
\mathbb{P}(|T|\geq2.5)
=2\mathbb{P}(T\geq2.5).
$$

In R,

```{r}
#| eval: false
2 * pt(2.5, df = 5, lower.tail = FALSE)
```

The probability is approximately $.0545$. Under the standard normal reference, the corresponding two-sided tail probability is about $.0124$.

## Interpreting degrees of freedom

In the sample-mean construction, the centered deviations satisfy the linear constraint

$$
\sum_{i=1}^n(X_i-\overline X)=0.
$$

Thus only $n-1$ centered deviations can vary freely, giving

$$
\nu=n-1.
$$

Other statistics impose different constraints. Thus $\nu$ should not be treated as another name for sample size.

Smaller $\nu$ represents greater uncertainty about the scale used in the standardized quantity. The reference distribution then places more probability in the tails. Larger $\nu$ reduces that extra tail mass.

## Inspecting several references in base R

```{r}
#| eval: false
degrees_freedom <- c(5, 10, 30)

data.frame(
  df = degrees_freedom,
  lower = qt(.025, df = degrees_freedom),
  upper = qt(.975, df = degrees_freedom)
)
```

These quantiles are properties of the reference distributions, not estimates of the lexical decision effect itself.

## Keeping two uses separate

The t family may be used as a reference distribution for a standardized estimator. It may also be used as a model for response deviations with heavier tails. These are different statistical roles.

Choosing a t reference for an estimator does not change the observed response values or make the sample mean resistant to an extreme measurement. A robust response model requires an explicit likelihood choice.

## Check your understanding

1. Which central 95 percent interval is wider, $t_5$ or $t_{30}$? Explain using the defining random denominator.
2. Why does estimating one mean from $n$ values leave $n-1$ freely varying deviations?
3. Would $|t|=2.5$ have more upper tail probability under $t_5$ or the standard normal reference?
4. What must be specified before claiming that a t family makes a response model robust?

The t family represents a signed standardized quantity. The [next page](inverse-transform-sampling.qmd) turns cumulative probability into simulated values.
