---
title: "Student's t distribution"
---

The [normal distribution page](normal-distribution.qmd) introduced standardization when the population scale is fixed. Suppose instead that a sample mean is centered at a reference value and divided by a scale estimated from the same finite normal sample. The resulting statistic has more tail uncertainty than a standard normal variable.

[**Student's t distribution**](https://appliedstatisticsforlinguists.org/bwinter_stats_proofs.pdf#page=173) is the heavier tailed symmetric reference family for that statistic. We write

$$
T\sim t(\nu),
$$

where $\nu>0$ is the degrees of freedom.

The distribution is centered at zero and symmetric:

$$
f_T(-t)=f_T(t).
$$

Its tails contain more probability than the standard normal tails when $\nu$ is finite.

## Comparing central intervals

The middle 95 percent of a standard normal distribution lies between approximately $-1.96$ and $1.96$. For $t(5)$, the middle 95 percent lies between approximately

$$
-2.57\text{ and }2.57.
$$

The wider interval reflects heavier tails. A value of 2.2 is outside the central normal interval but inside the central $t(5)$ interval.

As the degrees of freedom increase, the t family approaches the standard normal family.

| reference | lower $.025$ quantile | upper $.975$ quantile |
|---|---:|---:|
| $t(5)$ | about $-2.57$ | about $2.57$ |
| $t(10)$ | about $-2.23$ | about $2.23$ |
| $t(30)$ | about $-2.04$ | about $2.04$ |
| standard normal | about $-1.96$ | about $1.96$ |

The table makes the limiting pattern visible without treating degrees of freedom as a generic scale parameter.

## Calculating a two-sided tail probability

Suppose the observed standardized value is $|t|=2.5$ with 5 degrees of freedom. Symmetry gives

$$
\mathbb{P}(|T|\geq2.5)
=2\mathbb{P}(T\geq2.5).
$$

In R,

```{r}
#| eval: false
2 * pt(2.5, df = 5, lower.tail = FALSE)
```

The probability is approximately $.0545$. Under the standard normal reference, the corresponding two-sided tail probability is about $.0124$. Scale uncertainty changes the reference probability substantially.

## Interpreting degrees of freedom

Degrees of freedom record how much independent information remains for estimating variation after constraints have been imposed. In the simple case of estimating one mean from $n$ observations, deviations from that estimated mean must sum to zero. Only $n-1$ deviations can vary freely, giving

$$
\nu=n-1.
$$

Other designs and estimators create different constraints. Thus $\nu$ should not be treated as another name for sample size.

Smaller $\nu$ represents greater uncertainty about the scale used in the standardized quantity. The reference distribution then places more probability in the tails. Larger $\nu$ reduces that extra tail mass.

## Inspecting several references in base R

```{r}
#| eval: false
degrees_freedom <- c(5, 10, 30)

data.frame(
  df = degrees_freedom,
  lower = qt(.025, df = degrees_freedom),
  upper = qt(.975, df = degrees_freedom)
)
```

These quantiles are properties of the reference distributions, not estimates of the lexical decision effect itself.

## Keeping two uses separate

The t family may be used as a reference distribution for a standardized estimator. It may also be used as a model for response deviations with heavier tails. These are different statistical roles.

Choosing a t reference for an estimator does not change the observed response values or make the sample mean resistant to an extreme measurement. A robust response model requires an explicit likelihood choice.

## Heavy tails do not repair the measurement model

A heavier tailed reference distribution does not automatically make the underlying measurement analysis robust. The reference describes uncertainty in a standardized quantity. It does not rewrite the measurement model that produced that quantity.

Specify which object follows the t distribution: a reference statistic, an error term, or another standardized variable.

## Check your understanding

1. Which central 95 percent interval is wider, $t(5)$ or $t(30)$? Explain using scale uncertainty.
2. Why does estimating one mean from $n$ values leave $n-1$ freely varying deviations?
3. Would $|t|=2.5$ have more upper tail probability under $t(5)$ or the standard normal reference?
4. What must be specified before claiming that a t family makes a response model robust?

The t family represents a signed standardized quantity. The [next page](chi-squared-distribution.qmd) represents nonnegative sums of squared standardized quantities.
