---
title: "Problem Set 2: Bayesian hypothesis testing"
subtitle: "Posterior counterparts to paired, exact, and contingency-table tests"
format:
  html:
    toc: true
execute:
  enabled: false
---

## Bayesian counterparts to hypothesis tests

Problem Set 1 begins with a null hypothesis and calculates the probability of a test statistic under a null reference distribution. This assignment begins with a [likelihood and a prior distribution](../../statistical-inference/posterior-distributions.qmd). [Bayes' rule](../../foundations/bayes-rule.qmd) updates the prior to a posterior distribution. Posterior probabilities are probabilities over parameter values conditional on the observed data. They are not p values.

The exercises follow the same sequence as Problem Set 1. A normal location model corresponds to the paired t analysis. A robust Student t location model serves the same inferential role as the Wilcoxon analysis. A beta–Bernoulli model corresponds to the exact sign analysis. Separate beta–binomial models correspond to the two-row association analyzed by Fisher's exact test. A posterior predictive independence model corresponds to the chi-squared analysis. The reading-time data become relevant in Exercises 1–4. The syntactic data become relevant in Exercises 5–6.

All six exercises are required. The points sum to 100. Every parameter is scalar. Bold symbols denote collections of independent and identically distributed observations, not vector-valued parameters.

::: {.callout-note title="Material from the earlier PS2"}
The previous version of PS2 asked whether predictor sets improve first-fixation prediction for a passage omitted during model fitting. That material has moved to the course-note sequence on [prediction targets](../../model-criticism-and-prediction/prediction-targets.qmd), [predictive loss](../../model-criticism-and-prediction/predictive-loss.qmd), [training and test data](../../model-criticism-and-prediction/training-and-test-data.qmd), [grouped cross-validation](../../model-criticism-and-prediction/grouped-cross-validation.qmd), and [comparing predictive loss](../../model-criticism-and-prediction/comparing-predictive-loss.qmd).

The Provo data remain in this assignment, but the inferential target has changed. PS2 now supplies Bayesian counterparts to the procedures in PS1. The grouped cross-validation material has not been folded into these exercises because it concerns predictive evaluation rather than the paired and categorical parameters analyzed here.
:::

::: {.callout-note title="Assignment files"}
[Download the complete Problem Set 2 package](../../downloads/ps2-assignment.zip){download="ps2-assignment.zip"}. The package contains the assignment source, the README, the preparation scripts, and the redistributable teaching tables.

The files are also available separately:

- [README and setup instructions](README.md)
- [`prepare_data.R`](scripts/prepare_data.R){download="prepare_data.R"}
- [`fetch_wh_data.py`](scripts/fetch_wh_data.py){download="fetch_wh_data.py"}
- [`prepare_ud_wh_tables.R`](scripts/prepare_ud_wh_tables.R){download="prepare_ud_wh_tables.R"}
:::

Use `<-` for assignment, use `snake_case` object names, and record a seed before every random calculation. Every numerical answer must be produced by visible code.

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
```

## Exercise 1: Construct paired passage differences (10 points)

The data for Exercises 1–4 come from the [Provo Corpus](https://doi.org/10.3758/s13428-017-0908-4). One row represents one content-word position. The response `mean_log_ffd` is the mean log first-fixation duration over observed readers who fixated that position. Thus the analysis is conditional on fixation and targets an observed-reader mean, not individual reading behavior. `Text_ID` identifies the passage and `Word_POS` identifies the word class.

The comparison is paired by passage. Within each passage, average `mean_log_ffd` over noun positions and over verb positions. Then define

$$
\delta_p
=
\overline{\log(\operatorname{FFD})}_{p,\mathrm{noun}}
-
\overline{\log(\operatorname{FFD})}_{p,\mathrm{verb}}.
$$

Thus $\exp(\delta_p)$ is the ratio of the noun-position geometric mean FFD to the verb-position geometric mean FFD in passage $p$.

### 1.1 Prepare the paired table

Load `data/provo-content-word-positions.csv`. Retain noun and verb positions. Verify that each passage contains at least four positions from each class. Summarize within passage and class, then join by `Text_ID` to obtain one row per passage. Do not align the summaries by row position.

```{r}
# Load the Provo teaching table.

# Examine the column names, classes, and word-class levels.

# Filter to noun and verb positions.

# Count positions by Text_ID and Word_POS.

# Verify at least four positions in each class for every passage.

# Compute the mean response within passage and word class.

# Join the noun and verb summaries by Text_ID.

# Calculate delta_log_ffd as noun minus verb.
```

### 1.2 Summarize the passage differences

Calculate the number of passages, the mean, the standard deviation, the median, and the number of negative, zero, and positive differences. Plot the differences with zero marked.

```{r}
# Descriptive statistics.
# 1. Count the passages.
# 2. Calculate the mean, standard deviation, and median of delta_log_ffd.
# 3. Count negative, zero, and positive differences.

# Plot the passage differences.
# 1. Put one point at each observed difference.
# 2. Mark zero with a reference line.
# 3. Label the subtraction order and log-duration scale.
```

Report the requested values in the subtraction order noun minus verb.

::: {.callout-note title="Your result"}
Replace this sentence with the sample size and requested summaries.
:::

## Exercise 2: Normal location model (25 points)

Let $Y_p:\Omega\to\mathbb R$ be the random noun minus verb difference for passage $p$, and let $\delta_p$ be its observed value. Let $M:\Omega\to\mathbb R$ be the population mean difference and $\Sigma:\Omega\to\mathbb R_{>0}$ the among-passage standard deviation. Use

$$
Y_p\mid M=\mu,\Sigma=\sigma
\sim
\mathcal N(\mu,\sigma^2)
$$

with scalar priors

$$
M\sim\mathcal N(0,.05^2)
\qquad\text{and}\qquad
\Sigma\sim\operatorname{Exponential}(20).
$$

In the course notation, the second argument of $\mathcal N(\mu,\sigma^2)$ is the variance. In `brms`, `normal(0, .05)` uses the standard deviation `.05`.

::: {.callout-note title="Background in the notes"}
The construction of a posterior from a likelihood and prior is developed in [posterior distributions](../../statistical-inference/posterior-distributions.qmd). [Conjugate priors](../../statistical-inference/conjugate-priors.qmd) develops the beta–Bernoulli case. The normal–normal derivation below is not supplied there. You must carry out that derivation before fitting the model.
:::

### 2.1 Derive the conditional normal–normal posterior

This derivation establishes a [conjugate update](../../statistical-inference/conjugate-priors.qmd) that is not derived in the notes. Let $P$ be the number of passages and define

$$
\overline Y
=
\frac{1}{P}\sum_{p=1}^{P}Y_p,
\qquad
\overline\delta
=
\frac{1}{P}\sum_{p=1}^{P}\delta_p.
$$

Conditional on a fixed value $\Sigma=\sigma$,

$$
\overline Y\mid M=\mu,\Sigma=\sigma
\sim
\mathcal N\left(\mu,\frac{\sigma^2}{P}\right).
$$

Write $m_0=0$ and $s_0=.05$. The quantity to derive is the scalar conditional density

$$
f_{M\mid\Sigma,\overline Y}(\mu\mid\sigma,\overline\delta).
$$

#### Step 1: likelihood kernel

Remove from the normal density of $\overline Y$ only the factors that do not depend on $\mu$.

**Target:** $f_{\overline Y\mid M,\Sigma}(\overline\delta\mid\mu,\sigma)$ as a kernel in $\mu$.

**Your LaTeX**

$$
\begin{aligned}
f_{\overline Y\mid M,\Sigma}(\overline\delta\mid\mu,\sigma)
&\propto\text{enter the likelihood kernel in }\mu.
\end{aligned}
$$

#### Step 2: prior kernel

Remove from the $\mathcal N(m_0,s_0^2)$ density only the factors that do not depend on $\mu$.

**Target:** $f_M(\mu)$ as a kernel in $\mu$.

**Your LaTeX**

$$
\begin{aligned}
f_M(\mu)&\propto\text{enter the prior kernel in }\mu.
\end{aligned}
$$

#### Step 3: posterior quadratic

Multiply the two kernels. Expand $(\overline\delta-\mu)^2$ and $(\mu-m_0)^2$. Drop only terms that do not depend on $\mu$. Collect the result as

$$
\exp\left\{-\frac12(A\mu^2-2B\mu)\right\}.
$$

**Target:** derive the coefficients $A$ and $B$ as functions of $P$, $\sigma$, $\overline\delta$, $m_0$, and $s_0$.

**Your LaTeX**

$$
\begin{aligned}
A&=\text{enter the coefficient of }\mu^2,\\
B&=\text{enter the coefficient paired with }2\mu.
\end{aligned}
$$

#### Step 4: complete the square

Use $A\mu^2-2B\mu=A(\mu-B/A)^2-B^2/A$.

**Target:** derive the posterior mean $m_P=B/A$, the posterior variance $s_P^2=1/A$, and the conditional posterior distribution.

**Your LaTeX**

$$
\begin{aligned}
m_P&=\text{enter the posterior mean in the original quantities},\\
s_P^2&=\text{enter the posterior variance in the original quantities},\\
M\mid\Sigma=\sigma,\overline Y=\overline\delta
&\sim\mathcal N\left(\text{enter the mean},\text{ enter the variance}\right).
\end{aligned}
$$

#### Step 5: posterior event probability

Define the event that the population mean difference is positive. Use full set-builder notation. Then standardize the posterior normal distribution to express its probability with the standard normal cumulative distribution function $\Phi$.

**Target:** $B_M^+$ and $\mathbb P(B_M^+\mid\Sigma=\sigma,\overline Y=\overline\delta)$.

**Your LaTeX**

$$
\begin{aligned}
B_M^+&=\{\omega\in\Omega\mid\text{enter the condition on }M(\omega)\},\\
\mathbb P(B_M^+\mid\Sigma=\sigma,\overline Y=\overline\delta)
&=\text{enter the expression in }m_P,s_P,\text{ and }\Phi.
\end{aligned}
$$

### 2.2 Check the prior and fit the model

A [prior predictive distribution](../../statistical-inference/prior-predictive-distributions.qmd) is the distribution of replicated data obtained by drawing parameters from their priors and data from the likelihood. Draw at least 10,000 values of $\mu$, $\sigma$, and one replicated difference. Plot the replicated differences and summarize $\exp(Y_{\mathrm{rep}})$.

```{r}
# YOUR CODE HERE
# 1. Set and report one seed.
# 2. Draw at least 10,000 values of mu from N(0, .05^2).
# 3. Draw the same number of sigma values from Exponential(20).
# 4. Draw one replicated difference for every mu and sigma pair.
# 5. Plot the replicated differences.
# 6. Summarize exp(y_rep) with declared quantiles.
```

Fit the normal model with `brms`. Use four chains, at least 2,000 iterations per chain, the `cmdstanr` backend, and a recorded seed.

```{r}
library(brms)
library(posterior)

detected_cores <- parallel::detectCores()
core_count <- if (is.na(detected_cores)) 1L else min(4L, detected_cores)

# Fit the normal location model.
# 1. Use delta_log_ffd ~ 1.
# 2. Use family = gaussian().
# 3. State the priors for the intercept and sigma.
# 4. Use four chains and at least 2,000 iterations per chain.
# 5. Use backend = "cmdstanr" and report the seed.
```

Report [$\widehat R$](../../statistical-inference/rhat.qmd), [bulk and tail effective sample size](../../statistical-inference/effective-sample-size.qmd), and [divergent transitions](../../statistical-inference/divergent-transitions.qmd) for the intercept and `sigma`. The [Markov-chain checks](../../statistical-inference/checking-markov-chains.qmd) put these diagnostics in the order used here.

```{r}
# YOUR CODE HERE
# 1. Extract R-hat for the intercept and sigma.
# 2. Extract bulk and tail effective sample sizes for both parameters.
# 3. Count divergent transitions.
```

From the posterior draws, calculate the median and 95% equal-tailed [credible interval](../../statistical-inference/posterior-summaries.qmd) for $M$ and $\exp(M)$. Estimate $\mathbb P(B_M^+\mid\boldsymbol\delta)$ as the proportion of intercept draws greater than zero, where $\boldsymbol\delta=(\delta_1,\ldots,\delta_P)$.

```{r}
# YOUR CODE HERE
# 1. Extract posterior draws of the intercept.
# 2. Calculate the median and equal-tailed 95% interval for M.
# 3. Transform every draw with exp() and summarize exp(M).
# 4. Calculate the proportion of M draws greater than zero.
```

Report the computational checks, posterior summaries, and posterior event probability.

::: {.callout-note title="Your result"}
Replace this sentence with the requested normal-model result.
:::

## Exercise 3: Robust location model (10 points)

The Wilcoxon test reduces the influence of numerical spacing by using ranks. A Bayesian Student t model takes a different route: it retains the measured differences but assigns more probability to extreme observations than a normal likelihood does. It is not a Bayesian version of the signed-rank test. It is the corresponding robust location analysis in this assignment.

Replace the normal likelihood in Exercise 2 with

$$
Y_p\mid M=\mu,\Sigma=\sigma,\Nu=\nu
\sim
t_\nu(\mu,\sigma),
$$

where $M$ and $\Sigma$ retain their earlier priors and the scalar degrees-of-freedom parameter has prior $\Nu\sim\operatorname{Gamma}(2,.1)$. Here the gamma distribution uses shape 2 and rate $.1$, so its mean is $2/.1=20$. The [Student t distribution](../../random-variables-and-distributions/student-t-distribution.qmd) approaches a normal distribution as $\nu$ increases and has heavier tails for smaller $\nu$.

Fit this model with `family = student()` and `prior(gamma(2, .1), class = nu)`. Use the same chain settings and seed policy as Exercise 2.

```{r}
# YOUR CODE HERE
# 1. Replace gaussian() with student().
# 2. Retain the priors for the intercept and sigma.
# 3. Add prior(gamma(2, .1), class = nu).
# 4. Retain the chain, iteration, backend, and seed policy from Exercise 2.
# 5. Report R-hat, effective sample sizes, and divergent transitions.
```

Calculate the posterior median and 95% interval for $M$, $\exp(M)$, and $\Nu$. Calculate $\mathbb P(B_M^+\mid\boldsymbol\delta)$ from the posterior draws.

```{r}
# YOUR CODE HERE
# 1. Extract posterior draws of M and Nu.
# 2. Summarize M and exp(M).
# 3. Summarize Nu.
# 4. Calculate the posterior probability that M is greater than zero.
# 5. Put the normal and Student t summaries in one table.
```

Report the requested values beside the corresponding normal-model values.

::: {.callout-note title="Your result"}
Replace this sentence with the normal and robust location results.
:::

## Exercise 4: Beta–Bernoulli sign model (10 points)

Define $Z_p:\Omega\to\{0,1\}$ by

$$
Z_p(\omega)=
\begin{cases}
1 & \text{if }Y_p(\omega)>0,\\
0 & \text{if }Y_p(\omega)\le0.
\end{cases}
$$

Let $\Pi:\Omega\to[0,1]$ be the population probability of a positive passage difference. Use

$$
Z_p\mid\Pi=\pi\sim\operatorname{Bernoulli}(\pi)
\qquad\text{and}\qquad
\Pi\sim\operatorname{Beta}(2,2).
$$

The [conjugate-prior notes](../../statistical-inference/conjugate-priors.qmd) derive the beta–Bernoulli update. Apply that update here.

### 4.1 Derive the posterior parameters

Let $k_{\mathrm{obs}}=\sum_{p=1}^{P}z_p$.

**Target:** state the beta posterior first in terms of $P$ and $k_{\mathrm{obs}}$, then substitute their observed values.

**Your LaTeX**

$$
\begin{aligned}
\Pi\mid\mathbf Z=\mathbf z
&\sim\operatorname{Beta}(\text{enter the updated parameters})\\
&=\operatorname{Beta}(\text{enter the numerical parameters}).
\end{aligned}
$$

### 4.2 Calculate posterior summaries

Calculate the posterior mean, median, and 95% equal-tailed interval. Define $B_\Pi^+=\{\omega\in\Omega\mid\Pi(\omega)>.5\}$ and calculate $\mathbb P(B_\Pi^+\mid\mathbf z)$ with `pbeta(..., lower.tail = FALSE)`. Verify it with at least 50,000 independent posterior draws and report the [Monte Carlo standard error](../../statistical-inference/monte-carlo-integration.qmd).

```{r}
# YOUR CODE HERE
# 1. Count positive and nonpositive passage differences.
# 2. Insert the counts into the beta posterior parameters.
# 3. Calculate the posterior mean, median, and equal-tailed 95% interval.
# 4. Calculate P(Pi > .5 | z) using pbeta().
# 5. Set a seed and draw at least 50,000 beta values.
# 6. Estimate the same probability from the draws.
# 7. Calculate its Monte Carlo standard error.
```

Report $k_{\mathrm{obs}}$, the posterior summaries, the event probability, and its Monte Carlo standard error.

::: {.callout-note title="Your result"}
Replace this sentence with the requested sign-model result.
:::

## Exercise 5: Bayesian two-row association (20 points)

The data for Exercises 5–6 come from the training split of [Universal Dependencies English EWT](https://github.com/UniversalDependencies/UD_English-EWT). The preparation scripts retain sentences containing exactly one selected *what*, *which*, or *who* token annotated as `nsubj` or `obj`. Each row of `data/ud_wh_dependencies.csv` represents one retained sentence. The relevant columns are `sentence_id`, `wh_word`, and `deprel`.

The categorical analysis in PS1 used a conditional null distribution over 2×2 tables. The [Fisher's exact-test notes](../../statistical-inference/fishers-exact-test.qmd) develop that construction. Here we instead place a beta prior on each scalar row probability and compare the resulting posterior distributions. The calculation does not condition on both observed margins, and its posterior event probability is not Fisher's p value.

For this exercise, retain *what* and *who*. Let $N_w$ be the number of retained sentences with word $w$, and let $K_w$ be the number annotated `obj`. Use the two scalar models

$$
K_w\mid\Pi_w=\pi_w\sim\operatorname{Binomial}(N_w,\pi_w)
\qquad\text{and}\qquad
\Pi_w\sim\operatorname{Beta}(1,1),
$$

where $w\in\{\mathrm{what},\mathrm{who}\}$.

### 5.1 Construct the table and update the posteriors

Load the data, verify that `sentence_id` is unique, and construct the two-by-two table with rows `what`, `who` and columns `nsubj`, `obj`.

```{r}
# Load the selected UD dependencies.

# Examine the column classes and observed levels.

# Verify one selected dependency per sentence_id.

# Restrict to what and who.

# Set the row and column factor levels explicitly.

# Construct the two-by-two table and extract N_w and K_w.
```

Apply the beta–binomial update separately to the two scalar parameters.

**Target:** state both posterior distributions after substituting the observed counts.

**Your LaTeX**

$$
\begin{aligned}
\Pi_{\mathrm{what}}\mid\text{data}
&\sim\operatorname{Beta}(\text{enter the parameters}),\\
\Pi_{\mathrm{who}}\mid\text{data}
&\sim\operatorname{Beta}(\text{enter the parameters}).
\end{aligned}
$$

### 5.2 Calculate the association

Draw at least 100,000 independent values from each posterior. Pair draws by simulation index and calculate

$$
\Delta_\Pi=\Pi_{\mathrm{what}}-\Pi_{\mathrm{who}}
$$

and

$$
\operatorname{OR}
=
\frac{\Pi_{\mathrm{what}}/(1-\Pi_{\mathrm{what}})}
{\Pi_{\mathrm{who}}/(1-\Pi_{\mathrm{who}})}.
$$

Calculate the posterior median and 95% interval for both quantities. Define $B_\Delta^+=\{\omega\in\Omega\mid\Delta_\Pi(\omega)>0\}$ and estimate its posterior probability and Monte Carlo standard error.

```{r}
# YOUR CODE HERE
# 1. Set and report one seed.
# 2. Draw at least 100,000 values from each scalar beta posterior.
# 3. Pair draws by simulation index.
# 4. Calculate Delta_Pi for every draw.
# 5. Calculate the odds ratio for every draw.
# 6. Summarize both posterior quantities.
# 7. Estimate P(Delta_Pi > 0 | data) and its Monte Carlo standard error.
```

Report the two observed row counts and all requested posterior summaries.

::: {.callout-note title="Your result"}
Replace this sentence with the requested two-row association result.
:::

### 5.3 Compare a second two-row association

Repeat the calculation with *which* and *who*. Define the difference as $\Pi_{\mathrm{which}}-\Pi_{\mathrm{who}}$ and orient the odds ratio with the *which* odds in the numerator. Use a new pair of scalar beta posteriors rather than a vector-valued parameter.

```{r}
# YOUR CODE HERE
# 1. Construct the which by who table.
# 2. Update the two scalar beta posteriors.
# 3. Draw from both posteriors using a recorded seed.
# 4. Calculate the probability difference and odds ratio.
# 5. Calculate the posterior medians, 95% intervals, and directional probability.
```

Put the *what*–*who* and *which*–*who* posterior summaries in one table.

::: {.callout-note title="Your result"}
Replace this sentence with the requested comparison of the two associations.
:::

### 5.4 Estimate sampling distributions across sample sizes

Treat the retained *what* and *who* sentences as a finite population. For each $n\in\{20,40,80,160\}$, draw 500 simple random samples without replacement. For each sample, update the two beta posteriors and record the posterior mean of $\Delta_\Pi$ and its 95% interval width. Use one recorded seed.

```{r}
# YOUR CODE HERE
# 1. Write a function that accepts a sample size and replicate identifier.
# 2. Draw sentences without replacement.
# 3. Update the what and who beta posteriors.
# 4. Record the posterior mean of Delta_Pi.
# 5. Record the equal-tailed 95% interval width.
# 6. Return one tidy row.
# 7. Run 500 replicates at every requested sample size.
```

For each $n$, calculate the mean and standard deviation of the posterior-mean estimates and the median interval width. Plot the empirical sampling distributions of the posterior means by $n$.

```{r}
# 1. Group the replicate results by sample size.
# 2. Calculate every requested summary.
# 3. Plot the posterior-mean sampling distributions by sample size.
```

Report the sample-size summary table.

::: {.callout-note title="Your result"}
Replace this sentence with the sample-size summary table.
:::

## Exercise 6: Posterior predictive independence model (25 points)

Use all three word forms. Let $g\in\{\mathrm{what},\mathrm{which},\mathrm{who}\}$ index rows, $N_g$ the row total, and $K_g$ the object count. Under independence, all rows share the scalar object probability $\Pi_0$:

$$
K_g\mid\Pi_0=\pi_0
\sim
\operatorname{Binomial}(N_g,\pi_0),
\qquad
\Pi_0\sim\operatorname{Beta}(1,1).
$$

A [posterior predictive distribution](../../statistical-inference/posterior-predictive-checks.qmd) is obtained by drawing $\Pi_0$ from its posterior and then drawing a replicated table from the likelihood.

### 6.1 Derive the shared-probability posterior

Pool the object counts and non-object counts across rows.

**Target:** derive the beta posterior of $\Pi_0$ and substitute the observed totals.

**Your LaTeX**

$$
\begin{aligned}
\Pi_0\mid\mathbf K=\mathbf k
&\sim\operatorname{Beta}(\text{enter the updated parameters})\\
&=\operatorname{Beta}(\text{enter the numerical parameters}).
\end{aligned}
$$

### 6.2 Define the discrepancy

Construct the observed three-by-two table in row order `what`, `which`, `who` and column order `nsubj`, `obj`. Calculate its expected counts under independence and its Pearson statistic using the formula from the [chi-squared-test notes](../../statistical-inference/chi-squared-test.qmd).

```{r}
# YOUR CODE HERE
# 1. Set the word-form and dependency-relation levels explicitly.
# 2. Construct the observed three-by-two table.
# 3. Calculate its row totals, column totals, and grand total.
# 4. Calculate the complete expected table under independence.
# 5. Calculate each Pearson contribution and x_squared_observed.
```

### 6.3 Simulate replicated tables

Generate at least 100,000 posterior predictive tables. For each replicate, draw one value of $\Pi_0$, draw one object count for each fixed row total, construct the replicated table, and calculate its Pearson statistic.

```{r}
# YOUR CODE HERE
# 1. Set and report one seed.
# 2. Draw at least 100,000 values from the posterior of Pi_0.
# 3. For each draw, generate one object count for each fixed row total.
# 4. Construct the replicated table.
# 5. Recalculate its expected counts from its margins.
# 6. Calculate and store its Pearson statistic.
```

Define the posterior predictive tail event in full set-builder notation.

**Target:** $B_{X^2}^{\mathrm{rep}}$ and its posterior probability.

**Your LaTeX**

$$
\begin{aligned}
B_{X^2}^{\mathrm{rep}}
&=\{\omega\in\Omega\mid\text{enter the discrepancy condition}\},\\
\mathbb P(B_{X^2}^{\mathrm{rep}}\mid\mathbf K=\mathbf k)
&=\text{enter the Monte Carlo estimator}.
\end{aligned}
$$

Plot the replicated statistics with the observed statistic marked. Calculate the posterior predictive tail probability and its Monte Carlo standard error. Calculate the observed Pearson residuals and identify the two cells with the largest absolute residuals.

```{r}
# Plot the replicated Pearson statistics and mark x_squared_observed.

# Calculate the posterior predictive tail probability.

# Calculate its Monte Carlo standard error.

# Calculate and sort the observed Pearson residuals by absolute magnitude.
```

Report the observed statistic, posterior predictive tail probability, Monte Carlo standard error, and two largest absolute residuals with their signs.

::: {.callout-note title="Your result"}
Replace this sentence with the requested posterior predictive result.
:::

## Submission checklist

- The notebook runs from a fresh R session after the data-preparation steps.
- Every derivation is entered in its LaTeX block.
- Every computation is entered in the R block immediately following its instruction.
- Every requested result is entered in its result block.
- Every parameter is scalar.
- The passage table is joined by `Text_ID` rather than aligned by row position.
- Every posterior event is written in full set-builder notation.
- Every random calculation has a recorded seed.
- Markov-chain diagnostics are checked before posterior summaries are reported.
