---
title: "Problem Set 2: Prediction across passages"
subtitle: "Statistical Methods in Linguistics"
format:
  html:
    toc: true
execute:
  enabled: false
---

## The question

Accounts of prediction in comprehension differ in what they take readers to anticipate. One view emphasizes a particular upcoming word form. Another allows context to support partial information, such as the upcoming word's syntactic class or semantic content, even when its full form is uncertain.

Luke and Christianson tested these accounts using the Provo Corpus. Their coefficient analyses suggest that orthographic, morphosyntactic, and semantic predictability are associated with eye movements. We will ask a different question: do these predictor sets improve first-fixation prediction for a passage omitted during model fitting?

The focal method is [grouped cross-validation](../../model-criticism-and-prediction/grouped-cross-validation.qmd). The models are [linear regressions](../../regression-models/frequentist-linear-regression.qmd) already introduced in the notes.

::: {.callout-note title="Reading"}
Read Luke and Christianson (2016), [“Limits on lexical prediction during reading”](https://doi.org/10.1016/j.cogpsych.2016.06.002), together with the Provo Corpus data description in Luke and Christianson (2018), [“The Provo Corpus”](https://doi.org/10.3758/s13428-017-0908-4). The first reading supplies the lexical and graded prediction contrast. The second explains how the passages, readers, eye movements, and predictability norms were collected. Both are included because a passage-held-out score is interpretable only after the theoretical predictors and the sampling units have been named.
:::

## Data, source, and target

The original data come from the [Provo Corpus](https://doi.org/10.3758/s13428-017-0908-4), with the theoretical analysis in [Luke and Christianson (2016)](https://doi.org/10.1016/j.cogpsych.2016.06.002). The corpus and its files are available from [OSF](https://osf.io/sjefs/) under [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/).

`data/provo-content-word-positions.csv` is an adapted teaching table. One row represents one content-word position. The response `mean_log_ffd` is the mean log first-fixation duration over observed readers who fixated that position. Thus the analysis is conditional on fixation and targets an observed-reader mean, not individual reading behavior.

```{r}
items <- read.csv("data/provo-content-word-positions.csv")
dim(items)
head(items)
```

The principal columns are:

- `position_id`: repaired passage by position identifier;
- `Text_ID`: passage identifier;
- `Word_POS`: adjective, adverb, noun, or verb;
- `Word_Length` and `Word_In_Sentence_Number`: control predictors;
- `OrthoMatchModel`: smoothed orthographic cloze probability;
- `POSMatchModel`: smoothed part-of-speech match probability;
- `LSA_Response_Match_Score`: mean semantic similarity between cloze responses and the target; and
- `mean_log_ffd`: the response.

The published `Word_Unique_ID` is not a valid key by itself: `QID2687` occurs in Texts 18 and 55. The preparation script constructs `position_id` from passage and word position.

## Task 1: state what must be held out (15 points)

Write a four-sentence [prediction-target statement](../../model-criticism-and-prediction/prediction-targets.qmd). Name:

1. the response;
2. the unit that must be new;
3. the information available for a held-out unit; and
4. the population or use setting to which the score is intended to apply.

Then explain why random word-position folds and passage folds answer different questions, even if their numerical scores happen to be similar.

## Task 2: prespecify the model sequence (20 points)

Create these deterministic transforms:

```{r}
items$log_length <- log(items$Word_Length)
items$log_position <- log(items$Word_In_Sentence_Number)
items$log_ortho <- log(items$OrthoMatchModel)
items$log_pos <- log(items$POSMatchModel)
```

Fit the three full-data models below for interpretation only.

1. **Control:** log length, log sentence position, and word class.
2. **Lexical:** the control model plus log orthographic cloze probability.
3. **Graded:** the lexical model plus log part-of-speech match and semantic similarity between the response and target.

```{r}
control_fit <- lm(
  mean_log_ffd ~ log_length + log_position + Word_POS,
  data = items
)

# Fit lexical_fit and graded_fit.
```

Before scoring, state the expected sign of each predictability coefficient. Connect the orthographic predictor to lexical prediction and the two added predictors to graded prediction.

Do not select predictors using these fitted p values. The sequence is fixed by the theoretical comparison.

## Task 3: random position folds (20 points)

The helper below fits a linear model from a fixed design matrix and returns test predictions.

```{r}
fit_predict <- function(design, response, train, test) {
  coefficients <- qr.coef(qr(design[train, , drop = FALSE]), response[train])
  coefficients[is.na(coefficients)] <- 0
  drop(design[test, , drop = FALSE] %*% coefficients)
}

score <- function(observed, predicted) {
  c(
    RMSE = sqrt(mean((observed - predicted)^2)),
    MAE = mean(abs(observed - predicted))
  )
}
```

Construct one design matrix for each prespecified model with `model.matrix()`. Use seed `20260801` to assign positions to ten random folds. Obtain one out-of-fold prediction per row and report RMSE and MAE for the three models.

```{r}
# Construct design matrices, random folds, and out-of-fold predictions.
```

Repeat the ten-fold procedure for the 100 seeds `20260801` through `20260900`. Report how often the graded model has lower RMSE than the lexical model. Explain why these 100 results are split-sensitivity checks, not 100 independent corpus replications.

## Task 4: leave one passage out (25 points)

Now obtain predictions by leaving out one complete passage at a time. Every position in a held-out passage must be predicted by a model fit to the other 54 passages.

```{r}
# Produce leave-one-passage-out predictions for all three models.
```

Verify the split with code: for every fitted model, the held-out `Text_ID` must be absent from the training indices. Report aggregate RMSE and MAE for each model.

Then compute these quantities for each passage:

$$
\Delta_{p,\mathrm{lexical}}
=
\operatorname{MSE}_{p,\mathrm{control}}
-
\operatorname{MSE}_{p,\mathrm{lexical}}
$$

and

$$
\Delta_{p,\mathrm{graded}}
=
\operatorname{MSE}_{p,\mathrm{lexical}}
-
\operatorname{MSE}_{p,\mathrm{graded}}.
$$

Positive values favor the more theoretically specific model. Plot both distributions with a visible zero point, and report the proportion of passages with positive values.

## Task 5: interpret transfer, not just coefficients (20 points)

For each passage-level comparison, compute a 95% t interval for the mean loss difference, following the earlier treatment of [confidence intervals](../../statistical-inference/confidence-intervals.qmd). Use the 55 passage differences, not ten computational fold scores or 1,593 word positions.

Write a results argument of two short paragraphs.

1. Does orthographic cloze improve prediction beyond the control model for passages omitted during fitting?
2. Do the morphosyntactic and semantic measures establish a stable further improvement across passages?

Reconcile your answer with the full-data coefficient directions and the random-fold results. Do not turn an interval containing zero into the claim that graded prediction is false. State exactly which predictive claim is limited by the result.

## LING 414 extension: loss sensitivity

Repeat the passage-level comparison using mean absolute error. Compare the direction, passage improvement rate, and interval with the squared-loss analysis. Explain which passages drive any disagreement and whether your substantive conclusion depends on the primary loss.

## Submission checklist

- The model sequence is fixed before scoring.
- Every response receives exactly one held-out prediction in each validation procedure.
- Passage rows move together in the grouped analysis.
- Test responses do not enter preprocessing or fitting.
- Loss differences use one sign convention and identify which model positive values favor.
- Uncertainty is reported over 55 passages.
- The conclusion distinguishes association in the fitted corpus from transfer to omitted passages.
